### 2019-08-15 article.tex
initial commit

\documentclass[11pt]{article}

\title{Draft Article}
\author{Alexander W. Petros}
\date{September 2019}
\usepackage{style}

% \addbibresource{thesis.bib}

\begin{document}
\maketitle
\newpage

Some text
\end{document}

### 2019-08-16 article.tex
started introduction

@@
-\documentclass[11pt]{article}
+\documentclass[12pt]{article}
+\usepackage{style}
 
@@
 \date{September 2019}
-\usepackage{style}
 
-% \addbibresource{thesis.bib}
+% \addbibresource{article.bib}
 
@@
 
-Some text
+The most damaging cyberattack committed against the United States federal government by a foreign agent was not a dramatic affair.
+Starting sometime in 2014, a small probe lurked deep within the network of the US Office of Personnel Management (OPM), an office of about 5,000 employees tasked with overseeing background checks, payroll, and other human resources concerns for the federal government.
+At regular intervals, the probe would send updates to opmsecurity.org, a domain registered to look like an official website, but not actually affiliated with the US government or the OPM security team.
+By the time the breach was discovered in April 2015, the cyber-spy had already accomplished its mission---the probe had accessed crucial federal databases and transmitted the stolen data back to its country of origin.
+
+Internal investigators determined that OPM had been compromised by an advanced persistent threat (APT), a formally organized and typically state-sponsored group of hackers.
+US officials would only confirm that they believed a \enquote{foreign entity or government} had been behind the attack, but it was an open secret at the time that they suspected the involvement of the Chinese government.\footcite{spetalnick_china_2015} The APT made identification easy by leaving their calling-card: the trojan opmsecurity.org domain was registered in the name of \enquote{Steve Rogers,} the Marvel comics character better known as Captain America.
+This virtual taunt, along with Chinese-based email and IP addresses, all but guaranteed that the attack had come from an APT sponsored by the Chinese government, likely the Chinese military's cyber-espionage division.
+
+One of the compromised databases was a trove of applications for federal security clearance, compromising the
+
+In recent years, cyberattacks perpetrated by state-sponsored foreign agents have received ever-increasing levels of public scrutiny.
+
+As investigations continued, the news got worse; the scope of the attack greatly exceeded the initial estimates.
+The Chinese had accessed an OPM database of applications for security clearance---which exposed not just the personal data of the applicants themselves but also the detailed information they had supplied about their family and friends.\footcite{nakashima_hacks_2015} Millions of social security numbers, job applications, and home addresses had been compromised, including information about personnel at the highest levels of government.
+When OPM announced the full extent of the damage, it caused an uproar.
+The American Federation of Government Employees, the largest federal workers union, filed a class-action lawsuit against OPM, seeking damages under the Privacy Act.\footcite[The lawsuits were later dismissed.]{chalfant_court_2017} Mark Warner (D-Va.), a member of the Senate Intelligence Committee whose constituents included many of the employees affected, called for OPM Director Katherine Archuleta's resignation.
+Because the information might be used for identity theft, the US government offered all affected employees free credit and identity monitoring services for three years.\footcite{nakashima_hacks_2015}
+
+Not only was the OPM hack a public relations disaster, it had enormous implications for national security and counterintelligence.
+US officials feared that the stolen records could expose undercover CIA operatives working in embassies around the world, whose names would be suspiciously absent from the OPM records.\footcite{nakashima_hacks_2015} The information could be used by a foreign intelligence service to identify American officials who might be susceptible to pressure, or worse, recruitment.
+FBI Director James Comey personally took questions about the incident.
+``If you have my [application for security clearance],'' Comey said, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.
+Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015} Now the Chinese intelligence service did and, more disturbingly, there were no signs of what they intended to do with it.\footcite{koerner_inside_2016}
 \end{document}

### 2019-08-16 article.tex
added bib

@@
 
-% \addbibresource{article.bib}
+\addbibresource{article.bib}
 
@@
 
+One of the compromised databases was a trove of applications for federal security clearance, compromising the
+
 Internal investigators determined that OPM had been compromised by an advanced persistent threat (APT), a formally organized and typically state-sponsored group of hackers.
 US officials would only confirm that they believed a \enquote{foreign entity or government} had been behind the attack, but it was an open secret at the time that they suspected the involvement of the Chinese government.\footcite{spetalnick_china_2015} The APT made identification easy by leaving their calling-card: the trojan opmsecurity.org domain was registered in the name of \enquote{Steve Rogers,} the Marvel comics character better known as Captain America.
-This virtual taunt, along with Chinese-based email and IP addresses, all but guaranteed that the attack had come from an APT sponsored by the Chinese government, likely the Chinese military's cyber-espionage division.
-
-One of the compromised databases was a trove of applications for federal security clearance, compromising the
+This virtual taunt, along with the Chinese-based email and IP addresses, all but guaranteed that the attack had come from an APT sponsored by the Chinese government, likely the Chinese military's cyber-espionage division.\footcite{koerner_inside_2016} Nonetheless, the US government refused to name China as the culprit.
 

### 2019-08-20 article.tex
initial R script

--- 2019-08-15 article.tex
@@
-\documentclass[11pt]{article}
+\documentclass[14pt]{extarticle}
+\usepackage{style}
 
@@
 \date{September 2019}
-\usepackage{style}
 
-% \addbibresource{thesis.bib}
+\addbibresource{article.bib}
 
@@
 \maketitle
-\newpage
+% \newpage
 
-Some text
+The most damaging cyberattack committed against the United States federal government by a foreign agent was remarkably subtle.
+Starting sometime in 2014, a small probe embedded itself deep within the network of the US Office of Personnel Management (OPM), an office of about 5,000 employees tasked with overseeing background checks, payroll, and other human resources concerns for the federal government.
+At regular intervals, the probe would send updates to opmsecurity.org, a domain registered to look like an official website, but not actually affiliated with the US government or the OPM security team.
+Unobtrusive and observant, the probe functioned in all respects like a digital spy.
+
+By the time the breach was discovered in April 2015, the probe had accessed crucial government databases and transmitted the stolen information back to its country of origin---including a trove of applications for federal security clearance.
+These applications exposed not just the personal data of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses.
+Federal authorities determined that the hack ultimately affected 22.1 million people in total, including personnel at the highest levels of government.\footcite{nakashima_hacks_2015} When OPM announced the full extent of the damage, it caused an uproar.
+The American Federation of Government Employees, the largest federal workers union, filed a class-action lawsuit against OPM, seeking damages under the Privacy Act.\footcite{chalfant_court_2017} Senator Mark Warner (D-Va.) of the Senate Intelligence Committee called for OPM Director Katherine Archuleta's resignation.
+Because the information might be used for identity theft, the US government offered all affected employees free credit and identity monitoring services for three years.
+
+Not only was the OPM hack a public relations disaster, it had significant national security implications as well.
+US officials feared that the stolen records could compromise intelligence efforts by exposing undercover CIA officers working in embassies around the world, whose names would be suspiciously absent from the OPM records.
+The information could also be used to identify American officials who might be susceptible to pressure, or worse, recruitment.
+FBI Director James Comey personally took questions about the incident.
+``If you have my [application for security clearance],'' Comey said, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.
+Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015} With evidence that a foreign intelligence service had successfully obtained exact that, one might expect that the United States would respond with a series of diplomatic measures intended to signal that highly invasive acts of espionage would not be tolerated from its international adversaries.
+
+Instead, the opposite happened.
+US officials would only confirm that they believed a \enquote{foreign entity or government} had been behind the attack, even though it was widely known that they suspected the involvement of the Chinese government.\footcite{spetalnick_china_2015} Internal investigators quickly determined that OPM had been compromised by an Advanced Persistent Threat (APT), a formally organized and typically state-sponsored group of hackers that engage in long-term, targeted penetration operations.
+In this case, the APT made identification easy by leaving their calling-card: the trojan opmsecurity.org domain was registered in the name of \enquote{Steve Rogers,} the Marvel comics character better known as Captain America.
+This digital taunt, along with email and IP addresses that verified the country of origin, all but guaranteed that the attack had come from an APT sponsored by the Chinese government, likely the Chinese military's cyber-espionage division.\footcite{koerner_inside_2016} Sitting on conclusive digital forensics, the Obama administration nonetheless refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018.
+See \cite{sanger_trump_2018}}
+
+In recent years, cyberattacks perpetrated by state-sponsored foreign agents have received ever-increasing levels of public scrutiny---which makes the government's response to the OPM hack especially puzzling.
+Russian interference in the 2016 US Presidential election prompted the expulsion of 37 diplomats and series of follow-on sanctions.
+
+% This article seeks to explain why espionage efforts are not more actively discouraged by nations, even as the results of intelligence operations can oftentimes have a devastating effect on the targeted nation's security.
+
+% first it will lay out how espionage is considered an occupational hazard in the business of running a nation-state, one that is counteracted but never discourages.
+
+% next I will discuss the theoretical justifications for why this might be the case. security seeking states and types of greedy states derive a reational benefit from allowing espionage to continue.
+
+% finally I will look at historical moments in which new technologies were deployed for espionage-like purporses, and demonstrate how the norms governing the use of these technologies always evolved to permit espionage.
+
 \end{document}

### 2019-08-20 data/incidents-scripts.R

@@ -1,0 +1,39 @@
+# Scripts for cyber-espionage journal article
+# Autor: Alex Petros 
+library(tidyverse)
+library(ggthemes)
+
+# https://www.cfr.org/interactive/cyber-operations
+cfr_data <- read_csv('./cyber-operations-incidents.csv') 
+
+# added some useful columns and removed sources
+cyber_incidents <- cfr_data %>% 
+  mutate(us_target=if_else(grepl("United States", Victims), TRUE, FALSE)) %>% 
+  mutate(has_response=!is.na(Response)) %>% 
+  mutate(Response=str_extract(Response, '\\w+\\b')) %>% 
+  select(-Sources_1, -Sources_2, -Sources_3) %>% 
+  replace_na(list(Type = "Undefined"))
+
+# select cyber-incidents aginst the US
+us_targeted <- cyber_incidents %>% 
+  filter(us_target == TRUE) %>%
+  select(-us_target)
+
+# count of how often states respond to cyber-incidents by type
+# swap in us_targeted for for US-only count
+response_stats <- cyber_incidents %>% 
+  group_by(Type) %>% 
+  summarize(num_incidents = n(), 
+            num_responses = sum(has_response==TRUE), 
+            response_pct = mean(has_response==TRUE)*100) %>% 
+  ungroup() %>% 
+  arrange(desc(response_pct))
+response_stats
+  
+
+# plot workspace
+ggplot(data=filter(us_targeted, !is.na(Response))) +
+  geom_bar(mapping=aes(x=Type, fill=Response))
+
+ex_table <- filter(us_targeted, Response == "Criminal")
+ex_table

### 2019-08-21 article.tex
introductory transition to central questions about espionage

@@
 ``If you have my [application for security clearance],'' Comey said, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.
-Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015} With evidence that a foreign intelligence service had successfully obtained exact that, one might expect that the United States would respond with a series of diplomatic measures intended to signal that highly invasive acts of espionage would not be tolerated from its international adversaries.
+Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015} With evidence that a foreign intelligence service had successfully obtained exactly that, one might expect that the United States would respond with a series of diplomatic measures intended to signal that highly invasive acts of espionage would not be tolerated from its international adversaries.
 
@@
 In this case, the APT made identification easy by leaving their calling-card: the trojan opmsecurity.org domain was registered in the name of \enquote{Steve Rogers,} the Marvel comics character better known as Captain America.
-This digital taunt, along with email and IP addresses that verified the country of origin, all but guaranteed that the attack had come from an APT sponsored by the Chinese government, likely the Chinese military's cyber-espionage division.\footcite{koerner_inside_2016} Sitting on conclusive digital forensics, the Obama administration nonetheless refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018.
+This digital taunt, along with email and IP addresses that verified the country of origin was not a false flag, all but guaranteed that the attack had come from an APT sponsored by the Chinese government, likely the Chinese military's cyber-espionage division.\footcite{koerner_inside_2016} Sitting on conclusive digital forensics, the Obama administration nonetheless refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018.
 See \cite{sanger_trump_2018}}
 
-In recent years, cyberattacks perpetrated by state-sponsored foreign agents have received ever-increasing levels of public scrutiny---which makes the government's response to the OPM hack especially puzzling.
-Russian interference in the 2016 US Presidential election prompted the expulsion of 37 diplomats and series of follow-on sanctions.
+The US government's tepid response to the OPM hack illustrates a universally understood but rarely acknowledged fact about the relationship between cybersecurity and espionage: when a cyber operation appears to fall within the bounds of a traditional intelligence operation, it will merit the same diplomatic consequences that a traditional intelligence operation would have.
+Typically those consequences are minimal, rarely amounting to more than a pat denunciation of the activity and the necessary countermeasures to end it.
+Most importantly, those consequences never rise to the level where they impose additional costs on the state that attempted the espionage in the first place; in the rare cases when costs \emph{are} imposed for a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.
+
+At first glance this observation might seem almost tautological in its simplicity.
+Of course cyber-espionage is treated like traditional espionage, one could argue---they're both espionage.
+That formulation begs the question;  it doesn't explain how states determine what constitutes \enquote{traditional} espionage, what purpose that distinction servers, or whether it is prudent to continue making it.
+
+Despite the magnitude of the OPM hack, American officials had a hard time seeing it as anything other than a well-executed bit of espionage.
+A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].} For activities that fall under the umbrella of intelligence-gathering, counterintelligence, not prosecutions, are deemed to be the appropriate response.\footcite[In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets.]{finklea_cyber_2015} Director of the CIA John Brennan described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, committed a classic Kinsley gaffe—saying what every official knows to be true but would never admit out loud: \enquote{You have to kind of salute the Chinese for what they did.
+If we had the opportunity to do that, I don’t think we’d hesitate for a minute.}\footcite{pepitone_clapper_2015}
+
+% In recent years, cyberattacks perpetrated by state-sponsored foreign agents have received ever-increasing levels of public scrutiny---which makes the government's response to the OPM hack especially puzzling.
+Why did an incident of this magnitude not merit an official response?
 
 % This article seeks to explain why espionage efforts are not more actively discouraged by nations, even as the results of intelligence operations can oftentimes have a devastating effect on the targeted nation's security.
+
+% Russian interference in the 2016 US Presidential election prompted the expulsion of 37 diplomats and series of follow-on sanctions
 

### 2019-08-21 data/incidents-scripts.R

@@ -21,7 +21,7 @@ us_targeted <- cyber_incidents %>%
 
 # count of how often states respond to cyber-incidents by type
 # swap in us_targeted for for US-only count
-response_stats <- cyber_incidents %>% 
+response_stats <- us_targeted %>% 
   group_by(Type) %>% 
   summarize(num_incidents = n(), 
             num_responses = sum(has_response==TRUE), 
@@ -35,5 +35,5 @@ response_stats
 ggplot(data=filter(us_targeted, !is.na(Response))) +
   geom_bar(mapping=aes(x=Type, fill=Response))
 
-ex_table <- filter(us_targeted, Response == "Criminal")
-ex_table
+# ex_table <- filter(us_targeted, Response == "Criminal")
+# ex_table

### 2019-08-22 article.tex
introductory section structure

@@
 The US government's tepid response to the OPM hack illustrates a universally understood but rarely acknowledged fact about the relationship between cybersecurity and espionage: when a cyber operation appears to fall within the bounds of a traditional intelligence operation, it will merit the same diplomatic consequences that a traditional intelligence operation would have.
-Typically those consequences are minimal, rarely amounting to more than a pat denunciation of the activity and the necessary countermeasures to end it.
+This justification was explicitly deployed by senior US policymakers, albeit often through unofficial moments of candor.
+A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].}\footcite[In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets.]{finklea_cyber_2015} For activities that fall under the umbrella of intelligence-gathering, counterintelligence, not prosecution, is deemed to be the appropriate response.
+Director of the CIA John Brennan described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence said what every official knows to be true but would never admit out loud: \enquote{You have to kind of salute the Chinese for what they did.
+If we had the opportunity to do that, I don’t think we’d hesitate for a minute.}\footcite{pepitone_clapper_2015} Therefore, to understand why policymakers treat cyber-espionage as \enquote{fair game}, one needs to understand why traditional espionage is treated that way.
+
+This article seeks to establish a theoretical framework for understanding the role that espionage plays in international relations (IR) and the diplomatic consequences that are associated with it.
+Typically those consequences are minimal, rarely amounting to more than a pat denunciation of the activity and implementation of the necessary countermeasures to end it.
 Most importantly, those consequences never rise to the level where they impose additional costs on the state that attempted the espionage in the first place; in the rare cases when costs \emph{are} imposed for a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.
+Espionage retains a special status within the international community: the activity itself is expressly forbidden, but every state tacitly acknowledges that allies and adversaries alike are guaranteed to engage in it.
 
-At first glance this observation might seem almost tautological in its simplicity.
+Because the diplomatic protocols behind espionage are relatively understudied from a social science perspective, this article makes two main contributions to the academic literature on espionage.
+The first is that it advances a theory for why espionage continues to retain its liminal status in the international legal system, where it is disavowed but allowed.
+Public discourse on this subject often uses terms like \enquote{rules of the road} or \enquote{gentleman's agreement} to describe the understanding between governments that espionage should fall within certain parameters.
+By employing both realist and constructivist schools of IR, this article develops rigerous explanations for why rational states, whether they be security-seeking and greedy, might cooperate to preserve this dynamic.
+
+\section{Existing research on espionage}
+Previous scholarly work examines the curious liminal space in which espionage exists, but not the rationale that states have for perpetuating it.
+
+\section{Cyber-espionage as espionage}
+At first glance this observation might seem almost tautological.
 Of course cyber-espionage is treated like traditional espionage, one could argue---they're both espionage.
 That formulation begs the question; it doesn't explain how states determine what constitutes \enquote{traditional} espionage, what purpose that distinction servers, or whether it is prudent to continue making it.
-
-Despite the magnitude of the OPM hack, American officials had a hard time seeing it as anything other than a well-executed bit of espionage.
-A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].} For activities that fall under the umbrella of intelligence-gathering, counterintelligence, not prosecutions, are deemed to be the appropriate response.\footcite[In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets.]{finklea_cyber_2015} Director of the CIA John Brennan described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, committed a classic Kinsley gaffe—saying what every official knows to be true but would never admit out loud: \enquote{You have to kind of salute the Chinese for what they did.
-If we had the opportunity to do that, I don’t think we’d hesitate for a minute.}\footcite{pepitone_clapper_2015}
 

### 2019-08-22 data/incidents-scripts.R

@@ -3,6 +3,9 @@
 library(tidyverse)
 library(ggthemes)
 
+# comment this line if your file is elsewhere
+setwd('~/Documents/journal/src/data/')
+
 # https://www.cfr.org/interactive/cyber-operations
 cfr_data <- read_csv('./cyber-operations-incidents.csv') 
 
@@ -21,7 +24,7 @@ us_targeted <- cyber_incidents %>%
 
 # count of how often states respond to cyber-incidents by type
 # swap in us_targeted for for US-only count
-response_stats <- us_targeted %>% 
+response_stats <- cyber_incidents %>% 
   group_by(Type) %>% 
   summarize(num_incidents = n(), 
             num_responses = sum(has_response==TRUE), 

### 2019-08-25 article.tex
intro first draft

@@
 A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].}\footcite[In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets.]{finklea_cyber_2015} For activities that fall under the umbrella of intelligence-gathering, counterintelligence, not prosecution, is deemed to be the appropriate response.
-Director of the CIA John Brennan described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence said what every official knows to be true but would never admit out loud: \enquote{You have to kind of salute the Chinese for what they did.
-If we had the opportunity to do that, I don’t think we’d hesitate for a minute.}\footcite{pepitone_clapper_2015} Therefore, to understand why policymakers treat cyber-espionage as \enquote{fair game}, one needs to understand why traditional espionage is treated that way.
+Director of the CIA John Brennan described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, was especially honest: \enquote{You have to kind of salute the Chinese for what they did.
+If we had the opportunity to do that, I don’t think we’d hesitate for a minute.}\footcite{pepitone_clapper_2015}
 
-This article seeks to establish a theoretical framework for understanding the role that espionage plays in international relations (IR) and the diplomatic consequences that are associated with it.
+% latex table generated in R 3.4.3 by xtable 1.8-2 package
+% Sat Aug 24 13:46:46 2019
+\begin{table}[ht]
+\centering
+\begin{tabular}{lrrr}
+  \hline
+Type & Incidents & Official responses & Response \% \\
+  \hline
+Defacement &   3 &   2 & 66.67 \\
+  Data destruction &   8 &   5 & 62.50 \\
+  Doxing &   6 &   3 & 50.00 \\
+  Sabotage &  14 &   5 & 35.71 \\
+  Denial of service &  17 &   6 & 35.29 \\
+  Espionage & 231 &  44 & 19.05 \\
+   \hline
+\end{tabular}
+\caption{Cyber-incident response rates by type of incident}
+\label{response-pct}
+\end{table}
+
+In fact, cyber-espionage is routinely treated as a lesser category of offense by all nations, not just the United States.
+Of the hundreds of state-sponsored cyberattacks about which there is public knowledge, attacks classified as espionage are by the far most common---and the least likely to garner an official response from the victim government (Table \ref{response-pct}).\footnote{These numbers were compiled from the Council on Foreign Relations' Cyber Operations Tracker, which is a publicly-sourced record of state-sponsored cyberattacks, updated quarterly.
+Of the 323 listed incidents, 38 were excluded from this calculation because they lacked an identifiable Victim, Sponsor, or Type.
+For more, see \cite{council_on_foreign_relations_new_2019}} The routine minimization of espionage is most easily observable with the plethora of cyber-incidents that we deal with today, but it has its roots in norms that are much older.
+Therefore, to understand why policymakers treat cyber-espionage as \enquote{fair game,} one needs to understand why traditional espionage receives that treatment first.
+
+I posit that the universal tolerance of espionage in the international system can be explained through a rational calculation of each state's incentives.
+On balance, states derive a benefit from preserving a system in which the consequences for espionage are minimized, so they refrain from imposing particularly harsh punishments for its practice; even when an individual incident is especially damaging to a state's security, if it falls within the traditionally accepted boundaries of an espionage operation, then it will likely go unpunished.
+In the context of the OPM hack, that means that US policymakers decided that they could not impose harsh penalties on the Chinese because it was more valuable to them that the US retain its ability to practice cyber-espionage with relative impunity in the future.
+This basic calculation has remained consistent since the advent of the modern peacetime intelligence service, perhaps even the nation-state itself.
+
+This article seeks to establish a theoretical framework for analyzing the role that espionage plays in international relations (IR) and the diplomatic norms that are associated with it.
+Through this framework, I explain what constitutes traditional espionage and why rational states might calculate that it is within their best interests to continue tolerating it.
+I then examine two cases in which new technical means of gathering intelligence were introduced during the Cold War: aerial photography (spy planes) and satellite photography (spy satellites).
+In both cases, the norms of traditional espionage ultimately prevailed, minimizing the diplomatic consequences for their use as long as it was in service of traditional intelligence goals.
+Through process tracing, I demonstrate how the justifications that policymakers employed for their use of these intelligence technologies---and the resulting minimization of the diplomatic penalties associated with them---are consistent with the theoretical underpinnings for why an accord tacitly permitting the use of espionage could be mutually strategic.
+
+\subsection{Literature base}
+This article makes three main contributions to the academic literature on espionage.
+The first is that it advances a theory for why espionage continues to retain its liminal status in the international legal system, where it is disavowed but allowed.
+Public discourse on this subject often uses terms like \enquote{rules of the road} or \enquote{gentleman's agreement} to describe the understanding between governments that espionage should fall within certain parameters.
+By employing both realist and constructivist schools of IR, this article develops rigorous explanations for why rational states, whether they be security-seeking and greedy, might cooperate to preserve this dynamic.
+
+Previous academic literature on espionage and diplomacy has treated espionage as \emph{sui generis}, a legal outlier whose norms are entirely endogenous to the insular intelligence community.
+
+Rather than explain the constituent practices of espionage as historically contingent,
+
 Typically those consequences are minimal, rarely amounting to more than a pat denunciation of the activity and implementation of the necessary countermeasures to end it.
@@
 Espionage retains a special status within the international community: the activity itself is expressly forbidden, but every state tacitly acknowledges that allies and adversaries alike are guaranteed to engage in it.
-
-Because the diplomatic protocols behind espionage are relatively understudied from a social science perspective, this article makes two main contributions to the academic literature on espionage.
-The first is that it advances a theory for why espionage continues to retain its liminal status in the international legal system, where it is disavowed but allowed.
-Public discourse on this subject often uses terms like \enquote{rules of the road} or \enquote{gentleman's agreement} to describe the understanding between governments that espionage should fall within certain parameters.
-By employing both realist and constructivist schools of IR, this article develops rigerous explanations for why rational states, whether they be security-seeking and greedy, might cooperate to preserve this dynamic.
 
@@
 That formulation begs the question; it doesn't explain how states determine what constitutes \enquote{traditional} espionage, what purpose that distinction servers, or whether it is prudent to continue making it.
+
+\section{Theory}
 

### 2019-08-25 data/incidents-scripts.R

@@ -2,6 +2,8 @@
 # Autor: Alex Petros 
 library(tidyverse)
 library(ggthemes)
+library(xtable)
+library(clipr)
 
 # comment this line if your file is elsewhere
 setwd('~/Documents/journal/src/data/')
@@ -9,7 +11,7 @@ setwd('~/Documents/journal/src/data/')
 # https://www.cfr.org/interactive/cyber-operations
 cfr_data <- read_csv('./cyber-operations-incidents.csv') 
 
-# added some useful columns and removed sources
+# added some useful columns, cleaned data, and removed sources
 cyber_incidents <- cfr_data %>% 
   mutate(us_target=if_else(grepl("United States", Victims), TRUE, FALSE)) %>% 
   mutate(has_response=!is.na(Response)) %>% 
@@ -25,14 +27,25 @@ us_targeted <- cyber_incidents %>%
 # count of how often states respond to cyber-incidents by type
 # swap in us_targeted for for US-only count
 response_stats <- cyber_incidents %>% 
+  filter(Type !='Undefined' & !is.na(Sponsor) & !is.na(Victims)) %>% 
   group_by(Type) %>% 
   summarize(num_incidents = n(), 
             num_responses = sum(has_response==TRUE), 
             response_pct = mean(has_response==TRUE)*100) %>% 
   ungroup() %>% 
   arrange(desc(response_pct))
+
+# print the table and save it as an xtable
 response_stats
-  
+export_response_stats <- response_stats %>% 
+  rename('Incidents' = num_incidents,
+         'Public responses' = num_responses,
+         'Response %' = response_pct) %>% 
+  xtable()
+write_clip(print(export_response_stats, include.rownames = FALSE))
+
+
+
 
 # plot workspace
 ggplot(data=filter(us_targeted, !is.na(Response))) +

### 2019-08-29 article.tex
started literature review

@@
 On balance, states derive a benefit from preserving a system in which the consequences for espionage are minimized, so they refrain from imposing particularly harsh punishments for its practice; even when an individual incident is especially damaging to a state's security, if it falls within the traditionally accepted boundaries of an espionage operation, then it will likely go unpunished.
-In the context of the OPM hack, that means that US policymakers decided that they could not impose harsh penalties on the Chinese because it was more valuable to them that the US retain its ability to practice cyber-espionage with relative impunity in the future.
+In the case of the OPM hack, US policymakers decided that they could not impose harsh penalties on the Chinese because it was more valuable to them that the US retain its ability to practice cyber-espionage with relative impunity in the future.
 This basic calculation has remained consistent since the advent of the modern peacetime intelligence service, perhaps even the nation-state itself.
@@
 In both cases, the norms of traditional espionage ultimately prevailed, minimizing the diplomatic consequences for their use as long as it was in service of traditional intelligence goals.
-Through process tracing, I demonstrate how the justifications that policymakers employed for their use of these intelligence technologies---and the resulting minimization of the diplomatic penalties associated with them---are consistent with the theoretical underpinnings for why an accord tacitly permitting the use of espionage could be mutually strategic.
+Through process tracing, I demonstrate how the justifications that policymakers employed for their use of these intelligence technologies---and the resulting minimization of the diplomatic penalties associated with them---are consistent with the theoretical underpinnings for why a tacit accord permitting the use of espionage could be mutually strategic.
 
-\subsection{Literature base}
+\subsection{Existing literature}
+Before continuing, it is necessary to properly define what I mean by espionage norms.
+While there are a number of practices that make espionage unique in IR, the norms examined in this article are the ones that govern the diplomatic protocols for intelligence.
+What consequences does a state impose when it uncovers an espionage operation run by another state?
+Espionage norms dictate that the consequences remain minimal.
+Rarely does the response amount to more than a pat denunciation of the activity and implementation of the necessary countermeasures to end it.
+Most importantly, those consequences never rise to the level where they impose additional costs on the state that attempted the espionage in the first place; in the rare cases when costs \emph{are} imposed for a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.
+
 This article makes three main contributions to the academic literature on espionage.
-The first is that it advances a theory for why espionage continues to retain its liminal status in the international legal system, where it is disavowed but allowed.
-Public discourse on this subject often uses terms like \enquote{rules of the road} or \enquote{gentleman's agreement} to describe the understanding between governments that espionage should fall within certain parameters.
+The first is that it advances a novel theory for why espionage retains its often contradictory status in the legal system.
+In the absence of any significant treaties regarding espionage, scholars must develop their own theories about its place in customary international law.\footnote{Because it is universally agreed that international law does not deal with peacetime espionage directly, many journal articles that deal with this question frame their work almost Socratically, devoting two sections to cases for and against the legality of espionage.
+In fact, the first major work on law and peacetime espionage is a series of commentaries collected in a single publication that respond to each other's arguments.
+See \cite{wright_essays_1962}} Espionage is illegal everywhere, so how can every state acknowledge that allies and adversaries alike are guaranteed to engage in it?\footnote{Further complicating matters, espionage is the rare crime that might not be a crime depending on who the victim is.
+If an American government employee walks out of The Pentagon and hands classified documents to the KGB, they are at risk prosecuted as a traitor; if an American case officer hands classified KGB documents over to the CIA, they will be lauded as a hero.
+This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.} There is no legally consistent way to resolve that, so scholars have typically treated espionage as \emph{sui generis}---starting from the assumption that espionage norms exist in a class of their own, then attempting to fit a legal doctrine around that fact.
+Jared Beim's Comment determines that there while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so the prohibition on espionage merits further enforcement.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe.
+For more, see \cite{beim_enforcing_2018}} Beim acknowledges, however, that his conclusion will have little impact on the practice of espionage, where, at best, \enquote{it is perhaps possible to stop the more egregious violations.}\footcite[p.~672]{beim_enforcing_2018} A.
+John Radsan takes a more philosophical approach.
+Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{Beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage all come to the same conclusion: it would be almost impossible to argue that espionage is legal, but it will likely continue to be downplayed because the way it's been for decades.
+This literature cannot explain why espionage norms require this toleration.
+
+% espionage is an inherently unique category of IR, whose norms are endogenous to the insular intelligence community and therefore bear little relation to other legal or strategic considerations.
+
+Another branch of
+
+I break from this literature by arguing that far from being an irreplicable, historically contingent set of practices, espionage norms are actually preserved through an ongoing set of strategic decisions.
+Each time a new espionage technology is made available, policymakers are given a fresh chance to decide whether or not the norms of espionage out to extend to its use.
+Critically, the research will show that they were not obligated to do so by any pre-existing norms.
+When the US first introduced photographic satellites, for instance, it was not clear whether the Soviet Union would treat them as spies or as weapons---there were indications in both directions.
+Ultimately, the infrastructure of what we now call \enquote{spy satellites} was developed, and the norms of espionage prevailed.
+
+While not necessarily indicative of leading scholarship, it is revealing that popular discourse about espionage often uses terms like \enquote{rules of the game} or \enquote{gentleman's agreement} to describe the understanding between governments that espionage should fall within certain parameters.
+That these informal rules exist in some capacity is beyond a doubt.
+
+% espionage as deescalataory
+
+the second contribution is the role of espionage in deescalating conflict
+
+Because I argue that espionage norms influence cyber policy, it is worth mentioning scholarship on the role that espionage, outside of cyberspace, plays in the international system today.
+The intersection of espionage and international relations has been studied before---beginning, in the modern context, as far back as 1962---but usually in a historical context, not a political science one.
+The role of espionage in deescalating conflict is typically relegated to accounts of moments where Great People properly interpreted---or fatally misunderstood---the information that their intelligence operations provided them.
+Chapters 3 and 4 will primarily focus on decisions made by Eisenhower and Khrushchev, so to some extent I am guilty of that as well.
+Nonetheless, I will try to highlight the way in which both of those leaders set precedents designed to outlast them, across a variety of intelligence methods.
+I hope that this thesis starts the conversation about how espionage norms operate to deescalate conflict independent of the leaders that promote them.
+
 By employing both realist and constructivist schools of IR, this article develops rigorous explanations for why rational states, whether they be security-seeking and greedy, might cooperate to preserve this dynamic.
 
-Previous academic literature on espionage and diplomacy has treated espionage as \emph{sui generis}, a legal outlier whose norms are entirely endogenous to the insular intelligence community.
-
-Rather than explain the constituent practices of espionage as historically contingent,
-
-Typically those consequences are minimal, rarely amounting to more than a pat denunciation of the activity and implementation of the necessary countermeasures to end it.
-Most importantly, those consequences never rise to the level where they impose additional costs on the state that attempted the espionage in the first place; in the rare cases when costs \emph{are} imposed for a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.
-Espionage retains a special status within the international community: the activity itself is expressly forbidden, but every state tacitly acknowledges that allies and adversaries alike are guaranteed to engage in it.
-
-\section{Existing research on espionage}
-Previous scholarly work examines the curious liminal space in which espionage exists, but not the rationale that states have for perpetuating it.
-
-\section{Cyber-espionage as espionage}
+% incorporating cyber espionage into the intelligence literature, rather than the cyebr literature
+% espionage first, cyber second
 At first glance this observation might seem almost tautological.

### 2019-09-02 article.tex
first pass at literature review

@@
 For more, see \cite{council_on_foreign_relations_new_2019}} The routine minimization of espionage is most easily observable with the plethora of cyber-incidents that we deal with today, but it has its roots in norms that are much older.
-Therefore, to understand why policymakers treat cyber-espionage as \enquote{fair game,} one needs to understand why traditional espionage receives that treatment first.
+
+At first glance this observation might seem almost tautological.
+Of course cyber-espionage is treated like traditional espionage, one could argue---they're both espionage.
+That formulation begs the question; it doesn't explain how states determine what constitutes \enquote{traditional} espionage in cyberspace, what purpose that distinction servers, or whether it is prudent to continue making it.
+Therefore, to understand why the Director of the CIA treats cyber-espionage as \enquote{fair game,} one needs to understand why traditional espionage receives that treatment first.
 
 I posit that the universal tolerance of espionage in the international system can be explained through a rational calculation of each state's incentives.
-On balance, states derive a benefit from preserving a system in which the consequences for espionage are minimized, so they refrain from imposing particularly harsh punishments for its practice; even when an individual incident is especially damaging to a state's security, if it falls within the traditionally accepted boundaries of an espionage operation, then it will likely go unpunished.
+On balance, states derive a benefit from preserving a system in which the consequences for espionage are minimized, so they refrain from imposing particularly harsh punishments for its practice.
+Even when an individual incident is especially damaging to a state's security, if it falls within the traditionally accepted boundaries of an espionage operation, then it will likely go unpunished.
 In the case of the OPM hack, US policymakers decided that they could not impose harsh penalties on the Chinese because it was more valuable to them that the US retain its ability to practice cyber-espionage with relative impunity in the future.
@@
 Most importantly, those consequences never rise to the level where they impose additional costs on the state that attempted the espionage in the first place; in the rare cases when costs \emph{are} imposed for a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.
+In this article I present a theory for why those norms exist.
 
-This article makes three main contributions to the academic literature on espionage.
-The first is that it advances a novel theory for why espionage retains its often contradictory status in the legal system.
-In the absence of any significant treaties regarding espionage, scholars must develop their own theories about its place in customary international law.\footnote{Because it is universally agreed that international law does not deal with peacetime espionage directly, many journal articles that deal with this question frame their work almost Socratically, devoting two sections to cases for and against the legality of espionage.
+Prior literature has attempted to explain espionage norms through the grammar of international law, but in the absence of any significant treaties regarding espionage, scholars must work with in ambiguities.\footnote{Because it is universally agreed that international law does not deal with peacetime espionage directly, many journal articles that deal with this question frame their work almost Socratically, devoting two sections to cases for and against the legality of espionage.
 In fact, the first major work on law and peacetime espionage is a series of commentaries collected in a single publication that respond to each other's arguments.
@@
 This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.} There is no legally consistent way to resolve that, so scholars have typically treated espionage as \emph{sui generis}---starting from the assumption that espionage norms exist in a class of their own, then attempting to fit a legal doctrine around that fact.
-Jared Beim's Comment determines that there while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so the prohibition on espionage merits further enforcement.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe.
+Jared Beim determines that there while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so the prohibition on espionage merits further enforcement.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe.
 For more, see \cite{beim_enforcing_2018}} Beim acknowledges, however, that his conclusion will have little impact on the practice of espionage, where, at best, \enquote{it is perhaps possible to stop the more egregious violations.}\footcite[p.~672]{beim_enforcing_2018} A.
 John Radsan takes a more philosophical approach.
-Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{Beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage all come to the same conclusion: it would be almost impossible to argue that espionage is legal, but it will likely continue to be downplayed because the way it's been for decades.
-This literature cannot explain why espionage norms require this toleration.
+Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{Beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage all come to the same conclusion: it would be almost impossible to argue that espionage is legal, but it will likely continue to be downplayed.
+Beyond simple inertia, this literature cannot explain why espionage is tolerated.
 
@@
 
-Another branch of
+Another branch of scholarship sidesteps the legal question in favor of a \enquote{functional theory} of IR.
+This is closer to my work, as it attempts to explain \emph{why} states tolerate espionage rather than taking as a given that they will.
+A functional approach, for instance, might highlight the role of espionage in treaty verification, the way Christopher Baker does when he argues that \enquote{the advantages that espionage offers over legally-binding verification and assurance regimes tip the scales in favor of functional cooperation.}\footcite{baker_tolerance_2004} Baker correctly notes that the additional assurance provided by espionage enables states to enter into riskier agreements that they would otherwise have avoided.
+His analysis is limited, however, to arenas in which espionage may facilitate various forms of interstate cooperation.
+It doesn't resolve the gap between the absolute illegality of espionage in domestic law and its widespread practice internationally.
+Responding to Baker, Radsan notes that one-off deals and treaties \enquote{are a very slow and indirect way to achieve international consensus on the legality of espionage.
+The gap is still there.}\footcite[p.~607]{radsan_unresolved_2007} This article expands on Baker's basic premise---that there are situations in which espionage is mutually advantageous for both parties---and universalizes it to explain why the entire institution of espionage is mutually advantageous.
+Espionage norms are the glue that keep the institution in place, despite its legal ambiguity.
 
-I break from this literature by arguing that far from being an irreplicable, historically contingent set of practices, espionage norms are actually preserved through an ongoing set of strategic decisions.
+The key insight this article presents---and where it breaks from previous literature---is that far from being an irreplicable, historically contingent set of practices, espionage norms are actually preserved through an ongoing set of strategic decisions.
 Each time a new espionage technology is made available, policymakers are given a fresh chance to decide whether or not the norms of espionage out to extend to its use.
@@
 Ultimately, the infrastructure of what we now call \enquote{spy satellites} was developed, and the norms of espionage prevailed.
+
+% Because I argue that espionage norms influence cyber policy, it is worth mentioning scholarship on the role that espionage, outside of cyberspace, plays in the international system today.
+The intersection of espionage and international relations has been studied before---beginning, in the modern context, as far back as 1962---but usually in a historical context, not a political science one.
+The role of espionage in deescalating conflict is typically relegated to accounts of moments where Great People properly interpreted---or fatally misunderstood---the information that their intelligence operations provided them.
+Chapters 3 and 4 will primarily focus on decisions made by Eisenhower and Khrushchev, so to some extent I am guilty of that as well.
+Nonetheless, I will try to highlight the way in which both of those leaders set precedents designed to outlast them, across a variety of intelligence methods.
+I hope that this thesis starts the conversation about how espionage norms operate to deescalate conflict independent of the leaders that promote them.
 
@@
 
-the second contribution is the role of espionage in deescalating conflict
+% the second contribution is the role of espionage in deescalating conflict
 
-Because I argue that espionage norms influence cyber policy, it is worth mentioning scholarship on the role that espionage, outside of cyberspace, plays in the international system today.
-The intersection of espionage and international relations has been studied before---beginning, in the modern context, as far back as 1962---but usually in a historical context, not a political science one.
-The role of espionage in deescalating conflict is typically relegated to accounts of moments where Great People properly interpreted---or fatally misunderstood---the information that their intelligence operations provided them.
-Chapters 3 and 4 will primarily focus on decisions made by Eisenhower and Khrushchev, so to some extent I am guilty of that as well.
-Nonetheless, I will try to highlight the way in which both of those leaders set precedents designed to outlast them, across a variety of intelligence methods.
-I hope that this thesis starts the conversation about how espionage norms operate to deescalate conflict independent of the leaders that promote them.
-
-By employing both realist and constructivist schools of IR, this article develops rigorous explanations for why rational states, whether they be security-seeking and greedy, might cooperate to preserve this dynamic.
+% By employing both realist and constructivist schools of IR, this article develops rigorous explanations for why rational states, whether they be security-seeking and greedy, might cooperate to preserve this dynamic.
 
@@
 % espionage first, cyber second
-At first glance this observation might seem almost tautological.
-Of course cyber-espionage is treated like traditional espionage, one could argue---they're both espionage.
-That formulation begs the question; it doesn't explain how states determine what constitutes \enquote{traditional} espionage, what purpose that distinction servers, or whether it is prudent to continue making it.
 

### 2019-10-25 article.tex
dad edits and beginning of theory

@@
 
-The most damaging cyberattack committed against the United States federal government by a foreign agent was remarkably subtle.
+The most damaging cyberattack committed against the United States government by a foreign agent was remarkably subtle.
 Starting sometime in 2014, a small probe embedded itself deep within the network of the US Office of Personnel Management (OPM), an office of about 5,000 employees tasked with overseeing background checks, payroll, and other human resources concerns for the federal government.
@@
 
-By the time the breach was discovered in April 2015, the probe had accessed crucial government databases and transmitted the stolen information back to its country of origin---including a trove of applications for federal security clearance.
+By the time the breach was discovered in April 2015, the probe had accessed sensitive government databases and transmitted the stolen information back to its country of origin---including a trove of applications for federal security clearance.
 These applications exposed not just the personal data of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses.
-Federal authorities determined that the hack ultimately affected 22.1 million people in total, including personnel at the highest levels of government.\footcite{nakashima_hacks_2015} When OPM announced the full extent of the damage, it caused an uproar.
+Federal authorities determined that the hack ultimately affected 22.1 million people, including personnel at the highest levels of government.\footcite{nakashima_hacks_2015} When OPM announced the full extent of the damage, it caused an uproar.
 The American Federation of Government Employees, the largest federal workers union, filed a class-action lawsuit against OPM, seeking damages under the Privacy Act.\footcite{chalfant_court_2017} Senator Mark Warner (D-Va.) of the Senate Intelligence Committee called for OPM Director Katherine Archuleta's resignation.
@@
 Not only was the OPM hack a public relations disaster, it had significant national security implications as well.
-US officials feared that the stolen records could compromise intelligence efforts by exposing undercover CIA officers working in embassies around the world, whose names would be suspiciously absent from the OPM records.
+US officials feared that the stolen data could compromise intelligence efforts by exposing undercover CIA officers working in embassies around the world, whose names would be suspiciously absent from OPM records.
 The information could also be used to identify American officials who might be susceptible to pressure, or worse, recruitment.
@@
 In this case, the APT made identification easy by leaving their calling-card: the trojan opmsecurity.org domain was registered in the name of \enquote{Steve Rogers,} the Marvel comics character better known as Captain America.
-This digital taunt, along with email and IP addresses that verified the country of origin was not a false flag, all but guaranteed that the attack had come from an APT sponsored by the Chinese government, likely the Chinese military's cyber-espionage division.\footcite{koerner_inside_2016} Sitting on conclusive digital forensics, the Obama administration nonetheless refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018.
+This digital taunt, along with email and IP addresses that verified the country of origin was not a false flag, all but guaranteed that the attack had come from an APT sponsored by the Chinese government, likely the Chinese military's cyber-espionage division.\footcite{koerner_inside_2016} Though it possessed conclusive digital forensics, the Obama administration nonetheless refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018.
 See \cite{sanger_trump_2018}}
 
-The US government's tepid response to the OPM hack illustrates a universally understood but rarely acknowledged fact about the relationship between cybersecurity and espionage: when a cyber operation appears to fall within the bounds of a traditional intelligence operation, it will merit the same diplomatic consequences that a traditional intelligence operation would have.
-This justification was explicitly deployed by senior US policymakers, albeit often through unofficial moments of candor.
-A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].}\footcite[In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets.]{finklea_cyber_2015} For activities that fall under the umbrella of intelligence-gathering, counterintelligence, not prosecution, is deemed to be the appropriate response.
-Director of the CIA John Brennan described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, was especially honest: \enquote{You have to kind of salute the Chinese for what they did.
+The US government's tepid response to the OPM hack illustrates a crucial fact about the relationship between cybersecurity and espionage: when a cyberattack appears to fall within the bounds of a traditional intelligence operation, it will merit the diplomatic consequences of a traditional intelligence operation.
+US policymakers rarely acknowledge that this distinction exists---officially, all cyberattacks are illegal international behavior and will receive a strong response---but occasionally, in moments of candor, they will admit that cyber-espionage is different.
+A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].}\footnote{In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets.
+See \cite{finklea_cyber_2015}} For activities that fall under the umbrella of intelligence-gathering, counterintelligence, not prosecution, is deemed to be the appropriate response.
+Director of the CIA John Brennan described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, was particularly frank: \enquote{You have to kind of salute the Chinese for what they did.
 If we had the opportunity to do that, I don’t think we’d hesitate for a minute.}\footcite{pepitone_clapper_2015}
@@
 Of the 323 listed incidents, 38 were excluded from this calculation because they lacked an identifiable Victim, Sponsor, or Type.
-For more, see \cite{council_on_foreign_relations_new_2019}} The routine minimization of espionage is most easily observable with the plethora of cyber-incidents that we deal with today, but it has its roots in norms that are much older.
+For more, see \cite{council_on_foreign_relations_new_2019}} The routine minimization of espionage is most easily observed through the plethora of modern cyber-incidents, but it has its roots in norms that are much older.
 
@@
 Of course cyber-espionage is treated like traditional espionage, one could argue---they're both espionage.
-That formulation begs the question; it doesn't explain how states determine what constitutes \enquote{traditional} espionage in cyberspace, what purpose that distinction servers, or whether it is prudent to continue making it.
-Therefore, to understand why the Director of the CIA treats cyber-espionage as \enquote{fair game,} one needs to understand why traditional espionage receives that treatment first.
+That formulation, however, begs the question; it doesn't explain how states determine what constitutes \enquote{traditional} espionage in cyberspace, what purpose that distinction servers, or whether it is prudent to continue making it.
+Certainly the US should respond proportionally to attacks on its systems, but what if the damage done to national security by espionage is especially significant?
+``This is one of those cases where you have to ask, `Does the size of the operation change the nature of it?'\thinspace'' said one anonymous senior intelligence official of the OPM hack.
+``Clearly, it does.''\footcite{sanger_u.s._2015} As as far as the public knows, however, the size of operation did \emph{not} change the nature of the diplomatic response.
+Therefore, to understand why the Director of the CIA treats cyber-espionage as \enquote{fair game,} one first needs to understand why traditional espionage has historically that treatment as well.
 
-I posit that the universal tolerance of espionage in the international system can be explained through a rational calculation of each state's incentives.
+I posit that universal tolerance of espionage in the international system can be explained through a rational calculation of each state's incentives.
 On balance, states derive a benefit from preserving a system in which the consequences for espionage are minimized, so they refrain from imposing particularly harsh punishments for its practice.
-Even when an individual incident is especially damaging to a state's security, if it falls within the traditionally accepted boundaries of an espionage operation, then it will likely go unpunished.
-In the case of the OPM hack, US policymakers decided that they could not impose harsh penalties on the Chinese because it was more valuable to them that the US retain its ability to practice cyber-espionage with relative impunity in the future.
-This basic calculation has remained consistent since the advent of the modern peacetime intelligence service, perhaps even the nation-state itself.
+Even when an individual incident is especially damaging to a state's security, if it falls within the traditionally accepted boundaries of an espionage operation, it will likely go unpunished.
+In the case of the OPM hack, for example, US policymakers decided they could not impose harsh penalties on the Chinese because it was more valuable for the US to retain its ability to practice cyber-espionage with relative impunity in the future.
+This basic calculation---that espionage against one's own country is a tolerable price to pay for being allowed to practice espionage---has remained consistent since the advent of the modern peacetime intelligence service, perhaps even the nation-state itself.
+
+This raises a fascinating paradox: for espionage to be mutually permissible, both states must agree that it is so.
+But that requires \emph{both} states to determine they are getting more from practicing espionage against their adversary than their adversary is getting from practicing it against them.
+Otherwise, one would simply defect from that equilibrium and punish espionage at a level that imposes direct costs.
+Since this generally doesn't happen between major world powers, one of two things must be true: either a state is miscalculating the relative gains it gets from espionage, or the mutual tolerance of espionage has benefits that are sufficiently important to maintain even if one state gains less from it than its adversary.
+I argue that the latter is true.
 
 This article seeks to establish a theoretical framework for analyzing the role that espionage plays in international relations (IR) and the diplomatic norms that are associated with it.
-Through this framework, I explain what constitutes traditional espionage and why rational states might calculate that it is within their best interests to continue tolerating it.
+Through this framework, I explain what constitutes traditional espionage and why rational states might calculate that it is in their best interests to continue tolerating it.
 I then examine two cases in which new technical means of gathering intelligence were introduced during the Cold War: aerial photography (spy planes) and satellite photography (spy satellites).
 In both cases, the norms of traditional espionage ultimately prevailed, minimizing the diplomatic consequences for their use as long as it was in service of traditional intelligence goals.
-Through process tracing, I demonstrate how the justifications that policymakers employed for their use of these intelligence technologies---and the resulting minimization of the diplomatic penalties associated with them---are consistent with the theoretical underpinnings for why a tacit accord permitting the use of espionage could be mutually strategic.
+Through process tracing, I demonstrate how the justifications policymakers employed for the use of these intelligence technologies---and the resulting minimization of the diplomatic penalties associated with them---are consistent with the theoretical underpinnings for how a tacit accord permitting general espionage might be mutually strategic.
 
 \subsection{Existing literature}
+The benefits of espionage have never been underrated---\enquote{it is only the enlightened ruler and the wise general,} reads Sun Tzu, \enquote{who will use the highest intelligence of the army for purposes of spying and thereby they achieve great results}---but the inevitability of espionage has long been taken for granted.
+If you question that inevitably, the assertion that espionage is something \enquote{all counties do,} then the way that countries treat espionage becomes considerably more confusing.
+Until now, few, if any, scholars have analyzed why espionage norms exist and what strategic purpose they serve.
+
 Before continuing, it is necessary to properly define what I mean by espionage norms.
-While there are a number of practices that make espionage unique in IR, the norms examined in this article are the ones that govern the diplomatic protocols for intelligence.
-What consequences does a state impose when it uncovers an espionage operation run by another state?
-Espionage norms dictate that the consequences remain minimal.
-Rarely does the response amount to more than a pat denunciation of the activity and implementation of the necessary countermeasures to end it.
-Most importantly, those consequences never rise to the level where they impose additional costs on the state that attempted the espionage in the first place; in the rare cases when costs \emph{are} imposed for a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.
+Simply put, espionage norms dictate what constitutes espionage and how states are expected to respond when it is conducted against them.\footnote{There are, of course, other \enquote{norms} that involve espionage.
+The so-called \enquote{Moscow Rules} dictate how an agent or case officer is supposed to conduct themselves in hostile territory.
+In order to avoid constantly specifying, espionage norms are more strictly defined to include only the ones that affect IR.} What consequences does a state impose when it uncovers an espionage operation run by another state?
+Espionage norms prescribe that the consequences remain minimal.
+Rarely does the response amount to more than a pat denunciation of the activity and implementation of the necessary countermeasures to end it---the consequences for espionage are borne by the agents who commit it, not the state that sent them.
+In the rare cases when additional costs \emph{are} imposed on a state for conducting a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.
 In this article I present a theory for why those norms exist.
 
-Prior literature has attempted to explain espionage norms through the grammar of international law, but in the absence of any significant treaties regarding espionage, scholars must work with in ambiguities.\footnote{Because it is universally agreed that international law does not deal with peacetime espionage directly, many journal articles that deal with this question frame their work almost Socratically, devoting two sections to cases for and against the legality of espionage.
+Prior literature has attempted to explain espionage norms using international law, but in the absence of any significant treaties governing peacetime espionage, scholars are often left without definitive answers.\footnote{Because it is universally agreed that international law does not deal with peacetime espionage directly, many journal articles that deal with this question frame their work almost Socratically, devoting two sections to cases for and against the legality of espionage.
 In fact, the first major work on law and peacetime espionage is a series of commentaries collected in a single publication that respond to each other's arguments.
-See \cite{wright_essays_1962}} Espionage is illegal everywhere, so how can every state acknowledge that allies and adversaries alike are guaranteed to engage in it?\footnote{Further complicating matters, espionage is the rare crime that might not be a crime depending on who the victim is.
+See \cite{wright_essays_1962}} Since espionage is illegal everywhere, how can every state acknowledge that allies and adversaries alike are guaranteed to engage in it?\footnote{Further complicating matters, espionage is the rare crime that might not be a crime depending on who the victim is.
 If an American government employee walks out of The Pentagon and hands classified documents to the KGB, they are at risk prosecuted as a traitor; if an American case officer hands classified KGB documents over to the CIA, they will be lauded as a hero.
-This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.} There is no legally consistent way to resolve that, so scholars have typically treated espionage as \emph{sui generis}---starting from the assumption that espionage norms exist in a class of their own, then attempting to fit a legal doctrine around that fact.
+This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.} There is no legally consistent way to resolve that paradox, so scholars have typically treated espionage as \emph{sui generis}---starting from the assumption that espionage norms exist in a class of their own, then attempting to fit a legal doctrine around that fact.
 Jared Beim determines that there while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so the prohibition on espionage merits further enforcement.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe.
 For more, see \cite{beim_enforcing_2018}} Beim acknowledges, however, that his conclusion will have little impact on the practice of espionage, where, at best, \enquote{it is perhaps possible to stop the more egregious violations.}\footcite[p.~672]{beim_enforcing_2018} A.
-John Radsan takes a more philosophical approach.
-Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{Beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage all come to the same conclusion: it would be almost impossible to argue that espionage is legal, but it will likely continue to be downplayed.
-Beyond simple inertia, this literature cannot explain why espionage is tolerated.
-
-% espionage is an inherently unique category of IR, whose norms are endogenous to the insular intelligence community and therefore bear little relation to other legal or strategic considerations.
+John Radsan employs a more philosophical approach.
+Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage all reach the same conclusion: it would be almost impossible to argue that espionage is legal, but it will likely continue to be downplayed.
+Beyond simple inertia, however, this literature cannot explain why espionage is tolerated.
 
 Another branch of scholarship sidesteps the legal question in favor of a \enquote{functional theory} of IR.
-This is closer to my work, as it attempts to explain \emph{why} states tolerate espionage rather than taking as a given that they will.
+This is closer to my work, as it attempts to explain \emph{why} states tolerate espionage rather than creating \emph{post hoc} justifications for it.
 A functional approach, for instance, might highlight the role of espionage in treaty verification, the way Christopher Baker does when he argues that \enquote{the advantages that espionage offers over legally-binding verification and assurance regimes tip the scales in favor of functional cooperation.}\footcite{baker_tolerance_2004} Baker correctly notes that the additional assurance provided by espionage enables states to enter into riskier agreements that they would otherwise have avoided.
@@
 The gap is still there.}\footcite[p.~607]{radsan_unresolved_2007} This article expands on Baker's basic premise---that there are situations in which espionage is mutually advantageous for both parties---and universalizes it to explain why the entire institution of espionage is mutually advantageous.
-Espionage norms are the glue that keep the institution in place, despite its legal ambiguity.
+Espionage norms provide the glue that keep the institution in place, despite its legal ambiguity.
 
 The key insight this article presents---and where it breaks from previous literature---is that far from being an irreplicable, historically contingent set of practices, espionage norms are actually preserved through an ongoing set of strategic decisions.
-Each time a new espionage technology is made available, policymakers are given a fresh chance to decide whether or not the norms of espionage out to extend to its use.
-Critically, the research will show that they were not obligated to do so by any pre-existing norms.
-When the US first introduced photographic satellites, for instance, it was not clear whether the Soviet Union would treat them as spies or as weapons---there were indications in both directions.
-Ultimately, the infrastructure of what we now call \enquote{spy satellites} was developed, and the norms of espionage prevailed.
+Each time a new espionage technology is made available, policymakers are given a fresh chance to decide whether or not the norms of espionage ought to extend to its use.
+Critically, the research that follows will show that they were not obligated to do so by any pre-existing norms.
+When the US first introduced photographic satellites, for example, it was not clear whether the Soviet Union would treat the satellites as spies or as weapons---there were indications in both directions.
+Ultimately, the infrastructure of what we now call \enquote{spy satellites} was developed, and the norms of espionage prevailed to protect their use.
 
@@
 
-While not necessarily indicative of leading scholarship, it is revealing that popular discourse about espionage often uses terms like \enquote{rules of the game} or \enquote{gentleman's agreement} to describe the understanding between governments that espionage should fall within certain parameters.
-That these informal rules exist in some capacity is beyond a doubt.
-
-% espionage as deescalataory
-
-% the second contribution is the role of espionage in deescalating conflict
-
 % By employing both realist and constructivist schools of IR, this article develops rigorous explanations for why rational states, whether they be security-seeking and greedy, might cooperate to preserve this dynamic.
@@
 \section{Theory}
+One of the most prevalent clich\'es in spy literature and journalism is the so-called \enquote{rules of the game.} Media-friendly lists of rules like \enquote{You are never alone} and \enquote{Don't trust anyone} make the spy world seem insular, mysterious, and impenetrable.\footcite{myre_moscow_2019} They hint at a larger set of processes whose motivations the average citizen, not given access to this privileged information, could not possibly begin to understand.
+In reality, while the \emph{operations} of a modern intelligence service are incredibly opaque, the geopolitical conditions in which it operates are not.
+Every aspect of a nation-state's interactions with another ought to be subject to the same critical analysis.
+Scholars can use rationalist principles to analyze everything from trade policy to the space race; espionage, which by definition involves deeply invasive interactions with another state, ought to be no different.
 
-% In recent years, cyberattacks perpetrated by state-sponsored foreign agents have received ever-increasing levels of public scrutiny---which makes the government's response to the OPM hack especially puzzling.
-Why did an incident of this magnitude not merit an official response?
-
-% This article seeks to explain why espionage efforts are not more actively discouraged by nations, even as the results of intelligence operations can oftentimes have a devastating effect on the targeted nation's security.
-
-% Russian interference in the 2016 US Presidential election prompted the expulsion of 37 diplomats and series of follow-on sanctions
+Espionage certainly seems like something that states should be doing more to discourage.
+Then results of intelligence operations can have devastating, even existential effects on the targeted nation's security.
+One of the most traditional form of espionage is stealing military secrets from a foreign power, and in that arena China has been relentless.
+A US Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} ``Long-term, US future military advantage is being diminished by years of IP exfiltration from the DoD, DoN, and DIB,'' the report read, ``all with little to no adverse consequences to the thieves.''\footcite[p.~6]{bayer_cybersecurity_2019} The Navy's forecasts were dire: ``If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.''\footcite[p.~5]{bayer_cybersecurity_2019} The Navy is not a disinterested party, of course, and there are reasons to take its conclusions with a grain of salt; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see \cite{gilli_why_2019}} Nonetheless, the possibilities are alarming and the consequences for remain minimal.
 

### 2019-10-25 data/incidents-scripts.R

@@ -4,6 +4,7 @@ library(tidyverse)
 library(ggthemes)
 library(xtable)
 library(clipr)
+library(lubridate)
 
 # comment this line if your file is elsewhere
 setwd('~/Documents/journal/src/data/')
@@ -28,12 +29,15 @@ us_targeted <- cyber_incidents %>%
 # swap in us_targeted for for US-only count
 response_stats <- cyber_incidents %>% 
   filter(Type !='Undefined' & !is.na(Sponsor) & !is.na(Victims)) %>% 
-  group_by(Type) %>% 
+  filter(Date > as.Date('2010-01-01')) %>% 
+  # group_by(Type) %>% 
   summarize(num_incidents = n(), 
             num_responses = sum(has_response==TRUE), 
             response_pct = mean(has_response==TRUE)*100) %>% 
   ungroup() %>% 
   arrange(desc(response_pct))
+
+response_stats
 
 # print the table and save it as an xtable
 response_stats

### 2019-10-30 article.tex
theory basics and flow update

@@
 \section{Theory}
+\subsection{Establishing the puzzle}
 One of the most prevalent clich\'es in spy literature and journalism is the so-called \enquote{rules of the game.} Media-friendly lists of rules like \enquote{You are never alone} and \enquote{Don't trust anyone} make the spy world seem insular, mysterious, and impenetrable.\footcite{myre_moscow_2019} They hint at a larger set of processes whose motivations the average citizen, not given access to this privileged information, could not possibly begin to understand.
-In reality, while the \emph{operations} of a modern intelligence service are incredibly opaque, the geopolitical conditions in which it operates are not.
+In reality, while the \emph{operations} of a modern intelligence service are incredibly opaque, the geopolitical conditions in which one operates are not.
 Every aspect of a nation-state's interactions with another ought to be subject to the same critical analysis.
@@
 
-Espionage certainly seems like something that states should be doing more to discourage.
+% Certainly there are factors that the go into diplomatic decisions that the general public will likely never know.
+There are few reasons to believe, however, that this is more true for intelligence-related diplomatic matters than for generic ones.
+Every diplomatic decision is made with the backing of classified intelligence.
+
+Under traditional methods of analysis, espionage certainly seems like something that states should be doing more to discourage.
 Then results of intelligence operations can have devastating, even existential effects on the targeted nation's security.
@@
 A US Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} ``Long-term, US future military advantage is being diminished by years of IP exfiltration from the DoD, DoN, and DIB,'' the report read, ``all with little to no adverse consequences to the thieves.''\footcite[p.~6]{bayer_cybersecurity_2019} The Navy's forecasts were dire: ``If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.''\footcite[p.~5]{bayer_cybersecurity_2019} The Navy is not a disinterested party, of course, and there are reasons to take its conclusions with a grain of salt; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see \cite{gilli_why_2019}} Nonetheless, the possibilities are alarming and the consequences for remain minimal.
+
+Not only does espionage have potentially disastrous consequences but the system is set up so that the structure of civilian intelligence services is never fundamentally threatened.
+The low percentage of responses for cyber-espionage have their basis in a similarly low response rate for traditional espionage.
+The traditional punishment for espionage is to declare certain embassy officials \emph{persona non grata}, which often results in ``tit-for-tat'' diplomatic expulsions.
+The name is taken from a highly-successful strategy for the prisoner's dilemma wherein the player simply mimics the choice that that their opponent made in the last round.
+By employing the tit-for-tat strategy, the player never gets taken advantage of for more than one turn and can always return to cooperation if their opponent unilaterally signals a willingness to do the same.
+
+Diplomatic expulsions play out in exactly the same fashion: when one country expels the diplomats of another, the other country will often respond in kind.
+Each side will continue to expel each other's diplomats until one side signals that it would like for the reciprocal PNG-ing of diplomats to stop.
+Then, both sides return to a cooperative state, each preserving some of their intelligence operations.
+Neither side can really gain a decisive advantage, so the back-and-forth punishment never ends in a mutual defection.
+It might seem obvious to note, but \emph{both} states must have decided that preserving this system is valuable for them, otherwise just one of them refusing to cooperate would result in the mutual devastation of human intelligence operations.
+
+\subsection{Espionage norms as a discrete series of choices}
+Recall: the key feature of espionage norms is that the scope of the punishment is structurally incapable of discouraging a future, identical operation.
+Expelling suspected intelligence officers from an embassy temporarily makes it more difficult to conduct espionage, but it doesn't make a state less likely to try; the worst that can happen is a net-neutral result---going back to \enquote{square one.} By contrast, were it more common to respond to espionage by imposing financial sanctions or even inflicting a military cost, a state would have to consider the possibility that attempting espionage could adversely impact other aspects of their security, leaving it worse off than when it started.
+Such a response would be a violation of the well-established norms of espionage and it is essentially unheard-of.
+
+Therefore, I propose that an iterated model can describe how states cooperate to preserve the liminal legal status of espionage.
+My model changes the parameters from above so that the decision to cooperate or defect is based not in the amount of punishment received (essentially measured by the number of diplomats expelled), but rather the type.
+Each time a state is forced to respond to a successful intelligence operation by a rival---such as the reveal of a high-level spy or a major data breach---it can either limit its response to that which is traditionally associated with espionage, or it can impose unprecedented costs that might discourage such an operation in the future.
+Table \ref{espionage-matrix} describes a single instance of this choice.
+States signal their interest in cooperating by limiting their retaliation in response to espionage, no matter how damaging the operation is to their national security.
+Each instance is a chance to play the game with fresh information; each espionage case is an opportunity to cooperate, reaffirming the norm that states always respond to espionage in a specific way, or defect, adding an additional cost to the attempt.
+
+% https://tex.stackexchange.com/questions/249480/creating-a-payoff-matrix-using-latex-tabular-environment
+\begin{table}[ht]
+\centering
+\setlength{\extrarowheight}{2pt}
+\small
+\begin{tabular}{cc|c|c|}
+  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{USSR}\\
+  & \multicolumn{1}{c}{} & \multicolumn{1}{c}{Tolerate Espionage}  & \multicolumn{1}{c}{Punish Espionage} \\\cline{3-4}
+  \multirow{3}*{USA}  & Tolerate Espionage & \makecell{~\\Both sides gain intelligence \\~} & \makecell{~\\ Only the USSR continues spying \\ ~} \\\cline{3-4}
+  & Punish Espionage & \makecell{~\\ Only the US continues spying \\~} & \makecell{~\\ No intelligence either direction \\~} \\\cline{3-4}
+\end{tabular}
+\caption{Espionage consequences decision matrix}
+\label{espionage-matrix}
+\end{table}
+
+The purpose of analyzing espionage norms this way is to remove the mystique.
+
+Even though the US and the USSR have the opportunity to defect by imposing harsher punishments in response to espionage, the gains would be short-lived, because the other state is guaranteed to immediately do the same---tit-for-tat.
+Espionage consequences are an iterated game that both parties are playing the exactly the same way, so the only two possible outcomes long-term are mutual defection or mutual cooperation.
+
+Given those two options, it makes intuitive sense that in the case of Cold War espionage both the US and the USSR chose to cooperate.
+Both sides gained a great deal from their intelligence operations; Soviet intelligence had an easier time operating in Washington than the US did in the USSR, but American intelligence was still formidable and Soviet defections were frequent.\footcite[p.~380]{macrakis_technophilic_2010} When the relative capabilities are not too lopsided, there is a pleasing reciprocity to espionage that makes cooperation easier.
 

### 2019-11-08 article.tex
starting glazer theory

@@
 Then results of intelligence operations can have devastating, even existential effects on the targeted nation's security.
-One of the most traditional form of espionage is stealing military secrets from a foreign power, and in that arena China has been relentless.
-A US Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} ``Long-term, US future military advantage is being diminished by years of IP exfiltration from the DoD, DoN, and DIB,'' the report read, ``all with little to no adverse consequences to the thieves.''\footcite[p.~6]{bayer_cybersecurity_2019} The Navy's forecasts were dire: ``If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.''\footcite[p.~5]{bayer_cybersecurity_2019} The Navy is not a disinterested party, of course, and there are reasons to take its conclusions with a grain of salt; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see \cite{gilli_why_2019}} Nonetheless, the possibilities are alarming and the consequences for remain minimal.
+Arguably the most traditional form of espionage is stealing military secrets from a foreign power, and in that arena China has been relentless.
+A US Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} ``Long-term, US future military advantage is being diminished by years of IP exfiltration from the DoD, DoN, and DIB,'' the report read, ``all with little to no adverse consequences to the thieves.''\footcite[p.~6]{bayer_cybersecurity_2019} The Navy's forecasts were dire: ``If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.''\footcite[p.~5]{bayer_cybersecurity_2019} The Navy is not a disinterested party, of course, and there are reasons to take its conclusions with a grain of salt; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see \cite{gilli_why_2019}} Nonetheless, the opportunities for harming national security are alarming and the consequences for doing so remain minimal.
 
-Not only does espionage have potentially disastrous consequences but the system is set up so that the structure of civilian intelligence services is never fundamentally threatened.
-The low percentage of responses for cyber-espionage have their basis in a similarly low response rate for traditional espionage.
+Not only does espionage have potentially disastrous consequences but steps that might actually curb it are out of the question.
+The system is set up so that the structure of civilian intelligence services is never fundamentally threatened; the minimal responses for cyber-espionage have their roots in the similarly minimal responses for classical espionage.
 The traditional punishment for espionage is to declare certain embassy officials \emph{persona non grata}, which often results in ``tit-for-tat'' diplomatic expulsions.
@@
 
-Therefore, I propose that an iterated model can describe how states cooperate to preserve the liminal legal status of espionage.
-My model changes the parameters from above so that the decision to cooperate or defect is based not in the amount of punishment received (essentially measured by the number of diplomats expelled), but rather the type.
-Each time a state is forced to respond to a successful intelligence operation by a rival---such as the reveal of a high-level spy or a major data breach---it can either limit its response to that which is traditionally associated with espionage, or it can impose unprecedented costs that might discourage such an operation in the future.
-Table \ref{espionage-matrix} describes a single instance of this choice.
-States signal their interest in cooperating by limiting their retaliation in response to espionage, no matter how damaging the operation is to their national security.
-Each instance is a chance to play the game with fresh information; each espionage case is an opportunity to cooperate, reaffirming the norm that states always respond to espionage in a specific way, or defect, adding an additional cost to the attempt.
-
 % https://tex.stackexchange.com/questions/249480/creating-a-payoff-matrix-using-latex-tabular-environment
@@
 \begin{tabular}{cc|c|c|}
-  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{USSR}\\
+  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{Country 2}\\
   & \multicolumn{1}{c}{} & \multicolumn{1}{c}{Tolerate Espionage}  & \multicolumn{1}{c}{Punish Espionage} \\\cline{3-4}
-  \multirow{3}*{USA}  & Tolerate Espionage & \makecell{~\\Both sides gain intelligence \\~} & \makecell{~\\ Only the USSR continues spying \\ ~} \\\cline{3-4}
+  \multirow{3}*{Country 1}  & Tolerate Espionage & \makecell{~\\Both sides gain intelligence \\~} & \makecell{~\\ Only the USSR continues spying \\ ~} \\\cline{3-4}
   & Punish Espionage & \makecell{~\\ Only the US continues spying \\~} & \makecell{~\\ No intelligence either direction \\~} \\\cline{3-4}
@@
 
-The purpose of analyzing espionage norms this way is to remove the mystique.
+Therefore, I propose an iterated model to describe how states cooperate to preserve the liminal legal status of espionage.
+My model changes the parameters so that the decision to cooperate or defect is based not in the amount of punishment received, but rather the type.
+Each time a state is forced to respond to a successful intelligence operation by a rival---such as the reveal of a high-level spy or a major data breach---it can either limit its response to one which is traditionally associated with espionage, or it can impose unprecedented costs that might discourage such an operation in the future.
+Table \ref{espionage-matrix} describes a single instance of this choice.
+States signal their interest in cooperating by limiting their retaliation in response to espionage, no matter how damaging the operation is to their national security.
+Each instance is a chance to play the game with fresh information; each espionage case is an opportunity to cooperate, reaffirming the norm that states always respond to espionage in a specific way, or defect, adding an additional cost to the attempt.
 
-Even though the US and the USSR have the opportunity to defect by imposing harsher punishments in response to espionage, the gains would be short-lived, because the other state is guaranteed to immediately do the same---tit-for-tat.
-Espionage consequences are an iterated game that both parties are playing the exactly the same way, so the only two possible outcomes long-term are mutual defection or mutual cooperation.
+Through this model two important things become clear.
+First, even though either country has the opportunity to defect by imposing harsher punishments in response to espionage, its gains would be short-lived, because the other country is guaranteed to immediately do the same---tit-for-tat.
+Espionage consequences are an iterated game that both parties are playing the exact same way, so, in the long run, the only two possible outcomes are mutual defection or mutual cooperation.
+In the realm of traditional espionage, we've only ever seen mutual cooperation, but the key observation is that any country has the power to change that by defecting.
+Even if the defector's rival would have preferred to maintain the status quo, the rival has no option other than to defect as well, or else be left at a senseless disadvantage.
+The relatively open flow of information is only possible with a universal buy-in.
 
-Given those two options, it makes intuitive sense that in the case of Cold War espionage both the US and the USSR chose to cooperate.
+Second, while there is unquestionably a cultural norm informing the response to intelligence operations, that norm is doing a lot of work to maintain the status quo.
+So much work, in fact, that tradition could not possibly explain the entire phenomenon.
+Stripped of its mystique, the benefits of allowing intelligence operations to continue receiving light treatment are, like any other state decision, a simple cost-benefit analysis.
+And therefore, regardless of its correctness or actual explanatory power, there must be \emph{some} rational explanation that political leaders employ when they decide that preserving their ability to spy is worth suffering through the inevitable leaks that result from espionage conducted against them.
+
+\subsection{Explaining norms cooperation}
+
+% Given those two options, it makes intuitive sense that in the case of Cold War espionage both the US and the USSR chose to cooperate.
 Both sides gained a great deal from their intelligence operations; Soviet intelligence had an easier time operating in Washington than the US did in the USSR, but American intelligence was still formidable and Soviet defections were frequent.\footcite[p.~380]{macrakis_technophilic_2010} When the relative capabilities are not too lopsided, there is a pleasing reciprocity to espionage that makes cooperation easier.

### 2019-11-17 article.tex
wohlforth introduction edits and self theory edits

--- 2019-10-25 article.tex
@@
 \maketitle
-% \newpage
 
@@
 The information could also be used to identify American officials who might be susceptible to pressure, or worse, recruitment.
-FBI Director James Comey personally took questions about the incident.
-``If you have my [application for security clearance],'' Comey said, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.
-Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015} With evidence that a foreign intelligence service had successfully obtained exactly that, one might expect that the United States would respond with a series of diplomatic measures intended to signal that highly invasive acts of espionage would not be tolerated from its international adversaries.
+The backlash was so intense that FBI Director James Comey was forced to take questions about the incident.
+``If you have my [application for security clearance],'' Comey admitted, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.
+Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015} With evidence that a foreign intelligence service had successfully obtained exactly that, one might expect the United States to respond with a series of diplomatic measures intended to signal that highly invasive acts of espionage would not be tolerated from its international adversaries.
 
@@
 The US government's tepid response to the OPM hack illustrates a crucial fact about the relationship between cybersecurity and espionage: when a cyberattack appears to fall within the bounds of a traditional intelligence operation, it will merit the diplomatic consequences of a traditional intelligence operation.
-US policymakers rarely acknowledge that this distinction exists---officially, all cyberattacks are illegal international behavior and will receive a strong response---but occasionally, in moments of candor, they will admit that cyber-espionage is different.
+US policymakers rarely acknowledge that this distinction exists---officially, cyberattacks are illegal international behavior and will receive a strong response---but occasionally, in moments of candor, policymakers will admit that cyber-espionage is different.
 A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].}\footnote{In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets.
@@
 \end{tabular}
-\caption{Cyber-incident response rates by type of incident}
+\caption{Global cyber-incident response rates}
 \label{response-pct}
@@
 In fact, cyber-espionage is routinely treated as a lesser category of offense by all nations, not just the United States.
-Of the hundreds of state-sponsored cyberattacks about which there is public knowledge, attacks classified as espionage are by the far most common---and the least likely to garner an official response from the victim government (Table \ref{response-pct}).\footnote{These numbers were compiled from the Council on Foreign Relations' Cyber Operations Tracker, which is a publicly-sourced record of state-sponsored cyberattacks, updated quarterly.
+Of the hundreds of state-sponsored cyberattacks known to the public, attacks classified as espionage are by the far most common---and the least likely to garner an official response from the victim government (Table \ref{response-pct}).\footnote{These numbers were compiled from the Council on Foreign Relations' Cyber Operations Tracker, which is a publicly-sourced record of state-sponsored cyberattacks, updated quarterly.
 Of the 323 listed incidents, 38 were excluded from this calculation because they lacked an identifiable Victim, Sponsor, or Type.
@@
 ``Clearly, it does.''\footcite{sanger_u.s._2015} As as far as the public knows, however, the size of operation did \emph{not} change the nature of the diplomatic response.
-Therefore, to understand why the Director of the CIA treats cyber-espionage as \enquote{fair game,} one first needs to understand why traditional espionage has historically that treatment as well.
+Therefore, to understand why the Director of the CIA treats cyber-espionage as \enquote{fair game,} one first needs to understand why traditional espionage has historically received that treatment as well.
 
@@
 
-This raises a fascinating paradox: for espionage to be mutually permissible, both states must agree that it is so.
-But that requires \emph{both} states to determine they are getting more from practicing espionage against their adversary than their adversary is getting from practicing it against them.
-Otherwise, one would simply defect from that equilibrium and punish espionage at a level that imposes direct costs.
+This hypothesis raises a fascinating paradox: for espionage to be mutually permissible, both states must agree to treat it that way, but, in theory, that requires \emph{both} states to determine they are getting more from practicing espionage against their adversary than their adversary is getting from practicing it against them.
+Otherwise, the state with less to gain would simply defect from that equilibrium and punish espionage at a level that imposes direct costs.
 Since this generally doesn't happen between major world powers, one of two things must be true: either a state is miscalculating the relative gains it gets from espionage, or the mutual tolerance of espionage has benefits that are sufficiently important to maintain even if one state gains less from it than its adversary.
-I argue that the latter is true.
+I argue the latter.
 
@@
 The so-called \enquote{Moscow Rules} dictate how an agent or case officer is supposed to conduct themselves in hostile territory.
-In order to avoid constantly specifying, espionage norms are more strictly defined to include only the ones that affect IR.} What consequences does a state impose when it uncovers an espionage operation run by another state?
+In order to avoid constantly specifying, espionage norms are more strictly defined to include only the ones that affect IR.} What consequences does a state impose when it uncovers an intelligence operation run by another state?
 Espionage norms prescribe that the consequences remain minimal.
@@
 In fact, the first major work on law and peacetime espionage is a series of commentaries collected in a single publication that respond to each other's arguments.
-See \cite{wright_essays_1962}} Since espionage is illegal everywhere, how can every state acknowledge that allies and adversaries alike are guaranteed to engage in it?\footnote{Further complicating matters, espionage is the rare crime that might not be a crime depending on who the victim is.
+See \cite{wright_essays_1962}} Since espionage is universally illegal, how can every state acknowledge that allies and adversaries alike are guaranteed to engage in it?\footnote{Further complicating matters, espionage is the rare crime that might not be a crime depending on who the victim is.
 If an American government employee walks out of The Pentagon and hands classified documents to the KGB, they are at risk prosecuted as a traitor; if an American case officer hands classified KGB documents over to the CIA, they will be lauded as a hero.
-This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.} There is no legally consistent way to resolve that paradox, so scholars have typically treated espionage as \emph{sui generis}---starting from the assumption that espionage norms exist in a class of their own, then attempting to fit a legal doctrine around that fact.
-Jared Beim determines that there while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so the prohibition on espionage merits further enforcement.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe.
+This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.} As a result, scholars have typically treated espionage as \emph{sui generis}---starting from the assumption that espionage norms exist in a class of their own, then attempting to fit a legal doctrine around that fact.
+Jared Beim determines that while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so the prohibition on espionage merits further enforcement.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe.
 For more, see \cite{beim_enforcing_2018}} Beim acknowledges, however, that his conclusion will have little impact on the practice of espionage, where, at best, \enquote{it is perhaps possible to stop the more egregious violations.}\footcite[p.~672]{beim_enforcing_2018} A.
 John Radsan employs a more philosophical approach.
-Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage all reach the same conclusion: it would be almost impossible to argue that espionage is legal, but it will likely continue to be downplayed.
-Beyond simple inertia, however, this literature cannot explain why espionage is tolerated.
+Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage all reach the same conclusion: it would be almost impossible to argue that espionage is legal, but no one really seems to care.
+Taken to its logical endpoint, this branch of scholarship would simply argue that the espionage norms we observe today result from simple inertia.
 
 Another branch of scholarship sidesteps the legal question in favor of a \enquote{functional theory} of IR.
-This is closer to my work, as it attempts to explain \emph{why} states tolerate espionage rather than creating \emph{post hoc} justifications for it.
-A functional approach, for instance, might highlight the role of espionage in treaty verification, the way Christopher Baker does when he argues that \enquote{the advantages that espionage offers over legally-binding verification and assurance regimes tip the scales in favor of functional cooperation.}\footcite{baker_tolerance_2004} Baker correctly notes that the additional assurance provided by espionage enables states to enter into riskier agreements that they would otherwise have avoided.
-His analysis is limited, however, to arenas in which espionage may facilitate various forms of interstate cooperation.
+This approach is more informative because it attempts to explain why states tolerate espionage rather than creating \emph{post hoc} justifications for it.
+A functional approach, for instance, might highlight the role of espionage in treaty verification, the way Christopher Baker does when he argues \enquote{the advantages that espionage offers over legally-binding verification and assurance regimes tip the scales in favor of functional cooperation.}\footcite{baker_tolerance_2004} Baker correctly notes that the additional assurance provided by espionage enables states to enter into riskier agreements that they would otherwise have avoided.
+His analysis is limited, however, by its focus on enabling specific policy outcomes, like treaties.
 It doesn't resolve the gap between the absolute illegality of espionage in domestic law and its widespread practice internationally.
@@
 The gap is still there.}\footcite[p.~607]{radsan_unresolved_2007} This article expands on Baker's basic premise---that there are situations in which espionage is mutually advantageous for both parties---and universalizes it to explain why the entire institution of espionage is mutually advantageous.
-Espionage norms provide the glue that keep the institution in place, despite its legal ambiguity.
 
@@
 \section{Theory}
-One of the most prevalent clich\'es in spy literature and journalism is the so-called \enquote{rules of the game.} Media-friendly lists of rules like \enquote{You are never alone} and \enquote{Don't trust anyone} make the spy world seem insular, mysterious, and impenetrable.\footcite{myre_moscow_2019} They hint at a larger set of processes whose motivations the average citizen, not given access to this privileged information, could not possibly begin to understand.
-In reality, while the \emph{operations} of a modern intelligence service are incredibly opaque, the geopolitical conditions in which it operates are not.
+\subsection{Establishing the puzzle}
+One of the most prevalent clich\'es in spy literature and journalism is the so-called \enquote{rules of the game.} The supposed rules of the spy world world---\enquote{You are never alone}, \enquote{Don't trust anyone}---make the spy world seem insular, mysterious, and impenetrable.\footcite{myre_moscow_2019} They hint at a larger set of processes whose motivations the average citizen, not given access to this privileged information, could not possibly begin to understand.
+In reality, while the \emph{operation} of a modern intelligence service is incredibly opaque, the geopolitical situation in which one operates is not.
 Every aspect of a nation-state's interactions with another ought to be subject to the same critical analysis.
@@
 
-Espionage certainly seems like something that states should be doing more to discourage.
-Then results of intelligence operations can have devastating, even existential effects on the targeted nation's security.
-One of the most traditional form of espionage is stealing military secrets from a foreign power, and in that arena China has been relentless.
-A US Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} ``Long-term, US future military advantage is being diminished by years of IP exfiltration from the DoD, DoN, and DIB,'' the report read, ``all with little to no adverse consequences to the thieves.''\footcite[p.~6]{bayer_cybersecurity_2019} The Navy's forecasts were dire: ``If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.''\footcite[p.~5]{bayer_cybersecurity_2019} The Navy is not a disinterested party, of course, and there are reasons to take its conclusions with a grain of salt; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see \cite{gilli_why_2019}} Nonetheless, the possibilities are alarming and the consequences for remain minimal.
+% Certainly there are factors that the go into diplomatic decisions that the general public will likely never know.
+There are few reasons to believe, however, that this is more true for intelligence-related diplomatic matters than for generic ones.
+Every diplomatic decision is made with the backing of classified intelligence.
 
-% first it will lay out how espionage is considered an occupational hazard in the business of running a nation-state, one that is counteracted but never discourages.
+Using traditional methods of analysis, espionage certainly seems like something that states should be doing more to discourage.
+The results of intelligence operations can have devastating, even existential effects on the targeted nation's security.
+Arguably the most traditional form of espionage is stealing military secrets from a foreign power, and in that arena China has been relentless.
+A United States Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} ``Long-term, US future military advantage is being diminished by years of IP exfiltration from the DoD, DoN, and DIB,'' the report read, ``all with little to no adverse consequences to the thieves.''\footcite[p.~6]{bayer_cybersecurity_2019} The forecasts were dire: ``If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.''\footcite[p.~5]{bayer_cybersecurity_2019} The US Navy is not a disinterested party, of course, and there are reasons to take its conclusions with a grain of salt; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see \cite{gilli_why_2019}} Nonetheless, espionage presents alarming opportunities for rival states and the consequences for taking advantage of them remain minimal.
 
-% next I will discuss the theoretical justifications for why this might be the case. security seeking states and types of greedy states derive a reational benefit from allowing espionage to continue.
+Espionage has potentially disastrous consequences---and steps that might actually curb it are out of the question.
+The system of repercussions is set up so that the structure of civilian intelligence services is never fundamentally threatened.
+When an especially damaging espionage operation is uncovered, the traditional punishment is to declare certain embassy officials \emph{persona non grata}, which often results in \enquote{tit-for-tat} diplomatic expulsions.
+These expulsions make it more difficult for the affected state to conduct intelligence operations in the short-term, but they don't fundamentally punish the state for having attempted espionage in the first place.
 
-% finally I will look at historical moments in which new technologies were deployed for espionage-like purporses, and demonstrate how the norms governing the use of these technologies always evolved to permit espionage.
+The name \enquote{tit-for-tat} is taken from a highly-successful strategy for the prisoner's dilemma wherein the player simply mimics the choice that that their opponent made in the last round.
+By employing the tit-for-tat strategy, the player never gets taken advantage of for more than one turn and can always return to cooperation if their opponent unilaterally signals a willingness to do the same.
+Diplomatic expulsions play out in exactly the same fashion: when one country expels the diplomats of another, the other country will often respond in kind.
+Each side will continue to expel each other's diplomats until one side signals that it would like for the reciprocal PNG-ing of diplomats to stop.
+Then, both sides return to a cooperative state, each preserving some of their intelligence operations.
+Because neither side can gain an a decisive advantage, the back-and-forth punishment never ends in a mutual defection.
+
+\subsection{Espionage norms as a discrete series of choices}
+The key feature of \enquote{tit-for-tat} diplomatic expulsions is that the scope of the punishment is structurally incapable of discouraging a future, identical operation.
+Losing intelligence officers in an embassy temporarily makes it more difficult to conduct espionage, but it doesn't make a state less likely to try; the worst that can happen is a net-neutral result---going back to \enquote{square one,} so to speak.
+By contrast, were it more common to respond to espionage by imposing financial sanctions or even inflicting a military cost, a state would have to consider the possibility that attempting espionage could adversely impact other aspects of their security, leaving it worse off than when it started.
+These are the sorts of calculations that states need to make in other high-stakes arenas---pursuing a nuclear weapons program, for instance, could lead to severe economic retaliation, as could protectionist trade policies---but engaging in espionage has very little potential for negative externalities because espionage norms take the possibility of a severe response off the table.
+In this section, I present what could happen if they didn't.
+
+% https://tex.stackexchange.com/questions/249480/creating-a-payoff-matrix-using-latex-tabular-environment
+\begin{table}[ht]
+\centering
+\setlength{\extrarowheight}{2pt}
+\small
+\begin{tabular}{cc|c|c|}
+  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{Country 2}\\
+  & \multicolumn{1}{c}{} & \multicolumn{1}{c}{Tolerate Espionage}  & \multicolumn{1}{c}{Punish Espionage} \\\cline{3-4}
+  \multirow{3}*{Country 1}  & Tolerate Espionage & \makecell{~\\Both sides gain intelligence \\~} & \makecell{~\\ Only the USSR continues spying \\ ~} \\\cline{3-4}
+  & Punish Espionage & \makecell{~\\ Only the US continues spying \\~} & \makecell{~\\ No intelligence either direction \\~} \\\cline{3-4}
+\end{tabular}
+\caption{Espionage consequences decision matrix}
+\label{espionage-matrix}
+\end{table}
+
+% Possibly swap this with the previous paragraph
+
+To describe how states cooperate to preserve the liminal legal status of espionage, I propose an iterated decision-making model.
+My model changes the parameters so that the decision to cooperate or defect is based not in the amount of punishment received, but rather the type.
+Each time a state is forced to respond to a successful intelligence operation by a rival---such as the reveal of a high-level spy or a major data breach---it can either limit its response to one which is traditionally associated with espionage, or it can impose unprecedented costs that might discourage such an operation in the future.
+Table \ref{espionage-matrix} describes a single instance of this choice.
+In the status quo, State 1 signals its interest in cooperating by limiting its retaliation in response to espionage, no matter how damaging the operation is to their national security.
+State 2 responds by doing the same.
+Each iteration is a chance to play the game with fresh information; each espionage case is an opportunity to cooperate, reaffirming the norm that states always respond to espionage in a specific way, or defect, adding an additional cost to the attempt.
+
+Through this model two important things become clear.
+First, even though either state has the opportunity to defect by imposing harsher punishments in response to espionage, its gains would be short-lived, because the other state is guaranteed to immediately do the same---tit-for-tat.
+Espionage consequences are an iterated game that both parties are playing the exact same way, so, in the long run, the only two possible outcomes are mutual defection or mutual cooperation.
+In the realm of traditional espionage, we've only ever seen mutual cooperation, but the critical observation is that any state has the power to change that by defecting.
+Even if the defector's rival would have preferred to maintain the status quo, the rival has no option other than to defect as well or else be left at a senseless disadvantage.
+The relatively open flow of information is only possible with a universal buy-in.
+
+Second, while there is unquestionably a cultural norm informing the response to intelligence operations, that norm is doing a lot of work to maintain the status quo.
+So much work, in fact, that tradition could not possibly explain the entire phenomenon.
+Stripped of its mystique, the benefits of allowing intelligence operations to continue are, like any other state decision, a simple cost-benefit analysis; states decide that it best serves their interests to remain in the top left corner, rather than the bottom right.
+We can expect that there is \emph{some} rational explanation that political leaders employ to justify it, where they weigh the benefits of preserving their ability to spy against the leaks that inevitably result from espionage conducted against them.
+
+One possible explanation is that some states are just miscalculating.
+Though I have not found any scholarly works making this argument, the basic premise is persuasive.
+Every policymaker knows exactly how crucial espionage is to their job.
+With years of experience making difficult policy calls informed by intelligence, it is certainly much easier to imagine how one's decisions will be made worse without espionage than it is to imagine how one's enemies' decisions will be made better.
+It follows that there would be a strong psychological bias towards attempting to discover as much information as possible.
+While this is likely true to some extent, \enquote{everyone else is getting it wrong} is not an explanation that typically suffices in political science, and it will not suffice here.
+
+The surprisingly straightforward conclusion is that, if only two worlds are possible---one with widespread espionage and one without---then all the relevant nations must be concluding that the world without espionage is incredibly bad.
+So bad that even if one nation correctly concludes that its rival gains more from a world with a free flow of information than it does, the alternative, a world with significantly harsher punishments for espionage and a staunched flow of intelligence, still leaves that nation worse off.
+That calculation has a name: defensive realism.
+
+\subsection{Defensive realism}
+To explain why states might be willing to accept relative losses in security, I employ Charles Glaser's \emph{Rational Theory of International Politics}.
+Glaser describes his theory as integrating defensive realism with neoclassical realism.\footnote{With apologies to Mr.
+Glaser, I am going to refer to it as defensive realism, because they both put the security dilemma at the center of their explanation for international competition, which is the key observation for me.
+He would have you know that defensive realism ``can be viewed as a way station along the route to the full theory'' that he develops, which is ``significantly more general and complete than is defensive realism.'' (p.
+13-14)} Glaser begins from the core assumption that the reason for competition between security-seeking states is insecurity.
+Consider a theoretical security seeker interacting with its adversary, which it believes to be a security-seeking state as well.
+The first state feels insecure, so it builds up arms to increase its security.
+When the rival state, seeing that its adversary is building up arms, now feels less secure and builds up its own arsenal in response, and a classic security dilemma ensues.
+Even though both states were just security seekers, they're now both much less secure.
+
+% If both states are genuine security-seekers, then this situation is clearly sub-optimal, because now both states are heavily militarized and comparatively much less secure.
+
+Defensive realism introduces a counterintuitive observation: the key to interrupting the security dilemma is to make the adversary feel more secure, because it is the adversary's response to securitization that ultimately makes the first state insecure.
+Therefore, defensive realism posits that it is logical for a security seeker to take actions that increase not just its own security, but its \emph{adversary's} security as well.\footcite[p.~7]{glaser_rational_2010} These actions would signal to the adversary that the state is a security seeker, hoping to reach a mutually agreeable outcome.
+If the adversary is indeed a security seeker, then, no longer concerned about its security, it will respond in kind and resolve the security dilemma for both of them.
+
+The problem for a defensive realist is that unilaterally taking steps to increase the adversary's security is obviously quite risky.
+The state could be wrong that its adversary is a security seeker, in which case its optimistic overtures of cooperation will have exposed it to greater risk of attack.
+When states are unwilling to make those signaling measures because they fear compromising their own security, then a security dilemma results.
+This is where intelligence comes into play.
+
+Prior to making any decisions, states asses the security situation and determine which of their options are most likely to increase their own security.
+Glaser identifies two types of variables that mediate the magnitude of a security dilemma.
+The first are material variables, such as offense-defense balance and offense-defense distinguishability.
+Espionage plays a significant role in properly assessing material variables; during the Cold War, the United States placed a premium on generating quality estimates of how many bombers, missiles, and bases the Soviet Union had deployed.
+By knowing which of its adversary's capabilities are offensive, a state can better target its efforts towards reducing them.
+
+In addition to material variables, Glaser notes that information variables play an important and comparatively under-studied role in the security dilemma.
+Better information about its adversary's motives can give the state reason to trust that their risky signaling measures might succeed---or fail.
+Even if the state is uncertain, espionage can provide some information about its adversary's intent, which, in turn, can moderate the security dilemma enough to make cooperation possible.
+In the best case-scenario, the security dilemma is simply eliminated if both states are confident that the other is a security seeker.
+This is the optimal outcome for two genuine security-seekers, and it is made much more likely from the intelligence that espionage can provide.
+
+Consider a situation in which one state is a genuine security-seeker and so is its adversary.
+The state conducts espionage on its adversary, and the resulting intelligence increases its internal estimate of how likely the adversary is to be a security-seeker.
+As a result, the state decides that some form of arms control is no longer too risky to attempt in order to signal its benign intentions.
+Because the state receiving the signal is a security-seeker, not only will its own intelligence make it more likely to respond in kind, but it also benefited from the first state's intelligence operations, because that led to the initial exchange of signals.
+The two-way flow of information has a measurable impact on each state's decisionmaking, facilitating the optimal outcome for both parties.
+
+\subsection{Introducing the empirical examples}
+% I use defensive realism to formulate a theoretical foundation to explain the existence of espionage norms generally, and the Soviet Union's extension of espionage norms to spy satellites in particular.
+Between two security-seeking states, both states will be less secure in the event of an arms race, so preventing one is a mutually-agreeable outcome.
+Resolving the security dilemma requires that the states signal their benign intentions, which states are more likely to do if they have good intelligence---intelligence that informs them about their adversary's capabilities and intent.
+Therefore, cooperating to preserve the practice of espionage as a semi-legitimate state institution makes rational sense, both because of the benefits espionage offers the state doing the spying, and the benefits the same state derives \emph{from being spied upon}.
 

### 2019-11-18 article.tex
new script to parse citations

@@
 
+\section{Aerial espionage}
+\section{Introduction}
+According to \emph{The New York Times}, the very first reconnaissance flight took place in 1911, only eight years after Kitty Hawk, when ``Lieut.
+Piazzi today, for the first time in the history of warfare, made from this place an aerial reconnaissance against a hostile power'' surveilling the Turkish Infantry near Tripoli.\footcite{special_cable_to_the_new_york_times_air_1911} The ability to fly over an enemy's position to gain intelligence about their movements---in a craft orders of magnitude more versatile than a balloon---was so obviously valuable that countermeasures quickly followed.
+Early aerial dogfighting famously took the form of pilots firing handguns at each other out of open cockpits in mid-air (emphasis on attempted), but by the time of the First World War, planes were already outfitted with synchronized machine guns and ground regiments equipped with anti-aircraft weaponry.
+Crucial information about military movements was going to come at a cost.
+
+Even though these aircraft were performing vital intelligence-gathering functions, that did not make them spy planes.
+Aerial reconnaissance first appeared in the press in 1911, but the term ``spy plane'' appeared in the pages of \emph{The New York Times} just once before 1960.\footnote{According to my search of the archives.
+The instance mentioned here was in 1944, when the AP wrote that ``Germans tried to probe the secrets of the gathering Allied storm by E-boat patrol dashes [...] and by spy-plane coastal raids.''(\cite{the_associated_press_britons_1944}) In context of the article, though, it's clear that they are referring to military aerial reconnaissance, especially since it was written during the Second World War.
+This is a marked difference from a plane created for peacetime intelligence, which is what the term ``spy plane'' has since come to mean.} The image of a spy plane crashed into the public consciousness that year when Gary Powers crawled out of a U-2 class aircraft over Soviet Russia.
+The downed craft was clearly not a generic fighter plane that had flown too close to the border, and it suddenly became necessary to categorize this new type of plane, flying at previously unheard-of altitudes, that was purpose-built for clandestine photography.
+The United States first claimed that the lost plane was a ``weather craft.''\footcite{caruthers_soviet_1960} When that cover didn't hold up, a new story was born.
+The diplomatic fallout of the U-2 shootdown, which threatened to torpedo an upcoming NATO-Warsaw summit, had become the ``spy-plane case.''\footcite[This is the second time that ``spy plane'' as a term of art appeared in \emph{The New York Times}.
+There would be many more.]{reston_allies_1960}
+
+I emphasize the linguistic distinction between a military plane used for reconnaissance and a ``spy plane'' because that distinction has a major impact on how the flight is perceived and what diplomatic repercussions might result when it is caught.\footnote{That is not to say there is a hard boundary between military and non-military intelligence, either in purpose or collection.
+The overlap between military and civilian intelligence agencies---especially in the United States---is considerable.
+The Air Force was deeply involved with the U-2 Project, for example, even though it was technically under the jurisdiction of the CIA.
+The aerial photographs taken by U-2 missions were used to inform both military and civilian operations.} A military fight carries with it the implications of combat, aggression, and the possible prelude to an attack.
+A spy flight is instead a tool of civilian intelligence, whose purposes are often frustrated but never quite discouraged.
+The creation of the U-2 plane---a craft exclusively for civilian intelligence---is a pivotal moment in the history of espionage because the U-2 is the physical embodiment of the espionage norms it helped to define.
+Its \emph{raison d'\^etre} is an activity which is not permitted---or its altitude would be unnecessary---but its design emphasizes its peaceful reconnaissance purpose.
+
+\section{The Open Skies policy}
+From the beginning of the Cold War, aerial intelligence played a crucial role in American foreign policy.
+The United States considered aerial surveillance so important that it was willing to allow the Soviets to do it in return.
+President Eisenhower formally proposed a mutual inspection regime called Open Skies to the Soviet Union as a means of reducing security tensions.
+In his prepared remarks at the Geneva Summit of 1955, Eisenhower laid out the following plan:
+
+	\begin{quote}
+	\textbf{To give to each other a complete blueprint of our military establishments}, from beginning to end, from one end of our countries to the other, lay out the establishments and provide the blueprints to each other.
+	\newline
+
+	Next, to provide within our countries facilities for aerial photography to the other country---\textbf{we to provide you the facilities within our country, ample facilities for aerial reconnaissance}, where you can make all the pictures you choose and take them to your own country to study, you to provide exactly the same facilities for us and we to make these examinations, and by this step \textbf{to convince the world that we are providing as between ourselves against the possibility of great surprise attack}, thus lessening danger and relaxing tension.
+(emphasis mine)\footcite{eisenhower_president_1955}
+	\end{quote}
+
+The UN-sponsored Geneva Summit, formally the \emph{Conference on the Peaceful Uses of Atomic Energy}, was a comparatively hopeful period in the Cold War, a conference promoting scientific diplomacy where such a cooperative information exchange might have seemed feasible.\footcite[p.~27]{luscher_nuclear_2018} Nonetheless, the scale of this offer was astounding.
+Eisenhower offered to let the the Soviets install operational military bases in the United States from which they could surveil American national defense infrastructure, so long as they allowed the US to do the same.\footnote{Also included were reciprocal nuclear facility inspections, an immediate nonstarter in the USSR, but also controversial within the American administration back home.
+(\cite{prados_review_2015})} In effect, proposed a limited form of state-sanctioned espionage for explicitly defensive purposes---exactly the norm I am arguing exists in practice, albeit in a different form.
+
+The offer was quickly rejected.
+In conversation over buffet that evening, Khrushchev politely countered that the plan would not further disarmament at all because it would merely confirm the fragmentary information that their respective intelligence services already possessed.\footcite{department_of_state_memorandum_1955} The response was telling---though it was in no way an endorsement of the activity, it both acknowledged that espionage was taking place between the two superpowers, and that it served a potentially de-escalatory purpose.
+
+Khrushchev also claimed that the offer was primarily a propaganda move on the part of the US, but backed down when Eisenhower essentially dared him to call his bluff.\footcite[p.~2.
+This is a secondary source---the book cites a conversation by Eisenhower, quoted in NSC meeting memos that are stored in the Eisenhower library.]{bury_eisenhower_2014} The formal rejection came a few weeks later: ``While giving credit to this proposal's attempt to find a solution,'' USSR Council of Ministers chair Nikolai Bulganin wrote in his report to the Supreme Soviet, ``\textelp{} aerial photography cannot yield the expected results, since both our countries contain limitless expanses in which, if one desires, everything can be hidden.''\footcite{bulganin_meetings_1955} Though a lukewarm excuse, it was not a disdainful one.
+Earlier in the report, he had even said that the conference, as a whole, ``must be considered a definite success of the peace-loving forces.''
+
+Was the offer genuine, or was it the stunt that Khrushchev accused it of being?
+In a book about the relationship between aerial reconnaissance and Eisenhower's New Look foreign policy, historian Helen Bury argues that a willingness to negotiate acceptable and enforceable agreements with foreign powers was central to his Cold War strategy.\footcite[p.~69]{bury_eisenhower_2014} More importantly, the intelligence that an Open Skies policy promised would have been invaluable in his ongoing effort to control defense spending, where Eisenhower was embattled by deeply paranoid American institutions.\footcite[p.~212]{bury_eisenhower_2014} His commitment to Open Skies in practice, however, is still a point of some historical debate.
+In reviewing her argument, intelligence historian John Prados sees a contradiction between her portrayal of Eisenhower's genuine desire for arms control and Eisenhower's role in significant nuclear buildup.\footcite[p.~233-234]{prados_review_2015} In the following chapter, I will present my own evidence that these two existed as part of the same strategy.
+Eisenhower used photo-reconnaissance---of the kind that an Open Skies policy would have legally permitted---to create an accurate assessment the military capabilities required to sufficiently deter a Soviet nuclear strike; as a result, the United States built far fewer missiles than it otherwise would have.
+
+% Valentino: Actually, they were explicitly consistent.
+Ike thought that nukes were the cheapest way to counter the Soviet numerical advantage in conventional weapons (which were and are much more expensive).
+So Ike saw both as cost saving measures.
+
+Regardless of Eisenhower's intent, it is clear that the United States would have far more to gain from such a policy than the USSR, and it made strategic sense for the Soviets to reject the offer.
+Writing about the history of CIA's imaging program, former imagery analyst David Lindgren notes that Khrushchev had the advantage of huge amounts of publicly-available information about US military installations through newspapers, maps, and official aerial photographs, while Soviet restrictions on press freedom denied symmetrical data.\footcite[p.~38]{lindgren_trust_2000} Allowing for an Open Skies policy would also have revealed that Soviet air defenses were not nearly as far along as their leadership claimed.
+Khrushchev's son would later write that his father's overriding concern at the time was to conceal Soviet weakness.\footcite[p.~133]{brugioni_eyes_2010}
+
+Treaty or not, the United States was desperate for intelligence about the Soviet missile program, and aerial photo-reconnaissance was crucial to gaining it.
+The West's HUMINT capabilities within the Soviet Union were hamstrung by the intense surveillance under which their diplomats were placed, as well as the enormous size of the Russia itself.\footcite[p.~23]{lindgren_trust_2000} As the US increasingly relied on signals intelligence (SIGINT) to gather intelligence, the USSR began pursuing a more aggressive aerial defense policy, demonstrating their willingness to open fire on foreign reconnaissance aircraft.
+The increased cost associated with these missions did little to deter senior US policymakers from attempting to surveil the skies.\footcite[p.~4]{pedlow_central_1992}
+
+% The Soviet Union was able to successfully destroy US Military aircraft during purported peacetime because in many cases those US aircraft were not supposed to be there in the first place.
+In many cases their flight path was a clear violation of Soviet territorial boundaries---a fact of which the administration was aware.\footcite{goodpaster_memorandum_1956}
+
+In the rest of this chapter, I will argue that Eisenhower satisfied the need for intelligence that an Open Skies treaty would have provided by normalizing the collection of aerial intelligence and minimizing its diplomatic consequences.
+This process happened in two parts: first, through the routine resolution of military reconnaissance shootdowns, and second, by symbolically divorcing peacetime aerial espionage from military reconnaissance with the creation of a spy plane.
+For both the military and civilian missions, I will demonstrate how the even the most sensationalized incidents resulted in no more than ineffectual diplomatic protests, even though the spy plane could surveil locations that a ferret flight never could.
+
+\section{Military reconnaissance flights}
+\subsection{Timeline of early Cold War SIGINT missions}
+Well before the U-2 Incident in 1960, the United States had been routinely flying military aircraft near Soviet airspace, and the USSR had been routinely shooting them down.
+For a conflict defined by its lack of direct skirmishes between the two superpowers, a surprising number of Americans died in these incidents, which are mostly forgotten to history.
+A Smithsonian investigation found that over the course of the Cold War more than 200 American pilots were lost to Soviet shootdowns---126 of which remain unaccounted for to this day.\footcite{glenshaw_secret_2017}
+
+The purpose of these missions was to gather SIGINT identifying the location of critical radar installations along the Soviet border.
+Converted bombers---called ``ferrets'' by the Air Force---were outfitted with advanced radio equipment and sent to intercept as much information about Soviet radio transmissions as possible.\footcite[p.~4]{peterson_maybe_1993} According to the CIA, these missions began as early as 1947.\footcite[p.~4]{peterson_maybe_1993} Some of these flights only flew near the border, remaining over traditionally recognized international waters, while others were ``overflights,'' missions that deliberately violated Soviet airspace to collect intelligence at a much higher risk of failure.
+The earliest postwar overflight of the Soviet Union---of which I can find definitive record---took place on May 10, 1949, when two RF-80As surveilled the Kurile islands\footcite[p.~8]{peebles_shadow_2000}
+
+Determining the quantity of early Cold War reconnaissance flights is difficult because they were a closely guarded secret at the time.
+Some flights, however, were inadvertently declassified---when they were shot down by enemy air defenses.
+Since my goal is to understand the diplomatic repercussions of these flights, an effective place to start is a list of reconnaissance flights that attracted diplomatic attention after being shot down.
+The original shootdown stories are available in the newspaper archives, and the incidents have been helpfully compiled by US Air Force Academy historian Dr.
+John Farquhar, who specializes in aerial reconnaissance and wrote his master's thesis on how it shaped early Cold War diplomacy.
+My choice of incidents to analyze in this chapter is based in part on his work.
+% \newline
+
+Additionally, the CIA history magazine \emph{Cryptologic Quarterly} analyzed the Soviet shootdowns of US reconnaissance aircraft in depth, creating a comprehensive (as far as we know) list of each American overflight that was shot down by the USSR (Table \ref{soviet-shootdowns}).
+The article was written in 1993, but not declassified until 2009.
+The author, Michael Peterson, uses two limiting criteria that perfectly suit our purposes: the only flights listed are those that were (a) shot down by Soviet attacks, not Chinese, North Korean, etc., and (b) explicitly US reconnaissance aircraft.\footcite[p.~4]{peterson_maybe_1993} Note that not all of these flights are necessarily overflights; many took place over international waters when they were shot down.
+In effect, the table catalogs the 13 times in which the USSR shot down an American aircraft designed for collecting intelligence between 1950 and 1964---intercepted espionage attempts.\footcite[p.~5.
+In the original document, this table lists the first incident as having taken place over the Barents Sea, not the Baltic Sea.
+Because the description of the mission---including a map of its route in the same document---takes place entirely over the Baltic sea, I have concluded that this has to be a typographical error, and corrected it here.]{peterson_maybe_1993}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{llr}
+\textbf{Date}     & \textbf{\makecell[l]{U.S. Service \&\\ Aircraft Type}}   & \textbf{General Location} \\
+8 April 1950      & USN PB4Y2 Privateer           & Baltic Sea                           \\
+6 November 1951   & USN P2V Neptune               & Sea of Japan                         \\
+13 June 1952      & USAF RB-29                    & Sea of Japan                         \\
+7 October 1952    & USAF RB-29                    & East of Hokkaido/Kuril Is.           \\
+29 July 1953      & USAF RB-50                    & Sea of Japan                         \\
+4 September 1954  & USN P2V Neptune               & Sea of Japan                         \\
+7 November 1954   & USAD RB-29                    & East of Hokkaido.Kuril Is.           \\
+18 April 1955     & USAF RB-47                    & Off Kamchatka Peninsula              \\
+10 September 1956 & USAF RB-50                    & Sea of Japan                         \\
+2 September 1958  & USADF C-130                   & Soviet Armenia (near Turkish border) \\
+1 May 1960        & CIA U-2                       & Sverdolsk, USSR                      \\
+1 July 1960       & USAF RB-47                    & Barents Sea                          \\
+10 March 1964     & USAF RB-66                    & East Germany
+\end{tabular}
+\caption{Summary of Soviet Shootdowns, 1950-1964}
+\label{soviet-shootdowns}
+\end{table}
+
+I cannot confirm whether Peterson's record is comprehensive.
+Regardless of how thorough declassified CIA information appears, I cannot rule out the possibility that there were other, more secret missions which for whatever reason the United States government has chosen to continue protecting to this day.
+Nonetheless, a few aspects of the \emph{Cryptologic Quarterly} article are promising, and suggest that the information is relatively complete.
+First, the article was written for a classified audience as an overview of reconnaissance shootdowns during the 1950s, so it is intended to be comprehensive.
+Shootdowns are a sufficiently significant event such that it is unlikely the author would have just missed one.
+Additionally, when the article was declassified 16 years later, it appears to have been declassified without redactions or missing pages.
+And as with any analysis of diplomatic repercussions and espionage, we are aided by a simple logical trap---if there had been another mission so highly classified that it was omitted from this document, it is difficult to imagine how diplomatic consequences could have been imposed with a degree of secrecy that would prevent us from knowing about it today.
+
+In the following section, I cross-reference the shootdowns that the CIA says took place with Farquhar's analysis of their media and diplomatic impact.
+A shootdown is either present in both lists, or just one of the two---some incidents in the CIA list are missing from Farquhar's list because they did not have a diplomatic impact, and some incidents that Farquhar analyzes are not mentioned by Peterson because they were not SIGINT reconnaissance missions and likely (though not definitively) served little espionage function.
+Together, the full spectrum of aerial reconnaissance incidents demonstrates a clear pattern in which the shootdowns that involved espionage were quickly resolved by both parties, while the shootdown of a flight unrelated to espionage produced swift, significant backlash.
+
+\subsection{The Baltic Incident}
+The first of these shootdowns, on April 1950, brought the previously-hidden aerial reconnaissance program to the attention of the American people.
+An unarmed Navy Patrol plane, a PB4Y2 Privateer, was intercepted and downed by Soviet fighters over the Baltic sea.
+The Soviet ambassador to the US lodged a formal note of protest, incorrectly alleging that the US aircraft which had been shot down was a B-29 bomber.\footcite{kirk_ambassador_1950} Upon receiving the note, Secretary of State Dean Acheson remarked that Ambassador Vishinsky's manner was ``serious but not aggressive or antagonistic,'' and the Secretary ``recommended that publicity on our side should be avoided or, if unavoidable, minimized.''\footcite{kirk_ambassador_1950}
+
+A few days later, the American ambassador issued a formal response, in which he correctly identified the model of plane and claimed that at no point did it cross into any territorial waters.
+He blamed the Soviet Union directly, outlining the punishment that the Americans planned to impose:
+
+\begin{quote}
+	The United States Government demands that the the Soviet Government institute a prompt and thorough investigation of this matter in order that the facts set forth above my be confirmed to its satisfaction.
+The United States Government further demands that the most strict and categorical instructions be issued to the Soviet air force that there be no repetition, under whatever pretext, of incidents of this kind which are so clearly calculated to magnify the difficulties of maintaining peaceful and correct international relationships.
+
+	The United States Government fully expects that, when its investigation is completed, the Soviet Government will express its regret for the unlawful and provocative behavior of its aviators, will see to it that those responsible for this action are promptly and severely punished and will, in accordance with established custom among peace-loving nations, pay appropriate indemnity for the unprovoked destruction of American lives and property.\footcite{the_associated_press_text_1950}
+\end{quote}
+This was a weak response---it imposed no consequences, threatened no sanctions, and demanded only the most \emph{pro forma} of reparations.
+In essence, the note demanded an apology and a promise not to repeat.
+There was absolutely no chance that the parties responsible on the Soviet side would be held accountable---in the most serious of cases, some lower-level officers might take the fall, but even that was unlikely to happen here.
+If the United States had followed through on its demand for reparations---the only material repercussion mentioned---then one could have argued that this statement had some teeth.
+Instead, the demands were rejected in totality by the USSR.\footcite{salisbury_kremlin_1950} As far as I can tell, no reparations were ever made and no further consequences were imposed as a result.
+
+Both sides had reasons to make this a larger political issue.
+The Soviet Union had fired first on a US Navy aircraft---an aircraft which could not have been the aggressor because its only armament was a .45-cal pistol carried by one of the crewmen.\footcite[p.~7]{peterson_maybe_1993} Given that the plane was never found, all available evidence suggests that the flight likely did not violate the traditional 12-mile airspace boundary.
+Based on the last received position report, today the CIA claims that the plane was flying 20-25 miles off the coast of Latvia at the time it was shot down (Figure \ref{baltic-flight-path}).\footcite[p.~7]{peterson_maybe_1993} At the time, some unnamed Navy and Air Force officers, speaking to reporters in response to the Soviet protest note, conceded that it was possible that a complete breakdown in the plane's navigation system might have caused it to veer off-course into Latvia, but given its last known position, it would have taken ``a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.''\footcite[Technically, the  article credits the first statement about the possible electronic breakdown to ``Navy and Air Force officials'' and the 90 degrees statement to ``Aerial navigators.'' It is unclear whether the navigators in questions are with the US military, but it does not matter regardless.]{the_new_york_times_soviet_1950}
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.45]{baltic-flight-path.png}
+\caption{USN PB4Y2 Flight Path}
+\label{baltic-flight-path}
+\end{figure}
+
+Meanwhile, the United States knew that it hadn't lost a B-29, and that the missing plane probably hadn't ended up in Soviet airspace, but the Privateer in question \emph{had} been performing a questionably legal reconnaissance mission when it was shot down.
+Though the US was likely correct that the plane never flew over Latvia, it certainly got as close as it needed to in order to gather the necessary intelligence about Soviet military installations on the coast.
+No traditional boundaries were violated, but a repurposed Navy privateer loaded with electronic reconnaissance equipment flying close to Latvia looked, to use a less technical term, shady.
+Yet based on the initial exchange of diplomatic protests, neither side was inclined to magnify the issue if they could avoid it.
+
+The American press had no such reservations.
+\emph{Washington Post} columnist Walter Lippmann wrote that it could not have been ``a local incident brought on by a local commander but [rather] that it was an act of Soviet policy.
+The known facts indicate that Soviet intelligence \textelp{} believed it was carrying important electronic equipment and that orders were given to the Soviet fighter command to intercept it.''\footcite{lippmann_baltic_1950} He speculated that there was no way the plane could have violated Soviet airspace, because wreckage would have been recovered, the plane was too slow, and no commander would have sent it on such a dangerous mission intentionally.
+``If, when, and as the American command were reconnoitering the Soviet military establishments on the Baltic coast,'' Lippmann wrote, ``it would use a plane of a wholly different type.''\footcite{lippmann_baltic_1950}
+
+A newspaper columnist is not a policymaker, but his argument was likely reflective of the political opinion that Americans held regarding espionage at the time: of course we do these things, we just do them \emph{better} than that---and if the US had been doing the thing that the Soviets claimed it was, then it would have been justified of them to open fire.
+``Baltic Plane Mystery,'' a \emph{Post} article written a few days later, was almost sympathetic to the Soviet commanders.
+``Electronics make the old delimitations for border coastal flights ridiculous.
+A plane flying on a course perfectly legal by standards accepted today might still be engaged in reconnaissance of the first importance that an unfriendly power would try to frustrate.''\footcite{childs_baltic_1950} This is true, both in terms of the mission itself and the technological environment, but he went even further.
+``The only sound and safe assumption is that the Russians have a thorough and far-reaching espionage system.
+And at the same time we must hope that our system, and particularly on the side of counterespionage, is effective.'' Even the press, though outraged at the loss of American life, apparently took it as a given that this kind of espionage is necessary---and expected from both sides.
+
+The most important consequence of the Baltic Incident was not diplomatic.
+As a direct result of the shootdown, President Truman, who until that point was largely uninformed about these missions, ordered a thirty-day halt to all reconnaissance flights.
+In that time, his office drafted an operating procedure that would balance the strategic need for reconnaissance with its potential diplomatic volatility.\footcite[p.~41]{farquhar_aerial_2015} The resulting memo from the Joint Chiefs of Staff created, with Truman's approval, the Special Electronic Airborne Search Project (SESP), which set clear delineations of responsibility between Navy and Air Force reconnaissance, and established executive responsibility for approving the missions.
+
+Farquhar cites this as a landmark moment in the history of aerial reconnaissance.
+``No longer would ferret operations be conducted ad hoc by the military services;'' he writes, ``from 1950 onward, reconnaissance operations attracted Presidential attention and played a significant role in shaping U.S.
+foreign policy.''\footcite[p.~42]{farquhar_aerial_2015} The strict procedures that SESP established were crafted to avoid the risk of escalation, while still explicitly acknowledging the ``serious disadvantages accruing to the United States if the cessation of these operations were to be extended over an excessively long period.''\footcite{bradley_memorandum_1950} But the SESP procedures also contain a curious provision: ``Flights will not be made closer than twenty miles to the USSR or USSR- or satellite-controlled territory.''\footcite{bradley_memorandum_1950} This rule would not be followed---as is typical of espionage, the consequences of the activity never outweigh its strategic value.
+
+\subsection{Routine resolution}
+Between the shootdown of the Navy Privateer in 1950 and the famous U-2 shootdown in 1960, the United States lost nine more aircraft to Soviet air defenses.
+Without analyzing each one at the same level of detail, I can demonstrate how the general attitudes displayed in response to the Baltic Incident became normalized.
+In each of the incidents listed, the pattern is the same---each side released dueling statements of protest, and the situation continued more or less as before.
+
+Describe here are the next few peacetime reconnaissance shootdowns whose diplomatic consequences Farquhar analyzes.
+A B-29 that went missing on October 7, 1952 hit the front page of the \emph{New York Times} two days later.
+This mission is on Peterson's list as a ferret flight.
+Farquhar notes that the incident happened during the US election season and that the media thought the attack was intended to lower American prestige, but he does not mention any diplomatic consequences.\footcite[p.~43-44]{farquhar_aerial_2015} The official US response, however, is preserved on the State Department website.
+As with the Baltic Incident, the US simply demanded that the Soviets pay for the plane and return any survivors.\footcite{the_new_york_times_u.s._1952} Two more shootdowns in March 1953---one involving an American F–84 Thunderjet and the other an RAF Lincoln bomber---actually resulted in conciliatory responses from both sides, including a secret meeting to reduce aerial tensions ``of which little is known''.\footcite[p.~45]{farquhar_aerial_2015}
+
+In another incident on July 29, 1953, the United States protested the shootdown of a B-50 (reconnaissance flight) over the Sea of Japan.
+The Soviet Union responded by protesting the alleged American shootdown of a Soviet passenger plane.
+Despite press outrage, no further action was taken.\footcite[p.~47]{farquhar_aerial_2015} A September 4, 1954 shootdown of a Navy P2V Neptune actually led to a call by one US Senator to break off diplomatic ties with the Soviet Union: ``Just another note from our State Department to the Kremlin hierarchy will not impress these uncivilized rulers \textelp{} that this new attack upon an American plane confirms Communist arrogance and aggressiveness to a point where the breaking of diplomatic relations is justified.''\footcite{the_associated_press_ending_1954} Instead of a harsher punishment, Eisenhower brought the issue to the UN with the full knowledge that an unfavorable judgment would be vetoed by the Soviets.
+The move demonstrated Eisenhower's ability to resist the calls for harsher action while still placating the American public.\footcite[p.~47]{farquhar_aerial_2015}
+
+Without belaboring the point, suffice to say that Farquhar concludes that after this moment in 1954, the resolution of these incidents became less tense.
+``International incidents posed by shoot downs of reconnaissance aircraft still acted as a barrier in the path of d\'etente; but, by late 1954, overriding strategic concerns dictated a move toward breaking the cycle of hostility.''\footcite[p.~49]{farquhar_aerial_2015} In terms of media coverage, and diplomatic significance, no shootdown that followed would require more significant consequences than the ones that had already happened.
+
+One incident in the period between 1950-1954 does not fit the broader pattern.
+On November 20, 1951, an American C-47 transport was forced down in Hungary.
+This incident is omitted from Peterson's list because, despite Soviet cries of ``spies and saboteurs,'' there is no evidence that it was a reconnaissance flight.\footnote{It also wasn't exactly shot down so much as forced to land.} Of all the cases that Farquhar examines, this is the only one that provoked a sustained diplomatic response from the United States---and it is the only case where the plane in question definitively carried no espionage equipment.
+While the ferret flights were perfectly calibrated to fit as much radio surveillance equipment as possible, the only evidence of espionage that the Soviets were able to produce from the C-47---a plane that they recovered intact---was a portable radio, two extra parachutes, and some packets of warm blankets.\footcite{the_united_press_soviet_1951}
+
+I contend that the lack of espionage equipment is precisely the reason the United States made a sustained public effort to recover the four airmen, which is something that almost never happened with ferret flights.\footnote{While some of the flights were shot down overseas, making recovery difficult, I am referring here to situations where the US was aware that the Soviets likely held their airmen in captivity and still made little effort to get them back.} Farquhar summarizes the American response to this incident as follows: ``Responding to the press attention, the Truman Administration acted swiftly, attempting to gain the fliers release through diplomatic pressure.
+The President ordered the Hungarian consulates in New York and Cleveland closed and banned private travel to the country.
+Legislatively, Truman asked Congress to pass a \$100 million Mutual Security Act to aid `selected persons residing in Soviet bloc states or refugees who wanted to form armed units' in opposition to Communism.''\footcite[p.~43]{farquhar_aerial_2015} Truman immediately took action against material Hungarian interests in the United States and threatened to empower \emph{armed dissidents} if the prisoners were not released.
+The contrast between this response and the one from just a year prior, when the US demanded ``a prompt and thorough investigation'' of the Baltic Incident, is stark.
+
+We also know that the United States conducted many more overflights than the failed ones that are listed here, using a variety of retrofitted planes.
+In 2001, the Air Force held a symposium honoring and recollecting the pilots who flew the overflights on Soviet territory in the early Cold War, an effort in which they were joined by the British in 1954.\footcite[p.~v]{hall_early_2003} Many of the presenters spoke about entire reconnaissance projects that apparently didn't even register diplomatically.
+Project Heart Throb, which outfitted RB-57s with surveillance equipment, flew between 15 and 19 missions over Eastern Europe from 1955 to 1956, in the recollection of Gerald E.
+Cooke, a Air Force pilot assigned to the project.\footcite[p.~194]{hall_early_2003} Another project from around the same time, Slick Chick, set up F-100 jets with rapid-fire 20mm cameras.
+The pilot who presented this project at the symposium, Cecil Rigby, personally flew two such missions and estimated that there were six Slick Chick missions total.\footcite[p.~176]{hall_early_2003}
+
+The USSR knew about these missions, which were tracked by radar and often chased by MiG fighters.
+Unlike some of the flights that were shot down, these overflights were clear violation of their territorial sovereignty.
+We will likely never know exactly how many such missions were flown.
+The Truman and Eisenhower administrations were shockingly unsuccessful at recovering the pilots who were held as POWs by the Soviets.
+Dino Brugioni, a senior CIA officer involved in the aerial imagery program, paints a grim picture of the institutional accountability involved when the flights went bad.
+In meetings between the State department and high-level Soviet diplomats, the subject of captured pilots ``was broached only perfunctorily in relation to other things being discussed.''\footcite[p.~72]{brugioni_eyes_2010} Some family remembers of lost airmen received posthumous awards, but most simply received the pilot's personal effects and no explanation.
+``And because these were secret missions conducted by a field command,'' Brugioni writes, ``a change of field commander meant that the fate of the men was soon forgotten.''\footcite[p.~72]{brugioni_eyes_2010} With missions so secret that we abandoned the men who flew them, we can have no expectation of finding a complete record today.
+
+Even without a complete list of flights, it is clear that the territorial boundaries of the USSR were routinely violated to generate critical intelligence for NATO powers in an early, relatively volatile stage of the Cold War.
+As time passed these flights appear to have become more accepted, not less.
+Diplomats of both nations were, as a matter of policy, able to paper over these issues in service of advancing other interests.
+When the shootdown caused unavoidable diplomatic consequences---such as when it was well-publicized---the countries exchanged the necessary diplomatic cables and engaged in the requisite saber rattling.
+These incidents undoubtedly altered the climate of Cold War diplomacy, but the fact remains that in 10 years of reconnaissance, the only action the Soviets took to keep the United States from conducting these incredibly provocative reconnaissance flights was to shoot them down.
+
+% Valentino: Can you talk about other ways the Sovierts might have responded.
+What other kinds of actions might have seemed reasonable under different norms?
+Maybe give an example of a Soviet response to something apparently similar on its face, but clearly out of the espionage domain, so the response was much harsher.
+
+\section{Civilian spy flights}
+\subsection{Project Aquatone}
+While Eisenhower was able to defuse some of the tension surrounding the military reconnaissance flights, even the most aggressive overflights could only hope to skirt the borders of Soviet territory; the United States couldn't just fly a B-50 over Moscow.
+What US intelligence-gathering capabilities needed was some way of photographing suspicious installations deep within Russia while still being able to claim the same benefit-of-the-doubt status that prevented even the most contentious overflights from escalating into a larger conflict.
+What they needed was a spy plane.
+
+The key feature of a spy plane is that it is designed to be minimally provocative.
+Any plane can be retrofitted with cameras or radios that allow it to perform useful reconnaissance functions---what separates a spy plane is that there can can be no doubt (in the mind of the target) that it is being used for anything \emph{but} reconnaissance.
+For the spy plane to serve its intended purpose as a minimally provocative agent of espionage then, two things must be true: its operation must be understood by adversaries to be non-military, and that distinction must meaningfully alter their threat perception and associated response.
+To build such a aircraft, Eisenhower approved Project Aquatone, the code name for the creation of the U-2.
+
+From its conception, the U-2 was purpose-built to be read as a civilian aircraft.
+Eisenhower required that the pilots of U-2 aircraft be CIA employees, specifically forbidding uniformed Air Force officers.\footcite[p.~33, Though many of the pilots did have Air Force backgrounds.]{lindgren_trust_2000} The CIA also had operational control of the program instead of the Air Force.
+``I want this whole thing to be a civilian operation,'' Eisenhower wrote, when settling an operational dispute between the two departments.
+``If uniformed personnel of the armed services of the United States fly over Russia, it is an act of war---legally---and I don't want any part of it.''\footcite[p.~60.
+The original source for this quote is an \emph{OSA History} that requires codeword clearance.
+It is quoted here by the History Staff of the CIA.]{pedlow_central_1992} This allowed the United States to ``truthfully deny,'' in the phrasing of the CIA, that any US ``military planes'' had flown over the USSR---which they had to do when the inevitable Soviet protest notes were filed after U-2 overflights began in 1956.\footcite[p.~109]{pedlow_central_1992}
+
+Having the CIA run the program---ostensibly for the purpose of deception---changed the nature of the program itself.
+If the only goal had been to achieve a level of plausible deniability in the event of a shootdown, then simply not wearing an Air Force uniform during the mission likely would have been sufficient.
+The ``weather reconnaissance'' excuse had even been used before with the ferret flights.\footcite[p.~45]{farquhar_aerial_2015} At the highest levels of American government, however, policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation---even if the NASA weather-craft cover didn't hold.
+``It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft'' read a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} Both halves of that equation were equally important.
+The Russians must believe that these operations are peacetime civilian operations, and they must be correct about it.
+
+The United States was aware how provocative these reconnaissance flights might be, and Eisenhower was always concerned, correctly, about ``the terrible propaganda impact that would be occasioned if a reconnaissance plane were to fail.''\footcite[p.~162]{pedlow_central_1992} Starting in 1958, the American administration ordered a drastic decrease in the number of overflights.\footcite[p.~51]{powers_operation_2004} But Eisenhower was also under tremendous political pressure to learn more about the Soviet missile program, as hawkish senators amplified fears of a so-called ``missile gap,'' concerned that Soviet ballistic missile capabilities were advancing faster than their own.\footcite[Fears of this missile gap quickly followed earlier fears of a ``bomber gap,'' which ironically the U-2 had been critical in disproving.]{licklider_missile_1970} He especially did not want to affect the upcoming Four Powers diplomatic summit in Paris, scheduled for the summer of 1960.
+Nonetheless, Eisenhower ordered two more overflights in 1960, one on April \nth{9} and the other on May \nth{1}.
+Only one of them returned.
+
+% It turned out however, that the Soviets opted to lodge their diplomatic protests privately, and soon they stopped protesting them at all.\footcite[p.~42]{lindgren_trust_2000}
+
+\subsection{The U-2 Incident}
+On the morning of May 1, 1960, Soviet Air Defense forces detected a high-altitude aircraft flying over Soviet Tajikistan.
+Though they were were not yet able to conclusively identify it, the foreign aircraft was an American U-2 spy plane performing an reconnaissance overflight of the USSR---at a time when the Soviets were deeply embarrassed by their inability to prevent such flights.\footcite{orlov_u-2_2007} The previous U-2 overflight on April 9, less than a month prior, had lasted a full six hours and the aftermath caused an upheaval in the Soviet military.
+Leadership ordered an investigative commission to root out the shortcomings of the Air Force and Air Defense agencies, and a series of charts were drawn up anticipating the routes of future U-2 flights.
+The next time one came through, the Soviets were absolutely determined to shoot it down.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.3]{powers-flight-path.jpg}
+\caption{Operation Grand Slam Flight Path}
+\label{powers-flight-path}
+\end{figure}
+
+The Soviets still proved unable to immediately stop the flight on May \nth{1}.
+A missile battalion in the plane's path was not on alert duty.
+Armed fighter aircraft were in the wrong positions.
+A frantic attempt to have another fighter plane literally ram it out of the sky was scrapped when the pilot failed to make visual contact.\footcite{orlov_u-2_2007} The highest possible levels of Soviet leadership were actively involved with the mission as it was taking place.
+A Soviet Colonel in the USSR Air Defense later recalled that Khrushchev ``clearly viewed the violation of their nation's skies by a foreign reconnaissance aircraft on the day of a Soviet national holiday, and just two weeks before a summit conference in Paris, as a political provocation.'' \footcite{orlov_u-2_2007}
+
+Why did the generally cautious Eisenhower authorize two overflights in such close proximity to a crucial summit?
+For one, the secrecy of the program worked against him.
+Eisenhower was not willing to make the U-2 program public, so he couldn't convince his critics that he had extremely high quality intelligence that no such missile gap existed, because to reveal the existence of his aerial photography would prompt questions about its origins.\footcite[p.~51]{lindgren_trust_2000} And the missions had been such a tremendous success that his reluctance to authorize more overflights grated his staff, especially the influential Director of Central Intelligence (DCI), Allen Dulles.\footcite[p.~354]{brugioni_eyes_2010}
+
+Another crucial factor emboldened Eisenhower and his team---most of whom, like Dulles, were much more enthusiastic about overflights than he was---to conduct that second mission.
+The USSR had been aware of the April 9 flight, and the United States \emph{knew} that they were aware of it, thanks to an onboard computer that had detected the Soviet tracking from the beginnings of the operation.\footcite{pedlow_central_1992} From the perspective of the American administration, the lack of a protest note after the previous flight was a sign that the USSR was not particularly offended by the overflights.
+That was a mistake.
+``It is one of the many unfortunate aspects of the high degree of secrecy surrounding Moscow's actions,'' imagery analyst David Lindgren writes, ``that the United States never understood the how angered and frustrated Soviet leaders were made by the U-2 overflights.''\footcite[p.~52]{lindgren_trust_2000} After years of bluster about the superiority of Soviet military forces, Khrushchev was not willing to admit that his Air Defense had been unable to shoot an unarmed spy plane flying deep into his homeland.\footcite[p.~59]{powers_operation_2004}
+
+Almost four hours into the spy plane's flight on May 1, pilot Francis Gary Powers felt a dull ``thump'' and ``a tremendous orange flash lit the cockpit and sky'' as a Soviet surface-to-air (SAM) missile exploded behind him, ripping the wings off his plane and sending it into a tailspin.\footcite[p.~61]{powers_operation_2004} His subsequent capture by the Soviet military became an immediate international incident.
+Three days after NASA quietly reported that they had lost a U-2 type weather reconnaissance plane, Khrushchev announced that the USSR had shot down an American plane that had flown into their airspace, which he called ``an aggressive provocation aimed at wrecking the Summit Conference.''\footcite[p.~112]{powers_operation_2004} Once Khrushchev gleefully debunked the resulting American cover story by revealing that they had taken Powers alive, Eisenhower admitted that the Gary Powers had been flying a spy plane, in a press conference where Eisenhower memorably declared that espionage was a ``a distasteful but vital necessity'' to prevent another attack like Pearl Harbor.\footcite{eisenhower_news_1960}
+
+Publicly, the USSR was not ready to accept that US overflights of its territory were a ``distasteful but vital necessity.'' The Four Powers summit in Paris was a disaster.
+Khrushchev launched into an angry rant on the first day and stormed out, immediately squashing any hope that pressing issues like the fate of West Berlin would be resolved there.
+He also canceled a planned visit by Eisenhower to the USSR that was scheduled to take place the next month.
+A contemporary newsreel about the conference said that ``in the course of two hours, Khrushchev brought US-Soviet relations to their lowest point since the end of World War II.''\footcite{universal_studios_summit_1960}
+% The contrast between the severity of response Khrushchev wanted to portray publicly and what the USSR actually did is highlighted by his own son Sergei, who edited and annotated the 2007 edition with notes featuring additional context for his father's statements.
+
+\section{Espionage in the open skies}
+\subsection{Minimal consequences with maxmimal noise}
+The U-2 Incident is often cited as a major failure of the Eisenhower administration.
+In the traditional telling, it significantly worsened US-Soviet relations at time when it seemed like lasting peace might be possible.
+I disagree; instead, I argue that that the U-2 Incident should be understood as an embarrassing but otherwise typical failure of an intelligence operation, one consistent with the minimal consequences typically applied to espionage.
+
+No one disputes that the collapse of the Paris Summit was directly attributable to the Gary Powers shootdown but, frankly, summits just aren't that important.
+From 1953 to the end of the Cold War in 1991, there were 23 total US-USSR summits.\footcite{fain_chronology_2011} The main issue that did not get resolved in 1960 was the status of Berlin.
+It also did not get resolved at the next summit in 1961, a comparatively warmer affair between Khrushchev and Eisenhower's successor, John F.
+Kennedy.
+If the worst that happened because of the U-2 incident was an unproductive summit meeting, then that is a relatively minor punishment.
+With the late rollout of the captured pilot, the dramatic scene at the summit, and the cancellation of Eisenhower's visit, Khrushchev publicly embarrassed the United States without meaningfully altering Soviet policy towards it.
+
+What would a major punishment have looked like?
+One proportional response Khrushchev could have considered was to force the issue on regional air bases.
+Since the US was using nearby bases to launch invasive reconnaissance missions, the USSR could have demanded that the US either close some of them or agree to halt surveillance flights.
+This likely would have involved some sort of trade, but it is not an unreasonable hypothetical.
+The Soviets were incredibly preoccupied with removing the forward-deployed Jupiter missiles from Turkey.
+Although the missiles were deployed to bases well within NATO territory, and arguably less threatening to Soviet security than American overflights of the Soviet bloc, the existence of these missile became a central bargaining chip in the Cuban Missile Crisis.
+By contrast, the Soviets simply did not have the same level of interest in curtailing reconnaissance flights.
+That both sides of the Cold War had strong incentives to avoid military escalation is well-documented, but at least threatening military retaliation for launching provocative military operations would not seem to have been out of place.
+
+While I argue that the response to the U-2 flight was within the reasonable bounds for espionage, I acknowledge that it was still unprecedented in the context of reconnaissance flights.
+There are a number of reasonable explanations for why the U-2 shootdown escalated.
+Partially the Soviets felt humiliated by a technology for which they had no countermeasures and the uniquely insulting way the overflights flew deep into Soviet territory.
+Eisenhower thought that Khrushchev might have been looking for an excuse to cancel his visit to Russia.\footcite[p.~555]{eisenhower_waging_1965} Others in his administration believed that the U-2 Incident was an excuse to justify an intensification of hostilities that had already been decided by hard-liners in the Kremlin.\footcite[p.~328]{kistiakowsky_scientist_1976}
+
+Most scholars who study this period argue that the cumulative tension created by these flights contributed to an overall chilly tone in US-Soviet relations.
+Farquhar, for instance, refers to this as the ``cycle of hostility,'' in which the series of aerial incidents increased suspicions of military buildup on both sides, leading to even more aerial incidents as both powers attempted to verify their suspicions.\footcite[p.~43]{farquhar_aerial_2015} While it's true that the there were many aerial incidents, it is not at all clear how this affected any other aspect of US-Soviet diplomacy.
+I can't definitively say that there wasn't some diplomatic issue that would have gotten resolved at a summit absent these reconnaissance flights, but I can't locate an obvious one either.
+
+Consider the period of time over which these flights took place.
+If aerial reconnaissance had risen to anything more than a lingering irritant, then it should have to have been resolved in some way before 1960.
+The US-Soviet relationship saw huge shifts, including periods of optimism, over the 11 years preceding the U-2 incident.
+During this period, hundreds of flights took place in contentious territory---many of them purposefully violating Soviet boundaries, causing some of them to be shot down---and the most significant consequence was a failed summit meeting, after which the flights continued on almost as before.
+Both sides were so willing to contain the issue that many of the airmen never saw their families again.
+
+\subsection{The U-2 after Powers}
+When the US really needed intelligence that only an overflight could provide, the diplomatic consequences that the USSR had imposed were not sufficient to discourage it.
+Two months to the day after Powers was shot down, a Soviet MiG-19 opened fire on an American RB-47H and captured the plane's navigator and co-pilot.
+While Powers was serving a 10-year sentence in Soviet prison, the two airmen were returned to the US in less than six months with only mild fanfare.
+Khrushchev wrote in his memoir that he wanted to continue ``our general line of peaceful coexistence'' and cited the resolution of this incident as an example of such, because the US was forced to make a formal request for the return of the airmen.\footcite[p.~256-257]{khrushchev_memoirs_2007} The only concessions the US made in return were to announce the discontinuation of its U-2 overflights (which Kennedy was already committed to) and to not make an issue of the illegal detention of the pilots.\footcite{time_cold_1961} For spy flights after Powers, the USSR simply took the necessary immediate countermeasures and shot down the planes that it could.
+Just as with traditional espionage, the countermeasures did not punish the attempt.
+
+Until the day he died, Khrushchev greatly exaggerated the lasting effect of the U-2 Incident.
+He wrote in his memoir that ``the commander of American forces in West Germany gave the order not to fly any closer than 50 kilometers from the border between East and West Germany.
+And no more incidents of that kind occurred.''\footcite[p.~256]{khrushchev_memoirs_2007} His son Sergei, who annotated the memoir, admitted that this statement was false: ``In practice such incidents occurred again from 1961 onward, but instead of the U-2 the Americans now used various types of Phantom or SR-71.''\footcite[p.~258]{khrushchev_memoirs_2007} On at least two occasions after the U-2 Incident, an American overflight crossed over into Soviet territory and the Soviets responded with...
+a diplomatic protest.
+The US quickly apologized, and that was that.\footcite{orlov_u-2_2007}
+
+The U-2 did not disappear after the Powers flight.
+A recently declassified CIA report details the many locations in which U-2 photography was employed to gather intelligence after 1960, including China, India, Indonesia, Thailand, Tibet, Laos, North Vietnam, and Venezuela.
+Lyndon Johnson even clarified that his predecessor's order to end U-2 flights over the Soviet Bloc was not indefinitely binding---it was valid only until countermanded.\footcite[p.~195]{pedlow_central_1992} As it turned out, it never became necessary to resume Soviet Bloc overflights because satellite photo-reconnaissance soon replaced them; U-2 overflights were rendered mostly obsolete.
+There is, however, one use for which satellites will likely never displace aerial photo-reconnaissance---a situation like the Cuban Missile Crisis, where events are so time-sensitive that intelligence is needed in a matter of hours.
+
+On October 15, 1962, U-2 photography revealed that the Soviet Union was assembling missiles in Cuba that were capable of carrying nuclear warheads.
+As the crisis unfolded, the United States continued to send U-2 planes over Cuba in order to maintain a close watch on the development of the missile sites.
+Kennedy's team decided that if a U-2 were to be shot down, the most likely SAM site would be immediately fired upon.\footcite[p.~217]{caro_passage_2013} Then, on October 27, it happened---a U-2 was shot down flying over Cuba and the pilot was killed.
+Kennedy flinched; he had just sent Khrushchev an offer that might lead to peace, and he didn't want to jeopardize it.
+Against the advice of administration hard-liners, including Vice President Johnson, Kennedy refused to authorize the strike against the Cuban SAM site.\footcite[p.~220]{caro_passage_2013} The next day, Khrushchev agreed to Kennedy's terms, and the Cuban Missile Crisis ended with only a single casualty---the U-2 pilot, Major Rudolph Anderson.
+
+As his title indicates, Rudolph Anderson was an Air Force Officer, not a member of the CIA.
+During the crisis, the President transferred control of Cuban U-2 reconnaissance to the Air Force, and authorized it to fly as many missions as necessary over the island.\footcite[p.~208-209]{pedlow_central_1992} Khrushchev did not say much about the Cuban U-2 in his writing, but he was very clear on one point: he did not order the SAM strike, and was worried that Kennedy ``might not be able to absorb this blow.''\footcite[p.~338]{khrushchev_memoirs_2007} Shooting down an American aircraft risked escalation, and that was not something Khrushchev wanted.
+This was the second time in history that Soviet forces had shot down a U-2 plane---but even though the spy plane was flying over Cuban airspace and piloted by an US Air Force Major, it was the Soviets who were worried that shooting it down had been an unnecessarily aggressive act.
+
+\subsection{Bridge of spies}
+The incidents involving U-2 aircraft during the Cold War perfectly mirror the traditional consequences for espionage because the United States succeeded in applying espionage norms to its use.
+Eisenhower's insistence on having the CIA manage Project Aquatone had the intended effect: the spy plane was, in story as well as fact, a civilian intelligence operation.
+
+Even when a U-2 operation went poorly, it still played by the rules of the game.
+Powers opted against using his suicide capsule, but the macabre truth about the U-2 is that pilots weren't supposed to survive a crash at all.
+The United States originally publicized the weather craft story because Eisenhower had been assured that Gary Powers could not possibly be alive to tell the Russians the truth.\footcite[p.~35]{lindgren_trust_2000} When Powers confessed to his Soviet captors, he told them, correctly, that he was a civilian pilot with the CIA.
+His cover had been blown, but underneath it he was still a spy, and in pretty much all respects he and the U-2 were treated accordingly.
+
+Both heads of state made clear in their writings that they considered Powers to be spy.
+On capturing Gary Powers, Khrushchev wrote: ``This was a hostile act by the leaders of the U.S.
+government, and they made no attempt to conceal it.
+They didn't think we had the capability of \textelp{} acquiring irrefutable proof that the United States was using methods that were impermissible in peacetime.''\footcite[p.~239]{khrushchev_memoirs_2007} Sergei's annotations once again contradicted his father's claim that these methods were impermissible; more than forty American reconnaissance aircraft were shot down by the USSR during the Cold War.
+As is often the case with espionage, Khrushchev wanted to make it seem like he was standing up for the integrity of the Soviet state without endangering the espionage equilibrium.
+
+Eisenhower, always an insightful thinker on issues of intelligence, understood that beneath the grandstanding, the fundamentals of international espionage still applied. In his own words:
+
+\begin{quote}
+	Furthermore, it seemed to me that Mr.
+Khrushchev's outbursts were hypnotizing the world with a passionate but highly distoreted presentation of one particular phase of international espionage.
+His government had been so notoriously involved in spying, especially in the United States, as to dwarf our activities, but by separating this particular type of espionage from all others he hoped to make convincing his charge of ``warmongering.'' To claim that because the equipment employed was an airplane with a camera, and therefore provocative of war, was plain silly, and I felt it necessary that the matter be put in perspective.
+\\
+
+	The real issue at stake was not the fact that both sides conducted intelligence activities, but rather that the conduct and announced intentions of the Communists created the necessity for such clandestine maneuvers.
+As a consequence, the West---more specifically the Untied States as the major military power of the West---had to maintain constantly the capacity to detect any possible prelude to an infinitely more destructive Pearl Harbor.\footcite[p.~551]{eisenhower_waging_1965}
+\end{quote}
+The next chapter examines in depth how the fear of a nuclear Pearl Harbor affected the entire American intelligence strategy, but even in the limited context of the spy plane, it is clear that Eisenhower deployed espionage to de-escalatory effect.
+Khrushchev had convinced the world that the Soviet missile capabilities were far more advanced than they actually were.
+Eisenhower was under tremendous pressure from the public, Republicans in Congress, and even his own staff to increase the defense budget.
+The only thing standing between Eisenhower and an accelerated arms race were detailed photo-reconnaissance reports from the U-2 that showed the Soviets were developmentally far behind what their public statements suggested.
+
+The peaceful uses of aerial reconnaissance were formalized at the end of the Cold War, when members of NATO and the Warsaw Pact signed the ``Treaty on Open Skies,'' which established guidelines for unarmed aerial surveillance flights in the spirit of open information and de-escalation.\footcite{organization_for_security_and_co-operation_in_europe_treaty_1992} This treaty was almost identical to the one first suggested by Eisenhower in 1955.\footcite{center_for_arms_control_and_non-proliferation_fact_2017} With certain restrictions, the type of flight that altered the course of the Cold War is now officially routine and permissible.
+In February 1962, Francis Gary Powers walked across a bridge into West Berlin at the same time as Russian Colonel Rudolph Abel, an American prisoner.
+Powers' father had suggested to the State Department that they might be traded, spy for a spy.\footcite[p.~239]{powers_operation_2004} ``This,'' Eisenhower wrote, ``was a tacit admission by Khrushchev that our `outrageous' U-2 pilots have their opposite numbers operating within the borders of this country.''\footcite[p.~558]{eisenhower_waging_1965}
+
+The Powers flight and Project Aquatone channel the norms of espionage in one final, but crucial respect: their strategic value easily outweighed the consequences of being caught.
+No one understood this better than President Eisenhower, who, when he was occasionally asked about the wisdom of the U-2 flights, would always reply: ``Would you be ready to give back all of the information we secured from our U-2 flights over Russia if there had been no disaster to one of our planes in Russia?'' In his telling, he never received an affirmative response.\footcite[p.~559]{eisenhower_waging_1965}
+
+\section{Spy satellites}
+
+\section{Conclusion}
+
 \end{document}

### 2019-11-18 scripts/get-cites.sh

@@ -1,0 +1,50 @@
+#!/bin/bash
+# Name: get-cites.sh
+# Author: Alex Petros
+
+# This is a small script to list the tags of cites in the project
+# If you supply the -u argument, it will on print the unused cites
+
+usage() { echo "Usage: get-cites [-u] [rootDir]"; exit 1; }
+
+bibFile="article.bib"
+texFile="article.tex"
+unused=false
+
+# Process options
+# TODO add sort-by-year option
+while getopts ":u" opt; do
+    case "${opt}" in
+        u) # output only unused citations
+            unused=true
+            ;;
+        \?) 
+            usage
+            ;;
+    esac
+done
+shift $((OPTIND -1))
+
+# Get cites location
+if [[ "$#" -ne 1 ]]; then
+    usage
+fi
+
+# Get the citations in the .bib file
+rootDir=$1
+citesPath="$rootDir/$bibFile"
+bibCites=$( cat "$citesPath" | grep -E '@\w+{' | sed -E 's/@.*{(.*),/\1/g' )
+
+# If -u was not specificed then output cites
+if [[ "$unused" = false ]]; then
+    echo "$bibCites"
+    exit 0
+fi
+
+# Otherwise process the differnce between the two
+articleCites=$( cat article.tex | grep -oE '\\(foot)?cite(\[[^]]*\])?{[^}]*}' | sed -E 's/^.*{(.*)}/\1/g' | sort -u )
+bibCites=$( echo "$bibCites" | sort -u )
+
+# remove extra stuff from the div
+diff <( echo "$articleCites" ) <( echo "$bibCites" ) | grep '>' | sed 's/> //'
+

### 2019-11-26 article.tex
Theory draft sent to Wohlforth

--- 2019-11-17 article.tex
@@
 
+% States now recognize that some forms of espionage facilitate international cooperation.\footcite{baker_tolerance_2004}
+
+A complete theoretical justification for espionage---that maintaining a mutual exchange of illicit information is better for both states than the alternative---can be constructed by combining the work of previous scholars, and finding archival evidence to support their claims.
+Christopher Baker explains how espionage makes functional cooperation possible through treaties, allowing states to verify each others' compliance.\footcite[p.~6]{baker_tolerance_2004} Charles Glaser uses the 1972 Anti-Ballistic missile treaty as an example of precisely the type of signal that could lead states to resolve the security dilemma, but he does not acknowledge that the intelligence operations of both states are what make signals like it possible.\footnote{It is entirely possible that Glaser himself realizes this, it's just not written down; the purpose of the book is to defend the theory, not explain the mechanics.} And Julius Stone, a professor of international law, foresaw this exact role for espionage in 1962, well before the public had any real idea about the US spy satellite program: \enquote{If you do not have a system of international inspection and if you can't get one (and I think it is quite likely that we can't) then the function which international inspection is supposed to serve still needs fulfilling.
+You may not be able to fulfill it to the optimum extent by reciprocally tolerated espionage, but you may be able to reduce the dangers.}\footcite[p.~41]{stone_legal_1962}
+
+In 1962, Stone defended espionage from a practical perspective in exactly the same terms that I am now defending it from a theoretical one.
+Permitting at least some kinds of espionage is actually beneficial for both sides because it allows us to independently verify the intent and capabilities of other nations.
+Doing so actually reduces the likelihood that miscalculation will lead to open hostilities.
+As a result, both states allow it to happen, because they each benefit from a less-tense security environment.
+Stone aptly described this phenomenon as \enquote{reciprocally tolerated espionage.}
+
 \subsection{Introducing the empirical examples}
+
+The best way to prove that this theory has explanatory power for state decision-making is to find examples from the past where states made decisions that showed they were willing to accept the cost of enemy espionage in order to practice espionage themselves.
+In the ideal scenario, we would find evidence of policymakers justifying their decisions along these lines.
+Such decisions were made frequently during the Cold War, and with the explicit intent to increase the bilateral flow of information to create a more secure world environment.
+This article will present two such cases: aerial espionage and spy satellites.
+
+Both planes and satellites played an underexplored role in the formulation of some of the Cold War's landmark diplomatic achievements.
+SALT I (the ABM Treaty) is widely considered to be the landmark diplomatic achievement of Nixon's \emph{d\'etente} policy, which set the stage for a broader easing of relations between the US and the USSR, and spy satellites were crucial to reducing the cost that each side would pay to sign it.
+Limiting ballistic missile defense systems was a relatively minor ask for both sides, but in the highly risk-averse context of the Cold War, the signal was significant.\footcite[p.~66]{glaser_rational_2010} Underneath the omnipresent threat of nuclear annihilation, \emph{d\'etente} provided a rational strategic choice for both states---made possible by their reciprocal tolerance of espionage.
+
+% I have yet to find a quote from Nikita Khrushchev in which he states that it was good for the Soviet Union to be have its state secrets compromised.
+I don't expect that I ever will.
+The realist approach to espionage is difficult to acknowledge politically; it cuts against national pride and power in order to reach a pragmatic, somewhat emotionally unsatisfying conclusion.
+I asked a senior national security policymaker, Jake Sullivan, whether he thought American espionage was a net positive for global security.
+``I wouldn't work for the American government if I did not believe that it was a net positive actor,'' he told me, ``so I want the United States to be able to make the most informed decisions possible.'' When I asked whether it promoted global stability for a foreign nation---Russia for instance---to spy on the United States, he qualified his answer.
+
+% He split the types of intelligence that foreign state might gather on the United States into three types: intent, capabilities, and strategy.
+``I have no problem with the Russians knowing that we do not have aggressive intent,'' he told me.
+Capabilities, however, are more complicated.
+Some defense capabilities would probably be fine to have another country independently verify, and some we might want to keep secret, such as the atomic bomb before the end of the Second World War.
+And under almost no circumstances would a state want to have its strategies and warfighting plans revealed, so those should be kept as secret as possible.\footcite{sullivan_personal_2019} Without invoking the theory itself, Sullivan described exactly the type of intelligence that a defensive realist would need to solve the security dilemma.\footnote{At the point in my research when I conducted this interview, I had never even heard the term ``defensive realism,'' lest you think I led him to it.}
+
+% The problem is that a state will never be able to control what type of information is being leaked.
+Espionage is an all-or-nothing proposition; either a state is willing to risk compromising some government secrets---including the ones that it is always disadvantageous to leak---in order to preserve the right to attempt it itself, or the state defects and begins imposing severe diplomatic consequences for espionage, inevitably inviting the same consequences upon itself, at which point the information pipeline tightens drastically.
+Unfailingly, states will choose to continue cooperating, so they must all be making the same calculus that the information they lose is worth the security they gain from getting to practice espionage themselves.
+
+% If you believe that American intent is benign, then tacitly permitting espionage is one way out of the security dilemma.
+Sullivan and Glaser did not conclusively explain, however, why \emph{every} major power continues to cooperate.
+More likely than not, some states today are not pure security-seekers.
+Glaser calls these greedy states, and when a state's motives are mixed between greed and security, those get assigned the label of greedy as well.
+For a greedy state, having another state correctly evaluate its intent could prove disastrous.
+The state might be pretending to have a benign intent in order to gain an advantage and espionage might reveal the truth.
+
+% That is only partially true.
+I certainly believe that elements of espionage and its place in diplomacy are socially constructed.
+The practice of expelling diplomats, for instance, predates all the events discussed in this thesis, and almost certainly evolved from some sort of customary protocol, rather than a rational calculation about the most effective way of responding to espionage.
+
+% An institutional theory of politics cannot explain the endurance of espionage norms, because there are no international institutions dedicated to governing espionage.
+As the legal section notes, peacetime espionage in most traditional sense---activities associated with human spies---has virtually no legal recognition at all.
+The first institutional recognition of espionage for conflict-reduction purposes is the SALT I treaty in 1972, which protected ``national means of verification,'' but espionage-as-verification is limited to the specific treaties in which it is in included.
+Espionage is probably just too uncouth to formally acknowledge.
+
+% Likewise, the only legal regime that applied to a spy plane was the one that applied to planes in general, and that explicitly prohibited the U-2 overflights; the Treaty on Open Skies was signed in 1992, well after the events discussed here.
+The clear illegality of the U-2 overflights should have given the Soviet Union more leverage to gain concessions, not less.
+As for space, the Outer Space Treaty entered into effect in 1967, eight years after \emph{Discoverer 14}; the strange timing of that treaty, right as the USSR was developing ASAT capabilities, was noted in Chapter 4.
+
+% I posit that the desire from all states to continue the practice of espionage writ large, even as they seek to stop particular instances of it against themselves, has an important psychological component.
+Every policymaker would like to their decisions to be as well-informed as possible.
+Without the benefit of hindsight, it is very difficult objectively gauge how much ``security'' is lost from a successful intelligence operation.
+By contrast, it is very easy for someone in charge of making an important decision to notice how much better off they are for having had access to some illicitly-obtained information.
+I don't mean to suggest that these trade-offs are always impossible to evaluate.
+Khrushchev was keenly aware of how the satellite photography would affect his missile bluff; preserving the missile bluff was just not worth impacting Khrushchevs's own ability to make use of satellite photography.
+In situations where it is difficult to evaluate relative gains, there exists a strong psychological bias towards preserving ones own view of the situation, even if it means allowing the adversary to have its view as well.
+
+% If espionage often leads to mutually agreeable outcomes, then why is not just completely legal?
+Well, in some cases it is.
+A true Open Skies Treaty followed soon after the collapse of the Soviet Union.
+In just over a decade, spy satellites went from ``espionage in space'' to ``national technical means of verification.'' But even as espionage gained some legitimacy in statecraft, the vagueness of that wording, ``means of verification,'' is indicative of what keeps espionage from true public acknowledgment.
+The intelligence business is often looked at---not entirely unfairly---as being shady.
+Reframing espionage as treaty verification procedure can help mitigate that, but if formally acknowledging spy satellites feels uncouth, its hard to imagine that human spies will ever receive a legal regime of their own.
+
+% Regardless of theoretical explanation, the cooperative outcomes of the Cold War is remarkable.
+The USSR had reason to believe that it would have benefited from a mutual prohibition on the technology that the United States was using.
+The next two chapters will go in-depth on the process that led to this decision---from both sides---to make the case that espionage norms have a fundamental appeal that the United States was able to exploit for its own benefit, and induce the Soviet Union to cooperate.
+
 % I use defensive realism to formulate a theoretical foundation to explain the existence of espionage norms generally, and the Soviet Union's extension of espionage norms to spy satellites in particular.
@@
 
+% \section{Aerial espionage}
+% \section{Introduction}
+% According to \emph{The New York Times}, the very first reconnaissance flight took place in 1911, only eight years after Kitty Hawk, when ``Lieut.
+Piazzi today, for the first time in the history of warfare, made from this place an aerial reconnaissance against a hostile power'' surveilling the Turkish Infantry near Tripoli.\footcite{special_cable_to_the_new_york_times_air_1911} The ability to fly over an enemy's position to gain intelligence about their movements---in a craft orders of magnitude more versatile than a balloon---was so obviously valuable that countermeasures quickly followed.
+Early aerial dogfighting famously took the form of pilots firing handguns at each other out of open cockpits in mid-air (emphasis on attempted), but by the time of the First World War, planes were already outfitted with synchronized machine guns and ground regiments equipped with anti-aircraft weaponry.
+Crucial information about military movements was going to come at a cost.
+
+% Even though these aircraft were performing vital intelligence-gathering functions, that did not make them spy planes.
+Aerial reconnaissance first appeared in the press in 1911, but the term ``spy plane'' appeared in the pages of \emph{The New York Times} just once before 1960.\footnote{According to my search of the archives.
+The instance mentioned here was in 1944, when the AP wrote that ``Germans tried to probe the secrets of the gathering Allied storm by E-boat patrol dashes [...] and by spy-plane coastal raids.''(\cite{the_associated_press_britons_1944}) In context of the article, though, it's clear that they are referring to military aerial reconnaissance, especially since it was written during the Second World War.
+This is a marked difference from a plane created for peacetime intelligence, which is what the term ``spy plane'' has since come to mean.} The image of a spy plane crashed into the public consciousness that year when Gary Powers crawled out of a U-2 class aircraft over Soviet Russia.
+The downed craft was clearly not a generic fighter plane that had flown too close to the border, and it suddenly became necessary to categorize this new type of plane, flying at previously unheard-of altitudes, that was purpose-built for clandestine photography.
+The United States first claimed that the lost plane was a ``weather craft.''\footcite{caruthers_soviet_1960} When that cover didn't hold up, a new story was born.
+The diplomatic fallout of the U-2 shootdown, which threatened to torpedo an upcoming NATO-Warsaw summit, had become the ``spy-plane case.''\footcite[This is the second time that ``spy plane'' as a term of art appeared in \emph{The New York Times}.
+There would be many more.]{reston_allies_1960}
+
+% I emphasize the linguistic distinction between a military plane used for reconnaissance and a ``spy plane'' because that distinction has a major impact on how the flight is perceived and what diplomatic repercussions might result when it is caught.\footnote{That is not to say there is a hard boundary between military and non-military intelligence, either in purpose or collection.
+The overlap between military and civilian intelligence agencies---especially in the United States---is considerable.
+The Air Force was deeply involved with the U-2 Project, for example, even though it was technically under the jurisdiction of the CIA.
+The aerial photographs taken by U-2 missions were used to inform both military and civilian operations.} A military fight carries with it the implications of combat, aggression, and the possible prelude to an attack.
+A spy flight is instead a tool of civilian intelligence, whose purposes are often frustrated but never quite discouraged.
+The creation of the U-2 plane---a craft exclusively for civilian intelligence---is a pivotal moment in the history of espionage because the U-2 is the physical embodiment of the espionage norms it helped to define.
+Its \emph{raison d'\^etre} is an activity which is not permitted---or its altitude would be unnecessary---but its design emphasizes its peaceful reconnaissance purpose.
+
+% \section{The Open Skies policy}
+% From the beginning of the Cold War, aerial intelligence played a crucial role in American foreign policy.
+The United States considered aerial surveillance so important that it was willing to allow the Soviets to do it in return.
+President Eisenhower formally proposed a mutual inspection regime called Open Skies to the Soviet Union as a means of reducing security tensions.
+In his prepared remarks at the Geneva Summit of 1955, Eisenhower laid out the following plan:
+
+% 	% \begin{quote}
+% 	\textbf{To give to each other a complete blueprint of our military establishments}, from beginning to end, from one end of our countries to the other, lay out the establishments and provide the blueprints to each other.
+% 	\newline
+
+% 	Next, to provide within our countries facilities for aerial photography to the other country---\textbf{we to provide you the facilities within our country, ample facilities for aerial reconnaissance}, where you can make all the pictures you choose and take them to your own country to study, you to provide exactly the same facilities for us and we to make these examinations, and by this step \textbf{to convince the world that we are providing as between ourselves against the possibility of great surprise attack}, thus lessening danger and relaxing tension.
+(emphasis mine)\footcite{eisenhower_president_1955}
+% 	% \end{quote}
+
+% The UN-sponsored Geneva Summit, formally the \emph{Conference on the Peaceful Uses of Atomic Energy}, was a comparatively hopeful period in the Cold War, a conference promoting scientific diplomacy where such a cooperative information exchange might have seemed feasible.\footcite[p.~27]{luscher_nuclear_2018} Nonetheless, the scale of this offer was astounding.
+Eisenhower offered to let the the Soviets install operational military bases in the United States from which they could surveil American national defense infrastructure, so long as they allowed the US to do the same.\footnote{Also included were reciprocal nuclear facility inspections, an immediate nonstarter in the USSR, but also controversial within the American administration back home.
+(\cite{prados_review_2015})} In effect, proposed a limited form of state-sanctioned espionage for explicitly defensive purposes---exactly the norm I am arguing exists in practice, albeit in a different form.
+
+% The offer was quickly rejected.
+In conversation over buffet that evening, Khrushchev politely countered that the plan would not further disarmament at all because it would merely confirm the fragmentary information that their respective intelligence services already possessed.\footcite{department_of_state_memorandum_1955} The response was telling---though it was in no way an endorsement of the activity, it both acknowledged that espionage was taking place between the two superpowers, and that it served a potentially de-escalatory purpose.
+
+% Khrushchev also claimed that the offer was primarily a propaganda move on the part of the US, but backed down when Eisenhower essentially dared him to call his bluff.\footcite[p.~2.
+This is a secondary source---the book cites a conversation by Eisenhower, quoted in NSC meeting memos that are stored in the Eisenhower library.]{bury_eisenhower_2014} The formal rejection came a few weeks later: ``While giving credit to this proposal's attempt to find a solution,'' USSR Council of Ministers chair Nikolai Bulganin wrote in his report to the Supreme Soviet, ``\textelp{} aerial photography cannot yield the expected results, since both our countries contain limitless expanses in which, if one desires, everything can be hidden.''\footcite{bulganin_meetings_1955} Though a lukewarm excuse, it was not a disdainful one.
+Earlier in the report, he had even said that the conference, as a whole, ``must be considered a definite success of the peace-loving forces.''
+
+% Was the offer genuine, or was it the stunt that Khrushchev accused it of being?
+In a book about the relationship between aerial reconnaissance and Eisenhower's New Look foreign policy, historian Helen Bury argues that a willingness to negotiate acceptable and enforceable agreements with foreign powers was central to his Cold War strategy.\footcite[p.~69]{bury_eisenhower_2014} More importantly, the intelligence that an Open Skies policy promised would have been invaluable in his ongoing effort to control defense spending, where Eisenhower was embattled by deeply paranoid American institutions.\footcite[p.~212]{bury_eisenhower_2014} His commitment to Open Skies in practice, however, is still a point of some historical debate.
+In reviewing her argument, intelligence historian John Prados sees a contradiction between her portrayal of Eisenhower's genuine desire for arms control and Eisenhower's role in significant nuclear buildup.\footcite[p.~233-234]{prados_review_2015} In the following chapter, I will present my own evidence that these two existed as part of the same strategy.
+Eisenhower used photo-reconnaissance---of the kind that an Open Skies policy would have legally permitted---to create an accurate assessment the military capabilities required to sufficiently deter a Soviet nuclear strike; as a result, the United States built far fewer missiles than it otherwise would have.
+
+% % Valentino: Actually, they were explicitly consistent.
+Ike thought that nukes were the cheapest way to counter the Soviet numerical advantage in conventional weapons (which were and are much more expensive).
+So Ike saw both as cost saving measures.
+
+% Regardless of Eisenhower's intent, it is clear that the United States would have far more to gain from such a policy than the USSR, and it made strategic sense for the Soviets to reject the offer.
+Writing about the history of CIA's imaging program, former imagery analyst David Lindgren notes that Khrushchev had the advantage of huge amounts of publicly-available information about US military installations through newspapers, maps, and official aerial photographs, while Soviet restrictions on press freedom denied symmetrical data.\footcite[p.~38]{lindgren_trust_2000} Allowing for an Open Skies policy would also have revealed that Soviet air defenses were not nearly as far along as their leadership claimed.
+Khrushchev's son would later write that his father's overriding concern at the time was to conceal Soviet weakness.\footcite[p.~133]{brugioni_eyes_2010}
+
+% Treaty or not, the United States was desperate for intelligence about the Soviet missile program, and aerial photo-reconnaissance was crucial to gaining it.
+The West's HUMINT capabilities within the Soviet Union were hamstrung by the intense surveillance under which their diplomats were placed, as well as the enormous size of the Russia itself.\footcite[p.~23]{lindgren_trust_2000} As the US increasingly relied on signals intelligence (SIGINT) to gather intelligence, the USSR began pursuing a more aggressive aerial defense policy, demonstrating their willingness to open fire on foreign reconnaissance aircraft.
+The increased cost associated with these missions did little to deter senior US policymakers from attempting to surveil the skies.\footcite[p.~4]{pedlow_central_1992}
+
+% % The Soviet Union was able to successfully destroy US Military aircraft during purported peacetime because in many cases those US aircraft were not supposed to be there in the first place.
+In many cases their flight path was a clear violation of Soviet territorial boundaries---a fact of which the administration was aware.\footcite{goodpaster_memorandum_1956}
+
+% In the rest of this chapter, I will argue that Eisenhower satisfied the need for intelligence that an Open Skies treaty would have provided by normalizing the collection of aerial intelligence and minimizing its diplomatic consequences.
+This process happened in two parts: first, through the routine resolution of military reconnaissance shootdowns, and second, by symbolically divorcing peacetime aerial espionage from military reconnaissance with the creation of a spy plane.
+For both the military and civilian missions, I will demonstrate how the even the most sensationalized incidents resulted in no more than ineffectual diplomatic protests, even though the spy plane could surveil locations that a ferret flight never could.
+
+% \section{Military reconnaissance flights}
+% \subsection{Timeline of early Cold War SIGINT missions}
+% Well before the U-2 Incident in 1960, the United States had been routinely flying military aircraft near Soviet airspace, and the USSR had been routinely shooting them down.
+For a conflict defined by its lack of direct skirmishes between the two superpowers, a surprising number of Americans died in these incidents, which are mostly forgotten to history.
+A Smithsonian investigation found that over the course of the Cold War more than 200 American pilots were lost to Soviet shootdowns---126 of which remain unaccounted for to this day.\footcite{glenshaw_secret_2017}
+
+% The purpose of these missions was to gather SIGINT identifying the location of critical radar installations along the Soviet border.
+Converted bombers---called ``ferrets'' by the Air Force---were outfitted with advanced radio equipment and sent to intercept as much information about Soviet radio transmissions as possible.\footcite[p.~4]{peterson_maybe_1993} According to the CIA, these missions began as early as 1947.\footcite[p.~4]{peterson_maybe_1993} Some of these flights only flew near the border, remaining over traditionally recognized international waters, while others were ``overflights,'' missions that deliberately violated Soviet airspace to collect intelligence at a much higher risk of failure.
+The earliest postwar overflight of the Soviet Union---of which I can find definitive record---took place on May 10, 1949, when two RF-80As surveilled the Kurile islands\footcite[p.~8]{peebles_shadow_2000}
+
+% Determining the quantity of early Cold War reconnaissance flights is difficult because they were a closely guarded secret at the time.
+Some flights, however, were inadvertently declassified---when they were shot down by enemy air defenses.
+Since my goal is to understand the diplomatic repercussions of these flights, an effective place to start is a list of reconnaissance flights that attracted diplomatic attention after being shot down.
+The original shootdown stories are available in the newspaper archives, and the incidents have been helpfully compiled by US Air Force Academy historian Dr.
+John Farquhar, who specializes in aerial reconnaissance and wrote his master's thesis on how it shaped early Cold War diplomacy.
+My choice of incidents to analyze in this chapter is based in part on his work.
+% % \newline
+
+% Additionally, the CIA history magazine \emph{Cryptologic Quarterly} analyzed the Soviet shootdowns of US reconnaissance aircraft in depth, creating a comprehensive (as far as we know) list of each American overflight that was shot down by the USSR (Table \ref{soviet-shootdowns}).
+The article was written in 1993, but not declassified until 2009.
+The author, Michael Peterson, uses two limiting criteria that perfectly suit our purposes: the only flights listed are those that were (a) shot down by Soviet attacks, not Chinese, North Korean, etc., and (b) explicitly US reconnaissance aircraft.\footcite[p.~4]{peterson_maybe_1993} Note that not all of these flights are necessarily overflights; many took place over international waters when they were shot down.
+In effect, the table catalogs the 13 times in which the USSR shot down an American aircraft designed for collecting intelligence between 1950 and 1964---intercepted espionage attempts.\footcite[p.~5.
+In the original document, this table lists the first incident as having taken place over the Barents Sea, not the Baltic Sea.
+Because the description of the mission---including a map of its route in the same document---takes place entirely over the Baltic sea, I have concluded that this has to be a typographical error, and corrected it here.]{peterson_maybe_1993}
+
+% \begin{table}[ht]
+% \centering
+% \begin{tabular}{llr}
+% \textbf{Date}     & \textbf{\makecell[l]{U.S. Service \&\\ Aircraft Type}}   & \textbf{General Location} \\
+% 8 April 1950      & USN PB4Y2 Privateer           & Baltic Sea                           \\
+% 6 November 1951   & USN P2V Neptune               & Sea of Japan                         \\
+% 13 June 1952      & USAF RB-29                    & Sea of Japan                         \\
+% 7 October 1952    & USAF RB-29                    & East of Hokkaido/Kuril Is.           \\
+% 29 July 1953      & USAF RB-50                    & Sea of Japan                         \\
+% 4 September 1954  & USN P2V Neptune               & Sea of Japan                         \\
+% 7 November 1954   & USAD RB-29                    & East of Hokkaido.Kuril Is.           \\
+% 18 April 1955     & USAF RB-47                    & Off Kamchatka Peninsula              \\
+% 10 September 1956 & USAF RB-50                    & Sea of Japan                         \\
+% 2 September 1958  & USADF C-130                   & Soviet Armenia (near Turkish border) \\
+% 1 May 1960        & CIA U-2                       & Sverdolsk, USSR                      \\
+% 1 July 1960       & USAF RB-47                    & Barents Sea                          \\
+% 10 March 1964     & USAF RB-66                    & East Germany
+% \end{tabular}
+% \caption{Summary of Soviet Shootdowns, 1950-1964}
+% \label{soviet-shootdowns}
+% \end{table}
+
+% I cannot confirm whether Peterson's record is comprehensive.
+Regardless of how thorough declassified CIA information appears, I cannot rule out the possibility that there were other, more secret missions which for whatever reason the United States government has chosen to continue protecting to this day.
+Nonetheless, a few aspects of the \emph{Cryptologic Quarterly} article are promising, and suggest that the information is relatively complete.
+First, the article was written for a classified audience as an overview of reconnaissance shootdowns during the 1950s, so it is intended to be comprehensive.
+Shootdowns are a sufficiently significant event such that it is unlikely the author would have just missed one.
+Additionally, when the article was declassified 16 years later, it appears to have been declassified without redactions or missing pages.
+And as with any analysis of diplomatic repercussions and espionage, we are aided by a simple logical trap---if there had been another mission so highly classified that it was omitted from this document, it is difficult to imagine how diplomatic consequences could have been imposed with a degree of secrecy that would prevent us from knowing about it today.
+
+% In the following section, I cross-reference the shootdowns that the CIA says took place with Farquhar's analysis of their media and diplomatic impact.
+A shootdown is either present in both lists, or just one of the two---some incidents in the CIA list are missing from Farquhar's list because they did not have a diplomatic impact, and some incidents that Farquhar analyzes are not mentioned by Peterson because they were not SIGINT reconnaissance missions and likely (though not definitively) served little espionage function.
+Together, the full spectrum of aerial reconnaissance incidents demonstrates a clear pattern in which the shootdowns that involved espionage were quickly resolved by both parties, while the shootdown of a flight unrelated to espionage produced swift, significant backlash.
+
+% \subsection{The Baltic Incident}
+% The first of these shootdowns, on April 1950, brought the previously-hidden aerial reconnaissance program to the attention of the American people.
+An unarmed Navy Patrol plane, a PB4Y2 Privateer, was intercepted and downed by Soviet fighters over the Baltic sea.
+The Soviet ambassador to the US lodged a formal note of protest, incorrectly alleging that the US aircraft which had been shot down was a B-29 bomber.\footcite{kirk_ambassador_1950} Upon receiving the note, Secretary of State Dean Acheson remarked that Ambassador Vishinsky's manner was ``serious but not aggressive or antagonistic,'' and the Secretary ``recommended that publicity on our side should be avoided or, if unavoidable, minimized.''\footcite{kirk_ambassador_1950}
+
+% A few days later, the American ambassador issued a formal response, in which he correctly identified the model of plane and claimed that at no point did it cross into any territorial waters.
+He blamed the Soviet Union directly, outlining the punishment that the Americans planned to impose:
+
+% % \begin{quote}
+% 	The United States Government demands that the the Soviet Government institute a prompt and thorough investigation of this matter in order that the facts set forth above my be confirmed to its satisfaction.
+The United States Government further demands that the most strict and categorical instructions be issued to the Soviet air force that there be no repetition, under whatever pretext, of incidents of this kind which are so clearly calculated to magnify the difficulties of maintaining peaceful and correct international relationships.
+
+% 	The United States Government fully expects that, when its investigation is completed, the Soviet Government will express its regret for the unlawful and provocative behavior of its aviators, will see to it that those responsible for this action are promptly and severely punished and will, in accordance with established custom among peace-loving nations, pay appropriate indemnity for the unprovoked destruction of American lives and property.\footcite{the_associated_press_text_1950}
+% % \end{quote}
+% This was a weak response---it imposed no consequences, threatened no sanctions, and demanded only the most \emph{pro forma} of reparations.
+In essence, the note demanded an apology and a promise not to repeat.
+There was absolutely no chance that the parties responsible on the Soviet side would be held accountable---in the most serious of cases, some lower-level officers might take the fall, but even that was unlikely to happen here.
+If the United States had followed through on its demand for reparations---the only material repercussion mentioned---then one could have argued that this statement had some teeth.
+Instead, the demands were rejected in totality by the USSR.\footcite{salisbury_kremlin_1950} As far as I can tell, no reparations were ever made and no further consequences were imposed as a result.
+
+% Both sides had reasons to make this a larger political issue.
+The Soviet Union had fired first on a US Navy aircraft---an aircraft which could not have been the aggressor because its only armament was a .45-cal pistol carried by one of the crewmen.\footcite[p.~7]{peterson_maybe_1993} Given that the plane was never found, all available evidence suggests that the flight likely did not violate the traditional 12-mile airspace boundary.
+Based on the last received position report, today the CIA claims that the plane was flying 20-25 miles off the coast of Latvia at the time it was shot down.\footcite[p.~7]{peterson_maybe_1993} At the time, some unnamed Navy and Air Force officers, speaking to reporters in response to the Soviet protest note, conceded that it was possible that a complete breakdown in the plane's navigation system might have caused it to veer off-course into Latvia, but given its last known position, it would have taken ``a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.''\footcite[Technically, the  article credits the first statement about the possible electronic breakdown to ``Navy and Air Force officials'' and the 90 degrees statement to ``Aerial navigators.'' It is unclear whether the navigators in questions are with the US military, but it does not matter regardless.]{the_new_york_times_soviet_1950}
+
+% % \begin{figure}
+% % \centering
+% % \includegraphics[scale=0.45]{baltic-flight-path.png}
+% % \caption{USN PB4Y2 Flight Path}
+% % \label{baltic-flight-path}
+% % \end{figure}
+
+% Meanwhile, the United States knew that it hadn't lost a B-29, and that the missing plane probably hadn't ended up in Soviet airspace, but the Privateer in question \emph{had} been performing a questionably legal reconnaissance mission when it was shot down.
+Though the US was likely correct that the plane never flew over Latvia, it certainly got as close as it needed to in order to gather the necessary intelligence about Soviet military installations on the coast.
+No traditional boundaries were violated, but a repurposed Navy privateer loaded with electronic reconnaissance equipment flying close to Latvia looked, to use a less technical term, shady.
+Yet based on the initial exchange of diplomatic protests, neither side was inclined to magnify the issue if they could avoid it.
+
+% The American press had no such reservations.
+\emph{Washington Post} columnist Walter Lippmann wrote that it could not have been ``a local incident brought on by a local commander but [rather] that it was an act of Soviet policy.
+The known facts indicate that Soviet intelligence \textelp{} believed it was carrying important electronic equipment and that orders were given to the Soviet fighter command to intercept it.''\footcite{lippmann_baltic_1950} He speculated that there was no way the plane could have violated Soviet airspace, because wreckage would have been recovered, the plane was too slow, and no commander would have sent it on such a dangerous mission intentionally.
+``If, when, and as the American command were reconnoitering the Soviet military establishments on the Baltic coast,'' Lippmann wrote, ``it would use a plane of a wholly different type.''\footcite{lippmann_baltic_1950}
+
+% A newspaper columnist is not a policymaker, but his argument was likely reflective of the political opinion that Americans held regarding espionage at the time: of course we do these things, we just do them \emph{better} than that---and if the US had been doing the thing that the Soviets claimed it was, then it would have been justified of them to open fire.
+``Baltic Plane Mystery,'' a \emph{Post} article written a few days later, was almost sympathetic to the Soviet commanders.
+``Electronics make the old delimitations for border coastal flights ridiculous.
+A plane flying on a course perfectly legal by standards accepted today might still be engaged in reconnaissance of the first importance that an unfriendly power would try to frustrate.''\footcite{childs_baltic_1950} This is true, both in terms of the mission itself and the technological environment, but he went even further.
+``The only sound and safe assumption is that the Russians have a thorough and far-reaching espionage system.
+And at the same time we must hope that our system, and particularly on the side of counterespionage, is effective.'' Even the press, though outraged at the loss of American life, apparently took it as a given that this kind of espionage is necessary---and expected from both sides.
+
+% The most important consequence of the Baltic Incident was not diplomatic.
+As a direct result of the shootdown, President Truman, who until that point was largely uninformed about these missions, ordered a thirty-day halt to all reconnaissance flights.
+In that time, his office drafted an operating procedure that would balance the strategic need for reconnaissance with its potential diplomatic volatility.\footcite[p.~41]{farquhar_aerial_2015} The resulting memo from the Joint Chiefs of Staff created, with Truman's approval, the Special Electronic Airborne Search Project (SESP), which set clear delineations of responsibility between Navy and Air Force reconnaissance, and established executive responsibility for approving the missions.
+
+% Farquhar cites this as a landmark moment in the history of aerial reconnaissance.
+``No longer would ferret operations be conducted ad hoc by the military services;'' he writes, ``from 1950 onward, reconnaissance operations attracted Presidential attention and played a significant role in shaping U.S.
+foreign policy.''\footcite[p.~42]{farquhar_aerial_2015} The strict procedures that SESP established were crafted to avoid the risk of escalation, while still explicitly acknowledging the ``serious disadvantages accruing to the United States if the cessation of these operations were to be extended over an excessively long period.''\footcite{bradley_memorandum_1950} But the SESP procedures also contain a curious provision: ``Flights will not be made closer than twenty miles to the USSR or USSR- or satellite-controlled territory.''\footcite{bradley_memorandum_1950} This rule would not be followed---as is typical of espionage, the consequences of the activity never outweigh its strategic value.
+
+% \subsection{Routine resolution}
+% Between the shootdown of the Navy Privateer in 1950 and the famous U-2 shootdown in 1960, the United States lost nine more aircraft to Soviet air defenses.
+Without analyzing each one at the same level of detail, I can demonstrate how the general attitudes displayed in response to the Baltic Incident became normalized.
+In each of the incidents listed, the pattern is the same---each side released dueling statements of protest, and the situation continued more or less as before.
+
+% Describe here are the next few peacetime reconnaissance shootdowns whose diplomatic consequences Farquhar analyzes.
+A B-29 that went missing on October 7, 1952 hit the front page of the \emph{New York Times} two days later.
+This mission is on Peterson's list as a ferret flight.
+Farquhar notes that the incident happened during the US election season and that the media thought the attack was intended to lower American prestige, but he does not mention any diplomatic consequences.\footcite[p.~43-44]{farquhar_aerial_2015} The official US response, however, is preserved on the State Department website.
+As with the Baltic Incident, the US simply demanded that the Soviets pay for the plane and return any survivors.\footcite{the_new_york_times_u.s._1952} Two more shootdowns in March 1953---one involving an American F–84 Thunderjet and the other an RAF Lincoln bomber---actually resulted in conciliatory responses from both sides, including a secret meeting to reduce aerial tensions ``of which little is known''.\footcite[p.~45]{farquhar_aerial_2015}
+
+% In another incident on July 29, 1953, the United States protested the shootdown of a B-50 (reconnaissance flight) over the Sea of Japan.
+The Soviet Union responded by protesting the alleged American shootdown of a Soviet passenger plane.
+Despite press outrage, no further action was taken.\footcite[p.~47]{farquhar_aerial_2015} A September 4, 1954 shootdown of a Navy P2V Neptune actually led to a call by one US Senator to break off diplomatic ties with the Soviet Union: ``Just another note from our State Department to the Kremlin hierarchy will not impress these uncivilized rulers \textelp{} that this new attack upon an American plane confirms Communist arrogance and aggressiveness to a point where the breaking of diplomatic relations is justified.''\footcite{the_associated_press_ending_1954} Instead of a harsher punishment, Eisenhower brought the issue to the UN with the full knowledge that an unfavorable judgment would be vetoed by the Soviets.
+The move demonstrated Eisenhower's ability to resist the calls for harsher action while still placating the American public.\footcite[p.~47]{farquhar_aerial_2015}
+
+% Without belaboring the point, suffice to say that Farquhar concludes that after this moment in 1954, the resolution of these incidents became less tense.
+``International incidents posed by shoot downs of reconnaissance aircraft still acted as a barrier in the path of d\'etente; but, by late 1954, overriding strategic concerns dictated a move toward breaking the cycle of hostility.''\footcite[p.~49]{farquhar_aerial_2015} In terms of media coverage, and diplomatic significance, no shootdown that followed would require more significant consequences than the ones that had already happened.
+
+% One incident in the period between 1950-1954 does not fit the broader pattern.
+On November 20, 1951, an American C-47 transport was forced down in Hungary.
+This incident is omitted from Peterson's list because, despite Soviet cries of ``spies and saboteurs,'' there is no evidence that it was a reconnaissance flight.\footnote{It also wasn't exactly shot down so much as forced to land.} Of all the cases that Farquhar examines, this is the only one that provoked a sustained diplomatic response from the United States---and it is the only case where the plane in question definitively carried no espionage equipment.
+While the ferret flights were perfectly calibrated to fit as much radio surveillance equipment as possible, the only evidence of espionage that the Soviets were able to produce from the C-47---a plane that they recovered intact---was a portable radio, two extra parachutes, and some packets of warm blankets.\footcite{the_united_press_soviet_1951}
+
+% I contend that the lack of espionage equipment is precisely the reason the United States made a sustained public effort to recover the four airmen, which is something that almost never happened with ferret flights.\footnote{While some of the flights were shot down overseas, making recovery difficult, I am referring here to situations where the US was aware that the Soviets likely held their airmen in captivity and still made little effort to get them back.} Farquhar summarizes the American response to this incident as follows: ``Responding to the press attention, the Truman Administration acted swiftly, attempting to gain the fliers release through diplomatic pressure.
+The President ordered the Hungarian consulates in New York and Cleveland closed and banned private travel to the country.
+Legislatively, Truman asked Congress to pass a \$100 million Mutual Security Act to aid `selected persons residing in Soviet bloc states or refugees who wanted to form armed units' in opposition to Communism.''\footcite[p.~43]{farquhar_aerial_2015} Truman immediately took action against material Hungarian interests in the United States and threatened to empower \emph{armed dissidents} if the prisoners were not released.
+The contrast between this response and the one from just a year prior, when the US demanded ``a prompt and thorough investigation'' of the Baltic Incident, is stark.
+
+% We also know that the United States conducted many more overflights than the failed ones that are listed here, using a variety of retrofitted planes.
+In 2001, the Air Force held a symposium honoring and recollecting the pilots who flew the overflights on Soviet territory in the early Cold War, an effort in which they were joined by the British in 1954.\footcite[p.~v]{hall_early_2003} Many of the presenters spoke about entire reconnaissance projects that apparently didn't even register diplomatically.
+Project Heart Throb, which outfitted RB-57s with surveillance equipment, flew between 15 and 19 missions over Eastern Europe from 1955 to 1956, in the recollection of Gerald E.
+Cooke, a Air Force pilot assigned to the project.\footcite[p.~194]{hall_early_2003} Another project from around the same time, Slick Chick, set up F-100 jets with rapid-fire 20mm cameras.
+The pilot who presented this project at the symposium, Cecil Rigby, personally flew two such missions and estimated that there were six Slick Chick missions total.\footcite[p.~176]{hall_early_2003}
+
+% The USSR knew about these missions, which were tracked by radar and often chased by MiG fighters.
+Unlike some of the flights that were shot down, these overflights were clear violation of their territorial sovereignty.
+We will likely never know exactly how many such missions were flown.
+The Truman and Eisenhower administrations were shockingly unsuccessful at recovering the pilots who were held as POWs by the Soviets.
+Dino Brugioni, a senior CIA officer involved in the aerial imagery program, paints a grim picture of the institutional accountability involved when the flights went bad.
+In meetings between the State department and high-level Soviet diplomats, the subject of captured pilots ``was broached only perfunctorily in relation to other things being discussed.''\footcite[p.~72]{brugioni_eyes_2010} Some family remembers of lost airmen received posthumous awards, but most simply received the pilot's personal effects and no explanation.
+``And because these were secret missions conducted by a field command,'' Brugioni writes, ``a change of field commander meant that the fate of the men was soon forgotten.''\footcite[p.~72]{brugioni_eyes_2010} With missions so secret that we abandoned the men who flew them, we can have no expectation of finding a complete record today.
+
+% Even without a complete list of flights, it is clear that the territorial boundaries of the USSR were routinely violated to generate critical intelligence for NATO powers in an early, relatively volatile stage of the Cold War.
+As time passed these flights appear to have become more accepted, not less.
+Diplomats of both nations were, as a matter of policy, able to paper over these issues in service of advancing other interests.
+When the shootdown caused unavoidable diplomatic consequences---such as when it was well-publicized---the countries exchanged the necessary diplomatic cables and engaged in the requisite saber rattling.
+These incidents undoubtedly altered the climate of Cold War diplomacy, but the fact remains that in 10 years of reconnaissance, the only action the Soviets took to keep the United States from conducting these incredibly provocative reconnaissance flights was to shoot them down.
+
+% % Valentino: Can you talk about other ways the Sovierts might have responded.
+What other kinds of actions might have seemed reasonable under different norms?
+Maybe give an example of a Soviet response to something apparently similar on its face, but clearly out of the espionage domain, so the response was much harsher.
+
+% \section{Civilian spy flights}
+% \subsection{Project Aquatone}
+% While Eisenhower was able to defuse some of the tension surrounding the military reconnaissance flights, even the most aggressive overflights could only hope to skirt the borders of Soviet territory; the United States couldn't just fly a B-50 over Moscow.
+What US intelligence-gathering capabilities needed was some way of photographing suspicious installations deep within Russia while still being able to claim the same benefit-of-the-doubt status that prevented even the most contentious overflights from escalating into a larger conflict.
+What they needed was a spy plane.
+
+% The key feature of a spy plane is that it is designed to be minimally provocative.
+Any plane can be retrofitted with cameras or radios that allow it to perform useful reconnaissance functions---what separates a spy plane is that there can can be no doubt (in the mind of the target) that it is being used for anything \emph{but} reconnaissance.
+For the spy plane to serve its intended purpose as a minimally provocative agent of espionage then, two things must be true: its operation must be understood by adversaries to be non-military, and that distinction must meaningfully alter their threat perception and associated response.
+To build such a aircraft, Eisenhower approved Project Aquatone, the code name for the creation of the U-2.
+
+% From its conception, the U-2 was purpose-built to be read as a civilian aircraft.
+Eisenhower required that the pilots of U-2 aircraft be CIA employees, specifically forbidding uniformed Air Force officers.\footcite[p.~33, Though many of the pilots did have Air Force backgrounds.]{lindgren_trust_2000} The CIA also had operational control of the program instead of the Air Force.
+``I want this whole thing to be a civilian operation,'' Eisenhower wrote, when settling an operational dispute between the two departments.
+``If uniformed personnel of the armed services of the United States fly over Russia, it is an act of war---legally---and I don't want any part of it.''\footcite[p.~60.
+The original source for this quote is an \emph{OSA History} that requires codeword clearance.
+It is quoted here by the History Staff of the CIA.]{pedlow_central_1992} This allowed the United States to ``truthfully deny,'' in the phrasing of the CIA, that any US ``military planes'' had flown over the USSR---which they had to do when the inevitable Soviet protest notes were filed after U-2 overflights began in 1956.\footcite[p.~109]{pedlow_central_1992}
+
+% Having the CIA run the program---ostensibly for the purpose of deception---changed the nature of the program itself.
+If the only goal had been to achieve a level of plausible deniability in the event of a shootdown, then simply not wearing an Air Force uniform during the mission likely would have been sufficient.
+The ``weather reconnaissance'' excuse had even been used before with the ferret flights.\footcite[p.~45]{farquhar_aerial_2015} At the highest levels of American government, however, policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation---even if the NASA weather-craft cover didn't hold.
+``It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft'' read a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} Both halves of that equation were equally important.
+The Russians must believe that these operations are peacetime civilian operations, and they must be correct about it.
+
+% The United States was aware how provocative these reconnaissance flights might be, and Eisenhower was always concerned, correctly, about ``the terrible propaganda impact that would be occasioned if a reconnaissance plane were to fail.''\footcite[p.~162]{pedlow_central_1992} Starting in 1958, the American administration ordered a drastic decrease in the number of overflights.\footcite[p.~51]{powers_operation_2004} But Eisenhower was also under tremendous political pressure to learn more about the Soviet missile program, as hawkish senators amplified fears of a so-called ``missile gap,'' concerned that Soviet ballistic missile capabilities were advancing faster than their own.\footcite[Fears of this missile gap quickly followed earlier fears of a ``bomber gap,'' which ironically the U-2 had been critical in disproving.]{licklider_missile_1970} He especially did not want to affect the upcoming Four Powers diplomatic summit in Paris, scheduled for the summer of 1960.
+Nonetheless, Eisenhower ordered two more overflights in 1960, one on April \nth{9} and the other on May \nth{1}.
+Only one of them returned.
+
+% % It turned out however, that the Soviets opted to lodge their diplomatic protests privately, and soon they stopped protesting them at all.\footcite[p.~42]{lindgren_trust_2000}
+
+% \subsection{The U-2 Incident}
+% On the morning of May 1, 1960, Soviet Air Defense forces detected a high-altitude aircraft flying over Soviet Tajikistan.
+Though they were were not yet able to conclusively identify it, the foreign aircraft was an American U-2 spy plane performing an reconnaissance overflight of the USSR---at a time when the Soviets were deeply embarrassed by their inability to prevent such flights.\footcite{orlov_u-2_2007} The previous U-2 overflight on April 9, less than a month prior, had lasted a full six hours and the aftermath caused an upheaval in the Soviet military.
+Leadership ordered an investigative commission to root out the shortcomings of the Air Force and Air Defense agencies, and a series of charts were drawn up anticipating the routes of future U-2 flights.
+The next time one came through, the Soviets were absolutely determined to shoot it down.
+
+% % \begin{figure}
+% % \centering
+% % \includegraphics[scale=0.3]{powers-flight-path.jpg}
+% % \caption{Operation Grand Slam Flight Path}
+% % \label{powers-flight-path}
+% % \end{figure}
+
+% The Soviets still proved unable to immediately stop the flight on May \nth{1}.
+A missile battalion in the plane's path was not on alert duty.
+Armed fighter aircraft were in the wrong positions.
+A frantic attempt to have another fighter plane literally ram it out of the sky was scrapped when the pilot failed to make visual contact.\footcite{orlov_u-2_2007} The highest possible levels of Soviet leadership were actively involved with the mission as it was taking place.
+A Soviet Colonel in the USSR Air Defense later recalled that Khrushchev ``clearly viewed the violation of their nation's skies by a foreign reconnaissance aircraft on the day of a Soviet national holiday, and just two weeks before a summit conference in Paris, as a political provocation.'' \footcite{orlov_u-2_2007}
+
+% Why did the generally cautious Eisenhower authorize two overflights in such close proximity to a crucial summit?
+For one, the secrecy of the program worked against him.
+Eisenhower was not willing to make the U-2 program public, so he couldn't convince his critics that he had extremely high quality intelligence that no such missile gap existed, because to reveal the existence of his aerial photography would prompt questions about its origins.\footcite[p.~51]{lindgren_trust_2000} And the missions had been such a tremendous success that his reluctance to authorize more overflights grated his staff, especially the influential Director of Central Intelligence (DCI), Allen Dulles.\footcite[p.~354]{brugioni_eyes_2010}
+
+% Another crucial factor emboldened Eisenhower and his team---most of whom, like Dulles, were much more enthusiastic about overflights than he was---to conduct that second mission.
+The USSR had been aware of the April 9 flight, and the United States \emph{knew} that they were aware of it, thanks to an onboard computer that had detected the Soviet tracking from the beginnings of the operation.\footcite{pedlow_central_1992} From the perspective of the American administration, the lack of a protest note after the previous flight was a sign that the USSR was not particularly offended by the overflights.
+That was a mistake.
+``It is one of the many unfortunate aspects of the high degree of secrecy surrounding Moscow's actions,'' imagery analyst David Lindgren writes, ``that the United States never understood the how angered and frustrated Soviet leaders were made by the U-2 overflights.''\footcite[p.~52]{lindgren_trust_2000} After years of bluster about the superiority of Soviet military forces, Khrushchev was not willing to admit that his Air Defense had been unable to shoot an unarmed spy plane flying deep into his homeland.\footcite[p.~59]{powers_operation_2004}
+
+% Almost four hours into the spy plane's flight on May 1, pilot Francis Gary Powers felt a dull ``thump'' and ``a tremendous orange flash lit the cockpit and sky'' as a Soviet surface-to-air (SAM) missile exploded behind him, ripping the wings off his plane and sending it into a tailspin.\footcite[p.~61]{powers_operation_2004} His subsequent capture by the Soviet military became an immediate international incident.
+Three days after NASA quietly reported that they had lost a U-2 type weather reconnaissance plane, Khrushchev announced that the USSR had shot down an American plane that had flown into their airspace, which he called ``an aggressive provocation aimed at wrecking the Summit Conference.''\footcite[p.~112]{powers_operation_2004} Once Khrushchev gleefully debunked the resulting American cover story by revealing that they had taken Powers alive, Eisenhower admitted that the Gary Powers had been flying a spy plane, in a press conference where Eisenhower memorably declared that espionage was a ``a distasteful but vital necessity'' to prevent another attack like Pearl Harbor.\footcite{eisenhower_news_1960}
+
+% Publicly, the USSR was not ready to accept that US overflights of its territory were a ``distasteful but vital necessity.'' The Four Powers summit in Paris was a disaster.
+Khrushchev launched into an angry rant on the first day and stormed out, immediately squashing any hope that pressing issues like the fate of West Berlin would be resolved there.
+He also canceled a planned visit by Eisenhower to the USSR that was scheduled to take place the next month.
+A contemporary newsreel about the conference said that ``in the course of two hours, Khrushchev brought US-Soviet relations to their lowest point since the end of World War II.''\footcite{universal_studios_summit_1960}
+% % The contrast between the severity of response Khrushchev wanted to portray publicly and what the USSR actually did is highlighted by his own son Sergei, who edited and annotated the 2007 edition with notes featuring additional context for his father's statements.
+
+% \section{Espionage in the open skies}
+% \subsection{Minimal consequences with maxmimal noise}
+% The U-2 Incident is often cited as a major failure of the Eisenhower administration.
+In the traditional telling, it significantly worsened US-Soviet relations at time when it seemed like lasting peace might be possible.
+I disagree; instead, I argue that that the U-2 Incident should be understood as an embarrassing but otherwise typical failure of an intelligence operation, one consistent with the minimal consequences typically applied to espionage.
+
+% No one disputes that the collapse of the Paris Summit was directly attributable to the Gary Powers shootdown but, frankly, summits just aren't that important.
+From 1953 to the end of the Cold War in 1991, there were 23 total US-USSR summits.\footcite{fain_chronology_2011} The main issue that did not get resolved in 1960 was the status of Berlin.
+It also did not get resolved at the next summit in 1961, a comparatively warmer affair between Khrushchev and Eisenhower's successor, John F.
+Kennedy.
+If the worst that happened because of the U-2 incident was an unproductive summit meeting, then that is a relatively minor punishment.
+With the late rollout of the captured pilot, the dramatic scene at the summit, and the cancellation of Eisenhower's visit, Khrushchev publicly embarrassed the United States without meaningfully altering Soviet policy towards it.
+
+% What would a major punishment have looked like?
+One proportional response Khrushchev could have considered was to force the issue on regional air bases.
+Since the US was using nearby bases to launch invasive reconnaissance missions, the USSR could have demanded that the US either close some of them or agree to halt surveillance flights.
+This likely would have involved some sort of trade, but it is not an unreasonable hypothetical.
+The Soviets were incredibly preoccupied with removing the forward-deployed Jupiter missiles from Turkey.
+Although the missiles were deployed to bases well within NATO territory, and arguably less threatening to Soviet security than American overflights of the Soviet bloc, the existence of these missile became a central bargaining chip in the Cuban Missile Crisis.
+By contrast, the Soviets simply did not have the same level of interest in curtailing reconnaissance flights.
+That both sides of the Cold War had strong incentives to avoid military escalation is well-documented, but at least threatening military retaliation for launching provocative military operations would not seem to have been out of place.
+
+% While I argue that the response to the U-2 flight was within the reasonable bounds for espionage, I acknowledge that it was still unprecedented in the context of reconnaissance flights.
+There are a number of reasonable explanations for why the U-2 shootdown escalated.
+Partially the Soviets felt humiliated by a technology for which they had no countermeasures and the uniquely insulting way the overflights flew deep into Soviet territory.
+Eisenhower thought that Khrushchev might have been looking for an excuse to cancel his visit to Russia.\footcite[p.~555]{eisenhower_waging_1965} Others in his administration believed that the U-2 Incident was an excuse to justify an intensification of hostilities that had already been decided by hard-liners in the Kremlin.\footcite[p.~328]{kistiakowsky_scientist_1976}
+
+% Most scholars who study this period argue that the cumulative tension created by these flights contributed to an overall chilly tone in US-Soviet relations.
+Farquhar, for instance, refers to this as the ``cycle of hostility,'' in which the series of aerial incidents increased suspicions of military buildup on both sides, leading to even more aerial incidents as both powers attempted to verify their suspicions.\footcite[p.~43]{farquhar_aerial_2015} While it's true that the there were many aerial incidents, it is not at all clear how this affected any other aspect of US-Soviet diplomacy.
+I can't definitively say that there wasn't some diplomatic issue that would have gotten resolved at a summit absent these reconnaissance flights, but I can't locate an obvious one either.
+
+% Consider the period of time over which these flights took place.
+If aerial reconnaissance had risen to anything more than a lingering irritant, then it should have to have been resolved in some way before 1960.
+The US-Soviet relationship saw huge shifts, including periods of optimism, over the 11 years preceding the U-2 incident.
+During this period, hundreds of flights took place in contentious territory---many of them purposefully violating Soviet boundaries, causing some of them to be shot down---and the most significant consequence was a failed summit meeting, after which the flights continued on almost as before.
+Both sides were so willing to contain the issue that many of the airmen never saw their families again.
+
+% \subsection{The U-2 after Powers}
+% When the US really needed intelligence that only an overflight could provide, the diplomatic consequences that the USSR had imposed were not sufficient to discourage it.
+Two months to the day after Powers was shot down, a Soviet MiG-19 opened fire on an American RB-47H and captured the plane's navigator and co-pilot.
+While Powers was serving a 10-year sentence in Soviet prison, the two airmen were returned to the US in less than six months with only mild fanfare.
+Khrushchev wrote in his memoir that he wanted to continue ``our general line of peaceful coexistence'' and cited the resolution of this incident as an example of such, because the US was forced to make a formal request for the return of the airmen.\footcite[p.~256-257]{khrushchev_memoirs_2007} The only concessions the US made in return were to announce the discontinuation of its U-2 overflights (which Kennedy was already committed to) and to not make an issue of the illegal detention of the pilots.\footcite{time_cold_1961} For spy flights after Powers, the USSR simply took the necessary immediate countermeasures and shot down the planes that it could.
+Just as with traditional espionage, the countermeasures did not punish the attempt.
+
+% Until the day he died, Khrushchev greatly exaggerated the lasting effect of the U-2 Incident.
+He wrote in his memoir that ``the commander of American forces in West Germany gave the order not to fly any closer than 50 kilometers from the border between East and West Germany.
+And no more incidents of that kind occurred.''\footcite[p.~256]{khrushchev_memoirs_2007} His son Sergei, who annotated the memoir, admitted that this statement was false: ``In practice such incidents occurred again from 1961 onward, but instead of the U-2 the Americans now used various types of Phantom or SR-71.''\footcite[p.~258]{khrushchev_memoirs_2007} On at least two occasions after the U-2 Incident, an American overflight crossed over into Soviet territory and the Soviets responded with...
+a diplomatic protest.
+The US quickly apologized, and that was that.\footcite{orlov_u-2_2007}
+
+% The U-2 did not disappear after the Powers flight.
+A recently declassified CIA report details the many locations in which U-2 photography was employed to gather intelligence after 1960, including China, India, Indonesia, Thailand, Tibet, Laos, North Vietnam, and Venezuela.
+Lyndon Johnson even clarified that his predecessor's order to end U-2 flights over the Soviet Bloc was not indefinitely binding---it was valid only until countermanded.\footcite[p.~195]{pedlow_central_1992} As it turned out, it never became necessary to resume Soviet Bloc overflights because satellite photo-reconnaissance soon replaced them; U-2 overflights were rendered mostly obsolete.
+There is, however, one use for which satellites will likely never displace aerial photo-reconnaissance---a situation like the Cuban Missile Crisis, where events are so time-sensitive that intelligence is needed in a matter of hours.
+
+% On October 15, 1962, U-2 photography revealed that the Soviet Union was assembling missiles in Cuba that were capable of carrying nuclear warheads.
+As the crisis unfolded, the United States continued to send U-2 planes over Cuba in order to maintain a close watch on the development of the missile sites.
+Kennedy's team decided that if a U-2 were to be shot down, the most likely SAM site would be immediately fired upon.\footcite[p.~217]{caro_passage_2013} Then, on October 27, it happened---a U-2 was shot down flying over Cuba and the pilot was killed.
+Kennedy flinched; he had just sent Khrushchev an offer that might lead to peace, and he didn't want to jeopardize it.
+Against the advice of administration hard-liners, including Vice President Johnson, Kennedy refused to authorize the strike against the Cuban SAM site.\footcite[p.~220]{caro_passage_2013} The next day, Khrushchev agreed to Kennedy's terms, and the Cuban Missile Crisis ended with only a single casualty---the U-2 pilot, Major Rudolph Anderson.
+
+% As his title indicates, Rudolph Anderson was an Air Force Officer, not a member of the CIA.
+During the crisis, the President transferred control of Cuban U-2 reconnaissance to the Air Force, and authorized it to fly as many missions as necessary over the island.\footcite[p.~208-209]{pedlow_central_1992} Khrushchev did not say much about the Cuban U-2 in his writing, but he was very clear on one point: he did not order the SAM strike, and was worried that Kennedy ``might not be able to absorb this blow.''\footcite[p.~338]{khrushchev_memoirs_2007} Shooting down an American aircraft risked escalation, and that was not something Khrushchev wanted.
+This was the second time in history that Soviet forces had shot down a U-2 plane---but even though the spy plane was flying over Cuban airspace and piloted by an US Air Force Major, it was the Soviets who were worried that shooting it down had been an unnecessarily aggressive act.
+
+% \subsection{Bridge of spies}
+% The incidents involving U-2 aircraft during the Cold War perfectly mirror the traditional consequences for espionage because the United States succeeded in applying espionage norms to its use.
+Eisenhower's insistence on having the CIA manage Project Aquatone had the intended effect: the spy plane was, in story as well as fact, a civilian intelligence operation.
+
+% Even when a U-2 operation went poorly, it still played by the rules of the game.
+Powers opted against using his suicide capsule, but the macabre truth about the U-2 is that pilots weren't supposed to survive a crash at all.
+The United States originally publicized the weather craft story because Eisenhower had been assured that Gary Powers could not possibly be alive to tell the Russians the truth.\footcite[p.~35]{lindgren_trust_2000} When Powers confessed to his Soviet captors, he told them, correctly, that he was a civilian pilot with the CIA.
+His cover had been blown, but underneath it he was still a spy, and in pretty much all respects he and the U-2 were treated accordingly.
+
+% Both heads of state made clear in their writings that they considered Powers to be spy.
+On capturing Gary Powers, Khrushchev wrote: ``This was a hostile act by the leaders of the U.S.
+government, and they made no attempt to conceal it.
+They didn't think we had the capability of \textelp{} acquiring irrefutable proof that the United States was using methods that were impermissible in peacetime.''\footcite[p.~239]{khrushchev_memoirs_2007} Sergei's annotations once again contradicted his father's claim that these methods were impermissible; more than forty American reconnaissance aircraft were shot down by the USSR during the Cold War.
+As is often the case with espionage, Khrushchev wanted to make it seem like he was standing up for the integrity of the Soviet state without endangering the espionage equilibrium.
+
+% Eisenhower, always an insightful thinker on issues of intelligence, understood that beneath the grandstanding, the fundamentals of international espionage still applied. In his own words:
+
+% % \begin{quote}
+% 	Furthermore, it seemed to me that Mr.
+Khrushchev's outbursts were hypnotizing the world with a passionate but highly distoreted presentation of one particular phase of international espionage.
+His government had been so notoriously involved in spying, especially in the United States, as to dwarf our activities, but by separating this particular type of espionage from all others he hoped to make convincing his charge of ``warmongering.'' To claim that because the equipment employed was an airplane with a camera, and therefore provocative of war, was plain silly, and I felt it necessary that the matter be put in perspective.
+\\
+
+% 	The real issue at stake was not the fact that both sides conducted intelligence activities, but rather that the conduct and announced intentions of the Communists created the necessity for such clandestine maneuvers.
+As a consequence, the West---more specifically the Untied States as the major military power of the West---had to maintain constantly the capacity to detect any possible prelude to an infinitely more destructive Pearl Harbor.\footcite[p.~551]{eisenhower_waging_1965}
+% % \end{quote}
+% The next chapter examines in depth how the fear of a nuclear Pearl Harbor affected the entire American intelligence strategy, but even in the limited context of the spy plane, it is clear that Eisenhower deployed espionage to de-escalatory effect.
+Khrushchev had convinced the world that the Soviet missile capabilities were far more advanced than they actually were.
+Eisenhower was under tremendous pressure from the public, Republicans in Congress, and even his own staff to increase the defense budget.
+The only thing standing between Eisenhower and an accelerated arms race were detailed photo-reconnaissance reports from the U-2 that showed the Soviets were developmentally far behind what their public statements suggested.
+
+% The peaceful uses of aerial reconnaissance were formalized at the end of the Cold War, when members of NATO and the Warsaw Pact signed the ``Treaty on Open Skies,'' which established guidelines for unarmed aerial surveillance flights in the spirit of open information and de-escalation.\footcite{organization_for_security_and_co-operation_in_europe_treaty_1992} This treaty was almost identical to the one first suggested by Eisenhower in 1955.\footcite{center_for_arms_control_and_non-proliferation_fact_2017} With certain restrictions, the type of flight that altered the course of the Cold War is now officially routine and permissible.
+In February 1962, Francis Gary Powers walked across a bridge into West Berlin at the same time as Russian Colonel Rudolph Abel, an American prisoner.
+Powers' father had suggested to the State Department that they might be traded, spy for a spy.\footcite[p.~239]{powers_operation_2004} ``This,'' Eisenhower wrote, ``was a tacit admission by Khrushchev that our `outrageous' U-2 pilots have their opposite numbers operating within the borders of this country.''\footcite[p.~558]{eisenhower_waging_1965}
+
+% The Powers flight and Project Aquatone channel the norms of espionage in one final, but crucial respect: their strategic value easily outweighed the consequences of being caught.
+No one understood this better than President Eisenhower, who, when he was occasionally asked about the wisdom of the U-2 flights, would always reply: ``Would you be ready to give back all of the information we secured from our U-2 flights over Russia if there had been no disaster to one of our planes in Russia?'' In his telling, he never received an affirmative response.\footcite[p.~559]{eisenhower_waging_1965}
+
+% \section{Spy satellites}
+
+% \section{Conclusion}
+
 \end{document}

### 2019-12-19 article.tex
Clean up baltic incident

@@
 
-% \section{Aerial espionage}
-% \section{Introduction}
-% According to \emph{The New York Times}, the very first reconnaissance flight took place in 1911, only eight years after Kitty Hawk, when ``Lieut.
-Piazzi today, for the first time in the history of warfare, made from this place an aerial reconnaissance against a hostile power'' surveilling the Turkish Infantry near Tripoli.\footcite{special_cable_to_the_new_york_times_air_1911} The ability to fly over an enemy's position to gain intelligence about their movements---in a craft orders of magnitude more versatile than a balloon---was so obviously valuable that countermeasures quickly followed.
-Early aerial dogfighting famously took the form of pilots firing handguns at each other out of open cockpits in mid-air (emphasis on attempted), but by the time of the First World War, planes were already outfitted with synchronized machine guns and ground regiments equipped with anti-aircraft weaponry.
-Crucial information about military movements was going to come at a cost.
+\section{Aerial espionage}
+The line between a \enquote{cold} and a \enquote{hot} war is supposed to be obvious.
+A superpower's own uniformed military personnel opening fire on the other superpower's military or territory would be \enquote{hot,} everything else is \enquote{cold.} Broadly speaking, history remembers the Cold War as lacking direct military confrontation between the United States and Russia; though the two superpowers of the Cold War were open rivals, they supposedly competed for influence by funding, arming, or otherwise influencing proxies, rather than attacking each other directly.
+That impression is not exactly accurate.
+Not only did the US and USSR often come close to openly attacking each other, but a surprising number of times they actually did—--including one of the Cold War's most famous moments.
 
-% Even though these aircraft were performing vital intelligence-gathering functions, that did not make them spy planes.
-Aerial reconnaissance first appeared in the press in 1911, but the term ``spy plane'' appeared in the pages of \emph{The New York Times} just once before 1960.\footnote{According to my search of the archives.
-The instance mentioned here was in 1944, when the AP wrote that ``Germans tried to probe the secrets of the gathering Allied storm by E-boat patrol dashes [...] and by spy-plane coastal raids.''(\cite{the_associated_press_britons_1944}) In context of the article, though, it's clear that they are referring to military aerial reconnaissance, especially since it was written during the Second World War.
-This is a marked difference from a plane created for peacetime intelligence, which is what the term ``spy plane'' has since come to mean.} The image of a spy plane crashed into the public consciousness that year when Gary Powers crawled out of a U-2 class aircraft over Soviet Russia.
-The downed craft was clearly not a generic fighter plane that had flown too close to the border, and it suddenly became necessary to categorize this new type of plane, flying at previously unheard-of altitudes, that was purpose-built for clandestine photography.
-The United States first claimed that the lost plane was a ``weather craft.''\footcite{caruthers_soviet_1960} When that cover didn't hold up, a new story was born.
-The diplomatic fallout of the U-2 shootdown, which threatened to torpedo an upcoming NATO-Warsaw summit, had become the ``spy-plane case.''\footcite[This is the second time that ``spy plane'' as a term of art appeared in \emph{The New York Times}.
-There would be many more.]{reston_allies_1960}
+% citation needed on the definition of cold
 
-% I emphasize the linguistic distinction between a military plane used for reconnaissance and a ``spy plane'' because that distinction has a major impact on how the flight is perceived and what diplomatic repercussions might result when it is caught.\footnote{That is not to say there is a hard boundary between military and non-military intelligence, either in purpose or collection.
-The overlap between military and civilian intelligence agencies---especially in the United States---is considerable.
-The Air Force was deeply involved with the U-2 Project, for example, even though it was technically under the jurisdiction of the CIA.
-The aerial photographs taken by U-2 missions were used to inform both military and civilian operations.} A military fight carries with it the implications of combat, aggression, and the possible prelude to an attack.
-A spy flight is instead a tool of civilian intelligence, whose purposes are often frustrated but never quite discouraged.
-The creation of the U-2 plane---a craft exclusively for civilian intelligence---is a pivotal moment in the history of espionage because the U-2 is the physical embodiment of the espionage norms it helped to define.
-Its \emph{raison d'\^etre} is an activity which is not permitted---or its altitude would be unnecessary---but its design emphasizes its peaceful reconnaissance purpose.
+Pilot Francis Gary Powers knew that U-2 overflights of the Soviet Union were incredibly risky.
+The Soviets were deeply unhappy that the US was violating their airspace in such a brazen fashion, but for years they had kept quiet about the U-2 missions, embarrassed that they were technologically incapable of intercepting the high-flying aircraft.\footcite[p.~59]{powers_operation_2004} When Soviet air defenses finally managed to shoot one down in 1960, Soviet leadership was equal parts triumphant and furious.
+At a planned summit in Paris only a few weeks later, Khrushchev launched into a rant about the U-2 planes and stormed out on the first day, effectively ending the summit---and any hope that pressing issues like the fate of West Berlin would be resolved there.
+He also canceled a planned visit by Eisenhower to the USSR that was scheduled to take place the next month.
+A contemporary newsreel about the conference said that \enquote{in the course of two hours, Khrushchev brought US-Soviet relations to their lowest point since the end of World War II.}\footcite{universal_studios_summit_1960}
 
-% \section{The Open Skies policy}
-% From the beginning of the Cold War, aerial intelligence played a crucial role in American foreign policy.
-The United States considered aerial surveillance so important that it was willing to allow the Soviets to do it in return.
-President Eisenhower formally proposed a mutual inspection regime called Open Skies to the Soviet Union as a means of reducing security tensions.
-In his prepared remarks at the Geneva Summit of 1955, Eisenhower laid out the following plan:
+The shootdown of Gary Powers, what would later come to be known as \enquote{The U-2 Incident}, clearly violated the only \enquote{rule} of the Cold War.
+Official Soviet air defenses shot down a US aircraft.
+All parties involved had cause to escalate military---the Soviets for the violation of their airspace, the US for the attack on their \enquote{weather craft}---but instead they chose to treat the U-2 missions as an intelligence operation, not military action.
+The Soviets therefore applied the traditional consequences associated with espionage; Khrushchev engaged in some saber-rattling and denounced the violation of his airspace, but no material actions were taken, and the United States escaped the incident with few tangible consequences.
+The striking disparity between the diplomatic furor and the lack of material consequences can be explained by an apparent technicality: the U-2 was not a military aircraft, it was a spy plane.
 
-% 	% \begin{quote}
-% 	\textbf{To give to each other a complete blueprint of our military establishments}, from beginning to end, from one end of our countries to the other, lay out the establishments and provide the blueprints to each other.
-% 	\newline
+No single attribute defines a spy plane.
+A military aircraft can forgo weapons but, by virtue of its pilot, its mission, its timing, or its branding still remain in its operation unquestionably military.
+Inversely, intelligence missions can be flown by uniformed Air Force pilots in re-purposed B-52s and somehow still be read as espionage.
+A spy plane is an ontological designation, one that reveals itself to the adversary through a series of social and physical signifiers that are not replicable across time.
+Germany in 1937 would not have reacted to a U-2 overflight the same way that the Soviet Union did in 1960.
+But even if the way that different states handled espionage was perfectly consistent, there still wouldn't be a checklist that guarantees a mission will treated as espionage; it is not enough make a military mission look like an intelligence operation.
+The reality is both simpler and more frustrating---it has \emph{to be one}.
 
-% 	Next, to provide within our countries facilities for aerial photography to the other country---\textbf{we to provide you the facilities within our country, ample facilities for aerial reconnaissance}, where you can make all the pictures you choose and take them to your own country to study, you to provide exactly the same facilities for us and we to make these examinations, and by this step \textbf{to convince the world that we are providing as between ourselves against the possibility of great surprise attack}, thus lessening danger and relaxing tension.
-(emphasis mine)\footcite{eisenhower_president_1955}
-% 	% \end{quote}
+The U-2 incident was the most visible in a series of aerial intelligence operations that the United States undertook in the early Cold War.
+The US did not always employ civilian pilots nor it did always use obviously innocuous aircraft, yet it managed to run a shocking number of missions that violated Soviet airspace without significantly altering the relative level of tension between the two superpowers.
+On occasion, these missions would turn \enquote{hot} when Soviet air defenses shot them down.
+Not once did these shootdowns lead to retaliation from the US, or significant pressure from the Soviets to stop the flights.
+Instead, each one was resolved as if the pilots who survived were simply spies, captured in the normal course of counterintelligence.
+This article seeks to answers how and why states ensure that certain operations reveal themselves as espionage; the decade of covert aerial espionage was one such successful operation.
 
-% The UN-sponsored Geneva Summit, formally the \emph{Conference on the Peaceful Uses of Atomic Energy}, was a comparatively hopeful period in the Cold War, a conference promoting scientific diplomacy where such a cooperative information exchange might have seemed feasible.\footcite[p.~27]{luscher_nuclear_2018} Nonetheless, the scale of this offer was astounding.
-Eisenhower offered to let the the Soviets install operational military bases in the United States from which they could surveil American national defense infrastructure, so long as they allowed the US to do the same.\footnote{Also included were reciprocal nuclear facility inspections, an immediate nonstarter in the USSR, but also controversial within the American administration back home.
-(\cite{prados_review_2015})} In effect, proposed a limited form of state-sanctioned espionage for explicitly defensive purposes---exactly the norm I am arguing exists in practice, albeit in a different form.
+\subsection{Military reconnaissance flights}
+The tense battle between the US and USSR over surveillance of Soviet border installations was never formally acknowledged by either party---and neither were its combatants.
+Over the course of the Cold War, more than 200 American pilots were shot down while spying on the Soviet Union.\footcite{glenshaw_secret_2017} These missions were so highly classified that in many cases the families of the pilots never learned how they died.
+126 of these American airmen remain unaccounted for today, casualties of missions whose existence both the US government and the Soviet Union collaborated for decades to deny.
 
-% The offer was quickly rejected.
-In conversation over buffet that evening, Khrushchev politely countered that the plan would not further disarmament at all because it would merely confirm the fragmentary information that their respective intelligence services already possessed.\footcite{department_of_state_memorandum_1955} The response was telling---though it was in no way an endorsement of the activity, it both acknowledged that espionage was taking place between the two superpowers, and that it served a potentially de-escalatory purpose.
-
-% Khrushchev also claimed that the offer was primarily a propaganda move on the part of the US, but backed down when Eisenhower essentially dared him to call his bluff.\footcite[p.~2.
-This is a secondary source---the book cites a conversation by Eisenhower, quoted in NSC meeting memos that are stored in the Eisenhower library.]{bury_eisenhower_2014} The formal rejection came a few weeks later: ``While giving credit to this proposal's attempt to find a solution,'' USSR Council of Ministers chair Nikolai Bulganin wrote in his report to the Supreme Soviet, ``\textelp{} aerial photography cannot yield the expected results, since both our countries contain limitless expanses in which, if one desires, everything can be hidden.''\footcite{bulganin_meetings_1955} Though a lukewarm excuse, it was not a disdainful one.
-Earlier in the report, he had even said that the conference, as a whole, ``must be considered a definite success of the peace-loving forces.''
-
-% Was the offer genuine, or was it the stunt that Khrushchev accused it of being?
-In a book about the relationship between aerial reconnaissance and Eisenhower's New Look foreign policy, historian Helen Bury argues that a willingness to negotiate acceptable and enforceable agreements with foreign powers was central to his Cold War strategy.\footcite[p.~69]{bury_eisenhower_2014} More importantly, the intelligence that an Open Skies policy promised would have been invaluable in his ongoing effort to control defense spending, where Eisenhower was embattled by deeply paranoid American institutions.\footcite[p.~212]{bury_eisenhower_2014} His commitment to Open Skies in practice, however, is still a point of some historical debate.
-In reviewing her argument, intelligence historian John Prados sees a contradiction between her portrayal of Eisenhower's genuine desire for arms control and Eisenhower's role in significant nuclear buildup.\footcite[p.~233-234]{prados_review_2015} In the following chapter, I will present my own evidence that these two existed as part of the same strategy.
-Eisenhower used photo-reconnaissance---of the kind that an Open Skies policy would have legally permitted---to create an accurate assessment the military capabilities required to sufficiently deter a Soviet nuclear strike; as a result, the United States built far fewer missiles than it otherwise would have.
-
-% % Valentino: Actually, they were explicitly consistent.
-Ike thought that nukes were the cheapest way to counter the Soviet numerical advantage in conventional weapons (which were and are much more expensive).
-So Ike saw both as cost saving measures.
-
-% Regardless of Eisenhower's intent, it is clear that the United States would have far more to gain from such a policy than the USSR, and it made strategic sense for the Soviets to reject the offer.
-Writing about the history of CIA's imaging program, former imagery analyst David Lindgren notes that Khrushchev had the advantage of huge amounts of publicly-available information about US military installations through newspapers, maps, and official aerial photographs, while Soviet restrictions on press freedom denied symmetrical data.\footcite[p.~38]{lindgren_trust_2000} Allowing for an Open Skies policy would also have revealed that Soviet air defenses were not nearly as far along as their leadership claimed.
-Khrushchev's son would later write that his father's overriding concern at the time was to conceal Soviet weakness.\footcite[p.~133]{brugioni_eyes_2010}
-
-% Treaty or not, the United States was desperate for intelligence about the Soviet missile program, and aerial photo-reconnaissance was crucial to gaining it.
-The West's HUMINT capabilities within the Soviet Union were hamstrung by the intense surveillance under which their diplomats were placed, as well as the enormous size of the Russia itself.\footcite[p.~23]{lindgren_trust_2000} As the US increasingly relied on signals intelligence (SIGINT) to gather intelligence, the USSR began pursuing a more aggressive aerial defense policy, demonstrating their willingness to open fire on foreign reconnaissance aircraft.
-The increased cost associated with these missions did little to deter senior US policymakers from attempting to surveil the skies.\footcite[p.~4]{pedlow_central_1992}
-
-% % The Soviet Union was able to successfully destroy US Military aircraft during purported peacetime because in many cases those US aircraft were not supposed to be there in the first place.
-In many cases their flight path was a clear violation of Soviet territorial boundaries---a fact of which the administration was aware.\footcite{goodpaster_memorandum_1956}
-
-% In the rest of this chapter, I will argue that Eisenhower satisfied the need for intelligence that an Open Skies treaty would have provided by normalizing the collection of aerial intelligence and minimizing its diplomatic consequences.
-This process happened in two parts: first, through the routine resolution of military reconnaissance shootdowns, and second, by symbolically divorcing peacetime aerial espionage from military reconnaissance with the creation of a spy plane.
-For both the military and civilian missions, I will demonstrate how the even the most sensationalized incidents resulted in no more than ineffectual diplomatic protests, even though the spy plane could surveil locations that a ferret flight never could.
-
-% \section{Military reconnaissance flights}
-% \subsection{Timeline of early Cold War SIGINT missions}
-% Well before the U-2 Incident in 1960, the United States had been routinely flying military aircraft near Soviet airspace, and the USSR had been routinely shooting them down.
-For a conflict defined by its lack of direct skirmishes between the two superpowers, a surprising number of Americans died in these incidents, which are mostly forgotten to history.
-A Smithsonian investigation found that over the course of the Cold War more than 200 American pilots were lost to Soviet shootdowns---126 of which remain unaccounted for to this day.\footcite{glenshaw_secret_2017}
-
-% The purpose of these missions was to gather SIGINT identifying the location of critical radar installations along the Soviet border.
-Converted bombers---called ``ferrets'' by the Air Force---were outfitted with advanced radio equipment and sent to intercept as much information about Soviet radio transmissions as possible.\footcite[p.~4]{peterson_maybe_1993} According to the CIA, these missions began as early as 1947.\footcite[p.~4]{peterson_maybe_1993} Some of these flights only flew near the border, remaining over traditionally recognized international waters, while others were ``overflights,'' missions that deliberately violated Soviet airspace to collect intelligence at a much higher risk of failure.
+Though they flew Air Force planes, these men were actually conducting one of the earliest forms of Cold War peacetime espionage.
+Their purpose was to gather SIGINT (signals intelligence) identifying the location of critical radar installations along the Soviet border.
+Converted bombers---called \enquote{ferrets} by the Air Force---were outfitted with advanced radio equipment and sent to intercept as much information about Soviet radio transmissions as possible.\footcite[p.~4]{peterson_maybe_1993} According to the CIA, these missions began as early as 1947.\footcite[p.~4]{peterson_maybe_1993} Some of these flights only flew near the border, remaining over traditionally recognized international waters, while others were \enquote{overflights,} missions that deliberately violated Soviet airspace.
 The earliest postwar overflight of the Soviet Union---of which I can find definitive record---took place on May 10, 1949, when two RF-80As surveilled the Kurile islands\footcite[p.~8]{peebles_shadow_2000}
 
-% Determining the quantity of early Cold War reconnaissance flights is difficult because they were a closely guarded secret at the time.
-Some flights, however, were inadvertently declassified---when they were shot down by enemy air defenses.
-Since my goal is to understand the diplomatic repercussions of these flights, an effective place to start is a list of reconnaissance flights that attracted diplomatic attention after being shot down.
-The original shootdown stories are available in the newspaper archives, and the incidents have been helpfully compiled by US Air Force Academy historian Dr.
-John Farquhar, who specializes in aerial reconnaissance and wrote his master's thesis on how it shaped early Cold War diplomacy.
-My choice of incidents to analyze in this chapter is based in part on his work.
-% % \newline
+Without access to classified archives it is impossible to state for certain exactly how many of these flights were conducted.
+Even if one did have unlimited access, some of the necessary information was so highly classified that it is likely lost forever.
+Nonetheless, the civilian researcher is aided by the declassified research of military archivists before them.
+In this paper, I highlight the work of two scholars in particular.
+First, Air Force Academy historian Dr.
+John Farquhar specializes in aerial reconnaissance and its diplomatic consequences.
+Many of the incidents analyzed in this section are taken from newspaper clippings that he compiled and analyzed.
 
-% Additionally, the CIA history magazine \emph{Cryptologic Quarterly} analyzed the Soviet shootdowns of US reconnaissance aircraft in depth, creating a comprehensive (as far as we know) list of each American overflight that was shot down by the USSR (Table \ref{soviet-shootdowns}).
-The article was written in 1993, but not declassified until 2009.
-The author, Michael Peterson, uses two limiting criteria that perfectly suit our purposes: the only flights listed are those that were (a) shot down by Soviet attacks, not Chinese, North Korean, etc., and (b) explicitly US reconnaissance aircraft.\footcite[p.~4]{peterson_maybe_1993} Note that not all of these flights are necessarily overflights; many took place over international waters when they were shot down.
+Second, former intelligence analyst and section chief Michael Peterson wrote an article for the CIA history magazine \emph{Cryptologic Quarterly} that analyzed the Soviet shootdowns of US reconnaissance aircraft in depth.
+The article was written for a classified audience in 1993 and declassified in 2009, providing the public with a comprehensive (as far as we know) list of all American SIGINT flights shot down by the USSR (Table \ref{soviet-shootdowns}).
+Peterson uses two limiting criteria that perfectly suit the purposes of this article: the only flights he lists are those that were (a) shot down by Soviet attacks, not Chinese, North Korean, etc., and (b) explicitly US reconnaissance aircraft.\footcite[p.~4]{peterson_maybe_1993} Note that not all of these flights are necessarily overflights; many took place over international waters when they were shot down.
 In effect, the table catalogs the 13 times in which the USSR shot down an American aircraft designed for collecting intelligence between 1950 and 1964---intercepted espionage attempts.\footcite[p.~5.
@@
 
-% \begin{table}[ht]
-% \centering
-% \begin{tabular}{llr}
-% \textbf{Date}     & \textbf{\makecell[l]{U.S. Service \&\\ Aircraft Type}}   & \textbf{General Location} \\
-% 8 April 1950      & USN PB4Y2 Privateer           & Baltic Sea                           \\
-% 6 November 1951   & USN P2V Neptune               & Sea of Japan                         \\
-% 13 June 1952      & USAF RB-29                    & Sea of Japan                         \\
-% 7 October 1952    & USAF RB-29                    & East of Hokkaido/Kuril Is.           \\
-% 29 July 1953      & USAF RB-50                    & Sea of Japan                         \\
-% 4 September 1954  & USN P2V Neptune               & Sea of Japan                         \\
-% 7 November 1954   & USAD RB-29                    & East of Hokkaido.Kuril Is.           \\
-% 18 April 1955     & USAF RB-47                    & Off Kamchatka Peninsula              \\
-% 10 September 1956 & USAF RB-50                    & Sea of Japan                         \\
-% 2 September 1958  & USADF C-130                   & Soviet Armenia (near Turkish border) \\
-% 1 May 1960        & CIA U-2                       & Sverdolsk, USSR                      \\
-% 1 July 1960       & USAF RB-47                    & Barents Sea                          \\
-% 10 March 1964     & USAF RB-66                    & East Germany
-% \end{tabular}
-% \caption{Summary of Soviet Shootdowns, 1950-1964}
-% \label{soviet-shootdowns}
-% \end{table}
+\begin{table}[ht]
+\centering
+\begin{tabular}{llr}
+\textbf{Date}     & \textbf{\makecell[l]{U.S. Service \&\\ Aircraft Type}}   & \textbf{General Location} \\
+8 April 1950      & USN PB4Y2 Privateer           & Baltic Sea                           \\
+6 November 1951   & USN P2V Neptune               & Sea of Japan                         \\
+13 June 1952      & USAF RB-29                    & Sea of Japan                         \\
+7 October 1952    & USAF RB-29                    & East of Hokkaido/Kuril Is.           \\
+29 July 1953      & USAF RB-50                    & Sea of Japan                         \\
+4 September 1954  & USN P2V Neptune               & Sea of Japan                         \\
+7 November 1954   & USAD RB-29                    & East of Hokkaido.Kuril Is.           \\
+18 April 1955     & USAF RB-47                    & Off Kamchatka Peninsula              \\
+10 September 1956 & USAF RB-50                    & Sea of Japan                         \\
+2 September 1958  & USADF C-130                   & Soviet Armenia (near Turkish border) \\
+1 May 1960        & CIA U-2                       & Sverdolsk, USSR                      \\
+1 July 1960       & USAF RB-47                    & Barents Sea                          \\
+10 March 1964     & USAF RB-66                    & East Germany
+\end{tabular}
+\caption{Summary of Soviet Shootdowns, 1950-1964}
+\label{soviet-shootdowns}
+\end{table}
 
-% I cannot confirm whether Peterson's record is comprehensive.
-Regardless of how thorough declassified CIA information appears, I cannot rule out the possibility that there were other, more secret missions which for whatever reason the United States government has chosen to continue protecting to this day.
+Regardless of how thorough declassified CIA information appears, I cannot rule out the possibility that there were other, even more secret missions which for whatever reason the United States government has chosen to continue protecting to this day.
 Nonetheless, a few aspects of the \emph{Cryptologic Quarterly} article are promising, and suggest that the information is relatively complete.
-First, the article was written for a classified audience as an overview of reconnaissance shootdowns during the 1950s, so it is intended to be comprehensive.
+First, the article was written for a classified audience as an overview of reconnaissance shootdowns during the 1950s, so it is likely intended to be comprehensive.
 Shootdowns are a sufficiently significant event such that it is unlikely the author would have just missed one.
@@
 
-% In the following section, I cross-reference the shootdowns that the CIA says took place with Farquhar's analysis of their media and diplomatic impact.
+To comprehensively analyze the diplomatic effects of aerial espionage, I cross-referenced the shootdowns that the CIA says took place with Farquhar's analysis of their media and diplomatic impact.
 A shootdown is either present in both lists, or just one of the two---some incidents in the CIA list are missing from Farquhar's list because they did not have a diplomatic impact, and some incidents that Farquhar analyzes are not mentioned by Peterson because they were not SIGINT reconnaissance missions and likely (though not definitively) served little espionage function.
-Together, the full spectrum of aerial reconnaissance incidents demonstrates a clear pattern in which the shootdowns that involved espionage were quickly resolved by both parties, while the shootdown of a flight unrelated to espionage produced swift, significant backlash.
+Together, the full spectrum of aerial reconnaissance incidents demonstrates a clear pattern in which shootdowns that were interpreted as espionage were quickly resolved by both parties, while the shootdown of a flight unrelated to espionage produced swift, significant backlash.
 
-% \subsection{The Baltic Incident}
-% The first of these shootdowns, on April 1950, brought the previously-hidden aerial reconnaissance program to the attention of the American people.
+The first of these shootdowns, on April 1950, brought the previously-hidden aerial reconnaissance program to the attention of the American people.
 An unarmed Navy Patrol plane, a PB4Y2 Privateer, was intercepted and downed by Soviet fighters over the Baltic sea.
-The Soviet ambassador to the US lodged a formal note of protest, incorrectly alleging that the US aircraft which had been shot down was a B-29 bomber.\footcite{kirk_ambassador_1950} Upon receiving the note, Secretary of State Dean Acheson remarked that Ambassador Vishinsky's manner was ``serious but not aggressive or antagonistic,'' and the Secretary ``recommended that publicity on our side should be avoided or, if unavoidable, minimized.''\footcite{kirk_ambassador_1950}
+The Soviet ambassador to the US lodged a formal note of protest, incorrectly alleging that the US aircraft which had been shot down was a B-29 bomber.\footcite{kirk_ambassador_1950} A few days later, the American ambassador issued a formal response, in which he correctly identified the model of plane and claimed that at no point did it cross into any territorial waters.
+He demanded from the Soviets \enquote{a prompt and thorough investigation} and that the Soviet Government \enquote{see to it that those responsible for this action are promptly and severely punished.}\footcite{the_associated_press_text_1950} The Soviet Union recognized this for the toothless statement that it was, and rejected the totality of its demands.\footcite{salisbury_kremlin_1950} No further action was taken.
 
-% A few days later, the American ambassador issued a formal response, in which he correctly identified the model of plane and claimed that at no point did it cross into any territorial waters.
-He blamed the Soviet Union directly, outlining the punishment that the Americans planned to impose:
-
-% % \begin{quote}
-% 	The United States Government demands that the the Soviet Government institute a prompt and thorough investigation of this matter in order that the facts set forth above my be confirmed to its satisfaction.
-The United States Government further demands that the most strict and categorical instructions be issued to the Soviet air force that there be no repetition, under whatever pretext, of incidents of this kind which are so clearly calculated to magnify the difficulties of maintaining peaceful and correct international relationships.
-
-% 	The United States Government fully expects that, when its investigation is completed, the Soviet Government will express its regret for the unlawful and provocative behavior of its aviators, will see to it that those responsible for this action are promptly and severely punished and will, in accordance with established custom among peace-loving nations, pay appropriate indemnity for the unprovoked destruction of American lives and property.\footcite{the_associated_press_text_1950}
-% % \end{quote}
-% This was a weak response---it imposed no consequences, threatened no sanctions, and demanded only the most \emph{pro forma} of reparations.
-In essence, the note demanded an apology and a promise not to repeat.
-There was absolutely no chance that the parties responsible on the Soviet side would be held accountable---in the most serious of cases, some lower-level officers might take the fall, but even that was unlikely to happen here.
-If the United States had followed through on its demand for reparations---the only material repercussion mentioned---then one could have argued that this statement had some teeth.
-Instead, the demands were rejected in totality by the USSR.\footcite{salisbury_kremlin_1950} As far as I can tell, no reparations were ever made and no further consequences were imposed as a result.
-
-% Both sides had reasons to make this a larger political issue.
+Both sides had reasons to make this a larger political issue.
 The Soviet Union had fired first on a US Navy aircraft---an aircraft which could not have been the aggressor because its only armament was a .45-cal pistol carried by one of the crewmen.\footcite[p.~7]{peterson_maybe_1993} Given that the plane was never found, all available evidence suggests that the flight likely did not violate the traditional 12-mile airspace boundary.
-Based on the last received position report, today the CIA claims that the plane was flying 20-25 miles off the coast of Latvia at the time it was shot down.\footcite[p.~7]{peterson_maybe_1993} At the time, some unnamed Navy and Air Force officers, speaking to reporters in response to the Soviet protest note, conceded that it was possible that a complete breakdown in the plane's navigation system might have caused it to veer off-course into Latvia, but given its last known position, it would have taken ``a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.''\footcite[Technically, the  article credits the first statement about the possible electronic breakdown to ``Navy and Air Force officials'' and the 90 degrees statement to ``Aerial navigators.'' It is unclear whether the navigators in questions are with the US military, but it does not matter regardless.]{the_new_york_times_soviet_1950}
-
-% % \begin{figure}
-% % \centering
-% % \includegraphics[scale=0.45]{baltic-flight-path.png}
-% % \caption{USN PB4Y2 Flight Path}
-% % \label{baltic-flight-path}
-% % \end{figure}
-
-% Meanwhile, the United States knew that it hadn't lost a B-29, and that the missing plane probably hadn't ended up in Soviet airspace, but the Privateer in question \emph{had} been performing a questionably legal reconnaissance mission when it was shot down.
-Though the US was likely correct that the plane never flew over Latvia, it certainly got as close as it needed to in order to gather the necessary intelligence about Soviet military installations on the coast.
+Based on the last received position report, today the CIA claims that the plane was flying 20-25 miles off the coast of Latvia at the time it was shot down.\footcite[p.~7]{peterson_maybe_1993} At the time, it was theorized that a theoretical breakdown in the plane's navigation system could have caused it to veer off-course into Latvia, but given its last known position, it would have taken ``a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.''\footcite{the_new_york_times_soviet_1950} Meanwhile, the United States knew that the missing plane probably hadn't ended up in Soviet airspace, but the Privateer in question \emph{had} been performing a questionably legal reconnaissance mission when it was shot down.
 No traditional boundaries were violated, but a repurposed Navy privateer loaded with electronic reconnaissance equipment flying close to Latvia looked, to use a less technical term, shady.
-Yet based on the initial exchange of diplomatic protests, neither side was inclined to magnify the issue if they could avoid it.
+And still, based on the initial exchange of diplomatic protests, neither side was inclined to magnify the issue if they could avoid it.
 
@@
 
-% The most important consequence of the Baltic Incident was not diplomatic.
-As a direct result of the shootdown, President Truman, who until that point was largely uninformed about these missions, ordered a thirty-day halt to all reconnaissance flights.
-In that time, his office drafted an operating procedure that would balance the strategic need for reconnaissance with its potential diplomatic volatility.\footcite[p.~41]{farquhar_aerial_2015} The resulting memo from the Joint Chiefs of Staff created, with Truman's approval, the Special Electronic Airborne Search Project (SESP), which set clear delineations of responsibility between Navy and Air Force reconnaissance, and established executive responsibility for approving the missions.
-
 % Farquhar cites this as a landmark moment in the history of aerial reconnaissance.
@@
 
-% \subsection{Routine resolution}
-% Between the shootdown of the Navy Privateer in 1950 and the famous U-2 shootdown in 1960, the United States lost nine more aircraft to Soviet air defenses.
-Without analyzing each one at the same level of detail, I can demonstrate how the general attitudes displayed in response to the Baltic Incident became normalized.
-In each of the incidents listed, the pattern is the same---each side released dueling statements of protest, and the situation continued more or less as before.
-
-% Describe here are the next few peacetime reconnaissance shootdowns whose diplomatic consequences Farquhar analyzes.
+The general attitudes displayed in response to the Baltic Incident became the normal course of response to Soviet flights.
+The shootdown of the Navy Privateer was not an isolated incident; from then until the U-2 Incident in 1960, the United States lost nine aircraft to Soviet air defenses.
+Each time, the pattern of responses was the same---each side released dueling statements of protest, and the situation continued more or less as before.
 A B-29 that went missing on October 7, 1952 hit the front page of the \emph{New York Times} two days later.
@@
 
-% In another incident on July 29, 1953, the United States protested the shootdown of a B-50 (reconnaissance flight) over the Sea of Japan.
+In another incident on July 29, 1953, the United States protested the shootdown of a B-50 (reconnaissance flight) over the Sea of Japan.
 The Soviet Union responded by protesting the alleged American shootdown of a Soviet passenger plane.
@@
 
-% One incident in the period between 1950-1954 does not fit the broader pattern.
+One incident in the period between 1950-1954 does not fit the broader pattern.
 On November 20, 1951, an American C-47 transport was forced down in Hungary.
@@
 
-% I contend that the lack of espionage equipment is precisely the reason the United States made a sustained public effort to recover the four airmen, which is something that almost never happened with ferret flights.\footnote{While some of the flights were shot down overseas, making recovery difficult, I am referring here to situations where the US was aware that the Soviets likely held their airmen in captivity and still made little effort to get them back.} Farquhar summarizes the American response to this incident as follows: ``Responding to the press attention, the Truman Administration acted swiftly, attempting to gain the fliers release through diplomatic pressure.
+I contend that the lack of espionage equipment is precisely the reason the United States made a sustained public effort to recover the four airmen, which is something that almost never happened with ferret flights.\footnote{While some of the flights were shot down overseas, making recovery difficult, I am referring here to situations where the US was aware that the Soviets likely held their airmen in captivity and still made little effort to get them back.} Farquhar summarizes the American response to this incident as follows: ``Responding to the press attention, the Truman Administration acted swiftly, attempting to gain the fliers release through diplomatic pressure.
 The President ordered the Hungarian consulates in New York and Cleveland closed and banned private travel to the country.
@@
 
-% We also know that the United States conducted many more overflights than the failed ones that are listed here, using a variety of retrofitted planes.
+We also know that the United States conducted many more overflights than the failed ones that are listed here, using a variety of retrofitted planes.
 In 2001, the Air Force held a symposium honoring and recollecting the pilots who flew the overflights on Soviet territory in the early Cold War, an effort in which they were joined by the British in 1954.\footcite[p.~v]{hall_early_2003} Many of the presenters spoke about entire reconnaissance projects that apparently didn't even register diplomatically.
@@
 
-% The USSR knew about these missions, which were tracked by radar and often chased by MiG fighters.
+The USSR knew about these missions, which were tracked by radar and often chased by MiG fighters.
 Unlike some of the flights that were shot down, these overflights were clear violation of their territorial sovereignty.
@@
 
-% Even without a complete list of flights, it is clear that the territorial boundaries of the USSR were routinely violated to generate critical intelligence for NATO powers in an early, relatively volatile stage of the Cold War.
+Even without a complete list of flights, it is clear that the territorial boundaries of the USSR were routinely violated to generate critical intelligence for NATO powers in an early, relatively volatile stage of the Cold War.
 As time passed these flights appear to have become more accepted, not less.
@@
 These incidents undoubtedly altered the climate of Cold War diplomacy, but the fact remains that in 10 years of reconnaissance, the only action the Soviets took to keep the United States from conducting these incredibly provocative reconnaissance flights was to shoot them down.
-
-% % Valentino: Can you talk about other ways the Sovierts might have responded.
-What other kinds of actions might have seemed reasonable under different norms?
-Maybe give an example of a Soviet response to something apparently similar on its face, but clearly out of the espionage domain, so the response was much harsher.
 
@@
 
-% Another crucial factor emboldened Eisenhower and his team---most of whom, like Dulles, were much more enthusiastic about overflights than he was---to conduct that second mission.
-The USSR had been aware of the April 9 flight, and the United States \emph{knew} that they were aware of it, thanks to an onboard computer that had detected the Soviet tracking from the beginnings of the operation.\footcite{pedlow_central_1992} From the perspective of the American administration, the lack of a protest note after the previous flight was a sign that the USSR was not particularly offended by the overflights.
-That was a mistake.
-``It is one of the many unfortunate aspects of the high degree of secrecy surrounding Moscow's actions,'' imagery analyst David Lindgren writes, ``that the United States never understood the how angered and frustrated Soviet leaders were made by the U-2 overflights.''\footcite[p.~52]{lindgren_trust_2000} After years of bluster about the superiority of Soviet military forces, Khrushchev was not willing to admit that his Air Defense had been unable to shoot an unarmed spy plane flying deep into his homeland.\footcite[p.~59]{powers_operation_2004}
-
 % Almost four hours into the spy plane's flight on May 1, pilot Francis Gary Powers felt a dull ``thump'' and ``a tremendous orange flash lit the cockpit and sky'' as a Soviet surface-to-air (SAM) missile exploded behind him, ripping the wings off his plane and sending it into a tailspin.\footcite[p.~61]{powers_operation_2004} His subsequent capture by the Soviet military became an immediate international incident.
@@
 
-% Publicly, the USSR was not ready to accept that US overflights of its territory were a ``distasteful but vital necessity.'' The Four Powers summit in Paris was a disaster.
-Khrushchev launched into an angry rant on the first day and stormed out, immediately squashing any hope that pressing issues like the fate of West Berlin would be resolved there.
-He also canceled a planned visit by Eisenhower to the USSR that was scheduled to take place the next month.
-A contemporary newsreel about the conference said that ``in the course of two hours, Khrushchev brought US-Soviet relations to their lowest point since the end of World War II.''\footcite{universal_studios_summit_1960}
 % % The contrast between the severity of response Khrushchev wanted to portray publicly and what the USSR actually did is highlighted by his own son Sergei, who edited and annotated the 2007 edition with notes featuring additional context for his father's statements.

### 2019-12-19 article.tex
Replace (most) quotes with \enquote

@@
 The backlash was so intense that FBI Director James Comey was forced to take questions about the incident.
-``If you have my [application for security clearance],'' Comey admitted, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.
+\enquote{If you have my [application for security clearance],} Comey admitted, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.
 Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015} With evidence that a foreign intelligence service had successfully obtained exactly that, one might expect the United States to respond with a series of diplomatic measures intended to signal that highly invasive acts of espionage would not be tolerated from its international adversaries.
@@
 ``This is one of those cases where you have to ask, `Does the size of the operation change the nature of it?'\thinspace'' said one anonymous senior intelligence official of the OPM hack.
-``Clearly, it does.''\footcite{sanger_u.s._2015} As as far as the public knows, however, the size of operation did \emph{not} change the nature of the diplomatic response.
+\enquote{Clearly, it does.}\footcite{sanger_u.s._2015} As as far as the public knows, however, the size of operation did \emph{not} change the nature of the diplomatic response.
 Therefore, to understand why the Director of the CIA treats cyber-espionage as \enquote{fair game,} one first needs to understand why traditional espionage has historically received that treatment as well.
@@
 Arguably the most traditional form of espionage is stealing military secrets from a foreign power, and in that arena China has been relentless.
-A United States Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} ``Long-term, US future military advantage is being diminished by years of IP exfiltration from the DoD, DoN, and DIB,'' the report read, ``all with little to no adverse consequences to the thieves.''\footcite[p.~6]{bayer_cybersecurity_2019} The forecasts were dire: ``If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.''\footcite[p.~5]{bayer_cybersecurity_2019} The US Navy is not a disinterested party, of course, and there are reasons to take its conclusions with a grain of salt; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see \cite{gilli_why_2019}} Nonetheless, espionage presents alarming opportunities for rival states and the consequences for taking advantage of them remain minimal.
+A United States Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} \enquote{Long-term, US future military advantage is being diminished by years of IP exfiltration from the DoD, DoN, and DIB,} the report read, \enquote{all with little to no adverse consequences to the thieves.}\footcite[p.~6]{bayer_cybersecurity_2019} The forecasts were dire: \enquote{If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.}\footcite[p.~5]{bayer_cybersecurity_2019} The US Navy is not a disinterested party, of course, and there are reasons to take its conclusions with a grain of salt; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see \cite{gilli_why_2019}} Nonetheless, espionage presents alarming opportunities for rival states and the consequences for taking advantage of them remain minimal.
 
@@
 Glaser, I am going to refer to it as defensive realism, because they both put the security dilemma at the center of their explanation for international competition, which is the key observation for me.
-He would have you know that defensive realism ``can be viewed as a way station along the route to the full theory'' that he develops, which is ``significantly more general and complete than is defensive realism.'' (p.
+He would have you know that defensive realism \enquote{can be viewed as a way station along the route to the full theory} that he develops, which is \enquote{significantly more general and complete than is defensive realism.} (p.
 13-14)} Glaser begins from the core assumption that the reason for competition between security-seeking states is insecurity.
@@
 I asked a senior national security policymaker, Jake Sullivan, whether he thought American espionage was a net positive for global security.
-``I wouldn't work for the American government if I did not believe that it was a net positive actor,'' he told me, ``so I want the United States to be able to make the most informed decisions possible.'' When I asked whether it promoted global stability for a foreign nation---Russia for instance---to spy on the United States, he qualified his answer.
+``I wouldn't work for the American government if I did not believe that it was a net positive actor,'' he told me, \enquote{so I want the United States to be able to make the most informed decisions possible.} When I asked whether it promoted global stability for a foreign nation---Russia for instance---to spy on the United States, he qualified his answer.
 
 % He split the types of intelligence that foreign state might gather on the United States into three types: intent, capabilities, and strategy.
-``I have no problem with the Russians knowing that we do not have aggressive intent,'' he told me.
+\enquote{I have no problem with the Russians knowing that we do not have aggressive intent,} he told me.
 Capabilities, however, are more complicated.
 Some defense capabilities would probably be fine to have another country independently verify, and some we might want to keep secret, such as the atomic bomb before the end of the Second World War.
-And under almost no circumstances would a state want to have its strategies and warfighting plans revealed, so those should be kept as secret as possible.\footcite{sullivan_personal_2019} Without invoking the theory itself, Sullivan described exactly the type of intelligence that a defensive realist would need to solve the security dilemma.\footnote{At the point in my research when I conducted this interview, I had never even heard the term ``defensive realism,'' lest you think I led him to it.}
+And under almost no circumstances would a state want to have its strategies and warfighting plans revealed, so those should be kept as secret as possible.\footcite{sullivan_personal_2019} Without invoking the theory itself, Sullivan described exactly the type of intelligence that a defensive realist would need to solve the security dilemma.\footnote{At the point in my research when I conducted this interview, I had never even heard the term \enquote{defensive realism,} lest you think I led him to it.}
 
@@
 As the legal section notes, peacetime espionage in most traditional sense---activities associated with human spies---has virtually no legal recognition at all.
-The first institutional recognition of espionage for conflict-reduction purposes is the SALT I treaty in 1972, which protected ``national means of verification,'' but espionage-as-verification is limited to the specific treaties in which it is in included.
+The first institutional recognition of espionage for conflict-reduction purposes is the SALT I treaty in 1972, which protected \enquote{national means of verification,} but espionage-as-verification is limited to the specific treaties in which it is in included.
 Espionage is probably just too uncouth to formally acknowledge.
@@
 Every policymaker would like to their decisions to be as well-informed as possible.
-Without the benefit of hindsight, it is very difficult objectively gauge how much ``security'' is lost from a successful intelligence operation.
+Without the benefit of hindsight, it is very difficult objectively gauge how much \enquote{security} is lost from a successful intelligence operation.
 By contrast, it is very easy for someone in charge of making an important decision to notice how much better off they are for having had access to some illicitly-obtained information.
@@
 A true Open Skies Treaty followed soon after the collapse of the Soviet Union.
-In just over a decade, spy satellites went from ``espionage in space'' to ``national technical means of verification.'' But even as espionage gained some legitimacy in statecraft, the vagueness of that wording, ``means of verification,'' is indicative of what keeps espionage from true public acknowledgment.
+In just over a decade, spy satellites went from \enquote{espionage in space} to \enquote{national technical means of verification.} But even as espionage gained some legitimacy in statecraft, the vagueness of that wording, \enquote{means of verification,} is indicative of what keeps espionage from true public acknowledgment.
 The intelligence business is often looked at---not entirely unfairly---as being shady.
@@
 The Soviet Union had fired first on a US Navy aircraft---an aircraft which could not have been the aggressor because its only armament was a .45-cal pistol carried by one of the crewmen.\footcite[p.~7]{peterson_maybe_1993} Given that the plane was never found, all available evidence suggests that the flight likely did not violate the traditional 12-mile airspace boundary.
-Based on the last received position report, today the CIA claims that the plane was flying 20-25 miles off the coast of Latvia at the time it was shot down.\footcite[p.~7]{peterson_maybe_1993} At the time, it was theorized that a theoretical breakdown in the plane's navigation system could have caused it to veer off-course into Latvia, but given its last known position, it would have taken ``a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.''\footcite{the_new_york_times_soviet_1950} Meanwhile, the United States knew that the missing plane probably hadn't ended up in Soviet airspace, but the Privateer in question \emph{had} been performing a questionably legal reconnaissance mission when it was shot down.
+Based on the last received position report, today the CIA claims that the plane was flying 20-25 miles off the coast of Latvia at the time it was shot down.\footcite[p.~7]{peterson_maybe_1993} At the time, it was theorized that a theoretical breakdown in the plane's navigation system could have caused it to veer off-course into Latvia, but given its last known position, it would have taken \enquote{a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.}\footcite{the_new_york_times_soviet_1950} Meanwhile, the United States knew that the missing plane probably hadn't ended up in Soviet airspace, but the Privateer in question \emph{had} been performing a questionably legal reconnaissance mission when it was shot down.
 No traditional boundaries were violated, but a repurposed Navy privateer loaded with electronic reconnaissance equipment flying close to Latvia looked, to use a less technical term, shady.
@@
 % The American press had no such reservations.
-\emph{Washington Post} columnist Walter Lippmann wrote that it could not have been ``a local incident brought on by a local commander but [rather] that it was an act of Soviet policy.
-The known facts indicate that Soviet intelligence \textelp{} believed it was carrying important electronic equipment and that orders were given to the Soviet fighter command to intercept it.''\footcite{lippmann_baltic_1950} He speculated that there was no way the plane could have violated Soviet airspace, because wreckage would have been recovered, the plane was too slow, and no commander would have sent it on such a dangerous mission intentionally.
-``If, when, and as the American command were reconnoitering the Soviet military establishments on the Baltic coast,'' Lippmann wrote, ``it would use a plane of a wholly different type.''\footcite{lippmann_baltic_1950}
+\emph{Washington Post} columnist Walter Lippmann wrote that it could not have been \enquote{a local incident brought on by a local commander but [rather] that it was an act of Soviet policy.
+The known facts indicate that Soviet intelligence \textelp{} believed it was carrying important electronic equipment and that orders were given to the Soviet fighter command to intercept it.}\footcite{lippmann_baltic_1950} He speculated that there was no way the plane could have violated Soviet airspace, because wreckage would have been recovered, the plane was too slow, and no commander would have sent it on such a dangerous mission intentionally.
+\enquote{If, when, and as the American command were reconnoitering the Soviet military establishments on the Baltic coast,} Lippmann wrote, \enquote{it would use a plane of a wholly different type.}\footcite{lippmann_baltic_1950}
 
 % A newspaper columnist is not a policymaker, but his argument was likely reflective of the political opinion that Americans held regarding espionage at the time: of course we do these things, we just do them \emph{better} than that---and if the US had been doing the thing that the Soviets claimed it was, then it would have been justified of them to open fire.
-``Baltic Plane Mystery,'' a \emph{Post} article written a few days later, was almost sympathetic to the Soviet commanders.
-``Electronics make the old delimitations for border coastal flights ridiculous.
-A plane flying on a course perfectly legal by standards accepted today might still be engaged in reconnaissance of the first importance that an unfriendly power would try to frustrate.''\footcite{childs_baltic_1950} This is true, both in terms of the mission itself and the technological environment, but he went even further.
-``The only sound and safe assumption is that the Russians have a thorough and far-reaching espionage system.
-And at the same time we must hope that our system, and particularly on the side of counterespionage, is effective.'' Even the press, though outraged at the loss of American life, apparently took it as a given that this kind of espionage is necessary---and expected from both sides.
+\enquote{Baltic Plane Mystery,} a \emph{Post} article written a few days later, was almost sympathetic to the Soviet commanders.
+\enquote{Electronics make the old delimitations for border coastal flights ridiculous.
+A plane flying on a course perfectly legal by standards accepted today might still be engaged in reconnaissance of the first importance that an unfriendly power would try to frustrate.}\footcite{childs_baltic_1950} This is true, both in terms of the mission itself and the technological environment, but he went even further.
+\enquote{The only sound and safe assumption is that the Russians have a thorough and far-reaching espionage system.
+And at the same time we must hope that our system, and particularly on the side of counterespionage, is effective.} Even the press, though outraged at the loss of American life, apparently took it as a given that this kind of espionage is necessary---and expected from both sides.
 
 % Farquhar cites this as a landmark moment in the history of aerial reconnaissance.
-``No longer would ferret operations be conducted ad hoc by the military services;'' he writes, ``from 1950 onward, reconnaissance operations attracted Presidential attention and played a significant role in shaping U.S.
-foreign policy.''\footcite[p.~42]{farquhar_aerial_2015} The strict procedures that SESP established were crafted to avoid the risk of escalation, while still explicitly acknowledging the ``serious disadvantages accruing to the United States if the cessation of these operations were to be extended over an excessively long period.''\footcite{bradley_memorandum_1950} But the SESP procedures also contain a curious provision: ``Flights will not be made closer than twenty miles to the USSR or USSR- or satellite-controlled territory.''\footcite{bradley_memorandum_1950} This rule would not be followed---as is typical of espionage, the consequences of the activity never outweigh its strategic value.
+\enquote{No longer would ferret operations be conducted ad hoc by the military services;} he writes, \enquote{from 1950 onward, reconnaissance operations attracted Presidential attention and played a significant role in shaping U.S.
+foreign policy.}\footcite[p.~42]{farquhar_aerial_2015} The strict procedures that SESP established were crafted to avoid the risk of escalation, while still explicitly acknowledging the \enquote{serious disadvantages accruing to the United States if the cessation of these operations were to be extended over an excessively long period.}\footcite{bradley_memorandum_1950} But the SESP procedures also contain a curious provision: \enquote{Flights will not be made closer than twenty miles to the USSR or USSR- or satellite-controlled territory.}\footcite{bradley_memorandum_1950} This rule would not be followed---as is typical of espionage, the consequences of the activity never outweigh its strategic value.
 
@@
 Farquhar notes that the incident happened during the US election season and that the media thought the attack was intended to lower American prestige, but he does not mention any diplomatic consequences.\footcite[p.~43-44]{farquhar_aerial_2015} The official US response, however, is preserved on the State Department website.
-As with the Baltic Incident, the US simply demanded that the Soviets pay for the plane and return any survivors.\footcite{the_new_york_times_u.s._1952} Two more shootdowns in March 1953---one involving an American F–84 Thunderjet and the other an RAF Lincoln bomber---actually resulted in conciliatory responses from both sides, including a secret meeting to reduce aerial tensions ``of which little is known''.\footcite[p.~45]{farquhar_aerial_2015}
+As with the Baltic Incident, the US simply demanded that the Soviets pay for the plane and return any survivors.\footcite{the_new_york_times_u.s._1952} Two more shootdowns in March 1953---one involving an American F–84 Thunderjet and the other an RAF Lincoln bomber---actually resulted in conciliatory responses from both sides, including a secret meeting to reduce aerial tensions \enquote{of which little is known.}\footcite[p.~45]{farquhar_aerial_2015}
 
-In another incident on July 29, 1953, the United States protested the shootdown of a B-50 (reconnaissance flight) over the Sea of Japan.
+In another incident on July 29, 1953, the United States protested the shootdown of a B-50 (reconnaissance flight, per Peterson) over the Sea of Japan.
 The Soviet Union responded by protesting the alleged American shootdown of a Soviet passenger plane.
-Despite press outrage, no further action was taken.\footcite[p.~47]{farquhar_aerial_2015} A September 4, 1954 shootdown of a Navy P2V Neptune actually led to a call by one US Senator to break off diplomatic ties with the Soviet Union: ``Just another note from our State Department to the Kremlin hierarchy will not impress these uncivilized rulers \textelp{} that this new attack upon an American plane confirms Communist arrogance and aggressiveness to a point where the breaking of diplomatic relations is justified.''\footcite{the_associated_press_ending_1954} Instead of a harsher punishment, Eisenhower brought the issue to the UN with the full knowledge that an unfavorable judgment would be vetoed by the Soviets.
-The move demonstrated Eisenhower's ability to resist the calls for harsher action while still placating the American public.\footcite[p.~47]{farquhar_aerial_2015}
-
-% Without belaboring the point, suffice to say that Farquhar concludes that after this moment in 1954, the resolution of these incidents became less tense.
-``International incidents posed by shoot downs of reconnaissance aircraft still acted as a barrier in the path of d\'etente; but, by late 1954, overriding strategic concerns dictated a move toward breaking the cycle of hostility.''\footcite[p.~49]{farquhar_aerial_2015} In terms of media coverage, and diplomatic significance, no shootdown that followed would require more significant consequences than the ones that had already happened.
+Despite press outrage, no further action was taken.\footcite[p.~47]{farquhar_aerial_2015} A September 4, 1954 shootdown of a Navy P2V Neptune actually led to a call by one US Senator to break off diplomatic ties with the Soviet Union: \enquote{Just another note from our State Department to the Kremlin hierarchy will not impress these uncivilized rulers \textelp{} that this new attack upon an American plane confirms Communist arrogance and aggressiveness to a point where the breaking of diplomatic relations is justified.}\footcite{the_associated_press_ending_1954} Instead of a harsher punishment, Eisenhower brought the issue to the UN with the full knowledge that an unfavorable judgment would be vetoed by the Soviets.
+The move demonstrated Eisenhower's ability and commitment to resisting the calls for harsher action while still placating the American public.\footcite[p.~47]{farquhar_aerial_2015} Besides, by late 1954, the \enquote{cycle of hostility} over aerial reconnaissance was already breaking.\footcite[p.~49]{farquhar_aerial_2015} In terms of media coverage, and diplomatic significance, no shootdown that followed would require more significant consequences than the ones that had already happened.
 
@@
 On November 20, 1951, an American C-47 transport was forced down in Hungary.
-This incident is omitted from Peterson's list because, despite Soviet cries of ``spies and saboteurs,'' there is no evidence that it was a reconnaissance flight.\footnote{It also wasn't exactly shot down so much as forced to land.} Of all the cases that Farquhar examines, this is the only one that provoked a sustained diplomatic response from the United States---and it is the only case where the plane in question definitively carried no espionage equipment.
+This incident is omitted from Peterson's list because, despite Soviet cries of \enquote{spies and saboteurs,} there is no evidence that it was a reconnaissance flight.\footnote{It also wasn't exactly shot down so much as forced to land.} Of all the cases that Farquhar examines, this is the only one that provoked a sustained diplomatic response from the United States---and it is the only case where the plane in question definitively carried no espionage equipment.
 While the ferret flights were perfectly calibrated to fit as much radio surveillance equipment as possible, the only evidence of espionage that the Soviets were able to produce from the C-47---a plane that they recovered intact---was a portable radio, two extra parachutes, and some packets of warm blankets.\footcite{the_united_press_soviet_1951}
@@
 Legislatively, Truman asked Congress to pass a \$100 million Mutual Security Act to aid `selected persons residing in Soviet bloc states or refugees who wanted to form armed units' in opposition to Communism.''\footcite[p.~43]{farquhar_aerial_2015} Truman immediately took action against material Hungarian interests in the United States and threatened to empower \emph{armed dissidents} if the prisoners were not released.
-The contrast between this response and the one from just a year prior, when the US demanded ``a prompt and thorough investigation'' of the Baltic Incident, is stark.
+The contrast between this response and the one from just a year prior, when the US demanded \enquote{a prompt and thorough investigation} of the Baltic Incident, is stark.
 
@@
 Dino Brugioni, a senior CIA officer involved in the aerial imagery program, paints a grim picture of the institutional accountability involved when the flights went bad.
-In meetings between the State department and high-level Soviet diplomats, the subject of captured pilots ``was broached only perfunctorily in relation to other things being discussed.''\footcite[p.~72]{brugioni_eyes_2010} Some family remembers of lost airmen received posthumous awards, but most simply received the pilot's personal effects and no explanation.
-``And because these were secret missions conducted by a field command,'' Brugioni writes, ``a change of field commander meant that the fate of the men was soon forgotten.''\footcite[p.~72]{brugioni_eyes_2010} With missions so secret that we abandoned the men who flew them, we can have no expectation of finding a complete record today.
+In meetings between the State department and high-level Soviet diplomats, the subject of captured pilots \enquote{was broached only perfunctorily in relation to other things being discussed.}\footcite[p.~72]{brugioni_eyes_2010} Some family remembers of lost airmen received posthumous awards, but most simply received the pilot's personal effects and no explanation.
+\enquote{And because these were secret missions conducted by a field command,} Brugioni writes, \enquote{a change of field commander meant that the fate of the men was soon forgotten.}\footcite[p.~72]{brugioni_eyes_2010} With missions so secret that we abandoned the men who flew them, we can have no expectation of finding a complete record today.
 
@@
 Eisenhower required that the pilots of U-2 aircraft be CIA employees, specifically forbidding uniformed Air Force officers.\footcite[p.~33, Though many of the pilots did have Air Force backgrounds.]{lindgren_trust_2000} The CIA also had operational control of the program instead of the Air Force.
-``I want this whole thing to be a civilian operation,'' Eisenhower wrote, when settling an operational dispute between the two departments.
+\enquote{I want this whole thing to be a civilian operation,} Eisenhower wrote, when settling an operational dispute between the two departments.
 ``If uniformed personnel of the armed services of the United States fly over Russia, it is an act of war---legally---and I don't want any part of it.''\footcite[p.~60.
 The original source for this quote is an \emph{OSA History} that requires codeword clearance.
-It is quoted here by the History Staff of the CIA.]{pedlow_central_1992} This allowed the United States to ``truthfully deny,'' in the phrasing of the CIA, that any US ``military planes'' had flown over the USSR---which they had to do when the inevitable Soviet protest notes were filed after U-2 overflights began in 1956.\footcite[p.~109]{pedlow_central_1992}
+It is quoted here by the History Staff of the CIA.]{pedlow_central_1992} This allowed the United States to \enquote{truthfully deny,} in the phrasing of the CIA, that any US \enquote{military planes} had flown over the USSR---which they had to do when the inevitable Soviet protest notes were filed after U-2 overflights began in 1956.\footcite[p.~109]{pedlow_central_1992}
 
@@
 If the only goal had been to achieve a level of plausible deniability in the event of a shootdown, then simply not wearing an Air Force uniform during the mission likely would have been sufficient.
-The ``weather reconnaissance'' excuse had even been used before with the ferret flights.\footcite[p.~45]{farquhar_aerial_2015} At the highest levels of American government, however, policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation---even if the NASA weather-craft cover didn't hold.
-``It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft'' read a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} Both halves of that equation were equally important.
+The \enquote{weather reconnaissance} excuse had even been used before with the ferret flights.\footcite[p.~45]{farquhar_aerial_2015} At the highest levels of American government, however, policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation---even if the NASA weather-craft cover didn't hold.
+\enquote{It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft} read a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} Both halves of that equation were equally important.
 The Russians must believe that these operations are peacetime civilian operations, and they must be correct about it.
 
-% The United States was aware how provocative these reconnaissance flights might be, and Eisenhower was always concerned, correctly, about ``the terrible propaganda impact that would be occasioned if a reconnaissance plane were to fail.''\footcite[p.~162]{pedlow_central_1992} Starting in 1958, the American administration ordered a drastic decrease in the number of overflights.\footcite[p.~51]{powers_operation_2004} But Eisenhower was also under tremendous political pressure to learn more about the Soviet missile program, as hawkish senators amplified fears of a so-called ``missile gap,'' concerned that Soviet ballistic missile capabilities were advancing faster than their own.\footcite[Fears of this missile gap quickly followed earlier fears of a ``bomber gap,'' which ironically the U-2 had been critical in disproving.]{licklider_missile_1970} He especially did not want to affect the upcoming Four Powers diplomatic summit in Paris, scheduled for the summer of 1960.
+% The United States was aware how provocative these reconnaissance flights might be, and Eisenhower was always concerned, correctly, about \enquote{the terrible propaganda impact that would be occasioned if a reconnaissance plane were to fail.}\footcite[p.~162]{pedlow_central_1992} Starting in 1958, the American administration ordered a drastic decrease in the number of overflights.\footcite[p.~51]{powers_operation_2004} But Eisenhower was also under tremendous political pressure to learn more about the Soviet missile program, as hawkish senators amplified fears of a so-called \enquote{missile gap,} concerned that Soviet ballistic missile capabilities were advancing faster than their own.\footcite[Fears of this missile gap quickly followed earlier fears of a \enquote{bomber gap,} which ironically the U-2 had been critical in disproving.]{licklider_missile_1970} He especially did not want to affect the upcoming Four Powers diplomatic summit in Paris, scheduled for the summer of 1960.
 Nonetheless, Eisenhower ordered two more overflights in 1960, one on April \nth{9} and the other on May \nth{1}.
@@
 
-% Almost four hours into the spy plane's flight on May 1, pilot Francis Gary Powers felt a dull ``thump'' and ``a tremendous orange flash lit the cockpit and sky'' as a Soviet surface-to-air (SAM) missile exploded behind him, ripping the wings off his plane and sending it into a tailspin.\footcite[p.~61]{powers_operation_2004} His subsequent capture by the Soviet military became an immediate international incident.
-Three days after NASA quietly reported that they had lost a U-2 type weather reconnaissance plane, Khrushchev announced that the USSR had shot down an American plane that had flown into their airspace, which he called ``an aggressive provocation aimed at wrecking the Summit Conference.''\footcite[p.~112]{powers_operation_2004} Once Khrushchev gleefully debunked the resulting American cover story by revealing that they had taken Powers alive, Eisenhower admitted that the Gary Powers had been flying a spy plane, in a press conference where Eisenhower memorably declared that espionage was a ``a distasteful but vital necessity'' to prevent another attack like Pearl Harbor.\footcite{eisenhower_news_1960}
+% Almost four hours into the spy plane's flight on May 1, pilot Francis Gary Powers felt a dull \enquote{thump} and \enquote{a tremendous orange flash lit the cockpit and sky} as a Soviet surface-to-air (SAM) missile exploded behind him, ripping the wings off his plane and sending it into a tailspin.\footcite[p.~61]{powers_operation_2004} His subsequent capture by the Soviet military became an immediate international incident.
+Three days after NASA quietly reported that they had lost a U-2 type weather reconnaissance plane, Khrushchev announced that the USSR had shot down an American plane that had flown into their airspace, which he called \enquote{an aggressive provocation aimed at wrecking the Summit Conference.}\footcite[p.~112]{powers_operation_2004} Once Khrushchev gleefully debunked the resulting American cover story by revealing that they had taken Powers alive, Eisenhower admitted that the Gary Powers had been flying a spy plane, in a press conference where Eisenhower memorably declared that espionage was a \enquote{a distasteful but vital necessity} to prevent another attack like Pearl Harbor.\footcite{eisenhower_news_1960}
 
@@
 % Most scholars who study this period argue that the cumulative tension created by these flights contributed to an overall chilly tone in US-Soviet relations.
-Farquhar, for instance, refers to this as the ``cycle of hostility,'' in which the series of aerial incidents increased suspicions of military buildup on both sides, leading to even more aerial incidents as both powers attempted to verify their suspicions.\footcite[p.~43]{farquhar_aerial_2015} While it's true that the there were many aerial incidents, it is not at all clear how this affected any other aspect of US-Soviet diplomacy.
+Farquhar, for instance, refers to this as the \enquote{cycle of hostility,} in which the series of aerial incidents increased suspicions of military buildup on both sides, leading to even more aerial incidents as both powers attempted to verify their suspicions.\footcite[p.~43]{farquhar_aerial_2015} While it's true that the there were many aerial incidents, it is not at all clear how this affected any other aspect of US-Soviet diplomacy.
 I can't definitively say that there wasn't some diplomatic issue that would have gotten resolved at a summit absent these reconnaissance flights, but I can't locate an obvious one either.
@@
 While Powers was serving a 10-year sentence in Soviet prison, the two airmen were returned to the US in less than six months with only mild fanfare.
-Khrushchev wrote in his memoir that he wanted to continue ``our general line of peaceful coexistence'' and cited the resolution of this incident as an example of such, because the US was forced to make a formal request for the return of the airmen.\footcite[p.~256-257]{khrushchev_memoirs_2007} The only concessions the US made in return were to announce the discontinuation of its U-2 overflights (which Kennedy was already committed to) and to not make an issue of the illegal detention of the pilots.\footcite{time_cold_1961} For spy flights after Powers, the USSR simply took the necessary immediate countermeasures and shot down the planes that it could.
+Khrushchev wrote in his memoir that he wanted to continue \enquote{our general line of peaceful coexistence} and cited the resolution of this incident as an example of such, because the US was forced to make a formal request for the return of the airmen.\footcite[p.~256-257]{khrushchev_memoirs_2007} The only concessions the US made in return were to announce the discontinuation of its U-2 overflights (which Kennedy was already committed to) and to not make an issue of the illegal detention of the pilots.\footcite{time_cold_1961} For spy flights after Powers, the USSR simply took the necessary immediate countermeasures and shot down the planes that it could.
 Just as with traditional espionage, the countermeasures did not punish the attempt.
@@
 % Until the day he died, Khrushchev greatly exaggerated the lasting effect of the U-2 Incident.
-He wrote in his memoir that ``the commander of American forces in West Germany gave the order not to fly any closer than 50 kilometers from the border between East and West Germany.
-And no more incidents of that kind occurred.''\footcite[p.~256]{khrushchev_memoirs_2007} His son Sergei, who annotated the memoir, admitted that this statement was false: ``In practice such incidents occurred again from 1961 onward, but instead of the U-2 the Americans now used various types of Phantom or SR-71.''\footcite[p.~258]{khrushchev_memoirs_2007} On at least two occasions after the U-2 Incident, an American overflight crossed over into Soviet territory and the Soviets responded with...
+He wrote in his memoir that \enquote{the commander of American forces in West Germany gave the order not to fly any closer than 50 kilometers from the border between East and West Germany.
+And no more incidents of that kind occurred.}\footcite[p.~256]{khrushchev_memoirs_2007} His son Sergei, who annotated the memoir, admitted that this statement was false: \enquote{In practice such incidents occurred again from 1961 onward, but instead of the U-2 the Americans now used various types of Phantom or SR-71.}\footcite[p.~258]{khrushchev_memoirs_2007} On at least two occasions after the U-2 Incident, an American overflight crossed over into Soviet territory and the Soviets responded with...
 a diplomatic protest.
@@
 % As his title indicates, Rudolph Anderson was an Air Force Officer, not a member of the CIA.
-During the crisis, the President transferred control of Cuban U-2 reconnaissance to the Air Force, and authorized it to fly as many missions as necessary over the island.\footcite[p.~208-209]{pedlow_central_1992} Khrushchev did not say much about the Cuban U-2 in his writing, but he was very clear on one point: he did not order the SAM strike, and was worried that Kennedy ``might not be able to absorb this blow.''\footcite[p.~338]{khrushchev_memoirs_2007} Shooting down an American aircraft risked escalation, and that was not something Khrushchev wanted.
+During the crisis, the President transferred control of Cuban U-2 reconnaissance to the Air Force, and authorized it to fly as many missions as necessary over the island.\footcite[p.~208-209]{pedlow_central_1992} Khrushchev did not say much about the Cuban U-2 in his writing, but he was very clear on one point: he did not order the SAM strike, and was worried that Kennedy \enquote{might not be able to absorb this blow.}\footcite[p.~338]{khrushchev_memoirs_2007} Shooting down an American aircraft risked escalation, and that was not something Khrushchev wanted.
 This was the second time in history that Soviet forces had shot down a U-2 plane---but even though the spy plane was flying over Cuban airspace and piloted by an US Air Force Major, it was the Soviets who were worried that shooting it down had been an unnecessarily aggressive act.
@@
 Khrushchev's outbursts were hypnotizing the world with a passionate but highly distoreted presentation of one particular phase of international espionage.
-His government had been so notoriously involved in spying, especially in the United States, as to dwarf our activities, but by separating this particular type of espionage from all others he hoped to make convincing his charge of ``warmongering.'' To claim that because the equipment employed was an airplane with a camera, and therefore provocative of war, was plain silly, and I felt it necessary that the matter be put in perspective.
+His government had been so notoriously involved in spying, especially in the United States, as to dwarf our activities, but by separating this particular type of espionage from all others he hoped to make convincing his charge of \enquote{warmongering.} To claim that because the equipment employed was an airplane with a camera, and therefore provocative of war, was plain silly, and I felt it necessary that the matter be put in perspective.
 \\
@@
 
-% The peaceful uses of aerial reconnaissance were formalized at the end of the Cold War, when members of NATO and the Warsaw Pact signed the ``Treaty on Open Skies,'' which established guidelines for unarmed aerial surveillance flights in the spirit of open information and de-escalation.\footcite{organization_for_security_and_co-operation_in_europe_treaty_1992} This treaty was almost identical to the one first suggested by Eisenhower in 1955.\footcite{center_for_arms_control_and_non-proliferation_fact_2017} With certain restrictions, the type of flight that altered the course of the Cold War is now officially routine and permissible.
+% The peaceful uses of aerial reconnaissance were formalized at the end of the Cold War, when members of NATO and the Warsaw Pact signed the \enquote{Treaty on Open Skies,} which established guidelines for unarmed aerial surveillance flights in the spirit of open information and de-escalation.\footcite{organization_for_security_and_co-operation_in_europe_treaty_1992} This treaty was almost identical to the one first suggested by Eisenhower in 1955.\footcite{center_for_arms_control_and_non-proliferation_fact_2017} With certain restrictions, the type of flight that altered the course of the Cold War is now officially routine and permissible.
 In February 1962, Francis Gary Powers walked across a bridge into West Berlin at the same time as Russian Colonel Rudolph Abel, an American prisoner.
-Powers' father had suggested to the State Department that they might be traded, spy for a spy.\footcite[p.~239]{powers_operation_2004} ``This,'' Eisenhower wrote, ``was a tacit admission by Khrushchev that our `outrageous' U-2 pilots have their opposite numbers operating within the borders of this country.''\footcite[p.~558]{eisenhower_waging_1965}
+Powers' father had suggested to the State Department that they might be traded, spy for a spy.\footcite[p.~239]{powers_operation_2004} \enquote{This,} Eisenhower wrote, ``was a tacit admission by Khrushchev that our `outrageous' U-2 pilots have their opposite numbers operating within the borders of this country.''\footcite[p.~558]{eisenhower_waging_1965}
 
 % The Powers flight and Project Aquatone channel the norms of espionage in one final, but crucial respect: their strategic value easily outweighed the consequences of being caught.
-No one understood this better than President Eisenhower, who, when he was occasionally asked about the wisdom of the U-2 flights, would always reply: ``Would you be ready to give back all of the information we secured from our U-2 flights over Russia if there had been no disaster to one of our planes in Russia?'' In his telling, he never received an affirmative response.\footcite[p.~559]{eisenhower_waging_1965}
+No one understood this better than President Eisenhower, who, when he was occasionally asked about the wisdom of the U-2 flights, would always reply: \enquote{Would you be ready to give back all of the information we secured from our U-2 flights over Russia if there had been no disaster to one of our planes in Russia?} In his telling, he never received an affirmative response.\footcite[p.~559]{eisenhower_waging_1965}
 

### 2019-12-20 article.tex
Finish most of aerial espionage

@@
 10 September 1956 & USAF RB-50                    & Sea of Japan                         \\
-2 September 1958  & USADF C-130                   & Soviet Armenia (near Turkish border) \\
+2 September 1958  & USADF C-130                   & Soviet Armenia
+\\
 1 May 1960        & CIA U-2                       & Sverdolsk, USSR                      \\
@@
 The general attitudes displayed in response to the Baltic Incident became the normal course of response to Soviet flights.
-The shootdown of the Navy Privateer was not an isolated incident; from then until the U-2 Incident in 1960, the United States lost nine aircraft to Soviet air defenses.
+The shootdown of the Navy Privateer was not an isolated incident; from then until the U-2 Incident in 1960, the United States lost nine more aircraft to Soviet air defenses.
 Each time, the pattern of responses was the same---each side released dueling statements of protest, and the situation continued more or less as before.
@@
 
-I contend that the lack of espionage equipment is precisely the reason the United States made a sustained public effort to recover the four airmen, which is something that almost never happened with ferret flights.\footnote{While some of the flights were shot down overseas, making recovery difficult, I am referring here to situations where the US was aware that the Soviets likely held their airmen in captivity and still made little effort to get them back.} Farquhar summarizes the American response to this incident as follows: ``Responding to the press attention, the Truman Administration acted swiftly, attempting to gain the fliers release through diplomatic pressure.
+I contend that the lack of espionage equipment is precisely the reason the United States made a sustained public effort to recover the four airmen, which is something that almost never happened with ferret flights.\footnote{While some of the flights were shot down overseas, making recovery difficult, I am referring here to situations where the US was aware that the Soviets likely held their airmen in captivity and still made little effort to get them back.} Farquhar summarizes the American response to this incident as follows: \enquote{Responding to the press attention, the Truman Administration acted swiftly, attempting to gain the fliers release through diplomatic pressure.
 The President ordered the Hungarian consulates in New York and Cleveland closed and banned private travel to the country.
-Legislatively, Truman asked Congress to pass a \$100 million Mutual Security Act to aid `selected persons residing in Soviet bloc states or refugees who wanted to form armed units' in opposition to Communism.''\footcite[p.~43]{farquhar_aerial_2015} Truman immediately took action against material Hungarian interests in the United States and threatened to empower \emph{armed dissidents} if the prisoners were not released.
+Legislatively, Truman asked Congress to pass a \$100 million Mutual Security Act to aid \enquote{selected persons residing in Soviet bloc states or refugees who wanted to form armed units} in opposition to Communism.}\footcite[p.~43]{farquhar_aerial_2015} Truman immediately took action against material Hungarian interests in the United States and threatened to empower \emph{armed dissidents} if the prisoners were not released.
 The contrast between this response and the one from just a year prior, when the US demanded \enquote{a prompt and thorough investigation} of the Baltic Incident, is stark.
@@
 
-% \section{Civilian spy flights}
-% \subsection{Project Aquatone}
-% While Eisenhower was able to defuse some of the tension surrounding the military reconnaissance flights, even the most aggressive overflights could only hope to skirt the borders of Soviet territory; the United States couldn't just fly a B-50 over Moscow.
+\subsection{Inventing the spy plane}
+While Eisenhower was able to defuse some of the tension surrounding the military reconnaissance flights, even the most aggressive overflights could only hope to skirt the borders of Soviet territory; the United States couldn't just fly a B-50 over Moscow.
 What US intelligence-gathering capabilities needed was some way of photographing suspicious installations deep within Russia while still being able to claim the same benefit-of-the-doubt status that prevented even the most contentious overflights from escalating into a larger conflict.
@@
 
-% The key feature of a spy plane is that it is designed to be minimally provocative.
-Any plane can be retrofitted with cameras or radios that allow it to perform useful reconnaissance functions---what separates a spy plane is that there can can be no doubt (in the mind of the target) that it is being used for anything \emph{but} reconnaissance.
 For the spy plane to serve its intended purpose as a minimally provocative agent of espionage then, two things must be true: its operation must be understood by adversaries to be non-military, and that distinction must meaningfully alter their threat perception and associated response.
-To build such a aircraft, Eisenhower approved Project Aquatone, the code name for the creation of the U-2.
-
-% From its conception, the U-2 was purpose-built to be read as a civilian aircraft.
+To build such a aircraft, Eisenhower approved Project Aquatone---what became the U-2.
+From its conception, the U-2 was purpose-built to be read as a civilian aircraft.
 Eisenhower required that the pilots of U-2 aircraft be CIA employees, specifically forbidding uniformed Air Force officers.\footcite[p.~33, Though many of the pilots did have Air Force backgrounds.]{lindgren_trust_2000} The CIA also had operational control of the program instead of the Air Force.
 \enquote{I want this whole thing to be a civilian operation,} Eisenhower wrote, when settling an operational dispute between the two departments.
-``If uniformed personnel of the armed services of the United States fly over Russia, it is an act of war---legally---and I don't want any part of it.''\footcite[p.~60.
+\enquote{If uniformed personnel of the armed services of the United States fly over Russia, it is an act of war---legally---and I don't want any part of it.}\footcite[p.~60.
 The original source for this quote is an \emph{OSA History} that requires codeword clearance.
@@
 
-% Having the CIA run the program---ostensibly for the purpose of deception---changed the nature of the program itself.
+The steps Eisenhower took to ensure the U-2 civilian's character went beyond mere deception.
 If the only goal had been to achieve a level of plausible deniability in the event of a shootdown, then simply not wearing an Air Force uniform during the mission likely would have been sufficient.
-The \enquote{weather reconnaissance} excuse had even been used before with the ferret flights.\footcite[p.~45]{farquhar_aerial_2015} At the highest levels of American government, however, policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation---even if the NASA weather-craft cover didn't hold.
-\enquote{It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft} read a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} Both halves of that equation were equally important.
+The \enquote{weather reconnaissance} excuse had even been used before with the ferret flights.\footcite[p.~45]{farquhar_aerial_2015} At the highest levels of American government, however, policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation.\enquote{It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft} read a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} Both halves of that equation were equally important.
 The Russians must believe that these operations are peacetime civilian operations, and they must be correct about it.
 
-% The United States was aware how provocative these reconnaissance flights might be, and Eisenhower was always concerned, correctly, about \enquote{the terrible propaganda impact that would be occasioned if a reconnaissance plane were to fail.}\footcite[p.~162]{pedlow_central_1992} Starting in 1958, the American administration ordered a drastic decrease in the number of overflights.\footcite[p.~51]{powers_operation_2004} But Eisenhower was also under tremendous political pressure to learn more about the Soviet missile program, as hawkish senators amplified fears of a so-called \enquote{missile gap,} concerned that Soviet ballistic missile capabilities were advancing faster than their own.\footcite[Fears of this missile gap quickly followed earlier fears of a \enquote{bomber gap,} which ironically the U-2 had been critical in disproving.]{licklider_missile_1970} He especially did not want to affect the upcoming Four Powers diplomatic summit in Paris, scheduled for the summer of 1960.
-Nonetheless, Eisenhower ordered two more overflights in 1960, one on April \nth{9} and the other on May \nth{1}.
-Only one of them returned.
+At first, what made the U-2 flights \enquote{count} as espionage appears to be a deeply unsatisfying tautology.
+They counted as espionage because they were espionage!
+But Eisenhower's precautions were based in the reality of the United State's previous experience with reconnaissance flights.
+He had already been testing the limits of what the Soviet Union would be willing to accept.
+No finite list of attributes could ensure that the Russians wouldn't treat the U-2 plane---a massive leap in military surveillance technology---as a military development.
+But if the people in charge of Project Aquatone could convince \emph{themselves} that it was a civilian project, then they stood the best chance of having the Russian interpret it that way.
+The U-2 simply had to be a spy plane, in a manner that is both instantly recognizable and ontologically undefinable.
 
-% % It turned out however, that the Soviets opted to lodge their diplomatic protests privately, and soon they stopped protesting them at all.\footcite[p.~42]{lindgren_trust_2000}
+These precautions had the intended effect, even though at first it seemed like the Soviets were going to extract significant concessions.
+A Soviet Colonel in the USSR Air Defense later recalled that Khrushchev ``clearly viewed the violation of their nation's skies by a foreign reconnaissance aircraft \textelp{} as a political provocation'' and Khrushchev played it up.\footcite{orlov_u-2_2007} Three days after NASA quietly reported that they had lost a U-2 type weather reconnaissance plane, Khrushchev announced that the USSR had shot down an American plane that had flown into their airspace in \enquote{an aggressive provocation.}\footcite[p.~112]{powers_operation_2004} When the Americans stuck to their cover story, Khrushchev gleefully debunked it by revealing that they had taken Powers alive, Eisenhower admitted that the Gary Powers had been flying a spy plane, in a press conference where Eisenhower memorably declared that espionage was a \enquote{a distasteful but vital necessity} to prevent another attack like Pearl Harbor.\footcite{eisenhower_news_1960}
 
-% \subsection{The U-2 Incident}
-% On the morning of May 1, 1960, Soviet Air Defense forces detected a high-altitude aircraft flying over Soviet Tajikistan.
-Though they were were not yet able to conclusively identify it, the foreign aircraft was an American U-2 spy plane performing an reconnaissance overflight of the USSR---at a time when the Soviets were deeply embarrassed by their inability to prevent such flights.\footcite{orlov_u-2_2007} The previous U-2 overflight on April 9, less than a month prior, had lasted a full six hours and the aftermath caused an upheaval in the Soviet military.
-Leadership ordered an investigative commission to root out the shortcomings of the Air Force and Air Defense agencies, and a series of charts were drawn up anticipating the routes of future U-2 flights.
-The next time one came through, the Soviets were absolutely determined to shoot it down.
-
-% % \begin{figure}
-% % \centering
-% % \includegraphics[scale=0.3]{powers-flight-path.jpg}
-% % \caption{Operation Grand Slam Flight Path}
-% % \label{powers-flight-path}
-% % \end{figure}
-
-% The Soviets still proved unable to immediately stop the flight on May \nth{1}.
-A missile battalion in the plane's path was not on alert duty.
-Armed fighter aircraft were in the wrong positions.
-A frantic attempt to have another fighter plane literally ram it out of the sky was scrapped when the pilot failed to make visual contact.\footcite{orlov_u-2_2007} The highest possible levels of Soviet leadership were actively involved with the mission as it was taking place.
-A Soviet Colonel in the USSR Air Defense later recalled that Khrushchev ``clearly viewed the violation of their nation's skies by a foreign reconnaissance aircraft on the day of a Soviet national holiday, and just two weeks before a summit conference in Paris, as a political provocation.'' \footcite{orlov_u-2_2007}
-
-% Why did the generally cautious Eisenhower authorize two overflights in such close proximity to a crucial summit?
-For one, the secrecy of the program worked against him.
-Eisenhower was not willing to make the U-2 program public, so he couldn't convince his critics that he had extremely high quality intelligence that no such missile gap existed, because to reveal the existence of his aerial photography would prompt questions about its origins.\footcite[p.~51]{lindgren_trust_2000} And the missions had been such a tremendous success that his reluctance to authorize more overflights grated his staff, especially the influential Director of Central Intelligence (DCI), Allen Dulles.\footcite[p.~354]{brugioni_eyes_2010}
-
-% Almost four hours into the spy plane's flight on May 1, pilot Francis Gary Powers felt a dull \enquote{thump} and \enquote{a tremendous orange flash lit the cockpit and sky} as a Soviet surface-to-air (SAM) missile exploded behind him, ripping the wings off his plane and sending it into a tailspin.\footcite[p.~61]{powers_operation_2004} His subsequent capture by the Soviet military became an immediate international incident.
-Three days after NASA quietly reported that they had lost a U-2 type weather reconnaissance plane, Khrushchev announced that the USSR had shot down an American plane that had flown into their airspace, which he called \enquote{an aggressive provocation aimed at wrecking the Summit Conference.}\footcite[p.~112]{powers_operation_2004} Once Khrushchev gleefully debunked the resulting American cover story by revealing that they had taken Powers alive, Eisenhower admitted that the Gary Powers had been flying a spy plane, in a press conference where Eisenhower memorably declared that espionage was a \enquote{a distasteful but vital necessity} to prevent another attack like Pearl Harbor.\footcite{eisenhower_news_1960}
-
-% % The contrast between the severity of response Khrushchev wanted to portray publicly and what the USSR actually did is highlighted by his own son Sergei, who edited and annotated the 2007 edition with notes featuring additional context for his father's statements.
-
-% \section{Espionage in the open skies}
-% \subsection{Minimal consequences with maxmimal noise}
-% The U-2 Incident is often cited as a major failure of the Eisenhower administration.
-In the traditional telling, it significantly worsened US-Soviet relations at time when it seemed like lasting peace might be possible.
-I disagree; instead, I argue that that the U-2 Incident should be understood as an embarrassing but otherwise typical failure of an intelligence operation, one consistent with the minimal consequences typically applied to espionage.
-
-% No one disputes that the collapse of the Paris Summit was directly attributable to the Gary Powers shootdown but, frankly, summits just aren't that important.
+An yet few real consequences followed.
+No one disputes that the collapse of the Paris Summit was directly attributable to the Gary Powers shootdown but, frankly, summits just aren't that important.
 From 1953 to the end of the Cold War in 1991, there were 23 total US-USSR summits.\footcite{fain_chronology_2011} The main issue that did not get resolved in 1960 was the status of Berlin.
@@
 
-% What would a major punishment have looked like?
+\subsection{Evaluating the consequences of aerial espionage}
+The U-2 Incident is often cited as a seminal failure of the Eisenhower administration.
+In the traditional telling, it significantly worsened US-Soviet relations at time when it seemed like lasting peace might be possible.
+I disagree.
+Instead, the U-2 Incident should be understood as an embarrassing but otherwise typical failure of an intelligence operation, one consistent with the minimal consequences typically applied to espionage.\footnote{While I argue that the response to the U-2 flight was within the reasonable bounds for espionage, I acknowledge that it was still unprecedented in the context of reconnaissance flights.
+There are a number of reasonable explanations for why the U-2 shootdown escalated.
+Partially the Soviets felt humiliated by a technology for which they had no countermeasures and the uniquely insulting way the overflights flew deep into Soviet territory.
+Eisenhower thought that Khrushchev might have been looking for an excuse to cancel his visit to Russia.
+See: \cite[p.~555]{eisenhower_waging_1965}.
+Others in his administration believed that the U-2 Incident was an excuse to justify an intensification of hostilities that had already been decided by hard-liners in the Kremlin.
+See: \cite[p.~328]{kistiakowsky_scientist_1976}.} What would a major punishment have looked like?
 One proportional response Khrushchev could have considered was to force the issue on regional air bases.
@@
 
-% While I argue that the response to the U-2 flight was within the reasonable bounds for espionage, I acknowledge that it was still unprecedented in the context of reconnaissance flights.
-There are a number of reasonable explanations for why the U-2 shootdown escalated.
-Partially the Soviets felt humiliated by a technology for which they had no countermeasures and the uniquely insulting way the overflights flew deep into Soviet territory.
-Eisenhower thought that Khrushchev might have been looking for an excuse to cancel his visit to Russia.\footcite[p.~555]{eisenhower_waging_1965} Others in his administration believed that the U-2 Incident was an excuse to justify an intensification of hostilities that had already been decided by hard-liners in the Kremlin.\footcite[p.~328]{kistiakowsky_scientist_1976}
-
-% Most scholars who study this period argue that the cumulative tension created by these flights contributed to an overall chilly tone in US-Soviet relations.
+Most scholars who study this period argue that the cumulative tension created by these flights contributed to an overall chilly tone in US-Soviet relations.
 Farquhar, for instance, refers to this as the \enquote{cycle of hostility,} in which the series of aerial incidents increased suspicions of military buildup on both sides, leading to even more aerial incidents as both powers attempted to verify their suspicions.\footcite[p.~43]{farquhar_aerial_2015} While it's true that the there were many aerial incidents, it is not at all clear how this affected any other aspect of US-Soviet diplomacy.
@@
 
-% Consider the period of time over which these flights took place.
+Consider the period of time over which these flights took place.
 If aerial reconnaissance had risen to anything more than a lingering irritant, then it should have to have been resolved in some way before 1960.
@@
 
-% \subsection{The U-2 after Powers}
-% When the US really needed intelligence that only an overflight could provide, the diplomatic consequences that the USSR had imposed were not sufficient to discourage it.
+Until the day he died, Khrushchev greatly exaggerated the lasting effect of the U-2 Incident.
+He wrote in his memoir that \enquote{the commander of American forces in West Germany gave the order not to fly any closer than 50 kilometers from the border between East and West Germany.
+And no more incidents of that kind occurred.}\footcite[p.~256]{khrushchev_memoirs_2007} His son Sergei, who annotated the memoir, admitted that this statement was false: \enquote{In practice such incidents occurred again from 1961 onward, but instead of the U-2 the Americans now used various types of Phantom or SR-71.}\footcite[p.~258]{khrushchev_memoirs_2007} On at least two occasions after the U-2 Incident, an American overflight crossed over into Soviet territory and the Soviets responded with...
+a diplomatic protest.
+The US quickly apologized, and that was that.\footcite{orlov_u-2_2007}
+
+% As is characteristic of espionage, when the US really needed intelligence that only an overflight could provide, the diplomatic consequences that the USSR had imposed were not sufficient to discourage it.
 Two months to the day after Powers was shot down, a Soviet MiG-19 opened fire on an American RB-47H and captured the plane's navigator and co-pilot.
@@
 Just as with traditional espionage, the countermeasures did not punish the attempt.
-
-% Until the day he died, Khrushchev greatly exaggerated the lasting effect of the U-2 Incident.
-He wrote in his memoir that \enquote{the commander of American forces in West Germany gave the order not to fly any closer than 50 kilometers from the border between East and West Germany.
-And no more incidents of that kind occurred.}\footcite[p.~256]{khrushchev_memoirs_2007} His son Sergei, who annotated the memoir, admitted that this statement was false: \enquote{In practice such incidents occurred again from 1961 onward, but instead of the U-2 the Americans now used various types of Phantom or SR-71.}\footcite[p.~258]{khrushchev_memoirs_2007} On at least two occasions after the U-2 Incident, an American overflight crossed over into Soviet territory and the Soviets responded with...
-a diplomatic protest.
-The US quickly apologized, and that was that.\footcite{orlov_u-2_2007}
 
@@
 
-% \subsection{Bridge of spies}
 % The incidents involving U-2 aircraft during the Cold War perfectly mirror the traditional consequences for espionage because the United States succeeded in applying espionage norms to its use.
@@
 
-% Both heads of state made clear in their writings that they considered Powers to be spy.
+Both heads of state made clear in their writings that they considered Powers to be spy.
 On capturing Gary Powers, Khrushchev wrote: ``This was a hostile act by the leaders of the U.S.
@@
 They didn't think we had the capability of \textelp{} acquiring irrefutable proof that the United States was using methods that were impermissible in peacetime.''\footcite[p.~239]{khrushchev_memoirs_2007} Sergei's annotations once again contradicted his father's claim that these methods were impermissible; more than forty American reconnaissance aircraft were shot down by the USSR during the Cold War.
-As is often the case with espionage, Khrushchev wanted to make it seem like he was standing up for the integrity of the Soviet state without endangering the espionage equilibrium.
+Khrushchev wanted to make it seem like he was standing up for the integrity of the Soviet state without endangering the espionage equilibrium.
 
-% Eisenhower, always an insightful thinker on issues of intelligence, understood that beneath the grandstanding, the fundamentals of international espionage still applied. In his own words:
+Eisenhower, an insightful thinker on issues of intelligence, understood that beneath the grandstanding, the fundamentals of international espionage still applied.
+\enquote{[Khrushchev's] government,} he wrote, \enquote{had been so notoriously involved in spying, especially in the United States, as to dwarf our activities, but by separating this particular type of espionage from all others he hoped to make convincing his charge of \enquote{warmongering.} To claim that because the equipment employed was an airplane with a camera, and therefore provocative of war, was plain silly, and I felt it necessary that the matter be put in perspective.} \footcite[p.~551]{eisenhower_waging_1965}
 
-% % \begin{quote}
-% 	Furthermore, it seemed to me that Mr.
-Khrushchev's outbursts were hypnotizing the world with a passionate but highly distoreted presentation of one particular phase of international espionage.
-His government had been so notoriously involved in spying, especially in the United States, as to dwarf our activities, but by separating this particular type of espionage from all others he hoped to make convincing his charge of \enquote{warmongering.} To claim that because the equipment employed was an airplane with a camera, and therefore provocative of war, was plain silly, and I felt it necessary that the matter be put in perspective.
-\\
-
-% 	The real issue at stake was not the fact that both sides conducted intelligence activities, but rather that the conduct and announced intentions of the Communists created the necessity for such clandestine maneuvers.
-As a consequence, the West---more specifically the Untied States as the major military power of the West---had to maintain constantly the capacity to detect any possible prelude to an infinitely more destructive Pearl Harbor.\footcite[p.~551]{eisenhower_waging_1965}
-% % \end{quote}
-% The next chapter examines in depth how the fear of a nuclear Pearl Harbor affected the entire American intelligence strategy, but even in the limited context of the spy plane, it is clear that Eisenhower deployed espionage to de-escalatory effect.
-Khrushchev had convinced the world that the Soviet missile capabilities were far more advanced than they actually were.
-Eisenhower was under tremendous pressure from the public, Republicans in Congress, and even his own staff to increase the defense budget.
-The only thing standing between Eisenhower and an accelerated arms race were detailed photo-reconnaissance reports from the U-2 that showed the Soviets were developmentally far behind what their public statements suggested.
-
-% The peaceful uses of aerial reconnaissance were formalized at the end of the Cold War, when members of NATO and the Warsaw Pact signed the \enquote{Treaty on Open Skies,} which established guidelines for unarmed aerial surveillance flights in the spirit of open information and de-escalation.\footcite{organization_for_security_and_co-operation_in_europe_treaty_1992} This treaty was almost identical to the one first suggested by Eisenhower in 1955.\footcite{center_for_arms_control_and_non-proliferation_fact_2017} With certain restrictions, the type of flight that altered the course of the Cold War is now officially routine and permissible.
-In February 1962, Francis Gary Powers walked across a bridge into West Berlin at the same time as Russian Colonel Rudolph Abel, an American prisoner.
-Powers' father had suggested to the State Department that they might be traded, spy for a spy.\footcite[p.~239]{powers_operation_2004} \enquote{This,} Eisenhower wrote, ``was a tacit admission by Khrushchev that our `outrageous' U-2 pilots have their opposite numbers operating within the borders of this country.''\footcite[p.~558]{eisenhower_waging_1965}
-
-% The Powers flight and Project Aquatone channel the norms of espionage in one final, but crucial respect: their strategic value easily outweighed the consequences of being caught.
-No one understood this better than President Eisenhower, who, when he was occasionally asked about the wisdom of the U-2 flights, would always reply: \enquote{Would you be ready to give back all of the information we secured from our U-2 flights over Russia if there had been no disaster to one of our planes in Russia?} In his telling, he never received an affirmative response.\footcite[p.~559]{eisenhower_waging_1965}
+The Powers flight and Project Aquatone channel the norms of espionage in one final, but crucial respect: their strategic value easily outweighed the consequences of being caught.
+No one understood this better than President Eisenhower, who, when he was occasionally questioned about the wisdom of the U-2 flights, would always reply: \enquote{Would you be ready to give back all of the information we secured from our U-2 flights over Russia if there had been no disaster to one of our planes in Russia?} In his telling, he never received an affirmative response.\footcite[p.~559]{eisenhower_waging_1965}
 
@@
 \end{document}
+
+% The peaceful uses of aerial reconnaissance were formalized at the end of the Cold War, when members of NATO and the Warsaw Pact signed the \enquote{Treaty on Open Skies,} which established guidelines for unarmed aerial surveillance flights in the spirit of open information and de-escalation.\footcite{organization_for_security_and_co-operation_in_europe_treaty_1992} This treaty was almost identical to the one first suggested by Eisenhower in 1955.\footcite{center_for_arms_control_and_non-proliferation_fact_2017} With certain restrictions, the type of flight that altered the course of the Cold War is now officially routine and permissible.
+In February 1962, Francis Gary Powers walked across a bridge into West Berlin at the same time as Russian Colonel Rudolph Abel, an American prisoner.
+Powers' father had suggested to the State Department that they might be traded, spy for a spy.\footcite[p.~239]{powers_operation_2004} \enquote{This,} Eisenhower wrote, ``was a tacit admission by Khrushchev that our `outrageous' U-2 pilots have their opposite numbers operating within the borders of this country.''\footcite[p.~558]{eisenhower_waging_1965}

### 2019-12-23 article.tex
Complete draft of aerial espionage

@@
 \subsection{Inventing the spy plane}
-While Eisenhower was able to defuse some of the tension surrounding the military reconnaissance flights, even the most aggressive overflights could only hope to skirt the borders of Soviet territory; the United States couldn't just fly a B-50 over Moscow.
+While Eisenhower was able to defuse some of the tension surrounding the military reconnaissance flights, even the most aggressive overflights could only hope to skirt the borders of Soviet territory; the United States couldn't just simply fly a B-52 over Moscow.
 What US intelligence-gathering capabilities needed was some way of photographing suspicious installations deep within Russia while still being able to claim the same benefit-of-the-doubt status that prevented even the most contentious overflights from escalating into a larger conflict.
@@
 
-For the spy plane to serve its intended purpose as a minimally provocative agent of espionage then, two things must be true: its operation must be understood by adversaries to be non-military, and that distinction must meaningfully alter their threat perception and associated response.
-To build such a aircraft, Eisenhower approved Project Aquatone---what became the U-2.
+For the spy plane to serve its intended purpose as a minimally provocative agent of espionage two things must be true: its operation must be understood by adversaries to be non-military, and that distinction must meaningfully alter their associated response.
+To build such an aircraft, Eisenhower approved Project Aquatone---the codename for the U-2.
 From its conception, the U-2 was purpose-built to be read as a civilian aircraft.
@@
 The steps Eisenhower took to ensure the U-2 civilian's character went beyond mere deception.
-If the only goal had been to achieve a level of plausible deniability in the event of a shootdown, then simply not wearing an Air Force uniform during the mission likely would have been sufficient.
+If his only goal had been to achieve a level of plausible deniability in the event of a shootdown, then simply not wearing an Air Force uniform during the mission likely would have been sufficient.
 The \enquote{weather reconnaissance} excuse had even been used before with the ferret flights.\footcite[p.~45]{farquhar_aerial_2015} At the highest levels of American government, however, policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation.\enquote{It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft} read a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} Both halves of that equation were equally important.
@@
 But Eisenhower's precautions were based in the reality of the United State's previous experience with reconnaissance flights.
-He had already been testing the limits of what the Soviet Union would be willing to accept.
-No finite list of attributes could ensure that the Russians wouldn't treat the U-2 plane---a massive leap in military surveillance technology---as a military development.
-But if the people in charge of Project Aquatone could convince \emph{themselves} that it was a civilian project, then they stood the best chance of having the Russian interpret it that way.
+He had already been testing the limits of what the Soviet Union would be willing to accept; if he did everything possible to signal that this plane functioned like a spy, perhaps it would accept this new development as well.
+No finite list of attributes could ensure that the Soviet Union wouldn't treat the U-2 plane---a massive leap in military surveillance technology---as a military development.
+But if the people in charge of Project Aquatone could convince \emph{themselves} that it was a civilian project, then they stood the best chance of having the Russians interpret it that way as well.
 The U-2 simply had to be a spy plane, in a manner that is both instantly recognizable and ontologically undefinable.
 
-These precautions had the intended effect, even though at first it seemed like the Soviets were going to extract significant concessions.
+These precautions had the intended effect, even though it initially seemed like the Soviets were going to extract significant concessions.
 A Soviet Colonel in the USSR Air Defense later recalled that Khrushchev ``clearly viewed the violation of their nation's skies by a foreign reconnaissance aircraft \textelp{} as a political provocation'' and Khrushchev played it up.\footcite{orlov_u-2_2007} Three days after NASA quietly reported that they had lost a U-2 type weather reconnaissance plane, Khrushchev announced that the USSR had shot down an American plane that had flown into their airspace in \enquote{an aggressive provocation.}\footcite[p.~112]{powers_operation_2004} When the Americans stuck to their cover story, Khrushchev gleefully debunked it by revealing that they had taken Powers alive, Eisenhower admitted that the Gary Powers had been flying a spy plane, in a press conference where Eisenhower memorably declared that espionage was a \enquote{a distasteful but vital necessity} to prevent another attack like Pearl Harbor.\footcite{eisenhower_news_1960}
 
-An yet few real consequences followed.
+And yet few real consequences followed.
 No one disputes that the collapse of the Paris Summit was directly attributable to the Gary Powers shootdown but, frankly, summits just aren't that important.
@@
 With the late rollout of the captured pilot, the dramatic scene at the summit, and the cancellation of Eisenhower's visit, Khrushchev publicly embarrassed the United States without meaningfully altering Soviet policy towards it.
+
+Nor did the U-2 Incident end all reconnaissance flights.
+As is characteristic of espionage, when the US really needed intelligence that only an overflight could provide, the diplomatic consequences that the USSR had imposed were not sufficient to discourage it.
+Two months to the day after Powers was shot down, a Soviet MiG-19 opened fire on an American RB-47H and captured the plane's navigator and co-pilot.
+While Powers was serving a 10-year sentence in Soviet prison, the two airmen were returned to the US in less than six months with only mild fanfare.
+Khrushchev wrote in his memoir that he wanted to continue \enquote{our general line of peaceful coexistence} and cited the resolution of this incident as an example of such, because the US was forced to make a formal request for the return of the airmen.\footcite[p.~256-257]{khrushchev_memoirs_2007} The only concessions the US made in return were to announce the discontinuation of its U-2 overflights (which Kennedy was already committed to) and to not make an issue of the illegal detention of the pilots.\footcite{time_cold_1961} For spy flights after Powers, the USSR simply took the necessary immediate countermeasures and shot down the planes that it could.
+Just as with traditional espionage, the countermeasures did not punish the attempt.
 
@@
 The Soviets were incredibly preoccupied with removing the forward-deployed Jupiter missiles from Turkey.
-Although the missiles were deployed to bases well within NATO territory, and arguably less threatening to Soviet security than American overflights of the Soviet bloc, the existence of these missile became a central bargaining chip in the Cuban Missile Crisis.
-By contrast, the Soviets simply did not have the same level of interest in curtailing reconnaissance flights.
+\footnote{Although the missiles were deployed to bases well within NATO territory, and arguably less threatening to Soviet security than American overflights of the Soviet bloc, the existence of these missile became a central bargaining chip in the Cuban Missile Crisis.} By contrast, the Soviets simply did not have the same level of interest in curtailing reconnaissance flights.
 That both sides of the Cold War had strong incentives to avoid military escalation is well-documented, but at least threatening military retaliation for launching provocative military operations would not seem to have been out of place.
@@
 I can't definitively say that there wasn't some diplomatic issue that would have gotten resolved at a summit absent these reconnaissance flights, but I can't locate an obvious one either.
-
-Consider the period of time over which these flights took place.
-If aerial reconnaissance had risen to anything more than a lingering irritant, then it should have to have been resolved in some way before 1960.
+Consider the period of time over which these flights took place; if aerial reconnaissance had risen to anything more than a lingering irritant, then it should have to have been resolved in some way before 1960.
 The US-Soviet relationship saw huge shifts, including periods of optimism, over the 11 years preceding the U-2 incident.
@@
 The US quickly apologized, and that was that.\footcite{orlov_u-2_2007}
-
-% As is characteristic of espionage, when the US really needed intelligence that only an overflight could provide, the diplomatic consequences that the USSR had imposed were not sufficient to discourage it.
-Two months to the day after Powers was shot down, a Soviet MiG-19 opened fire on an American RB-47H and captured the plane's navigator and co-pilot.
-While Powers was serving a 10-year sentence in Soviet prison, the two airmen were returned to the US in less than six months with only mild fanfare.
-Khrushchev wrote in his memoir that he wanted to continue \enquote{our general line of peaceful coexistence} and cited the resolution of this incident as an example of such, because the US was forced to make a formal request for the return of the airmen.\footcite[p.~256-257]{khrushchev_memoirs_2007} The only concessions the US made in return were to announce the discontinuation of its U-2 overflights (which Kennedy was already committed to) and to not make an issue of the illegal detention of the pilots.\footcite{time_cold_1961} For spy flights after Powers, the USSR simply took the necessary immediate countermeasures and shot down the planes that it could.
-Just as with traditional espionage, the countermeasures did not punish the attempt.
 
@@
 
-% The incidents involving U-2 aircraft during the Cold War perfectly mirror the traditional consequences for espionage because the United States succeeded in applying espionage norms to its use.
-Eisenhower's insistence on having the CIA manage Project Aquatone had the intended effect: the spy plane was, in story as well as fact, a civilian intelligence operation.
-
 % Even when a U-2 operation went poorly, it still played by the rules of the game.
@@
 They didn't think we had the capability of \textelp{} acquiring irrefutable proof that the United States was using methods that were impermissible in peacetime.''\footcite[p.~239]{khrushchev_memoirs_2007} Sergei's annotations once again contradicted his father's claim that these methods were impermissible; more than forty American reconnaissance aircraft were shot down by the USSR during the Cold War.
-Khrushchev wanted to make it seem like he was standing up for the integrity of the Soviet state without endangering the espionage equilibrium.
-
-Eisenhower, an insightful thinker on issues of intelligence, understood that beneath the grandstanding, the fundamentals of international espionage still applied.
+Eisenhower, an insightful thinker on issues of intelligence, understood that beneath the grandstanding the fundamentals of international espionage still applied.
 \enquote{[Khrushchev's] government,} he wrote, \enquote{had been so notoriously involved in spying, especially in the United States, as to dwarf our activities, but by separating this particular type of espionage from all others he hoped to make convincing his charge of \enquote{warmongering.} To claim that because the equipment employed was an airplane with a camera, and therefore provocative of war, was plain silly, and I felt it necessary that the matter be put in perspective.} \footcite[p.~551]{eisenhower_waging_1965}
 
-The Powers flight and Project Aquatone channel the norms of espionage in one final, but crucial respect: their strategic value easily outweighed the consequences of being caught.
+% Add slight emphasis here that the U-2 was built as a spy plane because it would have a certain reaction, and it worked
+
+Project Aquatone channeled the norms of espionage in the most crucial respect: its strategic value easily outweighed the consequences of being caught.
 No one understood this better than President Eisenhower, who, when he was occasionally questioned about the wisdom of the U-2 flights, would always reply: \enquote{Would you be ready to give back all of the information we secured from our U-2 flights over Russia if there had been no disaster to one of our planes in Russia?} In his telling, he never received an affirmative response.\footcite[p.~559]{eisenhower_waging_1965}

### 2019-12-24 article.tex
Remove extra comments aerial ending

@@
 
-% Farquhar cites this as a landmark moment in the history of aerial reconnaissance.
-\enquote{No longer would ferret operations be conducted ad hoc by the military services;} he writes, \enquote{from 1950 onward, reconnaissance operations attracted Presidential attention and played a significant role in shaping U.S.
-foreign policy.}\footcite[p.~42]{farquhar_aerial_2015} The strict procedures that SESP established were crafted to avoid the risk of escalation, while still explicitly acknowledging the \enquote{serious disadvantages accruing to the United States if the cessation of these operations were to be extended over an excessively long period.}\footcite{bradley_memorandum_1950} But the SESP procedures also contain a curious provision: \enquote{Flights will not be made closer than twenty miles to the USSR or USSR- or satellite-controlled territory.}\footcite{bradley_memorandum_1950} This rule would not be followed---as is typical of espionage, the consequences of the activity never outweigh its strategic value.
-
 The general attitudes displayed in response to the Baltic Incident became the normal course of response to Soviet flights.
@@
 He wrote in his memoir that \enquote{the commander of American forces in West Germany gave the order not to fly any closer than 50 kilometers from the border between East and West Germany.
-And no more incidents of that kind occurred.}\footcite[p.~256]{khrushchev_memoirs_2007} His son Sergei, who annotated the memoir, admitted that this statement was false: \enquote{In practice such incidents occurred again from 1961 onward, but instead of the U-2 the Americans now used various types of Phantom or SR-71.}\footcite[p.~258]{khrushchev_memoirs_2007} On at least two occasions after the U-2 Incident, an American overflight crossed over into Soviet territory and the Soviets responded with...
+And no more incidents of that kind occurred.}\footcite[p.~256]{khrushchev_memoirs_2007} His son Sergei, who annotated the memoir, admitted that this statement was false: \enquote{In practice such incidents occurred again from 1961 onward, but instead of the U-2 the Americans now used various types of Phantom or SR-71.}\footcite[p.~258]{khrushchev_memoirs_2007} The United States continued to use the U-2 extensively in other regions as well.\footnote{A recently declassified CIA report details the many locations in which U-2 photography was employed to gather intelligence after 1960, including China, India, Indonesia, Thailand, Tibet, Laos, North Vietnam, and Venezuela.
+Lyndon Johnson even clarified that his predecessor's order to end U-2 flights over the Soviet Bloc was not indefinitely binding---it was valid only until countermanded.
+For more, see: \cite[p.~195]{pedlow_central_1992}}.
+On at least two occasions after the U-2 Incident, an American overflight crossed over into Soviet territory and the Soviets responded with...
 a diplomatic protest.
 The US quickly apologized, and that was that.\footcite{orlov_u-2_2007}
-
-% The U-2 did not disappear after the Powers flight.
-A recently declassified CIA report details the many locations in which U-2 photography was employed to gather intelligence after 1960, including China, India, Indonesia, Thailand, Tibet, Laos, North Vietnam, and Venezuela.
-Lyndon Johnson even clarified that his predecessor's order to end U-2 flights over the Soviet Bloc was not indefinitely binding---it was valid only until countermanded.\footcite[p.~195]{pedlow_central_1992} As it turned out, it never became necessary to resume Soviet Bloc overflights because satellite photo-reconnaissance soon replaced them; U-2 overflights were rendered mostly obsolete.
-There is, however, one use for which satellites will likely never displace aerial photo-reconnaissance---a situation like the Cuban Missile Crisis, where events are so time-sensitive that intelligence is needed in a matter of hours.
-
-% On October 15, 1962, U-2 photography revealed that the Soviet Union was assembling missiles in Cuba that were capable of carrying nuclear warheads.
-As the crisis unfolded, the United States continued to send U-2 planes over Cuba in order to maintain a close watch on the development of the missile sites.
-Kennedy's team decided that if a U-2 were to be shot down, the most likely SAM site would be immediately fired upon.\footcite[p.~217]{caro_passage_2013} Then, on October 27, it happened---a U-2 was shot down flying over Cuba and the pilot was killed.
-Kennedy flinched; he had just sent Khrushchev an offer that might lead to peace, and he didn't want to jeopardize it.
-Against the advice of administration hard-liners, including Vice President Johnson, Kennedy refused to authorize the strike against the Cuban SAM site.\footcite[p.~220]{caro_passage_2013} The next day, Khrushchev agreed to Kennedy's terms, and the Cuban Missile Crisis ended with only a single casualty---the U-2 pilot, Major Rudolph Anderson.
-
-% As his title indicates, Rudolph Anderson was an Air Force Officer, not a member of the CIA.
-During the crisis, the President transferred control of Cuban U-2 reconnaissance to the Air Force, and authorized it to fly as many missions as necessary over the island.\footcite[p.~208-209]{pedlow_central_1992} Khrushchev did not say much about the Cuban U-2 in his writing, but he was very clear on one point: he did not order the SAM strike, and was worried that Kennedy \enquote{might not be able to absorb this blow.}\footcite[p.~338]{khrushchev_memoirs_2007} Shooting down an American aircraft risked escalation, and that was not something Khrushchev wanted.
-This was the second time in history that Soviet forces had shot down a U-2 plane---but even though the spy plane was flying over Cuban airspace and piloted by an US Air Force Major, it was the Soviets who were worried that shooting it down had been an unnecessarily aggressive act.
-
-% Even when a U-2 operation went poorly, it still played by the rules of the game.
-Powers opted against using his suicide capsule, but the macabre truth about the U-2 is that pilots weren't supposed to survive a crash at all.
-The United States originally publicized the weather craft story because Eisenhower had been assured that Gary Powers could not possibly be alive to tell the Russians the truth.\footcite[p.~35]{lindgren_trust_2000} When Powers confessed to his Soviet captors, he told them, correctly, that he was a civilian pilot with the CIA.
-His cover had been blown, but underneath it he was still a spy, and in pretty much all respects he and the U-2 were treated accordingly.
 

### 2020-01-01 article.tex
Andrew upload

@@
 
-% \section{Spy satellites}
+\section{Spy satellites}
+Gary Powers had been in Soviet prison for less than three months when the United States launched an instrument of espionage far more advanced, and far more damaging, than the U-2 could ever be.
+Unlike the U-2, the first generation of spy satellite had limitless airspace, no countermeasures, and no technological antecedents; the world had never seen anything like it.
+In 1955, when the US and the USSR both announced their plans to build an artificial satellite, Outer Space was a true blank slate for international norms.
+
+Only four years after the first U-2 overflight, a familiar dynamic could have played out when the first spy satellite launched in August 1960: the US develops a new photo-reconnaissance technology which its main adversary is unable to prevent; the USSR diplomatically objects to the technology's use; the USSR finally develops a weapon capable of counteracting the new technology; and the USSR employs it.
+Each one of these steps came to pass except the last---even after developing a feasible anti-satellite (ASAT) weapon, the Soviet Union never made use of it.
+And the onus was on the Soviets to push back against American efforts to normalize satellite reconnaissance.
+\emph{Sputnik} briefly put the USSR at the forefront of the final frontier, but the satellite itself was totally innocuous; it collected data about the density of the atmosphere and propagation of radio waves.\footcite{nasa_sputnik_2019} It was the United States that would first launch a satellite capable of photo-reconnaissance---and it was the Soviet Union that allowed it to happen.
+
+Spy satellites are not just one of the best edge cases for norm against shooting down a satellite, they are the reason it exists in the first place.\footnote{I'm not suggesting that, absent spy satellites, nations would have just constantly shot down each other's weather satellites.
+But remember that the CIA planned a cover for the U-2 in which it was a ``weather plane'' that had gone off course.
+The USSR forced a cargo plane to land in Hungary because they suspected it of espionage.
+That never happens in space, because intelligence norms are strong enough that the accusation of dual-use is never even raised.
+There are no good and bad satellites; as long as a satellite doesn't have literal weapons, then there is no plausible reason to shoot it down.} The reconnaissance satellite succeeded because the Eisenhower administration purposefully designed its space program to establish a norm of peaceful development in space---with the single goal of legitimizing satellites for intelligence-gathering.
+By protecting the peaceful uses of outer space, the United States finally gained unrestricted access to aerial intelligence of the type that an Open Skies policy might have provided.
+That all satellites, regardless of purpose, are immune to hostilities as a result is just a happy byproduct.
+
+In this chapter I will explain how the United States developed a norm protecting the use of reconnaissance satellites.
+Unlike in previous chapters, it is not necessary to prove that there is a muted diplomatic response to satellite espionage because the lack of response is self-evident: no military action has ever been taken against satellites, and the Soviet Union only objected to reconnaissance satellites until it were able to make its own.
+The first section looks at the 10-year period prior to the first spy satellite, during which the United States built a diplomatic strategy that would legitimize its eventual launch.
+The section after analyzes the deployment of early spy satellites, and how it altered the security dynamic between both the US and the USSR.
+The chapter ends by comparing the possible explanations for why the peaceful uses of space norm endured in spite of the threat that it posed to Soviet security.
 

### 2020-01-06 article.tex
Add satellite security situation context

--- 2019-12-24 article.tex
@@
 
-% \section{Spy satellites}
+\section{Spy satellites}
+Gary Powers had been in Soviet prison for less than three months when the United States launched an instrument of espionage far more advanced, and far more damaging, than the U-2 could ever be.
+The first generation of spy satellite had limitless airspace access, no countermeasures, and no technological antecedents.
+It was impossible to predict how the the international order would react; in 1955, when the US and the USSR both announced their plans to build an artificial satellite, Outer Space was a true blank slate for international norms.
+
+Only four years after the first U-2 overflight, a familiar dynamic could have played out when the first spy satellite launched in August 1960: the US develops a new photo-reconnaissance technology which its main adversary is unable to prevent; the USSR diplomatically objects to the technology's use; the USSR finally develops a weapon capable of counteracting the new technology; and the USSR employs it.
+Each one of these steps came to pass except the last---even after developing a feasible anti-satellite (ASAT) weapon, the Soviet Union never made use of it.
+And the onus was on the Soviets to push back against American efforts to normalize satellite reconnaissance.
+\emph{Sputnik} briefly put the USSR at the forefront of the final frontier, but the satellite itself was totally innocuous; it collected data about the density of the atmosphere and propagation of radio waves.\footcite{nasa_sputnik_2019} It was the United States that would first launch a satellite capable of photo-reconnaissance---and it was the Soviet Union that allowed it to happen.
+
+\subsection{Freedom of Space}
+Spy satellites are not just one of the best edge cases for norm against shooting down a satellite, they are the reason it exists in the first place.
+The reconnaissance satellite succeeded in its mission because the Eisenhower administration purposefully designed its space program to establish a norm of peaceful development in space---with the single goal of legitimizing satellites for intelligence-gathering.
+No military action was ever been taken against satellites, and the Soviet Union only objected to reconnaissance satellites until it was able to make its own.
+By protecting the peaceful uses of outer space, the United States finally gained unrestricted access to aerial intelligence.
+That all satellites, regardless of purpose, are immune to hostilities as a result is just a happy byproduct.
+
+In 1954, President Eisenhower commissioned a task force which would tackle the problem of America's vulnerability to a surprise attack, the Technological Capabilities Panel (TCP).\footcite[p.~67.
+Also variously referred to as the \enquote{Killian Report,} \enquote{Surprise Attack Study,} or its formal title, \enquote{Meeting the Threat of a Surprise Attack.}]{killian_sputnik_1977} The TCP report concluded that any great power conflict would result in heavy American losses, so all necessary steps should be taken to avoid one.
+Its recommendations obviously included furthering American missile capabilities---traditional deterrence through force---but intelligence-gathering capabilities took comparable priority.
+``We \emph{must} find ways to increase the number of hard facts upon which our intelligence estimates are based,'' the TCP Report states, ``to provide better strategic warning, to minimize surprise in the kind of attack, and to reduce the danger of gross overestimation or gross underestimation of the threat.''\footcite{technological_capabilities_panel_meeting_1955} The TCP recommendations---reviewed personally by President Eisenhower---were hugely consequential across US government policy, but two projects that came out of it are especially relevant here: the U-2 plane and the artificial spy satellite.
+
+The U-2 plane made the leap from concept to reality much faster than the artificial satellite, but even before the Soviet Union became capable of shooting it down, the intelligence that the U-2 provided had limitations.
+While the U-2 provided CIA imagery analysts with impossibly high-resolution photos of specific locations, it was less helpful in finding unknown military bases throughout the vast expanse of Soviet territory.
+The first generation of artificial satellites promised to change that.
+In the time it takes a satellite to complete a full orbit, the Earth has rotated slightly underneath it, so with each pass a satellite launched into polar orbit photographs an area that has been longitudinally shifted.\footnote{Think of a winding a piece of string around a ball with your right hand, while slowly rotating the ball with your left; after a series of passes, you've covered a broad slice of the ball with string.} With enough orbits, a satellite can photograph a wide slice of Earth.
+Together, planes and satellites could operate for complementary photo-reconnaissance purposes: a plane dispatched when high quality localized intelligence was required, and a satellite to form a comprehensive picture of the Soviet military posture.
+And of course, satellites could surveil dangerous territory without risking the life of a pilot.
+
+Preventing a surprise nuclear attack could not be done without better intelligence and better intelligence could not be achieved without comprehensive photo-reconnaissance---using a technique that the Soviet Union would accept.
+The solution was to spy from space.
+\enquote{Camera-toting satellites,} historian Walter McDougall explains, \enquote{circling the earth south to north [sic] in a polar orbit, could view the entire surface of the earth as it rotated below, return to any location in a few days' time, hone in on suspicious areas, and do it all under the legal cover of freedom of space---if such legal cover could be established.}\footcite[p.~117]{mcdougall_heavens_1985} The key phrase here is ``if such a legal cover could be established.'' The Soviets considered  U-2 overflights to be a violation of their territorial sovereignty, and continued to register diplomatic protests about them; the overflight program was so risky that it was originally intended to last only a year or two.\footcite[p.~33]{lindgren_trust_2000} The undefined legal status of space presented an opportunity, but also a challenge.
+The TCP understood that the Soviet Union would likely recognize the enormous advantage that a photo-reconnaissance satellite would confer on the US, and object to their deployment using any international norms possible.
+It was therefore necessary to present reconnaissance satellites in a manner that would limit the grounds for objection.
+By the time that a reconnaissance satellite launched, it would have to have already been established that everyone had the unquestioned right to launch a satellite of any kind; this principle would come to be called \enquote{Freedom of Space.}
+
+\subsection{Spiking the space race}
+Everything about the administration's space policy was carefully crafted to set the stage for the eventual launch of a spy satellite.
+From the time that artificial satellites first appeared viable until the day he left office, Eisenhower used the civilian space program as a public means of diverting attention from the security-related programs he valued.\footcite[p.~119]{day_eye_2015} Acting on the recommendations of the TCP, the National Security Council (NSC) proposed a scientific satellite program in May 1955, which Eisenhower accepted.
+In this memo, NSC-5520, the NSC Planning Board argued that, when framed properly, a scientific satellite could be used as diplomatic tool to promote a norm establishing ``freedom of space.'' The proposal tracked closely with a similar one from the 1950 RAND report, which suggested that the US first launch an innocuous ``experimental satellite'' that did not cross over Russian territory, and gauge the diplomatic reaction.\footcite[p.~21]{kecskemetic_satellite_1950} The political framing of the satellite is discussed at the top of NSC-5520, in a section entitled \emph{General Considerations} (emphasis mine):
+
+% \begin{quote}
+	6.
+\textelp{} \textbf{Furthermore, a small scientific satellite will provide a test of the principle of “Freedom of Space”.
+The implications of this principle are being studied within the Executive Branch.} However, preliminary studies indicate that there is no obstacle under international law to the launching of such a satellite.
+
+	7.
+\textbf{It should be emphasized that a satellite would constitute no active military offensive threat to any country over which it might pass.} Although a large satellite might conceivably serve to launch a guided missile at a ground target, it will always be a poor choice for the purpose.
+A bomb could not be dropped from a satellite on a target below, because anything dropped from a satellite would simply continue alongside in the orbit.
+\footcite{nsc_planning_board_draft_1955}
+% \end{quote}
+Even though there was no international law prohibiting the use of satellites, a theoretical ``freedom of space'' principle depended on the first satellite not having military purposes.
+What NSC-5520 did not anticipate was that the US wouldn't have to establish that principle on their own.
+On October 4, 1957, the Soviet Union launched the world's first artificial satellite, a small metal ball named \emph{Sputnik}.
+
+Getting beaten to orbit by \emph{Sputnik} put the United States at an advantage.
+The Soviets could hardly denounce the American satellite program as suspicious when they very publicly made the first move---and without asking permission to orbit over the United States.\footcite[p.~40]{peebles_corona_1997} The US didn't \emph{try} to lose first leg of the space race, per se, but policymakers were aware well in advance that being second to launch a satellite had its advantages.
+Reviewing NSC-5520, Nelson Rockefeller wrote that since the Soviets had already announced plans to launch a satellite, the existence of their project could be used to good effect against inevitable Soviet protests about the American satellite program.\footcite[p.~120]{mcdougall_heavens_1985} Though ideally the US would be able to set the precedent for ``freedom of space'' itself, senior administration officials understood that if the USSR launched first, then the American satellite program would be almost automatically justified.
+
+Not only did the Eisenhower administration recognize that being second to orbit came with a significant consolation prize, they were not particularly focused on winning the space race to begin with.
+Launching the first satellite was assigned the lowest possible priority because the administration correctly determined that their other goals---finishing the ICBM, monitoring Soviet R\&D, and framing the satellite program as a civilian scientific endeavor---were of greater strategic importance.\footcite[p.~123-124]{mcdougall_heavens_1985} Eisenhower himself repeatedly dismissed concerns about the prestige of being the first nation to orbit an object in space, openly contemptuous of the idea that space was any kind of race at all.\footcite[p.~100]{lindgren_trust_2000} His complete disdain for what he called ``stunts'' would prove to be disastrous politically, but from the standpoint of international law, his restraint was crucial to establishing a strategically beneficial precedent.\footcite[p.~134]{day_eye_2015}
+
+In fact, the overriding American priority to legitimize reconnaissance satellites is probably the \emph{reason} that the Soviets started the Space Age instead.
+Just as Eisenhower limited the involvement of the Air Force in Aquatone to paint the U-2 as a civilian project, he made technical decisions that were intended to make the scientific satellite seem as non-threatening as possible, and none was more important than the vehicle that would launch it.
+The group tasked with implementing NSC-5520 had to choose between competing launch proposals from the Army, Navy, and Air Force.
+It made the somewhat surprising choice to go with the unproven Navy Research Laboratory (NRL) \emph{Vanguard} rocket, even though the Army Ballistic Missile Agency (ABMA) engineers, headed by famous rocketeer Wernher von Braun, had been developing their proposal for a year.
+Among the reasons the committee gave for their decision, two stand out.
+First, the NRL impressed the committee with its plans for scientific components and a radio tracking system.\footcite[p.~122]{mcdougall_heavens_1985} Second, the panel specifically wanted to avoid using a ballistic missile to launch a scientific satellite.\footcite[p.~129.
+Though the NRL is run by the Navy, it functions like a civilian scientific operation.
+Everything that I have read about this decision seems to take it for granted that despite the Navy involement, the NRL proposal would be seen as a civilian project.
+This makes sense to me, especially when you consider that it was competiting with the same agency that was building nuclear delivery systems.]{day_eye_2015}
+
+Launching the first satellite with a ballistic missile threatened to compromise both of the principal national security priorities.
+It might delay the development of the ICBM, an existential threat to American military superiority, and it would jeopardize the civilian character of the satellite program, which the committee members knew was necessary to establish ``freedom of space.''\footcite[p.~122]{mcdougall_heavens_1985} The Von Braun engineers vigorously protested the decision, because they believed---correctly---that their proposal would succeed first.
+But being first in space was not the goal, establishing a legal precedent was.
+Von Braun's appeals were continually rejected, and always for the same reason: launching a satellite with no scientific purpose other than demonstrating superiority would compromise the moral high ground which the US desperately needed to secure.\footcite[p.~131]{day_eye_2015}
+
+Publicly, \emph{Vanguard} was a disaster.
+The rocket experienced an embarrassing setback on December 6, 1957, when it failed to launch a three-and-a-half pound satellite, just after the USSR had successfully launched \emph{Sputnik II}\footcite[p.~119]{killian_sputnik_1977} This time, the satellite carried a dog---the Soviets put a dog in space before the Americans even managed to get a satellite up there.\footcite[Her name was Laika.]{george_sad_2018} When then United States finally succeeded in launching at satellite, it was not thanks to the \emph{Vanguard}, but rather the ABMA team that had been consistently rebuffed by the administration.
+Finally given the chance, Von Braun and his engineers quickly accomplished what they had always said they would: modify a Jupiter-C ballistic missile and send a satellite into orbit.
+The Army launched \emph{Explorer 1} on January 31, and it carried a pioneering array of miniaturized electronics, including two micrometeoroid detectors and a Geiger counter.\footcite[p.~168]{mcdougall_heavens_1985} \emph{Vanguard 1}, which launched about six weeks later, and \emph{Explorer 1} provided an impressive amount of scientific data in comparison to the Sputniks.\footcite[p.~168]{mcdougall_heavens_1985} More importantly, even though the US did end up launching their first satellite using a ballistic missile, its visible defeat in the space race and clear scientific purpose had the intended effect of blunting Soviet efforts to characterize the American satellite program as a military development.
+It was time to launch a spy satellite.
 
 % \section{Conclusion}
+\end{document}
 
-\end{document}
+% The breadth of information that a satellite could provide might also dispel the Eisenhower administration's political albatross---his supposedly dovish defense spending.
+Eisnhower faced constant pressure to increase the defense budget and close the alleged ``missile gap'' between the US and the USSR.
+While U-2 photography (codenamed ``Talent'') had been helpful for pushing back against these demands, the diplomatic volatility of each flight limited the amount of ground that it could cover.
+In July 1960, the CIA's Photographic Interpretation Center conducted a study to determine how much of the Soviet territory that it deemed suitable for ICBM deployment had been observed by Talent photography since January 1959.
+As it turned out, very little.
+The ``intelligence bonanza'' from Aquatone had still only observed roughly 13.6\% of the viable Soviet terrain.
+Visual observations from other sources added an additional 1.5\%.\footcite[A significant portion of this document is redacted, including, for some reason, the page numbers.]{cia_future_1960} Satellites showed enormous potential for covering that remaining 85\%, and in doing so, shut down discussion of the ``missile gap'' once and for all.
 

### 2020-01-06 article.tex
Edit up to satellite conclusion

@@
 
-\subsection{Evaluating the consequences of aerial espionage}
+\subsection{Espionage norms in aerial reconnaissance}
 The U-2 Incident is often cited as a seminal failure of the Eisenhower administration.
@@
 It was therefore necessary to present reconnaissance satellites in a manner that would limit the grounds for objection.
-By the time that a reconnaissance satellite launched, it would have to have already been established that everyone had the unquestioned right to launch a satellite of any kind; this principle would come to be called \enquote{Freedom of Space.}
+By the time that a reconnaissance satellite launched, it would have to have already been established that everyone had the unquestioned right to launch a satellite of any kind; this principle would come to be called \enquote{freedom of space.}
 
@@
 Everything about the administration's space policy was carefully crafted to set the stage for the eventual launch of a spy satellite.
-From the time that artificial satellites first appeared viable until the day he left office, Eisenhower used the civilian space program as a public means of diverting attention from the security-related programs he valued.\footcite[p.~119]{day_eye_2015} Acting on the recommendations of the TCP, the National Security Council (NSC) proposed a scientific satellite program in May 1955, which Eisenhower accepted.
+From the time that artificial satellites first appeared viable until the day he left office, Eisenhower used the civilian space program to divert attention from the security-related space programs he valued.\footcite[p.~119]{day_eye_2015} Acting on the recommendations of the TCP, the National Security Council (NSC) proposed a scientific satellite program in May 1955, which Eisenhower accepted.
 In this memo, NSC-5520, the NSC Planning Board argued that, when framed properly, a scientific satellite could be used as diplomatic tool to promote a norm establishing ``freedom of space.'' The proposal tracked closely with a similar one from the 1950 RAND report, which suggested that the US first launch an innocuous ``experimental satellite'' that did not cross over Russian territory, and gauge the diplomatic reaction.\footcite[p.~21]{kecskemetic_satellite_1950} The political framing of the satellite is discussed at the top of NSC-5520, in a section entitled \emph{General Considerations} (emphasis mine):
+\newline
 
-% \begin{quote}
+\begin{quoteblock}
 6.
@@
 \footcite{nsc_planning_board_draft_1955}
-% \end{quote}
-Even though there was no international law prohibiting the use of satellites, a theoretical ``freedom of space'' principle depended on the first satellite not having military purposes.
+\newline
+\end{quoteblock}
+
+Even though there existed no international law prohibiting the use of satellites, a theoretical ``freedom of space'' principle depended on the first satellite not having military purposes.
 What NSC-5520 did not anticipate was that the US wouldn't have to establish that principle on their own.
@@
 Getting beaten to orbit by \emph{Sputnik} put the United States at an advantage.
-The Soviets could hardly denounce the American satellite program as suspicious when they very publicly made the first move---and without asking permission to orbit over the United States.\footcite[p.~40]{peebles_corona_1997} The US didn't \emph{try} to lose first leg of the space race, per se, but policymakers were aware well in advance that being second to launch a satellite had its advantages.
-Reviewing NSC-5520, Nelson Rockefeller wrote that since the Soviets had already announced plans to launch a satellite, the existence of their project could be used to good effect against inevitable Soviet protests about the American satellite program.\footcite[p.~120]{mcdougall_heavens_1985} Though ideally the US would be able to set the precedent for ``freedom of space'' itself, senior administration officials understood that if the USSR launched first, then the American satellite program would be almost automatically justified.
+The Soviets could hardly denounce the American satellite program as suspicious when they very publicly made the first move---and without asking permission to orbit over the United States.\footcite[p.~40]{peebles_corona_1997} The US didn't \emph{try} to lose first leg of the space race exactly, but policymakers were aware well in advance that being the second nation to launch an artificial satellite would further their strategic goals.
+Reviewing NSC-5520, Nelson Rockefeller wrote that since the Soviets had already announced plans to launch a satellite, the existence of the Soviet project could be used to good effect against inevitable Soviet protests against American photo-reconnaissance satellites.\footcite[p.~120]{mcdougall_heavens_1985} Though ideally the US would set the precedent for ``freedom of space'' itself, senior administration officials understood that if the USSR launched first, then the American satellite program would be almost automatically justified.
 
-Not only did the Eisenhower administration recognize that being second to orbit came with a significant consolation prize, they were not particularly focused on winning the space race to begin with.
-Launching the first satellite was assigned the lowest possible priority because the administration correctly determined that their other goals---finishing the ICBM, monitoring Soviet R\&D, and framing the satellite program as a civilian scientific endeavor---were of greater strategic importance.\footcite[p.~123-124]{mcdougall_heavens_1985} Eisenhower himself repeatedly dismissed concerns about the prestige of being the first nation to orbit an object in space, openly contemptuous of the idea that space was any kind of race at all.\footcite[p.~100]{lindgren_trust_2000} His complete disdain for what he called ``stunts'' would prove to be disastrous politically, but from the standpoint of international law, his restraint was crucial to establishing a strategically beneficial precedent.\footcite[p.~134]{day_eye_2015}
+Not only did the Eisenhower administration recognize that being second to orbit came with a significant consolation prize, it was not particularly focused on winning the space race to begin with.
+Launching the world's first artificial satellite was assigned the lowest possible priority because the administration correctly determined that its other goals---finishing the ICBM, monitoring Soviet R\&D, and framing the satellite program as a civilian scientific endeavor---were of greater strategic importance.\footcite[p.~123-124]{mcdougall_heavens_1985} Eisenhower himself repeatedly dismissed concerns about the prestige of being the first nation to orbit an object in space, openly contemptuous of the idea that space was any kind of race at all.\footcite[p.~100]{lindgren_trust_2000} His complete disdain for what he called ``stunts'' would prove to be disastrous politically, but his restraint was crucial to establishing a strategically beneficial precedent.\footcite[p.~134]{day_eye_2015}
 
@@
 First, the NRL impressed the committee with its plans for scientific components and a radio tracking system.\footcite[p.~122]{mcdougall_heavens_1985} Second, the panel specifically wanted to avoid using a ballistic missile to launch a scientific satellite.\footcite[p.~129.
-Though the NRL is run by the Navy, it functions like a civilian scientific operation.
-Everything that I have read about this decision seems to take it for granted that despite the Navy involement, the NRL proposal would be seen as a civilian project.
-This makes sense to me, especially when you consider that it was competiting with the same agency that was building nuclear delivery systems.]{day_eye_2015}
+Though the NRL is run by the Navy, it functioned like a civilian scientific operation and all parties involved believed that the NRL proposal would be seen as a civilian project---especially in comparison with its competetor, the Army agency building nuclear delivery systems.]{day_eye_2015}
 
@@
 Publicly, \emph{Vanguard} was a disaster.
-The rocket experienced an embarrassing setback on December 6, 1957, when it failed to launch a three-and-a-half pound satellite, just after the USSR had successfully launched \emph{Sputnik II}\footcite[p.~119]{killian_sputnik_1977} This time, the satellite carried a dog---the Soviets put a dog in space before the Americans even managed to get a satellite up there.\footcite[Her name was Laika.]{george_sad_2018} When then United States finally succeeded in launching at satellite, it was not thanks to the \emph{Vanguard}, but rather the ABMA team that had been consistently rebuffed by the administration.
+The rocket experienced an embarrassing setback on December 6, 1957, when it failed to launch a three-and-a-half pound satellite, just after the USSR had successfully launched \emph{Sputnik II}.\footcite[p.~119]{killian_sputnik_1977} This time, the satellite carried a dog---the Soviets put a dog in space before the Americans even managed to get a satellite up there.\footcite[Her name was Laika.]{george_sad_2018} When then United States finally succeeded in launching at satellite, it was not thanks to the \emph{Vanguard}, but rather the ABMA team that had been originally shunned by the administration.
 Finally given the chance, Von Braun and his engineers quickly accomplished what they had always said they would: modify a Jupiter-C ballistic missile and send a satellite into orbit.
-The Army launched \emph{Explorer 1} on January 31, and it carried a pioneering array of miniaturized electronics, including two micrometeoroid detectors and a Geiger counter.\footcite[p.~168]{mcdougall_heavens_1985} \emph{Vanguard 1}, which launched about six weeks later, and \emph{Explorer 1} provided an impressive amount of scientific data in comparison to the Sputniks.\footcite[p.~168]{mcdougall_heavens_1985} More importantly, even though the US did end up launching their first satellite using a ballistic missile, its visible defeat in the space race and clear scientific purpose had the intended effect of blunting Soviet efforts to characterize the American satellite program as a military development.
+The Army launched \emph{Explorer 1} on January 31, and it carried a pioneering array of miniaturized electronics, including two micrometeoroid detectors and a Geiger counter.\footcite[p.~168]{mcdougall_heavens_1985} \emph{Vanguard 1}---which finally launched about six weeks later---and \emph{Explorer 1} provided an impressive amount of scientific data in comparison to the Sputniks.\footcite[p.~168]{mcdougall_heavens_1985} More importantly, even though the US did end up launching its first satellite using a ballistic missile, its visible defeat in the space race and clear scientific aims had the intended effect of blunting Soviet efforts to characterize the American satellite program as a military development.
 It was time to launch a spy satellite.
 
-% \section{Conclusion}
+\subsection{The Corona Project}
+The Soviet Union unintentionally set an early precedent for ``freedom of space,'' but that did not automatically justify using space for reconnaissance.
+As far as the international community was concerned, it had only been established that overflights were permissible for scientific purposes; military reconnaissance was still an open question.\footcite[p.~47-48]{peebles_corona_1997} Now that the space age was underway, the next step was to normalize photographic satellite overflights.
+Success was going to require tight discipline across all levels of government, but information about WS-117L, the Air Force reconnaissance satellite program, had already begun to leak.\footcite[p.~96]{lindgren_trust_2000}
+
+Like many of Eisenhower's decisions on space policy, his solution for reconnaissance satellites was a brilliant long-term success---but a complete mystery to the public and utterly baffling to most of the military establishment.
+WS-117L was originally intended to be a family of reconnaissance satellites.
+Eisenhower simply selected the most immediately promising project---a single-use film-retrieval satellite, codenamed Corona---and canceled it.
+Air Force personnel were ``thunderstruck.''\footcite[p.~45]{peebles_corona_1997} With public bewilderment instilled and American reconnaissance satellites now seemingly a few years out, the administration began a multi-pronged approach to legitimizing spy satellites.
+Publicly, the US government advocated for a ``freedom of space'' policy that explicitly included reconnaissance satellites; quietly, it resurrected the Corona project under deep-cover CIA administration.
+
+The public half of Eisenhower's strategy required international approval for space-based overflights, so the US advocated for a United Nations Ad Hoc Committee on the Peaceful Uses of Outer Space (COPUOUS) to deal with this issue.
+The Soviets, worried about having the limitations of their ICBM program exposed, refused to participate for years.\footcite[p.~140]{day_eye_2015} Meanwhile, the actual spy satellite program progressed at an accelerated pace, and with unmatched secrecy.
+The team started with only 30 personnel and eventually swelled to 300, almost none of whom knew the entire scope of the project.
+Team members had to take different routes to work to avoid being followed and were never allowed to say the word ``Corona'' on the telephone, or even the abbreviation ``C.''\footcite[p.~51]{peebles_shadow_2000} As far as the government was concerned, Corona's cover was bulletproof.
+When the satellite began making test launches, the US heavily promoted its new \emph{Discoverer} program as a scientific achievement---confident that its status as a front for the Corona project could not be proven false the way the U-2's cover had been.
+When it finally succeeded in recovering a \emph{Discoverer} satellite from orbit, the US government released photographs of the first successful capsule triumphantly floating in the water, and an American flag retrieved from within was presented to the president.\footcite[p.~83]{peebles_shadow_2000}
+
+Today's Earth-imaging satellites have enormous solar panels, radio their images down to command stations, and have mission durations in a scale of years.\footnote{This design simply did not work with film.
+A rival project that developed the film midair and transmitted the data back down failed to provide intelligence of sufficient quality.
+(\cite[p.~203-204]{brugioni_eyes_2010}) The last film-based orbiting reconnaissance camera developed for the U.S.
+Government was the KH-9 HEXAGON, which operated between 1971 and 1986.
+It was famously complicated, with four separate film-reentry pods (\cite{pressel_spy_2013}).
+I presume the rest since then have been digital, but because the HEXAGON was only declassified in September 2011, it will likely be quite a long time before we find out how they work.} Corona was nothing like that.
+Instead, each satellite was little more than a monstrously expensive disposable camera.
+The average Corona flight looked like this: the satellite would be sent into space carrying enough film for a 24-hour mission, and its camera would switch on every time the satellite flew over denied territory.
+After completing seventeen orbits, the satellite would pitch down 60 degrees and eject the film capsule back down to Earth.
+Various rocket systems would stabilize it, a heat shield would protect it, and a series of parachutes would decelerate it.
+What remained of the capsule would hopefully descend within a 200-by-60 mile recovery zone (``the ballpark''), but at the very least within the boundaries of an additional 400-mile section (``the outfield'').
+Ideally, it would be caught in midair by a cargo plane.\footcite[p.~56]{peebles_corona_1997}
+
+Why is that significant?
+Because at no point was satellite reconnaissance a \emph{fait accompli}---each launch was another opportunity for the Soviet Union to object to the continued use of a potentially illegal surveillance method.
+And there were a great many launches.
+From February to August 1960, right in the middle of the U-2 incident, twelve test launches failed before the US achieved the first successful recovery of a Corona satellite---of any object from orbit in the human history, for that matter---when it fished \emph{Discoverer 13} out of the sea.
+The US repeated that feat with \emph{Discoverer 14} a week later, and this time it carried a camera, bringing invaluable intelligence back from space that dealt a mortal blow to the myth of the missile gap.\footcite[p.~101-102.
+The previous failures had been so demoralizing that \emph{Discoverer 13} only carried diagnostic equipment, which made its success all the more exciting.
+Ships on standby were able to retrieve the satellite even though it fell outside the recovery zone---\emph{Discoverer 13} many not have landed in the ballpark or the outfield, but in many respects, the mission was a home run.]{lindgren_trust_2000} The launches kept coming---the next one to return high quality photos was \emph{Discoverer 18} in December.
+Eisenhower left office soon after, but between \emph{Discoverer 18} and \emph{Discoverer 38}, twenty launches were performed, two of which carried actual radiometry equipment (in support of Corona's cover story), and seven produced film of high enough quality for the mission to be considered a complete success.\footcite[p.~103]{lindgren_trust_2000}
+
+Like the U-2 overflights, this intricate process was immediately identified by the Soviets for what it was: a military reconnaissance effort.
+The Soviet Union protested the very first \emph{Discoverer} launch, correctly alleging its military utility even though the satellite had failed in its purpose, and would fail eleven more times.\footcite[p.~140]{day_eye_2015} Until 1963, Soviet propaganda continually railed against the Americans, in domestic media and international forums, for attempting to militarize space.\footcite[p.~271]{mcdougall_heavens_1985} The USSR sponsored a UN resolution to prohibit espionage from space, and Khrushchev vowed that his military would shoot down spy satellites, just as they had the U-2.\footcite[p.~166]{day_eye_2015} In the first few years after the United States developed satellite reconnaissance capability, it appeared that the USSR was acting exactly as the 1950 RAND report had predicted, and using the legal means at their disposal to end the satellite overflights.
+
+Khrushchev had every incentive to discourage satellite reconnaissance, because the intelligence that the US gained through Corona was absolutely devastating to Soviet claims about their military prowess.
+The U-2 overflights had sufficiently debunked the ``bomber gap,'' and seemed to suggest that no ``missile gap'' existed either.\footcite{goodpaster_cold_2003} But as the CIA report on U-2 photography showed, being unable to find ICBM sites was not conclusive proof that they did not exist, and the vast majority of possible missile sites remained un-photographed.
+Confusion about Soviet long-range attack capabilities still existed at various levels of government, confusion which ultimately led to the doomed flight of Gary Powers.\footcite[p.~344]{brugioni_eyes_2010} It was the \emph{Discoverer 14} and \emph{Discoverer 18} missions that successfully photographed the sites that intelligence analysts considered most likely to host operational ICBMs, and found nothing.\footcite[p.~379]{brugioni_eyes_2010}
+
+To make matters worse for the Soviets, the moment at which the US discovered that the USSR had far fewer missiles than they claimed was the exact moment at which that disparity was greatest.
+The missile gap actually went in the other direction---in 1962, the US had more than double the ICBMs, maintained a substantial lead in manned bombers, and had just begun to deploy nuclear submarines.\footcite[p.~251]{mcdougall_heavens_1985} By the time of the 1960 presidential election, both sides of the Cold War had a generally accurate read on each others' military capabilities, and they knew that their adversary was properly appraised of the strategic situation as well.
+If Khrushchev genuinely feared an American attack, absolutely nothing would have been a greater threat to Soviet security than the collapse of the USSR's most significant deterrent argument---a collapse brought about by American spy satellites.
+
+Not only did Corona make a theoretical offensive attack more likely, it made the plans for such an attack significantly more effective.
+The impact of the project on military planning is best summarized by how it affected the joint American-British plan for an offensive against the Soviet Union, which aerospace historian Curtis Peebles describes here:
+
+\begin{quoteblock}
+Starting in 1962, the Corona photos were also used for nuclear targeting.
+Corona photos provided the first accurate coordinates of Soviet military facilities.
+This resulted in a shift in U.S.
+policy.
+During the 1950s, little was known about the location of Soviet military bases.
+The lack of information was reflected in the first join SAC-RAF Bomber Command strike plan.
+The plan, which went into effect on October 1, 1958, assigned 106 Soviet targets to the RAF bomber force: 69 cities, 17 airfields, and 20 air defense targets.
+A full two-thirds of the targets were cities.
+This disparity was ever greater with the unilateral RAF strike plan.
+If Great Britain should be forced to attack alone, ninety-eight Soviet cities would be bombed.
+
+On August 1, 1962, a revised joint plan went into effect.
+The changes, which reflected a full two years of Corona operations, were profound.
+The RAF Thor IRBMs and bomber force would strike forty-four ``offensive capability'' targets such as airfields, ten ``defensive capability'' targets such as air defense control centers, twenty-eight Soviet IRBMs sites, and only sixteen cities.
+Urban targets had gone from the majority to next to last in number.
+The unilateral British plan also underwent a similar change.\footcite[p.~139]{peebles_corona_1997}
+\end{quoteblock}
+
+The Soviet Union lost its most significant military advantage over the West in a little under two years.
+No subsequent missile buildup or bunker hardening or second-strike capability could undo the damage done to the Soviet military posture when \emph{Discoverer XIV} and its successors sailed past the Iron Curtain.
+And yet the only thing the Soviet Union did in response was publicly protest the spy satellites, and refuse to accept American proposals for a legal regime that included the rights of reconnaissance overflights.
+Those protests continued for about 3 years, until the Soviets developed their own working spy satellites, ``freedom of space'' became a mutually-agreeable reality.\footcite[p.~271-275]{mcdougall_heavens_1985}
+
+% Start editing after here! So close!
+\subsection{Espionage norms in satellite reconnaissance}
+It seems obvious that the Soviet Union had almost everything to lose by legitimizing the use of spy satellites.
+Aerial reconnaissance neutralized one of Soviet Union's asymmetric structural advantages---the inherent secrecy of Soviet society---and put a dent in their claims to an arms advantage as well.
+What it gained in return is tough to say, but there is no possibility that it matched the intelligence windfall that befell the CIA.
+The legitimization of spy satellites dealt an indisputable blow to Soviet standing.
+So why did the Soviet Union let it happen?
+
+% Make these all one, maybe two paragraphs
+
+The brilliance of the American strategy goes a long way towards explaining why the USSR ultimately allowed satellites to fly over its territory.
+Both the overt and covert halves of the strategy played an important role.
+If the US had not been engaging in multinational talks about science and the freedom of space at the same time it was launching spy satellites, then it is very likely that upon developing ASAT capabilities, the USSR would have simply started shooting them down.
+That would have been an almost identical situation to the U-2 saga, and it was essential that when ASATs became available, the USSR not presume to have the unquestioned right to use them.
+
+But if the US had not been simultaneously running a covert program before the norm was established, then how would the norm have ever been established at all?
+At what point would the United States have determined that it was officially legal to launch a spy satellite?
+Soviet intransigence on COPUOS demonstrated that they were unwilling to formally acknowledge the use of spy satellites, so no legal solution was forthcoming.
+The Soviet Union was also unlikely to launch the first spy satellite; its HUMINT operations were able to provide most of the intelligence that a photo-reconnaissance operation would have, and launching the first spy satellite would have given the US the perfect excuse to close the intelligence gap once and for all.
+The situation that resulted was a perfect storm for the normalization of satellite reconnaissance: the Soviet leadership found that they had the ability to develop a reliable spy satellite much sooner than they could develop a reliable anti-satellite weapon, of which their right to use would still be uncertain.
+The path of least resistance, but greatest preservation of pride, was to simply start spying themselves.
+
+A few Khrushchev own-goals played a part in legitimizing satellite reconnaissance as well.
+At the infamous four-powers summit in 1960, after watching Khrushchev vent about the U-2 Incident, French president Charles de Gaulle calmly told him: ``You're making too big a fuss about the matter.
+There is probably a ton of Russian iron [referring to Russian satellites] coming through French space every day without my permissions.
+I have no idea what is inside those satellites and you have not told me.
+But I am not making a big fuss.'' Khrushchev admitted that the Soviets had put cameras in some satellites, but not necessarily any that flew over France.\footcite[p.~353]{brugioni_eyes_2010} He went on to say that he only considered airplanes to be out of bounds, not satellites.
+In Eisenhower's later paraphrasing, Khrushchev said that ``any nation in the world who wanted to photograph the Soviet areas by satellite was completely free to do so.''\footcite[p.~556]{eisenhower_waging_1965} Even being the first nation to space played a role.
+Of course, it cannot be said that launching the first satellite in human history was a \emph{mistake} on the part of Khrushchev---Eisenhower underestimated the symbolic importance of igniting the Space Age, to his significant political detriment---but it is indisputable that \emph{Sputnik} made it much harder for the USSR to argue that reconnaissance satellites were an imperialist invasion of territorial sovereignty.
+
+\subsection{The rational (mis)calculation explanation}
+And yet, American strategy and Khrushchev's missteps still do not fully explain why Soviet Union continued to permit photo-reconnaissance after ASATs were developed.
+The weapons were feasible; the USSR started testing its ASAT capabilities in 1968.\footcite[p.~149]{karas_new_1983} Given that the Soviet Union was using spy satellites as well, shooting down an American one would have been hard to justify, but it really only risked having some Soviet satellites shot down in response.
+Instead, the USSR choose to help write and ratify the Outer Space Treaty (OST) in 1967, the first legal regime codifying the peaceful uses of space.
+While the OST did not ban ASATs, it nonetheless set a precedent for the peaceful uses of space---and reflected the US policy stance that the main military value of space was as a harbor for reconnaissance satellites.\footcite[p~16]{mowthorpe_militarization_2004}
+
+% Took out Welton Chang, ask if he should come back?
+
+Some might argue that the USSR simply miscalculated the relative gains that they received from a norm protecting all satellites.
+This argument is unpersuasive.
+Khrushchev made it clear in his memoirs that he rejected the Open Skies treaty because he knew it would reveal the weakness at the heart of the Soviet state---that its missile program was nowhere near as developed as Khrushchev claimed.
+The USSR would not reach missile parity with the United States until 1969, after both sides had tested ASAT weapons.\footnote{There is obviously a lot more to relative military strength than just adding up the number of missiles that each side has access to---deployment vehicle, missile quality, and susceptibility to a first-strike are crucial.
+But number of missiles is an easily-digestible metric that politicians pay attention to, and helps visualize just how far behind the Soviet Union was when their actual strength was revealed.} After that, the USSR might have had less to hide, but the fact remains that it still had the opportunity to ``turn off the lights'' by taking action against American spy satellites, something the United States, with its free press and comparatively open borders, could not do.
+Instead, Open Skies began the moment the USSR launched its first spy satellite, and it never ended.
+
+The United States fully believed it had more to gain from a ``freedom of space'' norm---and I agree---but is it possible that the Soviet Union had some reason to believe the same?
+Perhaps, but given the indisputable advantage that the US gained from reciprocal reconnaissance, it is hard to imagine what non-reconnaissance military function a satellite could perform that would advantage the Soviet Union to the same degree.
+Satellites were also used to improve the accuracy of missile guidance systems, but since both sides relied on nuclear missiles as their main deterrent, satellite-based missile targeting likely provide no relative gains.
+Or perhaps the Soviet Union thought being the first nation to shoot down a satellite would adversely impact a key propaganda point: that the United States was responsible for the militarization of space.
+Both the US and the USSR were preoccupied with demonstrating that their respective ideology---capitalism or communism---had an inherent moral superiority.
+They each believed that this propaganda helped project their power in the third world, but that also raises a question: why would the world react adversely to Soviet defense of its own borders, if the United States did in fact break the glass on space militarization?
+
+Satellites are the purest demonstration of the appeal of espionage norms, because unlike with planes, there were no previously existing norms of territorial integrity.
+Therefore the only established norm at the dawn of the space age was one from the world of espionage---that intelligence-gathering entities are not inherently aggressive.
+Space became an environment where a strike against a satellite would constitute the first act of aggression, even if the satellite was operating for a military purpose, gathering information that could significantly compromise a state's security position.
+
+There is a fundamental appeal to a world with tacit mutual reconnaissance, an appeal that is rooted in defensive realism and manifested in espionage norms.
+Spy satellites are the single best proof that even when the relative gains are lopsided, states will choose to live in a world with more information, rather than less.
+The advent of mutual reconnaissance solved what had, until then, been the single most intractable problem of the Cold War---treaty verification.
+Rae Huffstuttler, later a director of the CIA's National Photographic Interpretation Center (the same institution that conducted the survey of U-2 photography), said that ``imagery set the stage for the arms limitation talks \textelp{} With imagery, we could go to the numbers-based strategic arms limitations negotiation with a high degree of confidence.''\footcite[p.~403]{brugioni_eyes_2010} With satellite imagery in hand, the United States could enter the negotiations with appropriate estimates and leave with a means of verifying that the treaty had been carried out.
+Khrushchev claimed for years to be in favor of arms control, but talks always collapsed because of the US demand for on-site inspections, which the Soviet Union always refused.\footcite[p.~255]{mcdougall_heavens_1985} Satellites served the same purpose, and the Strategic Arms Limitation Treaty of 1972 (otherwise known as The ABM Treaty, or SALT 1) enshrined in law that signatories are not permitted to interfere with each others' ``national technical means of verification,'' a tacit ban on ASAT weapons.\footcite[p.~431]{mcdougall_heavens_1985} Without having to say ``spy satellites,'' SALT 1 made them not just legal, but essential.
+
+Though there is no chance that Khrushchev intended this, the intelligence that satellite photo-reconnaissance provided resulted in the United States building \emph{fewer} missiles and resisting the temptation to engage in an arms race.
+The success of the SALT treaties was a significant factor in the stabilization of relations between the two countries---it created an atmosphere of trust and cooperation that, while strained by Reagan the and the invasion of Afghanistan, held for the remainder of the Cold War.\footcite[p.~179]{lindgren_trust_2000} ``Freedom of space'' exists because both sides preferred a world where they could each verify each others' intent over one where they were shooting blind.
+
+\section{Conclusion}
 \end{document}
@@
 Powers' father had suggested to the State Department that they might be traded, spy for a spy.\footcite[p.~239]{powers_operation_2004} \enquote{This,} Eisenhower wrote, ``was a tacit admission by Khrushchev that our `outrageous' U-2 pilots have their opposite numbers operating within the borders of this country.''\footcite[p.~558]{eisenhower_waging_1965}
+
+% While I make the claim that both the spy plane and the spy satellite were purposefully connected to an intelligence tradition, it cannot be denied that the first time each one was deployed resulted in very different diplomatic outcomes.
+The first series of U-2 flights led to diplomatic protests, then a shootdown, then a dramatic summit meeting; the first series of Corona launches led to diplomatic protests, then...
+nothing.
+In absolute terms, Corona was dramatically more successful.
+Eisenhower gained the right to orbit a satellite over the Soviet Union whenever he wanted, and after the Gary Powers incident the U-2 was only used sparingly.
+
+% U-2 overflights could have caused a much more damaging reaction, but their framing helped prevent it from becoming an even larger incident than it already way.
+The norms defining how states should respond to espionage are ambiguous---the principle that states should not fly planes deep into each others' territory is not.
+When the U-2 was first developed, it was not at all clear which norm would supersede the other.
+As it happened, neither did---the interaction of existing norms was unique to both the craft itself and the way that it was marketed by the administration that invented it.
+The U-2 is not 70\% spy and 30\% plane.
+Instead, it is a complicated intersection of the two.
+The Soviet Union---and frankly, any nation over which the US chooses to fly a plane without permission---had the undisputed right to shoot down the American spy plane, which they did, but they also chose not to treat as an act of war, despite Eisenhower's explicit concern that they would.

### 2020-01-08 article.tex
Finish draft satellites chapter

@@
 
-\subsection{Freedom of Space}
+\subsection{Freedom of space}
 Spy satellites are not just one of the best edge cases for norm against shooting down a satellite, they are the reason it exists in the first place.
@@
 
-In 1954, President Eisenhower commissioned a task force which would tackle the problem of America's vulnerability to a surprise attack, the Technological Capabilities Panel (TCP).\footcite[p.~67.
-Also variously referred to as the \enquote{Killian Report,} \enquote{Surprise Attack Study,} or its formal title, \enquote{Meeting the Threat of a Surprise Attack.}]{killian_sputnik_1977} The TCP report concluded that any great power conflict would result in heavy American losses, so all necessary steps should be taken to avoid one.
-Its recommendations obviously included furthering American missile capabilities---traditional deterrence through force---but intelligence-gathering capabilities took comparable priority.
-``We \emph{must} find ways to increase the number of hard facts upon which our intelligence estimates are based,'' the TCP Report states, ``to provide better strategic warning, to minimize surprise in the kind of attack, and to reduce the danger of gross overestimation or gross underestimation of the threat.''\footcite{technological_capabilities_panel_meeting_1955} The TCP recommendations---reviewed personally by President Eisenhower---were hugely consequential across US government policy, but two projects that came out of it are especially relevant here: the U-2 plane and the artificial spy satellite.
+In 1954, President Eisenhower established the Technological Capabilities Panel (TCP), a task force which would tackle the problem of America's vulnerability to a surprise attack.
+When the panel returned with its results, its report concluded that any great power conflict would result in heavy American losses, so all necessary steps should be taken to avoid one.\footcite[p.~67.
+The TCP report is also variously referred to as the \enquote{Killian Report,} \enquote{Surprise Attack Study,} or its formal title, \enquote{Meeting the Threat of a Surprise Attack.}]{killian_sputnik_1977} Its recommendations obviously included furthering American missile capabilities---traditional deterrence through force---but intelligence-gathering capabilities took comparable priority.
+``We \emph{must} find ways to increase the number of hard facts upon which our intelligence estimates are based,'' the TCP Report states, ``to provide better strategic warning, to minimize surprise in the kind of attack, and to reduce the danger of gross overestimation or gross underestimation of the threat.''\footcite{technological_capabilities_panel_meeting_1955} The TCP recommendations---reviewed personally by President Eisenhower---were hugely consequential across US government policy, but two projects that came out of them are especially relevant here: the U-2 plane and the artificial spy satellite.
 
@@
 The first generation of artificial satellites promised to change that.
-In the time it takes a satellite to complete a full orbit, the Earth has rotated slightly underneath it, so with each pass a satellite launched into polar orbit photographs an area that has been longitudinally shifted.\footnote{Think of a winding a piece of string around a ball with your right hand, while slowly rotating the ball with your left; after a series of passes, you've covered a broad slice of the ball with string.} With enough orbits, a satellite can photograph a wide slice of Earth.
+In the time it takes a satellite to complete a full orbit, the Earth has rotated slightly underneath it, altering the satellite's field of view---with each pass an artifical satellite launched into polar orbit will photograph an area that has been longitudinally shifted.\footnote{Think of a winding a piece of string around a ball with your right hand, while slowly rotating the ball with your left; after a series of passes, you've covered a broad slice of the ball with string.} With enough orbits, that satellite can photograph a wide slice of Earth.
 Together, planes and satellites could operate for complementary photo-reconnaissance purposes: a plane dispatched when high quality localized intelligence was required, and a satellite to form a comprehensive picture of the Soviet military posture.
@@
 
-Preventing a surprise nuclear attack could not be done without better intelligence and better intelligence could not be achieved without comprehensive photo-reconnaissance---using a technique that the Soviet Union would accept.
+Preventing a surprise nuclear attack could not be done without better intelligence and better intelligence could not be achieved without comprehensive photo-reconnaissance---reconnaissance performed using a technique that the Soviet Union would accept.
 The solution was to spy from space.
-\enquote{Camera-toting satellites,} historian Walter McDougall explains, \enquote{circling the earth south to north [sic] in a polar orbit, could view the entire surface of the earth as it rotated below, return to any location in a few days' time, hone in on suspicious areas, and do it all under the legal cover of freedom of space---if such legal cover could be established.}\footcite[p.~117]{mcdougall_heavens_1985} The key phrase here is ``if such a legal cover could be established.'' The Soviets considered  U-2 overflights to be a violation of their territorial sovereignty, and continued to register diplomatic protests about them; the overflight program was so risky that it was originally intended to last only a year or two.\footcite[p.~33]{lindgren_trust_2000} The undefined legal status of space presented an opportunity, but also a challenge.
-The TCP understood that the Soviet Union would likely recognize the enormous advantage that a photo-reconnaissance satellite would confer on the US, and object to their deployment using any international norms possible.
+\enquote{Camera-toting satellites,} historian Walter McDougall explains, \enquote{circling the earth south to north [sic] in a polar orbit, could view the entire surface of the earth as it rotated below, return to any location in a few days' time, hone in on suspicious areas, and do it all under the legal cover of freedom of space---if such legal cover could be established.}\footcite[p.~117]{mcdougall_heavens_1985}
+
+That legal cover could be established was not a given.
+The Soviets considered U-2 overflights to be a violation of their territorial sovereignty, and continued to register diplomatic protests about them; the overflight program was so risky that it was originally intended to last only a year or two.\footcite[p.~33]{lindgren_trust_2000} The undefined legal status of space presented an opportunity, but also a challenge.
+The TCP understood that the Soviet Union would recognize the enormous advantage a photo-reconnaissance satellite conferred on the US, and object to its deployment using any international norms possible.
 It was therefore necessary to present reconnaissance satellites in a manner that would limit the grounds for objection.
-By the time that a reconnaissance satellite launched, it would have to have already been established that everyone had the unquestioned right to launch a satellite of any kind; this principle would come to be called \enquote{freedom of space.}
+By the time that a reconnaissance satellite launched, it would have to have already been established that any nation had the unquestioned right to launch a satellite of any kind; this principle would come to be called \enquote{freedom of space.}
 
-\subsection{Spiking the space race}
+\subsection{Space is not a race}
 Everything about the administration's space policy was carefully crafted to set the stage for the eventual launch of a spy satellite.
@@
 In fact, the overriding American priority to legitimize reconnaissance satellites is probably the \emph{reason} that the Soviets started the Space Age instead.
-Just as Eisenhower limited the involvement of the Air Force in Aquatone to paint the U-2 as a civilian project, he made technical decisions that were intended to make the scientific satellite seem as non-threatening as possible, and none was more important than the vehicle that would launch it.
-The group tasked with implementing NSC-5520 had to choose between competing launch proposals from the Army, Navy, and Air Force.
-It made the somewhat surprising choice to go with the unproven Navy Research Laboratory (NRL) \emph{Vanguard} rocket, even though the Army Ballistic Missile Agency (ABMA) engineers, headed by famous rocketeer Wernher von Braun, had been developing their proposal for a year.
+Just as Eisenhower limited the involvement of the Air Force in Aquatone to paint the U-2 as a civilian project, he made technical decisions that were intended to make the scientific satellite seem as non-threatening as possible---decisions that directly impacted the project timeline.
+The group tasked with implementing NSC-5520 had to choose between competing proposals from the Army, Navy, and Air Force for which rocket would launch the satellite.
+It made the somewhat surprising choice to go with the unproven Navy Research Laboratory (NRL) \emph{Vanguard} rocket, even though the Army Ballistic Missile Agency (ABMA) engineers, headed by famous rocketeer (and ex-Nazi) Wernher von Braun, had been developing their proposal for a year.
 Among the reasons the committee gave for their decision, two stand out.
@@
 
-Launching the first satellite with a ballistic missile threatened to compromise both of the principal national security priorities.
+Launching the first artificial satellite with a ballistic missile threatened to compromise both of the principal national security priorities.
 It might delay the development of the ICBM, an existential threat to American military superiority, and it would jeopardize the civilian character of the satellite program, which the committee members knew was necessary to establish ``freedom of space.''\footcite[p.~122]{mcdougall_heavens_1985} The Von Braun engineers vigorously protested the decision, because they believed---correctly---that their proposal would succeed first.
@@
 
-Like many of Eisenhower's decisions on space policy, his solution for reconnaissance satellites was a brilliant long-term success---but a complete mystery to the public and utterly baffling to most of the military establishment.
+Like many of Eisenhower's space policy decisions, his solution for reconnaissance satellites was a brilliant long-term success---but a complete mystery to the public and utterly baffling to most of the military establishment.
 WS-117L was originally intended to be a family of reconnaissance satellites.
@@
 Air Force personnel were ``thunderstruck.''\footcite[p.~45]{peebles_corona_1997} With public bewilderment instilled and American reconnaissance satellites now seemingly a few years out, the administration began a multi-pronged approach to legitimizing spy satellites.
-Publicly, the US government advocated for a ``freedom of space'' policy that explicitly included reconnaissance satellites; quietly, it resurrected the Corona project under deep-cover CIA administration.
+Publicly, the US government advocated for a ``freedom of space'' policy that explicitly included reconnaissance satellites; secretly, it resurrected the Corona project under deep-cover CIA administration.
 
@@
 Instead, each satellite was little more than a monstrously expensive disposable camera.
-The average Corona flight looked like this: the satellite would be sent into space carrying enough film for a 24-hour mission, and its camera would switch on every time the satellite flew over denied territory.
+The satellite would be sent into space carrying enough film for a 24-hour mission and its camera would switch on every time the satellite flew over denied territory.
 After completing seventeen orbits, the satellite would pitch down 60 degrees and eject the film capsule back down to Earth.
@@
 No subsequent missile buildup or bunker hardening or second-strike capability could undo the damage done to the Soviet military posture when \emph{Discoverer XIV} and its successors sailed past the Iron Curtain.
+Aerial reconnaissance neutralized one of Soviet Union's asymmetric structural advantages---the inherent secrecy of Soviet society---and put a dent in their claims to an arms advantage as well.
 And yet the only thing the Soviet Union did in response was publicly protest the spy satellites, and refuse to accept American proposals for a legal regime that included the rights of reconnaissance overflights.
@@
 
-% Start editing after here! So close!
 \subsection{Espionage norms in satellite reconnaissance}
-It seems obvious that the Soviet Union had almost everything to lose by legitimizing the use of spy satellites.
-Aerial reconnaissance neutralized one of Soviet Union's asymmetric structural advantages---the inherent secrecy of Soviet society---and put a dent in their claims to an arms advantage as well.
-What it gained in return is tough to say, but there is no possibility that it matched the intelligence windfall that befell the CIA.
-The legitimization of spy satellites dealt an indisputable blow to Soviet standing.
-So why did the Soviet Union let it happen?
-
-% Make these all one, maybe two paragraphs
-
-The brilliance of the American strategy goes a long way towards explaining why the USSR ultimately allowed satellites to fly over its territory.
-Both the overt and covert halves of the strategy played an important role.
-If the US had not been engaging in multinational talks about science and the freedom of space at the same time it was launching spy satellites, then it is very likely that upon developing ASAT capabilities, the USSR would have simply started shooting them down.
+The legitimization of spy satellites dealt an indisputable blow to Soviet standing, so why did the Soviet Union let it happen?
+The partial explanation is that the US played a good hand very well.
+If the US had not been engaging in multinational talks about \enquote{scientific} freedom of space while secretly launching spy satellites, then it is very likely that upon developing ASAT capabilities, the USSR would have simply started shooting those satellites down.
 That would have been an almost identical situation to the U-2 saga, and it was essential that when ASATs became available, the USSR not presume to have the unquestioned right to use them.
-
-But if the US had not been simultaneously running a covert program before the norm was established, then how would the norm have ever been established at all?
-At what point would the United States have determined that it was officially legal to launch a spy satellite?
-Soviet intransigence on COPUOS demonstrated that they were unwilling to formally acknowledge the use of spy satellites, so no legal solution was forthcoming.
-The Soviet Union was also unlikely to launch the first spy satellite; its HUMINT operations were able to provide most of the intelligence that a photo-reconnaissance operation would have, and launching the first spy satellite would have given the US the perfect excuse to close the intelligence gap once and for all.
+It was likewise crucial that the US built a successful reconnaissance satellite in secret.
+The Soviet Union was unlikely to launch the first spy satellite and give the US a perfect excuse to neutralize the intelligence gap, so the US had to do it on their own.
 The situation that resulted was a perfect storm for the normalization of satellite reconnaissance: the Soviet leadership found that they had the ability to develop a reliable spy satellite much sooner than they could develop a reliable anti-satellite weapon, of which their right to use would still be uncertain.
@@
 
-A few Khrushchev own-goals played a part in legitimizing satellite reconnaissance as well.
+% But if the US had not been simultaneously running a covert program before the norm was established, then how would the norm have ever been established at all?
+At what point would the United States have determined that it was officially legal to launch a spy satellite?
+Soviet intransigence on COPUOS demonstrated that they were unwilling to formally acknowledge the use of spy satellites, so no legal solution was forthcoming.
+
+Nikita Khrushchev played an unwitting role in legitimizing satellite reconnaissance as well.
 At the infamous four-powers summit in 1960, after watching Khrushchev vent about the U-2 Incident, French president Charles de Gaulle calmly told him: ``You're making too big a fuss about the matter.
@@
 In Eisenhower's later paraphrasing, Khrushchev said that ``any nation in the world who wanted to photograph the Soviet areas by satellite was completely free to do so.''\footcite[p.~556]{eisenhower_waging_1965} Even being the first nation to space played a role.
-Of course, it cannot be said that launching the first satellite in human history was a \emph{mistake} on the part of Khrushchev---Eisenhower underestimated the symbolic importance of igniting the Space Age, to his significant political detriment---but it is indisputable that \emph{Sputnik} made it much harder for the USSR to argue that reconnaissance satellites were an imperialist invasion of territorial sovereignty.
+Of course, it cannot be said that launching the first satellite in human history was a \emph{mistake}---Eisenhower underestimated the symbolic importance of igniting the Space Age, to his significant political detriment---but it is indisputable that \emph{Sputnik} made it much harder for the USSR to argue that reconnaissance satellites were an imperialist invasion of territorial sovereignty.
 
-\subsection{The rational (mis)calculation explanation}
 And yet, American strategy and Khrushchev's missteps still do not fully explain why Soviet Union continued to permit photo-reconnaissance after ASATs were developed.
@@
 Instead, the USSR choose to help write and ratify the Outer Space Treaty (OST) in 1967, the first legal regime codifying the peaceful uses of space.
-While the OST did not ban ASATs, it nonetheless set a precedent for the peaceful uses of space---and reflected the US policy stance that the main military value of space was as a harbor for reconnaissance satellites.\footcite[p~16]{mowthorpe_militarization_2004}
+While the OST did not ban ASATs, it nonetheless set a precedent for the peaceful uses of space---one that validated the US policy stance that the main military value of space was as a harbor for reconnaissance satellites.\footcite[p~16]{mowthorpe_militarization_2004}
 
@@
 
-The United States fully believed it had more to gain from a ``freedom of space'' norm---and I agree---but is it possible that the Soviet Union had some reason to believe the same?
+% The United States fully believed it had more to gain from a ``freedom of space'' norm---and I agree---but is it possible that the Soviet Union had some reason to believe the same?
 Perhaps, but given the indisputable advantage that the US gained from reciprocal reconnaissance, it is hard to imagine what non-reconnaissance military function a satellite could perform that would advantage the Soviet Union to the same degree.

### 2020-01-09 article.tex
Made dad's edits (everything except satellite reconnaissance)

@@
 US officials feared that the stolen data could compromise intelligence efforts by exposing undercover CIA officers working in embassies around the world, whose names would be suspiciously absent from OPM records.
-The information could also be used to identify American officials who might be susceptible to pressure, or worse, recruitment.
-The backlash was so intense that FBI Director James Comey was forced to take questions about the incident.
+The stolen information could also be used to identify American officials who might be susceptible to pressure, or worse, recruitment.
+The backlash was so intense that FBI Director James Comey was forced to publicly take questions about the incident.
 \enquote{If you have my [application for security clearance],} Comey admitted, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.
@@
 Instead, the opposite happened.
-US officials would only confirm that they believed a \enquote{foreign entity or government} had been behind the attack, even though it was widely known that they suspected the involvement of the Chinese government.\footcite{spetalnick_china_2015} Internal investigators quickly determined that OPM had been compromised by an Advanced Persistent Threat (APT), a formally organized and typically state-sponsored group of hackers that engage in long-term, targeted penetration operations.
-In this case, the APT made identification easy by leaving their calling-card: the trojan opmsecurity.org domain was registered in the name of \enquote{Steve Rogers,} the Marvel comics character better known as Captain America.
+US officials would only confirm that they believed a \enquote{foreign entity or government} had been behind the attack, even though it was widely known that they suspected the Chinese government.\footcite{spetalnick_china_2015} Internal investigators quickly determined that OPM had been compromised by an Advanced Persistent Threat (APT), a formally organized and typically state-sponsored group of hackers that engage in long-term, targeted penetration operations.
+In this case, the APT made identification easy by leaving its calling-card: the trojan opmsecurity.org domain was registered in the name of \enquote{Steve Rogers,} the Marvel comics character better known as Captain America.
 This digital taunt, along with email and IP addresses that verified the country of origin was not a false flag, all but guaranteed that the attack had come from an APT sponsored by the Chinese government, likely the Chinese military's cyber-espionage division.\footcite{koerner_inside_2016} Though it possessed conclusive digital forensics, the Obama administration nonetheless refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018.
@@
 A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].}\footnote{In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets.
-See \cite{finklea_cyber_2015}} For activities that fall under the umbrella of intelligence-gathering, counterintelligence, not prosecution, is deemed to be the appropriate response.
+See \cite{finklea_cyber_2015}} For activities that fall under the umbrella of intelligence-gathering, counterintelligence---not prosecution---is deemed to be the appropriate response.
 Director of the CIA John Brennan described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, was particularly frank: \enquote{You have to kind of salute the Chinese for what they did.
@@
 In fact, cyber-espionage is routinely treated as a lesser category of offense by all nations, not just the United States.
-Of the hundreds of state-sponsored cyberattacks known to the public, attacks classified as espionage are by the far most common---and the least likely to garner an official response from the victim government (Table \ref{response-pct}).\footnote{These numbers were compiled from the Council on Foreign Relations' Cyber Operations Tracker, which is a publicly-sourced record of state-sponsored cyberattacks, updated quarterly.
+Of the hundreds of state-sponsored cyberattacks known to the public, attacks classified as espionage are by far the most common---and the least likely to garner an official response from the victim government (Table \ref{response-pct}).\footnote{These numbers were compiled from the Council on Foreign Relations' Cyber Operations Tracker, which is a publicly-sourced record of state-sponsored cyberattacks, updated quarterly.
 Of the 323 listed incidents, 38 were excluded from this calculation because they lacked an identifiable Victim, Sponsor, or Type.
@@
 At first glance this observation might seem almost tautological.
-Of course cyber-espionage is treated like traditional espionage, one could argue---they're both espionage.
-That formulation, however, begs the question; it doesn't explain how states determine what constitutes \enquote{traditional} espionage in cyberspace, what purpose that distinction servers, or whether it is prudent to continue making it.
+Of course cyber-espionage is treated like traditional espionage, one would argue---they're both espionage.
+That formulation, however, begs the question; it doesn't explain how states determine what constitutes \enquote{traditional} espionage in cyberspace, what purpose that distinction serves, or whether it is prudent to continue making it.
 Certainly the US should respond proportionally to attacks on its systems, but what if the damage done to national security by espionage is especially significant?
@@
 \enquote{Clearly, it does.}\footcite{sanger_u.s._2015} As as far as the public knows, however, the size of operation did \emph{not} change the nature of the diplomatic response.
-Therefore, to understand why the Director of the CIA treats cyber-espionage as \enquote{fair game,} one first needs to understand why traditional espionage has historically received that treatment as well.
+Therefore, to understand why the Director of the CIA treats cyber-espionage as \enquote{fair game,} one first needs to understand why traditional espionage has historically received the same treatment.
 
@@
 In the case of the OPM hack, for example, US policymakers decided they could not impose harsh penalties on the Chinese because it was more valuable for the US to retain its ability to practice cyber-espionage with relative impunity in the future.
-This basic calculation---that espionage against one's own country is a tolerable price to pay for being allowed to practice espionage---has remained consistent since the advent of the modern peacetime intelligence service, perhaps even the nation-state itself.
+This basic calculation---that espionage against one's own country is a tolerable price to pay for being allowed to practice espionage against one's rivals---has remained consistent since the advent of the modern peacetime intelligence service, perhaps even the nation-state itself.
 
@@
 Otherwise, the state with less to gain would simply defect from that equilibrium and punish espionage at a level that imposes direct costs.
-Since this generally doesn't happen between major world powers, one of two things must be true: either a state is miscalculating the relative gains it gets from espionage, or the mutual tolerance of espionage has benefits that are sufficiently important to maintain even if one state gains less from it than its adversary.
+Since this generally doesn't happen between major world powers, one of two things must be true: either a state is miscalculating the relative gains it achieves through espionage, or the mutual tolerance of espionage has benefits that are sufficiently important to maintain even if one state gains less from it than its adversary.
 I argue the latter.
@@
 The benefits of espionage have never been underrated---\enquote{it is only the enlightened ruler and the wise general,} reads Sun Tzu, \enquote{who will use the highest intelligence of the army for purposes of spying and thereby they achieve great results}---but the inevitability of espionage has long been taken for granted.
-If you question that inevitably, the assertion that espionage is something \enquote{all counties do,} then the way that countries treat espionage becomes considerably more confusing.
+If one questions the assertion that espionage is something \enquote{all countries do,} then the way that countries treat espionage becomes considerably more confusing.
 Until now, few, if any, scholars have analyzed why espionage norms exist and what strategic purpose they serve.
@@
 Simply put, espionage norms dictate what constitutes espionage and how states are expected to respond when it is conducted against them.\footnote{There are, of course, other \enquote{norms} that involve espionage.
-The so-called \enquote{Moscow Rules} dictate how an agent or case officer is supposed to conduct themselves in hostile territory.
+The so-called \enquote{Moscow Rules} dictate an intelligence officer are supposed to conduct themselves in hostile territory.
 In order to avoid constantly specifying, espionage norms are more strictly defined to include only the ones that affect IR.} What consequences does a state impose when it uncovers an intelligence operation run by another state?
@@
 Rarely does the response amount to more than a pat denunciation of the activity and implementation of the necessary countermeasures to end it---the consequences for espionage are borne by the agents who commit it, not the state that sent them.
-In the rare cases when additional costs \emph{are} imposed on a state for conducting a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.
+In rare cases when additional costs \emph{are} imposed on a state for conducting a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.
 In this article I present a theory for why those norms exist.
@@
 John Radsan employs a more philosophical approach.
-Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage all reach the same conclusion: it would be almost impossible to argue that espionage is legal, but no one really seems to care.
+Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage generally reach the same conclusion: it would be almost impossible to argue that espionage is legal, but no one really seems to care.
 Taken to its logical endpoint, this branch of scholarship would simply argue that the espionage norms we observe today result from simple inertia.
@@
 Responding to Baker, Radsan notes that one-off deals and treaties \enquote{are a very slow and indirect way to achieve international consensus on the legality of espionage.
-The gap is still there.}\footcite[p.~607]{radsan_unresolved_2007} This article expands on Baker's basic premise---that there are situations in which espionage is mutually advantageous for both parties---and universalizes it to explain why the entire institution of espionage is mutually advantageous.
+The gap is still there.}\footcite[p.~607]{radsan_unresolved_2007}
+
+% This article expands on Baker's basic premise---that there are situations in which espionage is mutually advantageous for both parties---and universalizes it to explain why the entire institution of espionage is mutually advantageous.
 
 The key insight this article presents---and where it breaks from previous literature---is that far from being an irreplicable, historically contingent set of practices, espionage norms are actually preserved through an ongoing set of strategic decisions.
-Each time a new espionage technology is made available, policymakers are given a fresh chance to decide whether or not the norms of espionage ought to extend to its use.
+Each time a new espionage technology is made available, policymakers are given a fresh opportunity to decide whether or not the norms of espionage ought to extend to its use.
 Critically, the research that follows will show that they were not obligated to do so by any pre-existing norms.
@@
 \subsection{Establishing the puzzle}
-One of the most prevalent clich\'es in spy literature and journalism is the so-called \enquote{rules of the game.} The supposed rules of the spy world world---\enquote{You are never alone}, \enquote{Don't trust anyone}---make the spy world seem insular, mysterious, and impenetrable.\footcite{myre_moscow_2019} They hint at a larger set of processes whose motivations the average citizen, not given access to this privileged information, could not possibly begin to understand.
+One of the most prevalent clich\'es in spy literature and journalism is the so-called \enquote{rules of the game.} The supposed rules of the spy world---\enquote{You are never alone}, \enquote{Don't trust anyone}---make the intelligence community seem insular, mysterious, and impenetrable.\footcite{myre_moscow_2019} They hint at a larger set of processes whose motivations the average citizen, not given access to this privileged information, could not possibly begin to understand.
 In reality, while the \emph{operation} of a modern intelligence service is incredibly opaque, the geopolitical situation in which one operates is not.
@@
 The results of intelligence operations can have devastating, even existential effects on the targeted nation's security.
-Arguably the most traditional form of espionage is stealing military secrets from a foreign power, and in that arena China has been relentless.
+Arguably the most traditional form of espionage is stealing military secrets from a foreign power, and in that arena China has been a relentless combatant.
 A United States Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} \enquote{Long-term, US future military advantage is being diminished by years of IP exfiltration from the DoD, DoN, and DIB,} the report read, \enquote{all with little to no adverse consequences to the thieves.}\footcite[p.~6]{bayer_cybersecurity_2019} The forecasts were dire: \enquote{If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.}\footcite[p.~5]{bayer_cybersecurity_2019} The US Navy is not a disinterested party, of course, and there are reasons to take its conclusions with a grain of salt; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see \cite{gilli_why_2019}} Nonetheless, espionage presents alarming opportunities for rival states and the consequences for taking advantage of them remain minimal.
 
-Espionage has potentially disastrous consequences---and steps that might actually curb it are out of the question.
+Espionage has potentially disastrous consequences but the steps that might actually curb it are out of the question.
 The system of repercussions is set up so that the structure of civilian intelligence services is never fundamentally threatened.
@@
 Diplomatic expulsions play out in exactly the same fashion: when one country expels the diplomats of another, the other country will often respond in kind.
-Each side will continue to expel each other's diplomats until one side signals that it would like for the reciprocal PNG-ing of diplomats to stop.
-Then, both sides return to a cooperative state, each preserving some of their intelligence operations.
+Each side will continue to expel each other's diplomats until one side signals that it would like for the reciprocal removal of diplomats to stop.
+Inevitably, both sides return to a cooperative state while they each still have \emph{some} diplomats left.
 Because neither side can gain an a decisive advantage, the back-and-forth punishment never ends in a mutual defection.
@@
 Losing intelligence officers in an embassy temporarily makes it more difficult to conduct espionage, but it doesn't make a state less likely to try; the worst that can happen is a net-neutral result---going back to \enquote{square one,} so to speak.
-By contrast, were it more common to respond to espionage by imposing financial sanctions or even inflicting a military cost, a state would have to consider the possibility that attempting espionage could adversely impact other aspects of their security, leaving it worse off than when it started.
-These are the sorts of calculations that states need to make in other high-stakes arenas---pursuing a nuclear weapons program, for instance, could lead to severe economic retaliation, as could protectionist trade policies---but engaging in espionage has very little potential for negative externalities because espionage norms take the possibility of a severe response off the table.
-In this section, I present what could happen if they didn't.
+By contrast, were it more common to respond to espionage by imposing financial sanctions or even inflicting a military cost, a state would have to consider the possibility that attempting espionage could adversely impact other aspects of its security, leaving it worse off than when it started.
+These are the sorts of calculations that states need to make in other high-stakes arenas.
+Pursuing a nuclear weapons program, for instance, could lead to severe economic retaliation, as could protectionist trade policies.
+By comparison, engaging in espionage has very little potential for negative externalities because espionage norms take the possibility of a severe response off the table.
 
@@
 Through this model two important things become clear.
-First, even though either state has the opportunity to defect by imposing harsher punishments in response to espionage, its gains would be short-lived, because the other state is guaranteed to immediately do the same---tit-for-tat.
-Espionage consequences are an iterated game that both parties are playing the exact same way, so, in the long run, the only two possible outcomes are mutual defection or mutual cooperation.
-In the realm of traditional espionage, we've only ever seen mutual cooperation, but the critical observation is that any state has the power to change that by defecting.
+First, even though either state has the opportunity to defect by imposing harsher punishments in response to espionage, its gains would be short-lived because the other state is guaranteed to immediately do the same---tit-for-tat.
+Espionage consequences are an iterated game in which both parties play the exact same way, so, in the long run, the only two possible outcomes are mutual defection or mutual cooperation.
+In the realm of traditional espionage, we've only seen states engage in mutual cooperation, but any state has the power to change that by defecting.
 Even if the defector's rival would have preferred to maintain the status quo, the rival has no option other than to defect as well or else be left at a senseless disadvantage.
@@
 Stripped of its mystique, the benefits of allowing intelligence operations to continue are, like any other state decision, a simple cost-benefit analysis; states decide that it best serves their interests to remain in the top left corner, rather than the bottom right.
-We can expect that there is \emph{some} rational explanation that political leaders employ to justify it, where they weigh the benefits of preserving their ability to spy against the leaks that inevitably result from espionage conducted against them.
+We can expect then that there is \emph{some} rational explanation that political leaders employ to justify it, where they weigh the benefits of preserving their ability to spy against the damage that results from espionage conducted against them.
 
-One possible explanation is that some states are just miscalculating.
-Though I have not found any scholarly works making this argument, the basic premise is persuasive.
+One possible explanation is that some states are simply miscalculating.
+Theoretically, the basic premise is persuasive.
 Every policymaker knows exactly how crucial espionage is to their job.
@@
 It follows that there would be a strong psychological bias towards attempting to discover as much information as possible.
-While this is likely true to some extent, \enquote{everyone else is getting it wrong} is not an explanation that typically suffices in political science, and it will not suffice here.
+While this is likely true to some extent, mass-miscalculation could not fully explain the widespread persistence of espionage.
 
-The surprisingly straightforward conclusion is that, if only two worlds are possible---one with widespread espionage and one without---then all the relevant nations must be concluding that the world without espionage is incredibly bad.
-So bad that even if one nation correctly concludes that its rival gains more from a world with a free flow of information than it does, the alternative, a world with significantly harsher punishments for espionage and a staunched flow of intelligence, still leaves that nation worse off.
+The surprisingly straightforward conclusion is that, if only two worlds are possible---one with widespread espionage and one without---then all the relevant nations must be concluding that the world without espionage is incredibly bad---so bad that even if one nation correctly concludes that its rival gains more from a world with a free flow of information than it does, the alternative, a world with significantly harsher punishments for espionage and a staunched flow of intelligence, still leaves that nation worse off.
 That calculation has a name: defensive realism.
@@
 13-14)} Glaser begins from the core assumption that the reason for competition between security-seeking states is insecurity.
-Consider a theoretical security seeker interacting with its adversary, which it believes to be a security-seeking state as well.
+Consider a hypothetical security seeker interacting with its adversary, which it believes to be a security-seeking state as well.
 The first state feels insecure, so it builds up arms to increase its security.
-When the rival state, seeing that its adversary is building up arms, now feels less secure and builds up its own arsenal in response, and a classic security dilemma ensues.
+The rival state, seeing that its adversary is building up arms, now feels less secure and builds up its own arsenal in response, and a classic security dilemma ensues.
 Even though both states were just security seekers, they're now both much less secure.
@@
 
-Prior to making any decisions, states asses the security situation and determine which of their options are most likely to increase their own security.
+Prior to making any decisions, states assess the security situation and determine which options are most likely to increase their own security.
 Glaser identifies two types of variables that mediate the magnitude of a security dilemma.
 The first are material variables, such as offense-defense balance and offense-defense distinguishability.
-Espionage plays a significant role in properly assessing material variables; during the Cold War, the United States placed a premium on generating quality estimates of how many bombers, missiles, and bases the Soviet Union had deployed.
+Espionage plays a significant role in properly assessing material variables; during the Cold War, for example, the United States placed a premium on generating quality estimates of how many bombers, missiles, and bases the Soviet Union had deployed.
 By knowing which of its adversary's capabilities are offensive, a state can better target its efforts towards reducing them.
@@
 In addition to material variables, Glaser notes that information variables play an important and comparatively under-studied role in the security dilemma.
-Better information about its adversary's motives can give the state reason to trust that their risky signaling measures might succeed---or fail.
+Better information about its adversary's motives can give the state reason to trust that risky signaling measures might succeed---or fail.
 Even if the state is uncertain, espionage can provide some information about its adversary's intent, which, in turn, can moderate the security dilemma enough to make cooperation possible.
@@
 As a result, the state decides that some form of arms control is no longer too risky to attempt in order to signal its benign intentions.
-Because the state receiving the signal is a security-seeker, not only will its own intelligence make it more likely to respond in kind, but it also benefited from the first state's intelligence operations, because that led to the initial exchange of signals.
-The two-way flow of information has a measurable impact on each state's decisionmaking, facilitating the optimal outcome for both parties.
+Because the state receiving the signal is also a security-seeker, this is a welcome development; it could be empowered to signal its benign intentions as well, based on its own intelligence.
+For both states, this is the optimal outcome.
+The counterintuitive observation here is that both states benefited not just from spying, but from being spied \emph{upon}; their adversary's intelligence made them more likely to make a risky signal for deescalation.
 
@@
 
-The best way to prove that this theory has explanatory power for state decision-making is to find examples from the past where states made decisions that showed they were willing to accept the cost of enemy espionage in order to practice espionage themselves.
-In the ideal scenario, we would find evidence of policymakers justifying their decisions along these lines.
+The best way to prove that this theory has explanatory power for state decision-making is to find historical examples where states made decisions that showed they were willing to accept the cost of enemy espionage in order to practice espionage themselves.
+In the ideal scenario, we would find evidence of policymakers justifying their decisions based on the value of mutual observation.
 Such decisions were made frequently during the Cold War, and with the explicit intent to increase the bilateral flow of information to create a more secure world environment.
@@
 
-Both planes and satellites played an underexplored role in the formulation of some of the Cold War's landmark diplomatic achievements.
+Both planes and satellites played an under-explored role in the formulation of some of the Cold War's landmark diplomatic achievements.
 SALT I (the ABM Treaty) is widely considered to be the landmark diplomatic achievement of Nixon's \emph{d\'etente} policy, which set the stage for a broader easing of relations between the US and the USSR, and spy satellites were crucial to reducing the cost that each side would pay to sign it.
-Limiting ballistic missile defense systems was a relatively minor ask for both sides, but in the highly risk-averse context of the Cold War, the signal was significant.\footcite[p.~66]{glaser_rational_2010} Underneath the omnipresent threat of nuclear annihilation, \emph{d\'etente} provided a rational strategic choice for both states---made possible by their reciprocal tolerance of espionage.
+Limiting ballistic missile defense systems was a relatively minor concession for both sides, but in the highly risk-averse context of the Cold War, the signal was significant.\footcite[p.~66]{glaser_rational_2010} Underneath the omnipresent threat of nuclear annihilation, \emph{d\'etente} provided a rational strategic choice for both states---made possible by their reciprocal tolerance of espionage.
 
@@
 The line between a \enquote{cold} and a \enquote{hot} war is supposed to be obvious.
-A superpower's own uniformed military personnel opening fire on the other superpower's military or territory would be \enquote{hot,} everything else is \enquote{cold.} Broadly speaking, history remembers the Cold War as lacking direct military confrontation between the United States and Russia; though the two superpowers of the Cold War were open rivals, they supposedly competed for influence by funding, arming, or otherwise influencing proxies, rather than attacking each other directly.
+A superpower's own uniformed military personnel opening fire on the other superpower's military or territory would be \enquote{hot,} everything else would be \enquote{cold.} Broadly speaking, the Cold War is remembered as lacking direct military confrontation between the United States and Russia; though the two superpowers of the Cold War were open rivals, they supposedly competed for influence by funding, arming, or otherwise influencing proxies, rather than attacking each other directly.
 That impression is not exactly accurate.
@@
 The Soviets were deeply unhappy that the US was violating their airspace in such a brazen fashion, but for years they had kept quiet about the U-2 missions, embarrassed that they were technologically incapable of intercepting the high-flying aircraft.\footcite[p.~59]{powers_operation_2004} When Soviet air defenses finally managed to shoot one down in 1960, Soviet leadership was equal parts triumphant and furious.
-At a planned summit in Paris only a few weeks later, Khrushchev launched into a rant about the U-2 planes and stormed out on the first day, effectively ending the summit---and any hope that pressing issues like the fate of West Berlin would be resolved there.
-He also canceled a planned visit by Eisenhower to the USSR that was scheduled to take place the next month.
+At a planned summit in Paris only a few weeks later, Soviet Premier Khrushchev launched into a rant about the U-2 planes and stormed out on the first day, effectively ending the summit---and any hope that pressing issues like the fate of West Berlin would be resolved there.
+He also canceled a planned visit by US President Dwight Eisenhower to the USSR that was scheduled to take place the next month.
 A contemporary newsreel about the conference said that \enquote{in the course of two hours, Khrushchev brought US-Soviet relations to their lowest point since the end of World War II.}\footcite{universal_studios_summit_1960}
 
-The shootdown of Gary Powers, what would later come to be known as \enquote{The U-2 Incident}, clearly violated the only \enquote{rule} of the Cold War.
+The shootdown of Gary Powers, in what would later come to be known as \enquote{The U-2 Incident}, clearly violated the only \enquote{rule} of the Cold War.
 Official Soviet air defenses shot down a US aircraft.
-All parties involved had cause to escalate military---the Soviets for the violation of their airspace, the US for the attack on their \enquote{weather craft}---but instead they chose to treat the U-2 missions as an intelligence operation, not military action.
+All parties involved had cause to escalate militarily---the Soviets for the violation of their airspace, the US for the attack on their \enquote{weather craft}---but instead they chose to treat the U-2 missions as an intelligence operation, not a military action.
 The Soviets therefore applied the traditional consequences associated with espionage; Khrushchev engaged in some saber-rattling and denounced the violation of his airspace, but no material actions were taken, and the United States escaped the incident with few tangible consequences.
@@
 No single attribute defines a spy plane.
-A military aircraft can forgo weapons but, by virtue of its pilot, its mission, its timing, or its branding still remain in its operation unquestionably military.
-Inversely, intelligence missions can be flown by uniformed Air Force pilots in re-purposed B-52s and somehow still be read as espionage.
+A military aircraft can forgo weapons but, by virtue of its pilot, its mission, its timing, or its branding still conduct what is unquestionably a military operation.
+Inversely, intelligence missions can be flown by uniformed Air Force pilots in re-purposed B-52s and somehow still be understood as conducting espionage.
 A spy plane is an ontological designation, one that reveals itself to the adversary through a series of social and physical signifiers that are not replicable across time.
-Germany in 1937 would not have reacted to a U-2 overflight the same way that the Soviet Union did in 1960.
-But even if the way that different states handled espionage was perfectly consistent, there still wouldn't be a checklist that guarantees a mission will treated as espionage; it is not enough make a military mission look like an intelligence operation.
+Germany in 1937 might not have reacted to a U-2 overflight the same way that the Soviet Union did in 1960.
+But even if the way different states handled espionage was perfectly consistent, there still wouldn't be a checklist to guarantee that a mission would be treated as espionage; it is not enough make a military mission look like an intelligence operation.
 The reality is both simpler and more frustrating---it has \emph{to be one}.
 
-The U-2 incident was the most visible in a series of aerial intelligence operations that the United States undertook in the early Cold War.
-The US did not always employ civilian pilots nor it did always use obviously innocuous aircraft, yet it managed to run a shocking number of missions that violated Soviet airspace without significantly altering the relative level of tension between the two superpowers.
-On occasion, these missions would turn \enquote{hot} when Soviet air defenses shot them down.
-Not once did these shootdowns lead to retaliation from the US, or significant pressure from the Soviets to stop the flights.
+The U-2 incident was the most visible in a series of aerial intelligence operations that the United States undertook early in the Cold War.
+The US did not always employ civilian pilots nor it did always use obviously weaponless aircraft, yet it managed to run a shocking number of missions that violated Soviet airspace without significantly altering the relative level of tension between the two superpowers.
+On a surprisingly frequent basis, these missions would turn \enquote{hot} when Soviet air defenses shot them down.
+Not once, however, did these shootdowns lead to retaliation from the US, or significant pressure from the Soviets to stop the flights.
 Instead, each one was resolved as if the pilots who survived were simply spies, captured in the normal course of counterintelligence.
@@
 Converted bombers---called \enquote{ferrets} by the Air Force---were outfitted with advanced radio equipment and sent to intercept as much information about Soviet radio transmissions as possible.\footcite[p.~4]{peterson_maybe_1993} According to the CIA, these missions began as early as 1947.\footcite[p.~4]{peterson_maybe_1993} Some of these flights only flew near the border, remaining over traditionally recognized international waters, while others were \enquote{overflights,} missions that deliberately violated Soviet airspace.
-The earliest postwar overflight of the Soviet Union---of which I can find definitive record---took place on May 10, 1949, when two RF-80As surveilled the Kurile islands\footcite[p.~8]{peebles_shadow_2000}
+The earliest definitive postwar overflight of the Soviet Union took place on May 10, 1949, when two RF-80As surveilled the Kurile islands\footcite[p.~8]{peebles_shadow_2000}
 
@@
 Even if one did have unlimited access, some of the necessary information was so highly classified that it is likely lost forever.
-Nonetheless, the civilian researcher is aided by the declassified research of military archivists before them.
-In this paper, I highlight the work of two scholars in particular.
+Nonetheless, the civilian researcher is aided by the declassified research of military archivists before them and the works of two scholars in particular are highlighted here.
 First, Air Force Academy historian Dr.
@@
 Many of the incidents analyzed in this section are taken from newspaper clippings that he compiled and analyzed.
+Second, former intelligence analyst and section chief Michael Peterson wrote an article for the CIA history magazine \emph{Cryptologic Quarterly} that analyzed the Soviet shootdowns of US reconnaissance aircraft in depth.
 
-Second, former intelligence analyst and section chief Michael Peterson wrote an article for the CIA history magazine \emph{Cryptologic Quarterly} that analyzed the Soviet shootdowns of US reconnaissance aircraft in depth.
-The article was written for a classified audience in 1993 and declassified in 2009, providing the public with a comprehensive (as far as we know) list of all American SIGINT flights shot down by the USSR (Table \ref{soviet-shootdowns}).
+Peterson's article was written for a classified audience in 1993 and declassified in 2009, providing the public with a comprehensive (as far as we know) list of all American SIGINT flights shot down by the USSR (Table \ref{soviet-shootdowns}).
 Peterson uses two limiting criteria that perfectly suit the purposes of this article: the only flights he lists are those that were (a) shot down by Soviet attacks, not Chinese, North Korean, etc., and (b) explicitly US reconnaissance aircraft.\footcite[p.~4]{peterson_maybe_1993} Note that not all of these flights are necessarily overflights; many took place over international waters when they were shot down.
-In effect, the table catalogs the 13 times in which the USSR shot down an American aircraft designed for collecting intelligence between 1950 and 1964---intercepted espionage attempts.\footcite[p.~5.
+In effect, the table catalogs the 13 times where the USSR shot down an American aircraft designed for collecting intelligence between 1950 and 1964---intercepted espionage attempts.\footcite[p.~5.
 In the original document, this table lists the first incident as having taken place over the Barents Sea, not the Baltic Sea.
@@
 
-Regardless of how thorough declassified CIA information appears, I cannot rule out the possibility that there were other, even more secret missions which for whatever reason the United States government has chosen to continue protecting to this day.
+Regardless of how thorough declassified CIA information appears, one cannot rule out the possibility that there were other, even more secret missions which, for whatever reason, the United States government has chosen to conceal to this day.
 Nonetheless, a few aspects of the \emph{Cryptologic Quarterly} article are promising, and suggest that the information is relatively complete.
@@
 Additionally, when the article was declassified 16 years later, it appears to have been declassified without redactions or missing pages.
-And as with any analysis of diplomatic repercussions and espionage, we are aided by a simple logical trap---if there had been another mission so highly classified that it was omitted from this document, it is difficult to imagine how diplomatic consequences could have been imposed with a degree of secrecy that would prevent us from knowing about it today.
+Lastly, if there had been another mission so highly classified that it was omitted from this document, it is difficult to imagine how diplomatic consequences could have been imposed with a degree of secrecy that would prevent us from knowing about it today.
 
@@
 The Soviet ambassador to the US lodged a formal note of protest, incorrectly alleging that the US aircraft which had been shot down was a B-29 bomber.\footcite{kirk_ambassador_1950} A few days later, the American ambassador issued a formal response, in which he correctly identified the model of plane and claimed that at no point did it cross into any territorial waters.
-He demanded from the Soviets \enquote{a prompt and thorough investigation} and that the Soviet Government \enquote{see to it that those responsible for this action are promptly and severely punished.}\footcite{the_associated_press_text_1950} The Soviet Union recognized this for the toothless statement that it was, and rejected the totality of its demands.\footcite{salisbury_kremlin_1950} No further action was taken.
+He demanded the Soviets conduct \enquote{a prompt and thorough investigation} and \enquote{see to it that those responsible for this action [were] promptly and severely punished.}\footcite{the_associated_press_text_1950} The Soviet Union recognized this for the toothless statement that it was, and rejected the demand in its entirety.\footcite{salisbury_kremlin_1950} No further action was taken.
 
 Both sides had reasons to make this a larger political issue.
-The Soviet Union had fired first on a US Navy aircraft---an aircraft which could not have been the aggressor because its only armament was a .45-cal pistol carried by one of the crewmen.\footcite[p.~7]{peterson_maybe_1993} Given that the plane was never found, all available evidence suggests that the flight likely did not violate the traditional 12-mile airspace boundary.
-Based on the last received position report, today the CIA claims that the plane was flying 20-25 miles off the coast of Latvia at the time it was shot down.\footcite[p.~7]{peterson_maybe_1993} At the time, it was theorized that a theoretical breakdown in the plane's navigation system could have caused it to veer off-course into Latvia, but given its last known position, it would have taken \enquote{a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.}\footcite{the_new_york_times_soviet_1950} Meanwhile, the United States knew that the missing plane probably hadn't ended up in Soviet airspace, but the Privateer in question \emph{had} been performing a questionably legal reconnaissance mission when it was shot down.
+The Soviet Union had fired first on a US Navy aircraft---which could not have been the aggressor because its only armament was a .45-cal pistol carried by one of the crewmen.\footcite[p.~7]{peterson_maybe_1993} Given that the plane was never found, all available evidence suggests that the flight likely did not violate the traditional 12-mile airspace boundary.
+Based on the last received position report, today the CIA claims that the plane was flying 20-25 miles off the coast of Latvia at the time it was shot down.\footcite[p.~7]{peterson_maybe_1993} At the time, it was theorized that a theoretical breakdown in the plane's navigation system could have caused it to veer off-course into Latvia, but given its last known position, it would have taken \enquote{a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.}\footcite{the_new_york_times_soviet_1950} Meanwhile, the United States knew that the missing plane probably hadn't ended up in Soviet airspace, but the Privateer in question \emph{had} been performing a reconnaissance mission of questionable legality when it was shot down.
 No traditional boundaries were violated, but a repurposed Navy privateer loaded with electronic reconnaissance equipment flying close to Latvia looked, to use a less technical term, shady.
@@
 The Soviet Union responded by protesting the alleged American shootdown of a Soviet passenger plane.
-Despite press outrage, no further action was taken.\footcite[p.~47]{farquhar_aerial_2015} A September 4, 1954 shootdown of a Navy P2V Neptune actually led to a call by one US Senator to break off diplomatic ties with the Soviet Union: \enquote{Just another note from our State Department to the Kremlin hierarchy will not impress these uncivilized rulers \textelp{} that this new attack upon an American plane confirms Communist arrogance and aggressiveness to a point where the breaking of diplomatic relations is justified.}\footcite{the_associated_press_ending_1954} Instead of a harsher punishment, Eisenhower brought the issue to the UN with the full knowledge that an unfavorable judgment would be vetoed by the Soviets.
-The move demonstrated Eisenhower's ability and commitment to resisting the calls for harsher action while still placating the American public.\footcite[p.~47]{farquhar_aerial_2015} Besides, by late 1954, the \enquote{cycle of hostility} over aerial reconnaissance was already breaking.\footcite[p.~49]{farquhar_aerial_2015} In terms of media coverage, and diplomatic significance, no shootdown that followed would require more significant consequences than the ones that had already happened.
+Despite press outrage, no further action was taken.\footcite[p.~47]{farquhar_aerial_2015} A September 4, 1954 shootdown of a Navy P2V Neptune actually led to a call by one US Senator to break off diplomatic ties with the Soviet Union: \enquote{Just another note from our State Department to the Kremlin hierarchy will not impress these uncivilized rulers \textelp{} that this new attack upon an American plane confirms Communist arrogance and aggressiveness to a point where the breaking of diplomatic relations is justified.}\footcite{the_associated_press_ending_1954} Instead of a harsher punishment, President Eisenhower brought the issue to the UN with the full knowledge that an unfavorable judgment would be vetoed by the Soviets.
+The move demonstrated Eisenhower's ability and commitment to resisting the calls for harsher action while still placating the American public.\footcite[p.~47]{farquhar_aerial_2015} Besides, by late 1954, the \enquote{cycle of hostility} over aerial reconnaissance was already breaking down.\footcite[p.~49]{farquhar_aerial_2015} In terms of both media coverage and diplomatic significance, no shootdown that followed would require more significant consequences than the ones that had already happened.
 
@@
 In 2001, the Air Force held a symposium honoring and recollecting the pilots who flew the overflights on Soviet territory in the early Cold War, an effort in which they were joined by the British in 1954.\footcite[p.~v]{hall_early_2003} Many of the presenters spoke about entire reconnaissance projects that apparently didn't even register diplomatically.
-Project Heart Throb, which outfitted RB-57s with surveillance equipment, flew between 15 and 19 missions over Eastern Europe from 1955 to 1956, in the recollection of Gerald E.
-Cooke, a Air Force pilot assigned to the project.\footcite[p.~194]{hall_early_2003} Another project from around the same time, Slick Chick, set up F-100 jets with rapid-fire 20mm cameras.
+Project Heart Throb, for example, outfitted RB-57s with surveillance equipment and flew between 15 and 19 missions over Eastern Europe from 1955 to 1956.\footcite[p.~194.
+Based on the recollection of Gerald E.
+Cooke, a Air Force pilot assigned to the project.]{hall_early_2003} Another project from around the same time, Slick Chick, set up F-100 jets with rapid-fire 20mm cameras.
 The pilot who presented this project at the symposium, Cecil Rigby, personally flew two such missions and estimated that there were six Slick Chick missions total.\footcite[p.~176]{hall_early_2003}
@@
 The original source for this quote is an \emph{OSA History} that requires codeword clearance.
-It is quoted here by the History Staff of the CIA.]{pedlow_central_1992} This allowed the United States to \enquote{truthfully deny,} in the phrasing of the CIA, that any US \enquote{military planes} had flown over the USSR---which they had to do when the inevitable Soviet protest notes were filed after U-2 overflights began in 1956.\footcite[p.~109]{pedlow_central_1992}
+It is quoted here by the History Staff of the CIA.]{pedlow_central_1992} This allowed the United States to, in the phrasing of the CIA, \enquote{truthfully deny} that any US \enquote{military planes} had flown over the USSR---which they had to do when the inevitable Soviet protest notes were filed after U-2 overflights began in 1956.\footcite[p.~109]{pedlow_central_1992}
 
@@
 If his only goal had been to achieve a level of plausible deniability in the event of a shootdown, then simply not wearing an Air Force uniform during the mission likely would have been sufficient.
-The \enquote{weather reconnaissance} excuse had even been used before with the ferret flights.\footcite[p.~45]{farquhar_aerial_2015} At the highest levels of American government, however, policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation.\enquote{It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft} read a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} Both halves of that equation were equally important.
+The classic \enquote{it's a weather reconnaissance craft} excuse had been used before with the ferret flights.\footcite[p.~45]{farquhar_aerial_2015} At the highest levels of American government, however, policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation.\enquote{It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft} read a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} Both halves of that equation were equally important.
 The Russians must believe that these operations are peacetime civilian operations, and they must be correct about it.
 
-At first, what made the U-2 flights \enquote{count} as espionage appears to be a deeply unsatisfying tautology.
-They counted as espionage because they were espionage!
-But Eisenhower's precautions were based in the reality of the United State's previous experience with reconnaissance flights.
+At first, what made the U-2 flights \enquote{count} as espionage appears to be a deeply unsatisfying tautology---they counted as espionage because they were espionage!
+But Eisenhower's precautions were based on the United State's previous experience with reconnaissance flights.
 He had already been testing the limits of what the Soviet Union would be willing to accept; if he did everything possible to signal that this plane functioned like a spy, perhaps it would accept this new development as well.
 No finite list of attributes could ensure that the Soviet Union wouldn't treat the U-2 plane---a massive leap in military surveillance technology---as a military development.
-But if the people in charge of Project Aquatone could convince \emph{themselves} that it was a civilian project, then they stood the best chance of having the Russians interpret it that way as well.
-The U-2 simply had to be a spy plane, in a manner that is both instantly recognizable and ontologically undefinable.
+But if the people in charge of Project Aquatone could convince \emph{themselves} that it was a civilian project, they stood a better chance of persuading the Russians to interpret it the same way.
+The U-2 simply had to be a spy plane, in a manner both instantly recognizable and ontologically undefinable.
 
 These precautions had the intended effect, even though it initially seemed like the Soviets were going to extract significant concessions.
-A Soviet Colonel in the USSR Air Defense later recalled that Khrushchev ``clearly viewed the violation of their nation's skies by a foreign reconnaissance aircraft \textelp{} as a political provocation'' and Khrushchev played it up.\footcite{orlov_u-2_2007} Three days after NASA quietly reported that they had lost a U-2 type weather reconnaissance plane, Khrushchev announced that the USSR had shot down an American plane that had flown into their airspace in \enquote{an aggressive provocation.}\footcite[p.~112]{powers_operation_2004} When the Americans stuck to their cover story, Khrushchev gleefully debunked it by revealing that they had taken Powers alive, Eisenhower admitted that the Gary Powers had been flying a spy plane, in a press conference where Eisenhower memorably declared that espionage was a \enquote{a distasteful but vital necessity} to prevent another attack like Pearl Harbor.\footcite{eisenhower_news_1960}
+A Soviet colonel in the USSR Air Defense later recalled that Khrushchev ``clearly viewed the violation of their nation's skies by a foreign reconnaissance aircraft \textelp{} as a political provocation.''\footcite{orlov_u-2_2007} Three days after NASA quietly reported that they had lost a U-2 type weather reconnaissance plane, Khrushchev announced that the USSR had shot down an American plane that had flown into its airspace in \enquote{an aggressive provocation.}\footcite[p.~112]{powers_operation_2004} When the Americans stuck to the \emph{weather craft} cover story, Khrushchev gleefully debunked it by revealing that Powers had been taken alive.
+In a press conference where Eisenhower memorably declared that espionage was a \enquote{a distasteful but vital necessity} to prevent another attack like Pearl Harbor, the President admitted that the Gary Powers had been flying a spy plane.\footcite{eisenhower_news_1960}
 
 And yet few real consequences followed.
-No one disputes that the collapse of the Paris Summit was directly attributable to the Gary Powers shootdown but, frankly, summits just aren't that important.
+No one disputes that the collapse of the Paris Summit was directly attributable to the Gary Powers shootdown but, in practice, summits aren't all that important.
 From 1953 to the end of the Cold War in 1991, there were 23 total US-USSR summits.\footcite{fain_chronology_2011} The main issue that did not get resolved in 1960 was the status of Berlin.
@@
 In the traditional telling, it significantly worsened US-Soviet relations at time when it seemed like lasting peace might be possible.
-I disagree.
+This interpretation is incomplete, if not inaccurate.
 Instead, the U-2 Incident should be understood as an embarrassing but otherwise typical failure of an intelligence operation, one consistent with the minimal consequences typically applied to espionage.\footnote{While I argue that the response to the U-2 flight was within the reasonable bounds for espionage, I acknowledge that it was still unprecedented in the context of reconnaissance flights.
@@
 See: \cite[p.~328]{kistiakowsky_scientist_1976}.} What would a major punishment have looked like?
-One proportional response Khrushchev could have considered was to force the issue on regional air bases.
+One proportional response Khrushchev could have considered was to press harder for the remove of regional air bases.
 Since the US was using nearby bases to launch invasive reconnaissance missions, the USSR could have demanded that the US either close some of them or agree to halt surveillance flights.
@@
 Farquhar, for instance, refers to this as the \enquote{cycle of hostility,} in which the series of aerial incidents increased suspicions of military buildup on both sides, leading to even more aerial incidents as both powers attempted to verify their suspicions.\footcite[p.~43]{farquhar_aerial_2015} While it's true that the there were many aerial incidents, it is not at all clear how this affected any other aspect of US-Soviet diplomacy.
-I can't definitively say that there wasn't some diplomatic issue that would have gotten resolved at a summit absent these reconnaissance flights, but I can't locate an obvious one either.
 Consider the period of time over which these flights took place; if aerial reconnaissance had risen to anything more than a lingering irritant, then it should have to have been resolved in some way before 1960.
@@
 During this period, hundreds of flights took place in contentious territory---many of them purposefully violating Soviet boundaries, causing some of them to be shot down---and the most significant consequence was a failed summit meeting, after which the flights continued on almost as before.
-Both sides were so willing to contain the issue that many of the airmen never saw their families again.
+Both sides were so willing to contain the issue that, tragically, many of the airmen never saw their families again.
 
@@
 a diplomatic protest.
-The US quickly apologized, and that was that.\footcite{orlov_u-2_2007}
+The US quickly apologized, and that settled the matter.\footcite{orlov_u-2_2007}
 

### 2020-01-10 article.tex
Make final edits before sending full draft to Wohlforth

@@
 At first glance this observation might seem almost tautological.
-Of course cyber-espionage is treated like traditional espionage, one would argue---they're both espionage.
+Of course cyber-espionage is treated like traditional espionage, one could argue---they're both espionage.
 That formulation, however, begs the question; it doesn't explain how states determine what constitutes \enquote{traditional} espionage in cyberspace, what purpose that distinction serves, or whether it is prudent to continue making it.
@@
 \enquote{Clearly, it does.}\footcite{sanger_u.s._2015} As as far as the public knows, however, the size of operation did \emph{not} change the nature of the diplomatic response.
-Therefore, to understand why the Director of the CIA treats cyber-espionage as \enquote{fair game,} one first needs to understand why traditional espionage has historically received the same treatment.
+Therefore, to understand why the Director of the CIA considers cyber-espionage to be \enquote{fair game,} one first needs to understand why traditional espionage has historically received the same treatment.
 
-I posit that universal tolerance of espionage in the international system can be explained through a rational calculation of each state's incentives.
+This article posits that universal tolerance of espionage in the international system can be explained through a rational calculation of each state's incentives.
 On balance, states derive a benefit from preserving a system in which the consequences for espionage are minimized, so they refrain from imposing particularly harsh punishments for its practice.
@@
 Since this generally doesn't happen between major world powers, one of two things must be true: either a state is miscalculating the relative gains it achieves through espionage, or the mutual tolerance of espionage has benefits that are sufficiently important to maintain even if one state gains less from it than its adversary.
-I argue the latter.
 
-This article seeks to establish a theoretical framework for analyzing the role that espionage plays in international relations (IR) and the diplomatic norms that are associated with it.
+This article argues the latter.
+To do so, I first establish a theoretical framework for analyzing the role that espionage plays in international relations (IR) and the diplomatic norms that are associated with it.
 Through this framework, I explain what constitutes traditional espionage and why rational states might calculate that it is in their best interests to continue tolerating it.
@@
 \subsection{Existing literature}
+Until now, few, if any, scholars have analyzed why espionage norms exist and what strategic purpose they serve.
 The benefits of espionage have never been underrated---\enquote{it is only the enlightened ruler and the wise general,} reads Sun Tzu, \enquote{who will use the highest intelligence of the army for purposes of spying and thereby they achieve great results}---but the inevitability of espionage has long been taken for granted.
 If one questions the assertion that espionage is something \enquote{all countries do,} then the way that countries treat espionage becomes considerably more confusing.
-Until now, few, if any, scholars have analyzed why espionage norms exist and what strategic purpose they serve.
 
@@
 In rare cases when additional costs \emph{are} imposed on a state for conducting a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.
-In this article I present a theory for why those norms exist.
 
@@
 If an American government employee walks out of The Pentagon and hands classified documents to the KGB, they are at risk prosecuted as a traitor; if an American case officer hands classified KGB documents over to the CIA, they will be lauded as a hero.
-This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.} As a result, scholars have typically treated espionage as \emph{sui generis}---starting from the assumption that espionage norms exist in a class of their own, then attempting to fit a legal doctrine around that fact.
+This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.} Because of that tension, scholars have typically treated espionage as \emph{sui generis}---starting from the assumption that espionage norms exist in a class of their own, then attempting to fit a legal doctrine around that fact.
 Jared Beim determines that while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so the prohibition on espionage merits further enforcement.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe.
@@
 Each time a new espionage technology is made available, policymakers are given a fresh opportunity to decide whether or not the norms of espionage ought to extend to its use.
-Critically, the research that follows will show that they were not obligated to do so by any pre-existing norms.
+Critically, the research that follows will demonstrate that they were not obligated to do so by any pre-existing norms.
 When the US first introduced photographic satellites, for example, it was not clear whether the Soviet Union would treat the satellites as spies or as weapons---there were indications in both directions.

### 2020-03-12 article.tex
Begin slicing introduction

@@
 \maketitle
+The Chinese Government's 2014 breach of the United States Office of Personnel Management is notable for two reasons.
+First, because it is, to date, the most damaging cyberattack committed against the United States government by a foreign agent.
+Federal authorities determined that the hack ultimately affected 22.1 million people, including personnel at the highest levels of government.\footcite{nakashima_hacks_2015} Among the stolen materials were applications for federal security clearances, which exposed not just the personal data of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses.
+US officials feared that the stolen data could compromise intelligence efforts by exposing undercover CIA officers, or identify the American officials who might be susceptible to blackmail, or worse, recruitment.
+The public backlash was so intense that FBI Director James Comey was forced to publicly take questions about the incident.
+\enquote{If you have my [application for security clearance],} Comey admitted, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.
+Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015}
 
-The most damaging cyberattack committed against the United States government by a foreign agent was remarkably subtle.
-Starting sometime in 2014, a small probe embedded itself deep within the network of the US Office of Personnel Management (OPM), an office of about 5,000 employees tasked with overseeing background checks, payroll, and other human resources concerns for the federal government.
-At regular intervals, the probe would send updates to opmsecurity.org, a domain registered to look like an official website, but not actually affiliated with the US government or the OPM security team.
-Unobtrusive and observant, the probe functioned in all respects like a digital spy.
-
-By the time the breach was discovered in April 2015, the probe had accessed sensitive government databases and transmitted the stolen information back to its country of origin---including a trove of applications for federal security clearance.
-These applications exposed not just the personal data of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses.
-Federal authorities determined that the hack ultimately affected 22.1 million people, including personnel at the highest levels of government.\footcite{nakashima_hacks_2015} When OPM announced the full extent of the damage, it caused an uproar.
-The American Federation of Government Employees, the largest federal workers union, filed a class-action lawsuit against OPM, seeking damages under the Privacy Act.\footcite{chalfant_court_2017} Senator Mark Warner (D-Va.) of the Senate Intelligence Committee called for OPM Director Katherine Archuleta's resignation.
-Because the information might be used for identity theft, the US government offered all affected employees free credit and identity monitoring services for three years.
-
-Not only was the OPM hack a public relations disaster, it had significant national security implications as well.
-US officials feared that the stolen data could compromise intelligence efforts by exposing undercover CIA officers working in embassies around the world, whose names would be suspiciously absent from OPM records.
-The stolen information could also be used to identify American officials who might be susceptible to pressure, or worse, recruitment.
-The backlash was so intense that FBI Director James Comey was forced to publicly take questions about the incident.
-\enquote{If you have my [application for security clearance],} Comey admitted, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.
-Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015} With evidence that a foreign intelligence service had successfully obtained exactly that, one might expect the United States to respond with a series of diplomatic measures intended to signal that highly invasive acts of espionage would not be tolerated from its international adversaries.
+The second reason it is notable is because the United States government did almost nothing in response.
 
@@
 See \cite{sanger_trump_2018}}
+
+With evidence that a foreign intelligence service had successfully obtained exactly that, one might expect the United States to respond with a series of diplomatic measures intended to signal that highly invasive acts of espionage would not be tolerated from its international adversaries.
+
+% When OPM announced the full extent of the damage, it caused an uproar.
+The American Federation of Government Employees, the largest federal workers union, filed a class-action lawsuit against OPM, seeking damages under the Privacy Act.\footcite{chalfant_court_2017} Senator Mark Warner (D-Va.) of the Senate Intelligence Committee called for OPM Director Katherine Archuleta's resignation.
+Because the information might be used for identity theft, the US government offered all affected employees free credit and identity monitoring services for three years.
+
+Not only was the OPM hack a public relations disaster, it had significant national security implications as well.
+
+% Starting sometime in 2014, a small probe embedded itself deep within the network of the US Office of Personnel Management (OPM), an office of about 5,000 employees tasked with overseeing background checks, payroll, and other human resources concerns for the federal government.
+At regular intervals, the probe would send updates to opmsecurity.org, a domain registered to look like an official website, but not actually affiliated with the US government or the OPM security team.
+Unobtrusive and observant, the probe functioned in all respects like a digital spy.
 

### 2020-04-17 article.tex
Rewrite up to Theory

@@
 \maketitle
-The Chinese Government's 2014 breach of the United States Office of Personnel Management is notable for two reasons.
-First, because it is, to date, the most damaging cyberattack committed against the United States government by a foreign agent.
-Federal authorities determined that the hack ultimately affected 22.1 million people, including personnel at the highest levels of government.\footcite{nakashima_hacks_2015} Among the stolen materials were applications for federal security clearances, which exposed not just the personal data of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses.
-US officials feared that the stolen data could compromise intelligence efforts by exposing undercover CIA officers, or identify the American officials who might be susceptible to blackmail, or worse, recruitment.
-The public backlash was so intense that FBI Director James Comey was forced to publicly take questions about the incident.
-\enquote{If you have my [application for security clearance],} Comey admitted, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.
-Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015}
 
-The second reason it is notable is because the United States government did almost nothing in response.
-
-Instead, the opposite happened.
-US officials would only confirm that they believed a \enquote{foreign entity or government} had been behind the attack, even though it was widely known that they suspected the Chinese government.\footcite{spetalnick_china_2015} Internal investigators quickly determined that OPM had been compromised by an Advanced Persistent Threat (APT), a formally organized and typically state-sponsored group of hackers that engage in long-term, targeted penetration operations.
-In this case, the APT made identification easy by leaving its calling-card: the trojan opmsecurity.org domain was registered in the name of \enquote{Steve Rogers,} the Marvel comics character better known as Captain America.
-This digital taunt, along with email and IP addresses that verified the country of origin was not a false flag, all but guaranteed that the attack had come from an APT sponsored by the Chinese government, likely the Chinese military's cyber-espionage division.\footcite{koerner_inside_2016} Though it possessed conclusive digital forensics, the Obama administration nonetheless refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018.
+In 2015, federal authorities discovered that malicious actors had for months maintained access to databases within the United States Office of Personnel Management (OPM).
+Among other types of sensitive data, these hackers stole a trove of applications for federal security clearances, securing not just the personal information of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses.
+The breach ultimately compromised the information of 22.1 million people, including personnel at the highest levels of government.\footcite{nakashima_hacks_2015} US officials feared that the stolen data could jeopardize intelligence efforts by exposing undercover CIA officers, or identify the American officials who might be susceptible to blackmail, or worse, recruitment.
+\enquote{if you have my [application for security clearance],} FBI Director James Comey admitted, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.
+Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015}.
+In the case of the OPM hack, imagination wasn't necessary; the US had already established that the Chinese military was likely behind the sophisticated attack.\footcite{koerner_inside_2016}.
+But even though it possessed conclusive digital forensics, the Obama administration refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018.
 See \cite{sanger_trump_2018}}
-
-With evidence that a foreign intelligence service had successfully obtained exactly that, one might expect the United States to respond with a series of diplomatic measures intended to signal that highly invasive acts of espionage would not be tolerated from its international adversaries.
 
@@
 
-Not only was the OPM hack a public relations disaster, it had significant national security implications as well.
+The US government's tepid response to the OPM hack illustrates a crucial fact about the relationship between cybersecurity and espionage: when a cyberattack looks like a traditional intelligence operation, it merits little to no diplomatic consequences.
+To the extent that phenomenon has been studied politically or academically, it is often explained away as a logical extension of existing policy: cyber-espionage is treated like \enquote{traditional} espionage because they are both forms of espionage.
+This article seeks to unravel that tautology by explaining how states determine what constitutes \enquote{traditional} espionage, what purpose that distinction serves, and whether it is prudent to continue making it.
+The transfer of espionage into the digital realm has made that task increasingly urgent, as the internet offers new vectors of attack that magnify the potential of espionage to do harm.
+``This is one of those cases where you have to ask, `Does the size of the operation change the nature of it?'\thinspace'' said one anonymous senior intelligence official of the OPM hack.
+\enquote{Clearly, it does.}\footcite{sanger_u.s._2015} As far as the public knows, however, the size of operation did \emph{not} change the nature the diplomatic response.
+Espionage, regardless of the damage it does, is always handled more lightly than other forms of international aggression.
 
-% Starting sometime in 2014, a small probe embedded itself deep within the network of the US Office of Personnel Management (OPM), an office of about 5,000 employees tasked with overseeing background checks, payroll, and other human resources concerns for the federal government.
-At regular intervals, the probe would send updates to opmsecurity.org, a domain registered to look like an official website, but not actually affiliated with the US government or the OPM security team.
-Unobtrusive and observant, the probe functioned in all respects like a digital spy.
+% Therefore, to understand why the Director of the CIA considers cyber-espionage to be \enquote{fair game,} one first needs to understand why traditional espionage has historically received the same treatment.
 
-The US government's tepid response to the OPM hack illustrates a crucial fact about the relationship between cybersecurity and espionage: when a cyberattack appears to fall within the bounds of a traditional intelligence operation, it will merit the diplomatic consequences of a traditional intelligence operation.
-US policymakers rarely acknowledge that this distinction exists---officially, cyberattacks are illegal international behavior and will receive a strong response---but occasionally, in moments of candor, policymakers will admit that cyber-espionage is different.
+This article posits that the universal tolerance of espionage in the international system can be explained through a rational calculation of each state's incentives.
+On balance, all states derive benefits from preserving a system in which the consequences for espionage are minimized, so they refrain from imposing particularly harsh punishments for its practice.
+Even when an individual incident is especially damaging to a state's security, if it falls within the traditionally accepted boundaries of an espionage operation, it will likely go unpunished.
+In the case of the OPM hack, for example, US policymakers decided they could not impose harsh penalties on the Chinese because it was more valuable for the US to retain its ability to practice cyber-espionage with relative impunity in the future.
+This basic calculation---that espionage against one's own country is a tolerable price to pay for being allowed to practice espionage against one's rivals---has remained consistent since the advent of the modern peacetime intelligence service, perhaps even the nation-state itself.
+
+To evaluate each state's incentives I first establish a theoretical framework for analyzing the role that espionage plays in international relations (IR) and the diplomatic norms that are associated with it.
+Through this framework, I define the bounds of traditional espionage and why rational states might calculate that it is in their best interests to continue tolerating it.
+I then examine two cases in which new technical means of gathering intelligence were introduced during the Cold War: aerial photography (spy planes) and satellite photography (spy satellites).
+In both cases, the norms of traditional espionage ultimately prevailed, minimizing the diplomatic consequences for their use as long as it was in service of traditional intelligence goals.
+Through process tracing, I demonstrate how the justifications policymakers employed for the use of these intelligence technologies---and the resulting minimization of the diplomatic penalties associated with them---are consistent with the theoretical underpinnings for how a tacit accord permitting general espionage might be mutually strategic.
+
+% Add information about my conclusions here
+
+\subsection{Espionage norms}
+There are few extant definitions of peacetime espionage, and no satisfying ones.
+The problem that aspiring scholars of espionage inevitably encounter is that \enquote{peacetime espionage} is universally understood as a term of art but almost entirely missing from any legal codes or case law.
+Most scholars who tackle this question settle on a definition like the one proposed by I\~{n}aki Navarrete and Russell Buchan: \enquote{\enquote{Peacetime espionage} is a colloquial rather than a legal term [used] to describe different methods of collecting confidential information from closed as opposed to open sources \textelp{} generally conducted by States, against States.}\footcite[p.~901-902]{navarrete_out_2019} This definition purposefully excludes forms of extralegal government intervention that are often conflated with espionage, such as assassinations, election interference, and covert operations.\footnote{Often papers won't attempt to \enquote{define} espionage as such, but rather offer a working definition that bounds the scope of the paper.
+See \cite[p.~600]{radsan_unresolved_2007},\cite[p.~2]{kapp_spying_2007})} Excluding this activities is important; because assassinations and similar covert operations are not considered espionage they do not typically receive the protections of espionage norms.
+
+For the purposes of this paper, I extend that definition to describe espionage \emph{norms} as the common understanding of how states are expected to respond to respond to activities that remain within the bounds of espionage.\footnote{There are, of course, other \enquote{norms} that involve espionage.
+The so-called \enquote{Moscow Rules} dictate an intelligence officer are supposed to conduct themselves in hostile territory.
+In order to avoid constantly specifying, espionage norms are more strictly defined to include only the ones that affect IR.} Espionage norms prescribe that when one state uncovers an espionage operation conducted against them, the consequences imposed on the spier remain minimal.
+Rarely does the response amount to more than a pat denunciation of the activity and implementation of the necessary countermeasures to end it---the consequences for espionage are borne by the agents who commit it, not the state that sent them.
+In rare cases when additional costs \emph{are} imposed on a state for conducting a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.
+
+Once again, we can use cyber-espionage as an analogue for traditional espionage writ large.
+Though not codified in US policy, the informal regime of espionage norms is an open secret.
 A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].}\footnote{In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets.
-See \cite{finklea_cyber_2015}} For activities that fall under the umbrella of intelligence-gathering, counterintelligence---not prosecution---is deemed to be the appropriate response.
-Director of the CIA John Brennan described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, was particularly frank: \enquote{You have to kind of salute the Chinese for what they did.
+See \cite{finklea_cyber_2015}} For activities that fall under the umbrella of intelligence-gathering, counterintelligence rather than prosecution is considered the appropriate response.
+CIA Director John Brennan described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, was particularly frank: \enquote{You have to kind of salute the Chinese for what they did.
 If we had the opportunity to do that, I don’t think we’d hesitate for a minute.}\footcite{pepitone_clapper_2015}
+
+There is solid data to support the presumption that cyber-espionage is routinely treated as a lesser category of offense by all nations, not just the United States.
+Of the hundreds of state-sponsored cyberattacks known to the public, attacks classified as espionage are by far the most common.
+By percentages, however, cyber-espionage is by far the least likely form of cyberattack to garner an official response from the victim (Table \ref{response-pct}).\footnote{These numbers were compiled from the Council on Foreign Relations' Cyber Operations Tracker, which is a publicly-sourced record of state-sponsored cyberattacks, updated quarterly.
+Of the 323 listed incidents, 38 were excluded from this calculation because they lacked an identifiable Victim, Sponsor, or Type.
+For more, see \cite{council_on_foreign_relations_new_2019}}
 
@@
 
-In fact, cyber-espionage is routinely treated as a lesser category of offense by all nations, not just the United States.
-Of the hundreds of state-sponsored cyberattacks known to the public, attacks classified as espionage are by far the most common---and the least likely to garner an official response from the victim government (Table \ref{response-pct}).\footnote{These numbers were compiled from the Council on Foreign Relations' Cyber Operations Tracker, which is a publicly-sourced record of state-sponsored cyberattacks, updated quarterly.
-Of the 323 listed incidents, 38 were excluded from this calculation because they lacked an identifiable Victim, Sponsor, or Type.
-For more, see \cite{council_on_foreign_relations_new_2019}} The routine minimization of espionage is most easily observed through the plethora of modern cyber-incidents, but it has its roots in norms that are much older.
-
-At first glance this observation might seem almost tautological.
-Of course cyber-espionage is treated like traditional espionage, one could argue---they're both espionage.
-That formulation, however, begs the question; it doesn't explain how states determine what constitutes \enquote{traditional} espionage in cyberspace, what purpose that distinction serves, or whether it is prudent to continue making it.
-Certainly the US should respond proportionally to attacks on its systems, but what if the damage done to national security by espionage is especially significant?
-``This is one of those cases where you have to ask, `Does the size of the operation change the nature of it?'\thinspace'' said one anonymous senior intelligence official of the OPM hack.
-\enquote{Clearly, it does.}\footcite{sanger_u.s._2015} As as far as the public knows, however, the size of operation did \emph{not} change the nature of the diplomatic response.
-Therefore, to understand why the Director of the CIA considers cyber-espionage to be \enquote{fair game,} one first needs to understand why traditional espionage has historically received the same treatment.
-
-This article posits that universal tolerance of espionage in the international system can be explained through a rational calculation of each state's incentives.
-On balance, states derive a benefit from preserving a system in which the consequences for espionage are minimized, so they refrain from imposing particularly harsh punishments for its practice.
-Even when an individual incident is especially damaging to a state's security, if it falls within the traditionally accepted boundaries of an espionage operation, it will likely go unpunished.
-In the case of the OPM hack, for example, US policymakers decided they could not impose harsh penalties on the Chinese because it was more valuable for the US to retain its ability to practice cyber-espionage with relative impunity in the future.
-This basic calculation---that espionage against one's own country is a tolerable price to pay for being allowed to practice espionage against one's rivals---has remained consistent since the advent of the modern peacetime intelligence service, perhaps even the nation-state itself.
-
-This hypothesis raises a fascinating paradox: for espionage to be mutually permissible, both states must agree to treat it that way, but, in theory, that requires \emph{both} states to determine they are getting more from practicing espionage against their adversary than their adversary is getting from practicing it against them.
-Otherwise, the state with less to gain would simply defect from that equilibrium and punish espionage at a level that imposes direct costs.
-Since this generally doesn't happen between major world powers, one of two things must be true: either a state is miscalculating the relative gains it achieves through espionage, or the mutual tolerance of espionage has benefits that are sufficiently important to maintain even if one state gains less from it than its adversary.
-
-This article argues the latter.
-To do so, I first establish a theoretical framework for analyzing the role that espionage plays in international relations (IR) and the diplomatic norms that are associated with it.
-Through this framework, I explain what constitutes traditional espionage and why rational states might calculate that it is in their best interests to continue tolerating it.
-I then examine two cases in which new technical means of gathering intelligence were introduced during the Cold War: aerial photography (spy planes) and satellite photography (spy satellites).
-In both cases, the norms of traditional espionage ultimately prevailed, minimizing the diplomatic consequences for their use as long as it was in service of traditional intelligence goals.
-Through process tracing, I demonstrate how the justifications policymakers employed for the use of these intelligence technologies---and the resulting minimization of the diplomatic penalties associated with them---are consistent with the theoretical underpinnings for how a tacit accord permitting general espionage might be mutually strategic.
+% The benefits of espionage have never been underrated---\enquote{it is only the enlightened ruler and the wise general,} reads Sun Tzu, \enquote{who will use the highest intelligence of the army for purposes of spying and thereby they achieve great results}---but the inevitability of espionage has long been taken for granted.
+If one questions the assertion that espionage is something \enquote{all countries do,} then the way that countries treat espionage becomes considerably more confusing.
 
 \subsection{Existing literature}
-Until now, few, if any, scholars have analyzed why espionage norms exist and what strategic purpose they serve.
-The benefits of espionage have never been underrated---\enquote{it is only the enlightened ruler and the wise general,} reads Sun Tzu, \enquote{who will use the highest intelligence of the army for purposes of spying and thereby they achieve great results}---but the inevitability of espionage has long been taken for granted.
-If one questions the assertion that espionage is something \enquote{all countries do,} then the way that countries treat espionage becomes considerably more confusing.
-
-Before continuing, it is necessary to properly define what I mean by espionage norms.
-Simply put, espionage norms dictate what constitutes espionage and how states are expected to respond when it is conducted against them.\footnote{There are, of course, other \enquote{norms} that involve espionage.
-The so-called \enquote{Moscow Rules} dictate an intelligence officer are supposed to conduct themselves in hostile territory.
-In order to avoid constantly specifying, espionage norms are more strictly defined to include only the ones that affect IR.} What consequences does a state impose when it uncovers an intelligence operation run by another state?
-Espionage norms prescribe that the consequences remain minimal.
-Rarely does the response amount to more than a pat denunciation of the activity and implementation of the necessary countermeasures to end it---the consequences for espionage are borne by the agents who commit it, not the state that sent them.
-In rare cases when additional costs \emph{are} imposed on a state for conducting a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.
-
 Prior literature has attempted to explain espionage norms using international law, but in the absence of any significant treaties governing peacetime espionage, scholars are often left without definitive answers.\footnote{Because it is universally agreed that international law does not deal with peacetime espionage directly, many journal articles that deal with this question frame their work almost Socratically, devoting two sections to cases for and against the legality of espionage.
 In fact, the first major work on law and peacetime espionage is a series of commentaries collected in a single publication that respond to each other's arguments.
-See \cite{wright_essays_1962}} Since espionage is universally illegal, how can every state acknowledge that allies and adversaries alike are guaranteed to engage in it?\footnote{Further complicating matters, espionage is the rare crime that might not be a crime depending on who the victim is.
+See \cite{wright_essays_1962}} Espionage is universally illegal but every state acknowledges that allies and adversaries alike are guaranteed to engage in it.\footnote{Further complicating matters, espionage is the rare crime that might not be a crime depending on who the victim is.
 If an American government employee walks out of The Pentagon and hands classified documents to the KGB, they are at risk prosecuted as a traitor; if an American case officer hands classified KGB documents over to the CIA, they will be lauded as a hero.
@@
 Scholars can use rationalist principles to analyze everything from trade policy to the space race; espionage, which by definition involves deeply invasive interactions with another state, ought to be no different.
+
+% Of course, for espionage to be mutually permissible, both states must agree to treat it that way, but, in a world with perfect information, that requires \emph{both} states to determine they are getting more from practicing espionage against their adversary than their adversary is getting from practicing it against them.
+Otherwise, the state with less to gain would simply defect from that equilibrium and punish espionage at a level that imposes direct costs.
+
+% Since this generally doesn't happen between major world powers, one of two things must be true: either a state is miscalculating the relative gains it achieves through espionage, or the mutual tolerance of espionage has benefits that are sufficiently important to maintain even if one state gains less from it than its adversary.
 

### 2020-04-19 article.tex
Made clean-up edits from dad

@@
 \author{Alexander W. Petros}
-\date{September 2019}
+\date{April 2020}
 
@@
 In 2015, federal authorities discovered that malicious actors had for months maintained access to databases within the United States Office of Personnel Management (OPM).
-Among other types of sensitive data, these hackers stole a trove of applications for federal security clearances, securing not just the personal information of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses.
-The breach ultimately compromised the information of 22.1 million people, including personnel at the highest levels of government.\footcite{nakashima_hacks_2015} US officials feared that the stolen data could jeopardize intelligence efforts by exposing undercover CIA officers, or identify the American officials who might be susceptible to blackmail, or worse, recruitment.
-\enquote{if you have my [application for security clearance],} FBI Director James Comey admitted, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.
+The hackers downloaded reams of sensitive data, including a cache of applications for federal security clearance.
+These applications exposed not just the personal information of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses.
+The breach ultimately compromised the information of 22.1 million people, including personnel at the highest levels of government.\footcite{nakashima_hacks_2015} US officials feared that the stolen data could help foreign intelligence services unmask undercover CIA officers, or identify American officials who might be susceptible to blackmail, or worse, recruitment.
+\enquote{[I]f you have my [application for security clearance],} FBI Director James Comey admitted, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.
 Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015}.
-In the case of the OPM hack, imagination wasn't necessary; the US had already established that the Chinese military was likely behind the sophisticated attack.\footcite{koerner_inside_2016}.
+In the case of the OPM hack, a vivid imagination wasn't necessary; the US had already established that the Chinese military was likely behind the sophisticated attack.\footcite{koerner_inside_2016}.
 But even though it possessed conclusive digital forensics, the Obama administration refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018.
@@
 The US government's tepid response to the OPM hack illustrates a crucial fact about the relationship between cybersecurity and espionage: when a cyberattack looks like a traditional intelligence operation, it merits little to no diplomatic consequences.
-To the extent that phenomenon has been studied politically or academically, it is often explained away as a logical extension of existing policy: cyber-espionage is treated like \enquote{traditional} espionage because they are both forms of espionage.
-This article seeks to unravel that tautology by explaining how states determine what constitutes \enquote{traditional} espionage, what purpose that distinction serves, and whether it is prudent to continue making it.
-The transfer of espionage into the digital realm has made that task increasingly urgent, as the internet offers new vectors of attack that magnify the potential of espionage to do harm.
-``This is one of those cases where you have to ask, `Does the size of the operation change the nature of it?'\thinspace'' said one anonymous senior intelligence official of the OPM hack.
-\enquote{Clearly, it does.}\footcite{sanger_u.s._2015} As far as the public knows, however, the size of operation did \emph{not} change the nature the diplomatic response.
-Espionage, regardless of the damage it does, is always handled more lightly than other forms of international aggression.
+Within both political and academic circles, this phenomenon is often rationalized as a logical extension of existing policy; cyber-espionage is treated like \enquote{traditional} espionage because they are both forms of espionage.
+This article seeks to unravel that tautology by first by explaining how states determine what constitutes \enquote{traditional} espionage and then by analyzing what strategic purpose that distinction serves.
+The practice of espionage in the digital realm has made this task increasingly urgent, as the internet offers new vectors of attack that magnify the potential of espionage to do harm, and governments re-evaluate whether to employ decades-old norms.
+Nations like Brazil are openly pushing for espionage to be regulated through larger regimes of Internet governance.\footcite[p.~464]{abdenur_triggering_2015} ``This is one of those cases where you have to ask, `Does the size of the operation change the nature of it?'\thinspace'', said one anonymous senior intelligence official of the OPM hack.
+\enquote{Clearly, it does.}\footcite{sanger_u.s._2015} For now, however, even operations on the scale of China's OPM infiltration still do not merit significant retaliation.
+
+% Espionage, regardless of the damage it does, is always handled more lightly than other forms of international aggression.
 
@@
 In both cases, the norms of traditional espionage ultimately prevailed, minimizing the diplomatic consequences for their use as long as it was in service of traditional intelligence goals.
-Through process tracing, I demonstrate how the justifications policymakers employed for the use of these intelligence technologies---and the resulting minimization of the diplomatic penalties associated with them---are consistent with the theoretical underpinnings for how a tacit accord permitting general espionage might be mutually strategic.
+Through process tracing, I demonstrate how the justifications policymakers employed for the use of these intelligence technologies---and the resulting minimization of the diplomatic penalties associated with them---are consistent with the theoretical underpinnings for a mutually strategic tacit accord permitting peacetime espionage.
 
@@
 There are few extant definitions of peacetime espionage, and no satisfying ones.
-The problem that aspiring scholars of espionage inevitably encounter is that \enquote{peacetime espionage} is universally understood as a term of art but almost entirely missing from any legal codes or case law.
-Most scholars who tackle this question settle on a definition like the one proposed by I\~{n}aki Navarrete and Russell Buchan: \enquote{\enquote{Peacetime espionage} is a colloquial rather than a legal term [used] to describe different methods of collecting confidential information from closed as opposed to open sources \textelp{} generally conducted by States, against States.}\footcite[p.~901-902]{navarrete_out_2019} This definition purposefully excludes forms of extralegal government intervention that are often conflated with espionage, such as assassinations, election interference, and covert operations.\footnote{Often papers won't attempt to \enquote{define} espionage as such, but rather offer a working definition that bounds the scope of the paper.
-See \cite[p.~600]{radsan_unresolved_2007},\cite[p.~2]{kapp_spying_2007})} Excluding this activities is important; because assassinations and similar covert operations are not considered espionage they do not typically receive the protections of espionage norms.
+The problem that aspiring espionage scholars inevitably encounter is that \enquote{peacetime espionage} is universally understood as a term of art but is almost entirely missing from any legal codes or case law.
+Most scholars who tackle this question settle on a definition like the one proposed by I\~{n}aki Navarrete and Russell Buchan: \enquote{\enquote{Peacetime espionage} is a colloquial rather than a legal term [used] to describe different methods of collecting confidential information from closed as opposed to open sources \textelp{} generally conducted by States, against States.}\footcite[p.~901-902]{navarrete_out_2019} This definition purposefully excludes forms of extralegal government intervention that are often conflated with espionage, such as assassinations, election interference, and covert operations.\footnote{Often papers won't take a stand on defining espionage, but rather offer a working definition just to bound the scope of that particular paper.
+See \cite[p.~600]{radsan_unresolved_2007}, \cite[p.~2]{kapp_spying_2007})} Excluding these activities is important; because assassinations and similar covert operations are not considered espionage, they do not typically receive the protections of espionage norms.
 
-For the purposes of this paper, I extend that definition to describe espionage \emph{norms} as the common understanding of how states are expected to respond to respond to activities that remain within the bounds of espionage.\footnote{There are, of course, other \enquote{norms} that involve espionage.
+For the purposes of this paper, I extend that definition to describe espionage \emph{norms} as the common understanding of how states are expected to respond to activities that remain within the bounds of espionage.\footnote{There are, of course, other \enquote{norms} that involve espionage.
 The so-called \enquote{Moscow Rules} dictate an intelligence officer are supposed to conduct themselves in hostile territory.
-In order to avoid constantly specifying, espionage norms are more strictly defined to include only the ones that affect IR.} Espionage norms prescribe that when one state uncovers an espionage operation conducted against them, the consequences imposed on the spier remain minimal.
+In order to avoid constantly specifying, espionage norms are more strictly defined to include only the ones that affect IR.} Espionage norms prescribe that when one state uncovers an espionage operation conducted against it, the consequences imposed on the spying country remain minimal.
 Rarely does the response amount to more than a pat denunciation of the activity and implementation of the necessary countermeasures to end it---the consequences for espionage are borne by the agents who commit it, not the state that sent them.
@@
 
-Once again, we can use cyber-espionage as an analogue for traditional espionage writ large.
+Once again, we can use cyber-espionage as an analogue for traditional espionage.
 Though not codified in US policy, the informal regime of espionage norms is an open secret.
@@
 See \cite{finklea_cyber_2015}} For activities that fall under the umbrella of intelligence-gathering, counterintelligence rather than prosecution is considered the appropriate response.
-CIA Director John Brennan described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, was particularly frank: \enquote{You have to kind of salute the Chinese for what they did.
+CIA Director John Brennan once described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, was particularly frank: \enquote{You have to kind of salute the Chinese for what they did [to OPM].
 If we had the opportunity to do that, I don’t think we’d hesitate for a minute.}\footcite{pepitone_clapper_2015}
@@
 This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.} Because of that tension, scholars have typically treated espionage as \emph{sui generis}---starting from the assumption that espionage norms exist in a class of their own, then attempting to fit a legal doctrine around that fact.
-Jared Beim determines that while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so the prohibition on espionage merits further enforcement.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe.
-For more, see \cite{beim_enforcing_2018}} Beim acknowledges, however, that his conclusion will have little impact on the practice of espionage, where, at best, \enquote{it is perhaps possible to stop the more egregious violations.}\footcite[p.~672]{beim_enforcing_2018} A.
+Jared Beim concludes that while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so there ought to be stronger enforcement mechanisms in place.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe.
+For more, see \cite{beim_enforcing_2018}} Beim acknowledges, however, that at best \enquote{it is perhaps possible to stop the more egregious violations.}\footcite[p.~672]{beim_enforcing_2018} A.
 John Radsan employs a more philosophical approach.
-Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage generally reach the same conclusion: it would be almost impossible to argue that espionage is legal, but no one really seems to care.
+Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage generally reach the same conclusion: it would be almost impossible to argue that espionage is legal, but the international system happily leaves this tension in a state of salutary neglect.
 Taken to its logical endpoint, this branch of scholarship would simply argue that the espionage norms we observe today result from simple inertia.

### 2020-06-02 article.tex
Finish theory rewrite

@@
 
-% When OPM announced the full extent of the damage, it caused an uproar.
-The American Federation of Government Employees, the largest federal workers union, filed a class-action lawsuit against OPM, seeking damages under the Privacy Act.\footcite{chalfant_court_2017} Senator Mark Warner (D-Va.) of the Senate Intelligence Committee called for OPM Director Katherine Archuleta's resignation.
-Because the information might be used for identity theft, the US government offered all affected employees free credit and identity monitoring services for three years.
-
 The US government's tepid response to the OPM hack illustrates a crucial fact about the relationship between cybersecurity and espionage: when a cyberattack looks like a traditional intelligence operation, it merits little to no diplomatic consequences.
@@
 \enquote{Clearly, it does.}\footcite{sanger_u.s._2015} For now, however, even operations on the scale of China's OPM infiltration still do not merit significant retaliation.
-
-% Espionage, regardless of the damage it does, is always handled more lightly than other forms of international aggression.
-
-% Therefore, to understand why the Director of the CIA considers cyber-espionage to be \enquote{fair game,} one first needs to understand why traditional espionage has historically received the same treatment.
 
@@
 
-% This article expands on Baker's basic premise---that there are situations in which espionage is mutually advantageous for both parties---and universalizes it to explain why the entire institution of espionage is mutually advantageous.
-
 The key insight this article presents---and where it breaks from previous literature---is that far from being an irreplicable, historically contingent set of practices, espionage norms are actually preserved through an ongoing set of strategic decisions.
@@
 
-% Because I argue that espionage norms influence cyber policy, it is worth mentioning scholarship on the role that espionage, outside of cyberspace, plays in the international system today.
-The intersection of espionage and international relations has been studied before---beginning, in the modern context, as far back as 1962---but usually in a historical context, not a political science one.
-The role of espionage in deescalating conflict is typically relegated to accounts of moments where Great People properly interpreted---or fatally misunderstood---the information that their intelligence operations provided them.
-Chapters 3 and 4 will primarily focus on decisions made by Eisenhower and Khrushchev, so to some extent I am guilty of that as well.
-Nonetheless, I will try to highlight the way in which both of those leaders set precedents designed to outlast them, across a variety of intelligence methods.
-I hope that this thesis starts the conversation about how espionage norms operate to deescalate conflict independent of the leaders that promote them.
+\section{Theory}
+\subsection{The nuclear option}
 
-% By employing both realist and constructivist schools of IR, this article develops rigorous explanations for why rational states, whether they be security-seeking and greedy, might cooperate to preserve this dynamic.
+% Add a poll about American knowledge of the CIA?
 
-% incorporating cyber espionage into the intelligence literature, rather than the cyebr literature
-% espionage first, cyber second
+Civilian study of espionage is sometimes treated with skepticism because espionage agencies, by definition, operate with maximum possible secrecy.
+That secrecy does not inoculate these agencies from critical analysis.
+While the operations of a modern intelligence service are opaque, the decision to maintain one is not.
+American citizens, as well as their international counterparts, are broadly aware that their nations conduct extensive intelligence operations abroad.
+The Central Intelligence Agency (CIA) is funded by the United States Congress, reports directly to the President, and offers internship positions for college students.
+Keeping these institutions intact is partly a function of inertia, but it is also an ongoing argument that espionage is a legitimate, even critical, component of statecraft.
 
-\section{Theory}
-\subsection{Establishing the puzzle}
-One of the most prevalent clich\'es in spy literature and journalism is the so-called \enquote{rules of the game.} The supposed rules of the spy world---\enquote{You are never alone}, \enquote{Don't trust anyone}---make the intelligence community seem insular, mysterious, and impenetrable.\footcite{myre_moscow_2019} They hint at a larger set of processes whose motivations the average citizen, not given access to this privileged information, could not possibly begin to understand.
-In reality, while the \emph{operation} of a modern intelligence service is incredibly opaque, the geopolitical situation in which one operates is not.
-Every aspect of a nation-state's interactions with another ought to be subject to the same critical analysis.
-Scholars can use rationalist principles to analyze everything from trade policy to the space race; espionage, which by definition involves deeply invasive interactions with another state, ought to be no different.
+% Scholars can use rationalist principles to analyze everything from trade policy to the space race; espionage, which by definition involves deeply invasive interactions with another state, ought to be no different.
 
-% Of course, for espionage to be mutually permissible, both states must agree to treat it that way, but, in a world with perfect information, that requires \emph{both} states to determine they are getting more from practicing espionage against their adversary than their adversary is getting from practicing it against them.
-Otherwise, the state with less to gain would simply defect from that equilibrium and punish espionage at a level that imposes direct costs.
+Using traditional methods of analysis, espionage certainly seems like something that states should be doing more to discourage.
+The results of intelligence operations can have devastating, even existential effects on national security.
+In recent years China has relentlessly targeted American military secrets.
+A United States Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} \enquote{Long-term, US future military advantage is being diminished by years of IP exfiltration from the DoD, DoN, and DIB,} the report read, \enquote{all with little to no adverse consequences to the thieves.}\footcite[p.~6]{bayer_cybersecurity_2019} The forecasts were dire: \enquote{If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.}\footcite[p.~5]{bayer_cybersecurity_2019} The US Navy is not a disinterested party, of course, and we should be cautious about its conclusions; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see: \cite{gilli_why_2019}} Nonetheless, espionage presents alarming opportunities for rival states to gain a comparative advantage and the consequences for taking advantage of them remain minimal.
 
@@
 
-Using traditional methods of analysis, espionage certainly seems like something that states should be doing more to discourage.
-The results of intelligence operations can have devastating, even existential effects on the targeted nation's security.
-Arguably the most traditional form of espionage is stealing military secrets from a foreign power, and in that arena China has been a relentless combatant.
-A United States Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} \enquote{Long-term, US future military advantage is being diminished by years of IP exfiltration from the DoD, DoN, and DIB,} the report read, \enquote{all with little to no adverse consequences to the thieves.}\footcite[p.~6]{bayer_cybersecurity_2019} The forecasts were dire: \enquote{If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.}\footcite[p.~5]{bayer_cybersecurity_2019} The US Navy is not a disinterested party, of course, and there are reasons to take its conclusions with a grain of salt; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see \cite{gilli_why_2019}} Nonetheless, espionage presents alarming opportunities for rival states and the consequences for taking advantage of them remain minimal.
+Not only are the risks associated with practicing espionage minimal, espionage norms limit the acceptable repercussions to ones that cannot possibly threaten the structure of civilian intelligence services.
+Were it more common to respond to espionage by imposing financial sanctions or even inflicting a military cost, a state would have to consider the possibility that attempting espionage could adversely impact other aspects of its security, leaving it worse off than when it started.
+States often need to make these calculations in other high-stakes arenas: pursuing a nuclear weapons program, for instance, or pursuing protectionist trade policies, could lead to severe economic retaliation.
+Instead, when an especially damaging espionage operation is uncovered, the traditional punishment is to declare certain embassy officials \emph{persona non grata}, ejecting them from the country.
+This action often results in an equal and opposite retaliation: \enquote{tit-for-tat} diplomatic expulsions.
+Both states will expel a portion of the other's embassy staff that they suspect of being spies.
+Losing intelligence officers in an embassy temporarily makes it more difficult to conduct espionage, but it doesn't make a state less likely to try.
+According to John Sipher, former head of the CIA's Russia operations, Western expulsions have never crippled Russian intelligence collection.
+\enquote{If you have 150-200 intelligence officers in-country, losing 50 is painful, but hardly debilitating.} \footcite{dettmer_united_2018}
 
-Espionage has potentially disastrous consequences but the steps that might actually curb it are out of the question.
-The system of repercussions is set up so that the structure of civilian intelligence services is never fundamentally threatened.
-When an especially damaging espionage operation is uncovered, the traditional punishment is to declare certain embassy officials \emph{persona non grata}, which often results in \enquote{tit-for-tat} diplomatic expulsions.
-These expulsions make it more difficult for the affected state to conduct intelligence operations in the short-term, but they don't fundamentally punish the state for having attempted espionage in the first place.
+If removing a quarter of the officers didn't do lasting damage, then why not remove the full 200?
+The swift expulsion of all suspected intelligence officers from a country---and the inevitable removal of the country's own intelligence officers in retaliation---would mortally injure both their espionage capabilities indefinitely.
+This nuclear option is always available to both parties and yet, in modern history, it has never been exercised.
+We can reasonably infer that each nation's leadership is making an identical calculation: the benefits of continuing to practice espionage in my adversary's backyard outweigh the risks associated with allowing my adversary to do the same to me.
+The question remains how they can both be correct.
 
-The name \enquote{tit-for-tat} is taken from a highly-successful strategy for the prisoner's dilemma wherein the player simply mimics the choice that that their opponent made in the last round.
-By employing the tit-for-tat strategy, the player never gets taken advantage of for more than one turn and can always return to cooperation if their opponent unilaterally signals a willingness to do the same.
+\subsection{A model for maintaining espionage norms}
+To describe how states cooperate to preserve the liminal legal status of espionage, I propose an iterated decision-making model based on the same tit-for-tat strategy associated with diplomatic expulsions.
+In game theory, tit-for-tat is highly-successful strategy wherein the player simply mimics the choice that that their opponent made in the last round.
+By employing this strategy, the player never gets taken advantage of for more than one turn and can always return to cooperation if their opponent unilaterally signals a willingness to do the same.
 Diplomatic expulsions play out in exactly the same fashion: when one country expels the diplomats of another, the other country will often respond in kind.
 Each side will continue to expel each other's diplomats until one side signals that it would like for the reciprocal removal of diplomats to stop.
-Inevitably, both sides return to a cooperative state while they each still have \emph{some} diplomats left.
+
+% Inevitably, both sides return to a cooperative state while they each still have \emph{some} diplomats left.
 Because neither side can gain an a decisive advantage, the back-and-forth punishment never ends in a mutual defection.
 
-\subsection{Espionage norms as a discrete series of choices}
-The key feature of \enquote{tit-for-tat} diplomatic expulsions is that the scope of the punishment is structurally incapable of discouraging a future, identical operation.
-Losing intelligence officers in an embassy temporarily makes it more difficult to conduct espionage, but it doesn't make a state less likely to try; the worst that can happen is a net-neutral result---going back to \enquote{square one,} so to speak.
-By contrast, were it more common to respond to espionage by imposing financial sanctions or even inflicting a military cost, a state would have to consider the possibility that attempting espionage could adversely impact other aspects of its security, leaving it worse off than when it started.
-These are the sorts of calculations that states need to make in other high-stakes arenas.
-Pursuing a nuclear weapons program, for instance, could lead to severe economic retaliation, as could protectionist trade policies.
-By comparison, engaging in espionage has very little potential for negative externalities because espionage norms take the possibility of a severe response off the table.
+My model changes the parameters so that the decision to cooperate or defect is based not in the amount of punishment received, but rather the type.
+Each time a state is forced to respond to a successful intelligence operation by a rival---such as the reveal of a high-level spy or a major data breach---it can either limit its response to one which is traditionally associated with espionage, or it can impose unprecedented costs that might discourage such an operation in the future.
+For instance, it might choose to impose strict financial sanctions or expel every suspected intelligence officer from its embassy, both of which would be considered punishments well in excess of the norm.
+Table \ref{espionage-decision-matrix} describes a single instance of this choice.
+In the status quo, State 1 signals its interest in cooperating by limiting its retaliation in response to espionage, no matter how damaging the operation is to their national security.
+State 2 responds by doing the same.
+Each iteration is a chance to play the game with fresh information; each espionage case is an opportunity to cooperate, reaffirming the norm that states always respond to espionage in a specific way, or defect, adding an additional cost to the attempt.
 
@@
 \begin{tabular}{cc|c|c|}
-  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{Country 2}\\
+  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{State 2}\\
   & \multicolumn{1}{c}{} & \multicolumn{1}{c}{Tolerate Espionage}  & \multicolumn{1}{c}{Punish Espionage} \\\cline{3-4}
-  \multirow{3}*{Country 1}  & Tolerate Espionage & \makecell{~\\Both sides gain intelligence \\~} & \makecell{~\\ Only the USSR continues spying \\ ~} \\\cline{3-4}
-  & Punish Espionage & \makecell{~\\ Only the US continues spying \\~} & \makecell{~\\ No intelligence either direction \\~} \\\cline{3-4}
+  \multirow{3}*{State 1}  & Tolerate Espionage & \makecell{~\\Both sides gain intelligence \\~} & \makecell{~\\ Only State 2 continues spying \\ ~} \\\cline{3-4}
+  & Punish Espionage & \makecell{~\\ Only State 1 continues spying \\~} & \makecell{~\\ No intelligence either direction \\~} \\\cline{3-4}
 \end{tabular}
 \caption{Espionage consequences decision matrix}
-\label{espionage-matrix}
+\label{espionage-decision-matrix}
 \end{table}
 
-% Possibly swap this with the previous paragraph
-
-To describe how states cooperate to preserve the liminal legal status of espionage, I propose an iterated decision-making model.
-My model changes the parameters so that the decision to cooperate or defect is based not in the amount of punishment received, but rather the type.
-Each time a state is forced to respond to a successful intelligence operation by a rival---such as the reveal of a high-level spy or a major data breach---it can either limit its response to one which is traditionally associated with espionage, or it can impose unprecedented costs that might discourage such an operation in the future.
-Table \ref{espionage-matrix} describes a single instance of this choice.
-In the status quo, State 1 signals its interest in cooperating by limiting its retaliation in response to espionage, no matter how damaging the operation is to their national security.
-State 2 responds by doing the same.
-Each iteration is a chance to play the game with fresh information; each espionage case is an opportunity to cooperate, reaffirming the norm that states always respond to espionage in a specific way, or defect, adding an additional cost to the attempt.
-
-Through this model two important things become clear.
-First, even though either state has the opportunity to defect by imposing harsher punishments in response to espionage, its gains would be short-lived because the other state is guaranteed to immediately do the same---tit-for-tat.
-Espionage consequences are an iterated game in which both parties play the exact same way, so, in the long run, the only two possible outcomes are mutual defection or mutual cooperation.
+While there is unquestionably a cultural norm informing espionage norms, tradition cannot fully explain why every state remains in a state of mutual cooperation.
+Espionage consequences are an iterated game that both parties play tit-for-tat, so, in the long run, the only two possible equilibria are mutual defection or mutual cooperation.
 In the realm of traditional espionage, we've only seen states engage in mutual cooperation, but any state has the power to change that by defecting.
@@
 The relatively open flow of information is only possible with a universal buy-in.
+Stripped of its mystique, the benefits of allowing intelligence operations to continue are, like any other state decision, a simple cost-benefit analysis; states decide that it best serves their interests to remain in the top left corner, rather than the bottom right.
+We can expect that there is \emph{some} rational explanation that political leaders employ to justify it, where they weigh the benefits of preserving their ability to spy against the damage that results from espionage conducted against them.
 
-Second, while there is unquestionably a cultural norm informing the response to intelligence operations, that norm is doing a lot of work to maintain the status quo.
-So much work, in fact, that tradition could not possibly explain the entire phenomenon.
-Stripped of its mystique, the benefits of allowing intelligence operations to continue are, like any other state decision, a simple cost-benefit analysis; states decide that it best serves their interests to remain in the top left corner, rather than the bottom right.
-We can expect then that there is \emph{some} rational explanation that political leaders employ to justify it, where they weigh the benefits of preserving their ability to spy against the damage that results from espionage conducted against them.
+% First, even though either state has the opportunity to defect by imposing harsher punishments in response to espionage, its gains would be short-lived because the other state is guaranteed to immediately do the same---tit-for-tat.
 
@@
 
-The surprisingly straightforward conclusion is that, if only two worlds are possible---one with widespread espionage and one without---then all the relevant nations must be concluding that the world without espionage is incredibly bad---so bad that even if one nation correctly concludes that its rival gains more from a world with a free flow of information than it does, the alternative, a world with significantly harsher punishments for espionage and a staunched flow of intelligence, still leaves that nation worse off.
+The surprisingly straightforward conclusion is that, if only two worlds are possible---one with widespread espionage and one without---then all the relevant nations must be concluding that the world without espionage is incredibly bad.
+Even if one nation correctly concludes that its rival gains more from a world with a free flow of information than it does, the alternative, a world with significantly harsher punishments for espionage and a staunched flow of intelligence, still leaves that nation worse off.
 That calculation has a name: defensive realism.
@@
 To explain why states might be willing to accept relative losses in security, I employ Charles Glaser's \emph{Rational Theory of International Politics}.
-Glaser describes his theory as integrating defensive realism with neoclassical realism.\footnote{With apologies to Mr.
-Glaser, I am going to refer to it as defensive realism, because they both put the security dilemma at the center of their explanation for international competition, which is the key observation for me.
+Glaser describes his theory as integrating defensive realism with neoclassical realism.\footnote{For the sake of simplicity, I am going to refer to it as defensive realism, because they both put the security dilemma at the center of their explanation for international competition and cooperation, which I believe to be the key observation.
 He would have you know that defensive realism \enquote{can be viewed as a way station along the route to the full theory} that he develops, which is \enquote{significantly more general and complete than is defensive realism.} (p.
@@
 Consider a hypothetical security seeker interacting with its adversary, which it believes to be a security-seeking state as well.
-The first state feels insecure, so it builds up arms to increase its security.
-The rival state, seeing that its adversary is building up arms, now feels less secure and builds up its own arsenal in response, and a classic security dilemma ensues.
-Even though both states were just security seekers, they're now both much less secure.
+If the first state feels insecure, it will build up arms to increase its security.
+The rival state, seeing that its adversary is building up arms, now feels less secure and builds up its own arsenal in response, a classic security dilemma.
 
-% If both states are genuine security-seekers, then this situation is clearly sub-optimal, because now both states are heavily militarized and comparatively much less secure.
+Like espionage norms, competition between security-seeking states has two natural equilibria: mutual cooperation and mutual defection.
+The ideal outcome for both is when neither state builds up its armaments---with their reciprocal lack of ability to deal damage, both states are confident that no war will result.
+Unilateral build-up leads to short-term relative gains in security, but in the next turn the other state will doubtless respond with a defection.
+The resulting mutual defection is worse than a mutual cooperation.
+There might be peace, but it will be an uneasy peace; both states are heavily militarized and comparatively much less secure.
+When written out in the form of a payoff matrix, this game-theoretic choice is known as a Stag Hunt.\footnote{Just like the "Prisoner's Dilemma" takes its name from a hypothetical choice to snitch on a fellow prisoner, the "Stag Hunt" simulates a two-person hunting party.
+If both hunters cooperate they can take down a stag, but each also has the choice of hunting a hare on their own.
+If one person hunts a hare while the other hunts a stag, the hunter who went after the stag gets nothing.
+The hare is a smaller payoff, but doesn't rely on the other hunter's cooperation.} Mutual cooperation is the preferred outcome, but achieving it relies on both parties' expectations about what the other will choose.
+The incentives for a security dilemma are actually skewed in favor of cooperation slightly more than the traditional Stag Hunt, because mutual defection leaves both states slightly worse off than a single betrayal.
+A state that builds up its arms will be rewarded with an adversary threatening the same---at least in a Stag Hunt the two distrustful hunters will each end up with a hare.
 
-Defensive realism introduces a counterintuitive observation: the key to interrupting the security dilemma is to make the adversary feel more secure, because it is the adversary's response to securitization that ultimately makes the first state insecure.
-Therefore, defensive realism posits that it is logical for a security seeker to take actions that increase not just its own security, but its \emph{adversary's} security as well.\footcite[p.~7]{glaser_rational_2010} These actions would signal to the adversary that the state is a security seeker, hoping to reach a mutually agreeable outcome.
+\begin{table}[ht]
+\centering
+\setlength{\extrarowheight}{2pt}
+\small
+\begin{tabular}{cc|c|c|}
+  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{State 2}\\
+  & \multicolumn{1}{c}{} & \multicolumn{1}{c}{Retain current arms}  & \multicolumn{1}{c}{Build up arms} \\\cline{3-4}
+  \multirow{3}*{State 1}  & Retain current arms & \makecell{~\\(5,5)\\~} & \makecell{~\\ (1,3) \\ ~} \\\cline{3-4}
+  & Build up arms & \makecell{~\\ (3,1) \\~} & \makecell{~\\ (2,2) \\~} \\\cline{3-4}
+\end{tabular}
+\caption{Defensive Realism payoff matrix}
+\label{defensive-realism-payoff-matrix}
+\end{table}
+
+To arrive at mutual cooperation, a security-seeking state needs to not only determine that its rival is a security seeker, it needs its rival to make that exact same judgment, because it is the adversary's response to securitization that ultimately makes the first state insecure.
+Defensive realism therefore introduces a counterintuitive conclusion: it is often logical for a security seeker to take actions that increase not only its own security, but its \emph{adversary's} security as well.\footcite[p.~7]{glaser_rational_2010} Such actions would make the adversary less likely to securitize, and also signal to the adversary that the state is a security seeker hoping to reach a mutually agreeable outcome.
 If the adversary is indeed a security seeker, then, no longer concerned about its security, it will respond in kind and resolve the security dilemma for both of them.
+The problem for a defensive realist is that unilaterally taking steps to increase the adversary's security is obviously quite risky.
+The state could be wrong that its adversary is a security seeker---what Glaser calls a \enquote{greedy state}---in which case its optimistic overtures of cooperation will have exposed it to greater risk of attack.
+When states are unwilling to make those signaling measures because they fear compromising their own security, then a security dilemma results.
 
-The problem for a defensive realist is that unilaterally taking steps to increase the adversary's security is obviously quite risky.
-The state could be wrong that its adversary is a security seeker, in which case its optimistic overtures of cooperation will have exposed it to greater risk of attack.
-When states are unwilling to make those signaling measures because they fear compromising their own security, then a security dilemma results.
-This is where intelligence comes into play.
-
-Prior to making any decisions, states assess the security situation and determine which options are most likely to increase their own security.
-Glaser identifies two types of variables that mediate the magnitude of a security dilemma.
-The first are material variables, such as offense-defense balance and offense-defense distinguishability.
-Espionage plays a significant role in properly assessing material variables; during the Cold War, for example, the United States placed a premium on generating quality estimates of how many bombers, missiles, and bases the Soviet Union had deployed.
-By knowing which of its adversary's capabilities are offensive, a state can better target its efforts towards reducing them.
-
-In addition to material variables, Glaser notes that information variables play an important and comparatively under-studied role in the security dilemma.
-Better information about its adversary's motives can give the state reason to trust that risky signaling measures might succeed---or fail.
-Even if the state is uncertain, espionage can provide some information about its adversary's intent, which, in turn, can moderate the security dilemma enough to make cooperation possible.
-In the best case-scenario, the security dilemma is simply eliminated if both states are confident that the other is a security seeker.
-This is the optimal outcome for two genuine security-seekers, and it is made much more likely from the intelligence that espionage can provide.
-
+Espionage creates the conditions that make solving the security dilemma possible.
 Consider a situation in which one state is a genuine security-seeker and so is its adversary.
@@
 Because the state receiving the signal is also a security-seeker, this is a welcome development; it could be empowered to signal its benign intentions as well, based on its own intelligence.
-For both states, this is the optimal outcome.
-The counterintuitive observation here is that both states benefited not just from spying, but from being spied \emph{upon}; their adversary's intelligence made them more likely to make a risky signal for deescalation.
+For both states, this is the optimal outcome, contributing to both the information variables that Glaser identifies as key to the security dilemma: \enquote{the state’s estimate of the adversary’s motives, and the state’s estimate of the adversary’s estimate of the state’s motives.}\footcite[p.~34-35]{glaser_rational_2010} Both states therefore benefit not just from spying, but from being spied \emph{upon}; their adversary's intelligence made that adversary more likely to propose or accept a risky signal for deescalation.
 
-% States now recognize that some forms of espionage facilitate international cooperation.\footcite{baker_tolerance_2004}
+States take the risk that espionage will reveal more than just a state's (hopefully benign) intent---it might also reveal a state's capabilities and strategy.
+According to senior Obama Administration official Jake Sullivan, the United States would have no problem with adversaries being able to independently verify American intent since, as he sees it, American intent is benign.
+Capabilities, however, are more complicated.
+Some defense capabilities would probably be fine to have another country independently verify, and some the US might want to keep secret because their secrecy confers a strategic advantage.
+And under almost no circumstances would a state want to have its strategies and warfighting plans revealed, so those should be kept as secret as possible.\footcite{sullivan_personal_2019} States would prefer to allow only the \enquote{good} parts of espionage, but they can't---the credibility of the information obtained depends on the fact that the espionage subverted its target's access controls.
+If you knew your target had let you find something, you wouldn't trust it.
 
-A complete theoretical justification for espionage---that maintaining a mutual exchange of illicit information is better for both states than the alternative---can be constructed by combining the work of previous scholars, and finding archival evidence to support their claims.
-Christopher Baker explains how espionage makes functional cooperation possible through treaties, allowing states to verify each others' compliance.\footcite[p.~6]{baker_tolerance_2004} Charles Glaser uses the 1972 Anti-Ballistic missile treaty as an example of precisely the type of signal that could lead states to resolve the security dilemma, but he does not acknowledge that the intelligence operations of both states are what make signals like it possible.\footnote{It is entirely possible that Glaser himself realizes this, it's just not written down; the purpose of the book is to defend the theory, not explain the mechanics.} And Julius Stone, a professor of international law, foresaw this exact role for espionage in 1962, well before the public had any real idea about the US spy satellite program: \enquote{If you do not have a system of international inspection and if you can't get one (and I think it is quite likely that we can't) then the function which international inspection is supposed to serve still needs fulfilling.
-You may not be able to fulfill it to the optimum extent by reciprocally tolerated espionage, but you may be able to reduce the dangers.}\footcite[p.~41]{stone_legal_1962}
+\begin{table}[ht]
+\centering
+\setlength{\extrarowheight}{2pt}
+\small
+\begin{tabular}{cc|c|c|}
+  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{State 2}\\
+  & \multicolumn{1}{c}{} & \multicolumn{1}{c}{Tolerate Espionage}  & \multicolumn{1}{c}{Punish Espionage} \\\cline{3-4}
+  \multirow{3}*{State 1}  & Tolerate Espionage & \makecell{~\\(5,5) \\~} & \makecell{~\\ (0,4) \\ ~} \\\cline{3-4}
+  & Punish Espionage & \makecell{~\\ (4,0) \\~} & \makecell{~\\ (2,2) \\~} \\\cline{3-4}
+\end{tabular}
+\caption{Espionage payoff matrix for two security-seeking states}
+\label{espionage-payoff-matrix}
+\end{table}
 
-In 1962, Stone defended espionage from a practical perspective in exactly the same terms that I am now defending it from a theoretical one.
-Permitting at least some kinds of espionage is actually beneficial for both sides because it allows us to independently verify the intent and capabilities of other nations.
+Therefore, states face a binary choice whether to tolerate espionage or crack down.
+We can translate that conclusion to a payoff matrix (for two security-seeking states) that looks almost exactly like the  security dilemma Stag Hunt: the greatest rewards for both are when they cooperate.
+Non-reciprocal espionage would generate significant relative advantages, but wouldn't confer the benefit of easing an adversary's fears.
+And the worst situation for both is when neither allows access to each others' information.
+Yes, that equilibrium minimizes the risk of losing of military secrets, but it significantly ratchets up the potential for conflict and miscalculation.
+
+%%% %%%
+% A second thing that is a little bit confusing to me about the theory is whether the theory is intelligence revealing information about intentions, or capabilities, or about the other actors "type."  Most of the examples you have in the empirical section seem to be really about gathering intelligence on capabilities rather than intentions.
+When you get to the cases it's never clear whether the US is discovering that the Soviet Union is truly a security seeker and hence promises escaping from the security dilemma, or whether the United States is actually gaining intelligence on Soviet capabilities that might, at the extreme, be used aggressively to “win” in a war with the Soviet Union.
+It is unclear from your theory why actors would tolerate this espionage if it is unclear whether it is in fact making both secure or actually is making one side better able to plan an offensive operation against the other.
+%%% %%%
+
+\subsection{What about greed?}
+The analysis in this section relies on one crucial assumption: that the states in question are security-seekers.
+Analyses of the risks of espionage must contend with possibility that one of the states in question is \enquote{fundamentally dissatisfied with the status quo, desiring additional territory even when it is not required for security}---a greedy state.\footcite[p.~4]{glaser_rational_2010} Such a state would obviously want to protect both its motives and its plans, and therefore derive no benefit from having espionage conducted against it.
+Nonetheless, we know that essentially every major nation-state abides by the accepted norms of espionagem and I have already addressed the possibility that states---including greedy states---are bound to these norms for reasons that extend beyond rational calculus, and do not consider that explanation to be sufficient.
+To fully evaluate the implications of greedy states on this theory of mutual cooperation would require an entirely different paper, but I will propose potential answers to the question of why greedy states would tolerate espionage as a starting point for future analysis.
+
+Perhaps greedy states simply do not exist.
+This is a very provocative assertion---one that is, in its most stringent sense, almost certainly wrong---but it is worth taking seriously the possibility that no states which abide by espionage norms are greedy \emph{to a destabilizing degree.} The concept of a greedy state is one that naturally operates on a spectrum.\footcite[p.~39.
+Throughout his book, Glaser simplifies and complicates his theory as is necessary for the particular analysis he is performing.
+Greedy states can be understood as the opposite of security-seeking states, or it can be its own independant variable, allowing for the possibility of states that are greedy \emph{and} security-seeking, or just one of the two.]{glaser_rational_2010} The quintessential greedy state, the \enquote{greediest} state, is 1937 Germany, which was planning to dominate the European continent.
+In 2020, maybe no states are worried about espionage revealing that they intend to launch a significant campaign of conquest because there are no states intending to launch a significant campaign of conquest.
+To use a modern example, the CIA warned that Russia might try to move into the Crimean peninsula shortly before it happened in 2014.\footcite{hosenball_ukraine_2014} Russia practiced good operational security to protect its intent from espionage, but it still took a risk that its operation would be discovered and counteracted.
+If in fact the annexation of Crimea is the act of a greedy state, that level of greed is likely tolerable to Russia's adversaries in a way that a nuclear first-strike operation would not.
+
+The international community is likely to view a sudden withdrawal from international espionage norms as suspicious.
+States that are farther along on the greed spectrum might decide that it is better to risk the chance that they'll be discovered as greedy states rather send a signal that will all but confirm it.
+And it cannot be discounted that sometimes states might not even be entirely confident of whether its \emph{own} intent is greedy.
+So far this paper has used the nation-state as its base level of analysis.
+As I present the empirical evidence for this theory however, it becomes necessary to analyze the motives of individual actors that comprise the state.
+At this level, even officials within the same administration have conflicting ideas about how to use intelligence, and those collecting the intelligence can't be confident whose purposes they are serving.
+
+% States that allow espionage are operating with faith that a freer flow of information is going to lead to a better outcome.
+
+% because it allows us to independently verify the intent and capabilities of other nations.
 Doing so actually reduces the likelihood that miscalculation will lead to open hostilities.
@@
 \subsection{Introducing the empirical examples}
+Previous intelligence scholars have noted specific benefits of espionage that support the abstract theoretical justification I build here.
+Christopher Baker explains how espionage makes functional cooperation possible through treaties.\footcite[p.~6]{baker_tolerance_2004} Charles Glaser uses the 1972 Anti-Ballistic missile treaty as an example of precisely the type of signal that could lead states to resolve the security dilemma Julius Stone, a professor of international law, foresaw this exact role for espionage in 1962: \enquote{If you do not have a system of international inspection and if you can't get one (and I think it is quite likely that we can't) then the function which international inspection is supposed to serve still needs fulfilling.
+You may not be able to fulfill it to the optimum extent by reciprocally tolerated espionage, but you may be able to reduce the dangers.}\footcite[p.~41]{stone_legal_1962}
 
-The best way to prove that this theory has explanatory power for state decision-making is to find historical examples where states made decisions that showed they were willing to accept the cost of enemy espionage in order to practice espionage themselves.
-In the ideal scenario, we would find evidence of policymakers justifying their decisions based on the value of mutual observation.
+The best way to demonstrate that this theory has explanatory power for state decision-making is to find historical examples where states made decisions that showed they were willing to accept the cost of enemy espionage in order to practice espionage themselves.
+If the theory is correct, we would find evidence of policymakers justifying their decisions based on the value of mutual observation.
 Such decisions were made frequently during the Cold War, and with the explicit intent to increase the bilateral flow of information to create a more secure world environment.
@@
 
+\newpage
 \section{Aerial espionage}

### 2020-06-02 article.tex
Made edits from dad to the first half

@@
 
-The US government's tepid response to the OPM hack illustrates a crucial fact about the relationship between cybersecurity and espionage: when a cyberattack looks like a traditional intelligence operation, it merits little to no diplomatic consequences.
+The US government's tepid response to the OPM hack illustrates a crucial fact about the relationship between cybersecurity and espionage: when a cyberattack looks like a traditional intelligence operation, little to no diplomatic consequences will follow.
 Within both political and academic circles, this phenomenon is often rationalized as a logical extension of existing policy; cyber-espionage is treated like \enquote{traditional} espionage because they are both forms of espionage.
@@
 There are few extant definitions of peacetime espionage, and no satisfying ones.
-The problem that aspiring espionage scholars inevitably encounter is that \enquote{peacetime espionage} is universally understood as a term of art but is almost entirely missing from any legal codes or case law.
+The problem that aspiring espionage scholars inevitably encounter is that \enquote{peacetime espionage} is universally understood as a term of art but almost never defined by legal codes or interpreted by case law.
 Most scholars who tackle this question settle on a definition like the one proposed by I\~{n}aki Navarrete and Russell Buchan: \enquote{\enquote{Peacetime espionage} is a colloquial rather than a legal term [used] to describe different methods of collecting confidential information from closed as opposed to open sources \textelp{} generally conducted by States, against States.}\footcite[p.~901-902]{navarrete_out_2019} This definition purposefully excludes forms of extralegal government intervention that are often conflated with espionage, such as assassinations, election interference, and covert operations.\footnote{Often papers won't take a stand on defining espionage, but rather offer a working definition just to bound the scope of that particular paper.
@@
 
-For the purposes of this paper, I extend that definition to describe espionage \emph{norms} as the common understanding of how states are expected to respond to activities that remain within the bounds of espionage.\footnote{There are, of course, other \enquote{norms} that involve espionage.
+For the purposes of this paper, I define espionage \emph{norms} as the common understanding of how states are expected to respond to activities that remain within the bounds of espionage.\footnote{There are, of course, other \enquote{norms} that involve espionage.
 The so-called \enquote{Moscow Rules} dictate an intelligence officer are supposed to conduct themselves in hostile territory.
 In order to avoid constantly specifying, espionage norms are more strictly defined to include only the ones that affect IR.} Espionage norms prescribe that when one state uncovers an espionage operation conducted against it, the consequences imposed on the spying country remain minimal.
-Rarely does the response amount to more than a pat denunciation of the activity and implementation of the necessary countermeasures to end it---the consequences for espionage are borne by the agents who commit it, not the state that sent them.
+Rarely does the response amount to more than a \emph{pro forma} denunciation of the activity and implementation of the necessary countermeasures to end it---the consequences for espionage are borne by the agents who commit it, not the state that sent them.
 In rare cases when additional costs \emph{are} imposed on a state for conducting a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.
@@
 Critically, the research that follows will demonstrate that they were not obligated to do so by any pre-existing norms.
-When the US first introduced photographic satellites, for example, it was not clear whether the Soviet Union would treat the satellites as spies or as weapons---there were indications in both directions.
+When the US first introduced photoreconnaissance satellites, for example, it was not clear whether the Soviet Union would treat the satellites as spies or as weapons.
 Ultimately, the infrastructure of what we now call \enquote{spy satellites} was developed, and the norms of espionage prevailed to protect their use.
@@
 \subsection{The nuclear option}
-
 % Add a poll about American knowledge of the CIA?
@@
 American citizens, as well as their international counterparts, are broadly aware that their nations conduct extensive intelligence operations abroad.
-The Central Intelligence Agency (CIA) is funded by the United States Congress, reports directly to the President, and offers internship positions for college students.
+The Central Intelligence Agency (CIA) is funded by the United States Congress and reports directly to the President---it even promises the possibility of a career in espionage by offering internship positions for college students.
 Keeping these institutions intact is partly a function of inertia, but it is also an ongoing argument that espionage is a legitimate, even critical, component of statecraft.
-
-% Scholars can use rationalist principles to analyze everything from trade policy to the space race; espionage, which by definition involves deeply invasive interactions with another state, ought to be no different.
 
@@
 In recent years China has relentlessly targeted American military secrets.
-A United States Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} \enquote{Long-term, US future military advantage is being diminished by years of IP exfiltration from the DoD, DoN, and DIB,} the report read, \enquote{all with little to no adverse consequences to the thieves.}\footcite[p.~6]{bayer_cybersecurity_2019} The forecasts were dire: \enquote{If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.}\footcite[p.~5]{bayer_cybersecurity_2019} The US Navy is not a disinterested party, of course, and we should be cautious about its conclusions; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see: \cite{gilli_why_2019}} Nonetheless, espionage presents alarming opportunities for rival states to gain a comparative advantage and the consequences for taking advantage of them remain minimal.
+A United States Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} \enquote{Long-term, US future military advantage is being diminished by years of IP exfiltration from [US military agencies],} the report read, \enquote{all with little to no adverse consequences to the thieves.}\footcite[p.~6]{bayer_cybersecurity_2019} The forecasts were dire: \enquote{If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.}\footcite[p.~5]{bayer_cybersecurity_2019} The US Navy is not a disinterested party, of course, and we should be cautious about its conclusions; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see: \cite{gilli_why_2019}} Nonetheless, espionage presents alarming opportunities for rival states to gain a comparative advantage and the consequences for taking advantage of them remain minimal.
 
@@
 If removing a quarter of the officers didn't do lasting damage, then why not remove the full 200?
-The swift expulsion of all suspected intelligence officers from a country---and the inevitable removal of the country's own intelligence officers in retaliation---would mortally injure both their espionage capabilities indefinitely.
-This nuclear option is always available to both parties and yet, in modern history, it has never been exercised.
+The swift expulsion of all suspected intelligence officers from a country---and the inevitable removal of the country's own intelligence officers in retaliation---would deeply injure both their espionage capabilities indefinitely.
+This nuclear option is always available to both of parties and yet, in modern history, it has never been exercised.
 We can reasonably infer that each nation's leadership is making an identical calculation: the benefits of continuing to practice espionage in my adversary's backyard outweigh the risks associated with allowing my adversary to do the same to me.
@@
 \subsection{A model for maintaining espionage norms}
-To describe how states cooperate to preserve the liminal legal status of espionage, I propose an iterated decision-making model based on the same tit-for-tat strategy associated with diplomatic expulsions.
+I propose an iterated decision-making model to describe how states cooperate to preserve the liminal legal status of espionage, based on the same tit-for-tat strategy associated with diplomatic expulsions.
 In game theory, tit-for-tat is highly-successful strategy wherein the player simply mimics the choice that that their opponent made in the last round.
@@
 
-My model changes the parameters so that the decision to cooperate or defect is based not in the amount of punishment received, but rather the type.
+My model changes the parameters so that the decision to cooperate or defect is based not on the amount of punishment received, but rather the type.
 Each time a state is forced to respond to a successful intelligence operation by a rival---such as the reveal of a high-level spy or a major data breach---it can either limit its response to one which is traditionally associated with espionage, or it can impose unprecedented costs that might discourage such an operation in the future.
@@
 The relatively open flow of information is only possible with a universal buy-in.
-Stripped of its mystique, the benefits of allowing intelligence operations to continue are, like any other state decision, a simple cost-benefit analysis; states decide that it best serves their interests to remain in the top left corner, rather than the bottom right.
+Stripped of its mystique, the benefits of allowing intelligence operations to continue are, like any other state decision, a simple cost-benefit analysis; As illustrated above, states decide that it best serves their interests to remain in the top left corner, rather than the bottom right.
 We can expect that there is \emph{some} rational explanation that political leaders employ to justify it, where they weigh the benefits of preserving their ability to spy against the damage that results from espionage conducted against them.
@@
 Every policymaker knows exactly how crucial espionage is to their job.
-With years of experience making difficult policy calls informed by intelligence, it is certainly much easier to imagine how one's decisions will be made worse without espionage than it is to imagine how one's enemies' decisions will be made better.
+With years of experience making difficult policy choices informed by intelligence, it is certainly much easier to imagine how one's decisions will be made worse without espionage than it is to imagine how one's enemies' decisions will be made better.
 It follows that there would be a strong psychological bias towards attempting to discover as much information as possible.
@@
 
-The surprisingly straightforward conclusion is that, if only two worlds are possible---one with widespread espionage and one without---then all the relevant nations must be concluding that the world without espionage is incredibly bad.
-Even if one nation correctly concludes that its rival gains more from a world with a free flow of information than it does, the alternative, a world with significantly harsher punishments for espionage and a staunched flow of intelligence, still leaves that nation worse off.
-That calculation has a name: defensive realism.
+The surprisingly straightforward conclusion is that if only two worlds are possible---one with widespread espionage and one without---then all the relevant nations must be concluding that the world without espionage is completely unacceptable.
+Even if one nation correctly concludes that its rival gains more from a world with a free flow of information than it does, the alternative---a world with significantly harsher punishments for espionage and a staunched flow of intelligence---still leaves the former worse off.
 
 \subsection{Defensive realism}
-To explain why states might be willing to accept relative losses in security, I employ Charles Glaser's \emph{Rational Theory of International Politics}.
+To explain why states might be willing to accept relative losses in security, I employ a theory called defensive realism, particularly the version formulated in Charles Glaser's \emph{Rational Theory of International Politics}.
 Glaser describes his theory as integrating defensive realism with neoclassical realism.\footnote{For the sake of simplicity, I am going to refer to it as defensive realism, because they both put the security dilemma at the center of their explanation for international competition and cooperation, which I believe to be the key observation.
@@
 Unilateral build-up leads to short-term relative gains in security, but in the next turn the other state will doubtless respond with a defection.
-The resulting mutual defection is worse than a mutual cooperation.
+The resulting mutual defection is worse than mutual cooperation.
 There might be peace, but it will be an uneasy peace; both states are heavily militarized and comparatively much less secure.
-When written out in the form of a payoff matrix, this game-theoretic choice is known as a Stag Hunt.\footnote{Just like the "Prisoner's Dilemma" takes its name from a hypothetical choice to snitch on a fellow prisoner, the "Stag Hunt" simulates a two-person hunting party.
+When examined in the form of a payoff matrix, this game-theoretic choice is known as a Stag Hunt.\footnote{Just like the "Prisoner's Dilemma" takes its name from a hypothetical choice to snitch on a fellow prisoner, the "Stag Hunt" simulates a two-person hunting party.
 If both hunters cooperate they can take down a stag, but each also has the choice of hunting a hare on their own.
@@
 To arrive at mutual cooperation, a security-seeking state needs to not only determine that its rival is a security seeker, it needs its rival to make that exact same judgment, because it is the adversary's response to securitization that ultimately makes the first state insecure.
-Defensive realism therefore introduces a counterintuitive conclusion: it is often logical for a security seeker to take actions that increase not only its own security, but its \emph{adversary's} security as well.\footcite[p.~7]{glaser_rational_2010} Such actions would make the adversary less likely to securitize, and also signal to the adversary that the state is a security seeker hoping to reach a mutually agreeable outcome.
+Defensive realism therefore introduces a counterintuitive conclusion: it is often logical for a security seeker to take actions that increase not only its own security, but its \emph{adversary's} security as well.\footcite[p.~7]{glaser_rational_2010} Such actions would make the adversary less likely to securitize, and also signal to the adversary that it is a security seeker hoping to reach a mutually agreeable outcome.
 If the adversary is indeed a security seeker, then, no longer concerned about its security, it will respond in kind and resolve the security dilemma for both of them.
@@
 Because the state receiving the signal is also a security-seeker, this is a welcome development; it could be empowered to signal its benign intentions as well, based on its own intelligence.
-For both states, this is the optimal outcome, contributing to both the information variables that Glaser identifies as key to the security dilemma: \enquote{the state’s estimate of the adversary’s motives, and the state’s estimate of the adversary’s estimate of the state’s motives.}\footcite[p.~34-35]{glaser_rational_2010} Both states therefore benefit not just from spying, but from being spied \emph{upon}; their adversary's intelligence made that adversary more likely to propose or accept a risky signal for deescalation.
+For both states, this is the optimal outcome, contributing to both information variables that Glaser identifies as key to the security dilemma: \enquote{the state’s estimate of the adversary’s motives, and the state’s estimate of the adversary’s estimate of the state’s motives.}\footcite[p.~34-35]{glaser_rational_2010} Both states therefore benefit not just from spying, but from being spied \emph{upon}; their adversary's intelligence made that adversary more likely to propose or accept a risky signal for deescalation.
 
 States take the risk that espionage will reveal more than just a state's (hopefully benign) intent---it might also reveal a state's capabilities and strategy.
-According to senior Obama Administration official Jake Sullivan, the United States would have no problem with adversaries being able to independently verify American intent since, as he sees it, American intent is benign.
+For example, I interviewed senior Obama administration official Jake Sullivan about the benefits of allowing foreign states to spy on the US.
+According to him, the United States would have no problem with adversaries being able to independently verify American intent because, as he sees it, American intent is benign.
 Capabilities, however, are more complicated.
-Some defense capabilities would probably be fine to have another country independently verify, and some the US might want to keep secret because their secrecy confers a strategic advantage.
-And under almost no circumstances would a state want to have its strategies and warfighting plans revealed, so those should be kept as secret as possible.\footcite{sullivan_personal_2019} States would prefer to allow only the \enquote{good} parts of espionage, but they can't---the credibility of the information obtained depends on the fact that the espionage subverted its target's access controls.
+Some defense capabilities would probably be fine to have an adversary independently verify, and some the US might want to keep secret because their secrecy confers a strategic advantage.
+And under almost no circumstances would the US want to have its strategies and warfighting plans revealed, so every measure should be taken to ensure their secrecy.\footcite{sullivan_personal_2019} States would prefer to reveal only the \enquote{good} information, but they can't---the credibility of the information obtained depends on the fact that the espionage subverted its target's access controls.
 If you knew your target had let you find something, you wouldn't trust it.
@@
 
-Therefore, states face a binary choice whether to tolerate espionage or crack down.
+Therefore, states face a binary choice whether to tolerate espionage or punish it.
 We can translate that conclusion to a payoff matrix (for two security-seeking states) that looks almost exactly like the  security dilemma Stag Hunt: the greatest rewards for both are when they cooperate.
 Non-reciprocal espionage would generate significant relative advantages, but wouldn't confer the benefit of easing an adversary's fears.
-And the worst situation for both is when neither allows access to each others' information.
-Yes, that equilibrium minimizes the risk of losing of military secrets, but it significantly ratchets up the potential for conflict and miscalculation.
+And the worst situation for both is when neither allows access to the other's information.
+Yes, that equilibrium minimizes the risk of losing of military secrets, but it significantly heightens the potential for conflict and miscalculation.
 
@@
 Analyses of the risks of espionage must contend with possibility that one of the states in question is \enquote{fundamentally dissatisfied with the status quo, desiring additional territory even when it is not required for security}---a greedy state.\footcite[p.~4]{glaser_rational_2010} Such a state would obviously want to protect both its motives and its plans, and therefore derive no benefit from having espionage conducted against it.
-Nonetheless, we know that essentially every major nation-state abides by the accepted norms of espionagem and I have already addressed the possibility that states---including greedy states---are bound to these norms for reasons that extend beyond rational calculus, and do not consider that explanation to be sufficient.
-To fully evaluate the implications of greedy states on this theory of mutual cooperation would require an entirely different paper, but I will propose potential answers to the question of why greedy states would tolerate espionage as a starting point for future analysis.
+Nonetheless, we know that essentially every major nation-state abides by the accepted norms of espionage and I have already addressed the possibility that states---including greedy states---are bound to these norms for reasons that extend beyond rational calculus, and do not consider that explanation to be sufficient.
+To fully evaluate the implications of greedy states on this theory of mutual cooperation is beyond the scope of this paper, but I will propose potential answers to the question of why greedy states would tolerate espionage as a starting point for future analysis.
 
 Perhaps greedy states simply do not exist.
-This is a very provocative assertion---one that is, in its most stringent sense, almost certainly wrong---but it is worth taking seriously the possibility that no states which abide by espionage norms are greedy \emph{to a destabilizing degree.} The concept of a greedy state is one that naturally operates on a spectrum.\footcite[p.~39.
+This is a very provocative assertion---one that is, in its most stringent sense, almost certainly wrong---but consider the possibility that no states which abide by espionage norms are greedy \emph{to a destabilizing degree.} The concept of a greedy state is one that naturally operates on a spectrum.\footcite[p.~39.
 Throughout his book, Glaser simplifies and complicates his theory as is necessary for the particular analysis he is performing.
-Greedy states can be understood as the opposite of security-seeking states, or it can be its own independant variable, allowing for the possibility of states that are greedy \emph{and} security-seeking, or just one of the two.]{glaser_rational_2010} The quintessential greedy state, the \enquote{greediest} state, is 1937 Germany, which was planning to dominate the European continent.
-In 2020, maybe no states are worried about espionage revealing that they intend to launch a significant campaign of conquest because there are no states intending to launch a significant campaign of conquest.
+Greedy states can be understood as the opposite of security-seeking states, or it can be its own independant variable, allowing for the possibility of states that are greedy \emph{and} security-seeking, or just one of the two.]{glaser_rational_2010} The quintessential \enquote{greediest} state, 1937 Germany, was planning to dominate the European continent.
+In 2020, if no states are worried about espionage revealing that they intend to launch a significant campaign of conquest, perhaps that is because there are no states intending to launch a significant campaign of conquest.
 To use a modern example, the CIA warned that Russia might try to move into the Crimean peninsula shortly before it happened in 2014.\footcite{hosenball_ukraine_2014} Russia practiced good operational security to protect its intent from espionage, but it still took a risk that its operation would be discovered and counteracted.
@@
 SALT I (the ABM Treaty) is widely considered to be the landmark diplomatic achievement of Nixon's \emph{d\'etente} policy, which set the stage for a broader easing of relations between the US and the USSR, and spy satellites were crucial to reducing the cost that each side would pay to sign it.
-Limiting ballistic missile defense systems was a relatively minor concession for both sides, but in the highly risk-averse context of the Cold War, the signal was significant.\footcite[p.~66]{glaser_rational_2010} Underneath the omnipresent threat of nuclear annihilation, \emph{d\'etente} provided a rational strategic choice for both states---made possible by their reciprocal tolerance of espionage.
+Limiting ballistic missile defense systems was a relatively minor concession for both sides, but in the highly risk-averse context of the Cold War, the signal was significant.\footcite[p.~66]{glaser_rational_2010} Facing the omnipresent threat of nuclear annihilation, \emph{d\'etente} provided a rational strategic choice for both states---made possible by their reciprocal tolerance of espionage.
 

### 2020-07-26 article.tex
Make Aerial Espionage and Spy Satellites subheaders of Case Studies

@@
 
-If removing a quarter of the officers didn't do lasting damage, then why not remove the full 200?
+If removing a quarter of the officers doesn't do lasting damage, then why not remove the full 200?
 The swift expulsion of all suspected intelligence officers from a country---and the inevitable removal of the country's own intelligence officers in retaliation---would deeply injure both their espionage capabilities indefinitely.
-This nuclear option is always available to both of parties and yet, in modern history, it has never been exercised.
+This nuclear option is always available to both parties and yet, in modern history, it has never been exercised.
 We can reasonably infer that each nation's leadership is making an identical calculation: the benefits of continuing to practice espionage in my adversary's backyard outweigh the risks associated with allowing my adversary to do the same to me.
@@
 The relatively open flow of information is only possible with a universal buy-in.
-Stripped of its mystique, the benefits of allowing intelligence operations to continue are, like any other state decision, a simple cost-benefit analysis; As illustrated above, states decide that it best serves their interests to remain in the top left corner, rather than the bottom right.
+Stripped of its mystique, the benefits of allowing intelligence operations to continue are, like any other state decision, a simple cost-benefit analysis; as illustrated above, states decide that it best serves their interests to remain in the top left corner, rather than the bottom right.
 We can expect that there is \emph{some} rational explanation that political leaders employ to justify it, where they weigh the benefits of preserving their ability to spy against the damage that results from espionage conducted against them.
@@
 States take the risk that espionage will reveal more than just a state's (hopefully benign) intent---it might also reveal a state's capabilities and strategy.
-For example, I interviewed senior Obama administration official Jake Sullivan about the benefits of allowing foreign states to spy on the US.
+For example, I interviewed senior Obama Administration official Jake Sullivan about the benefits of allowing foreign states to spy on the US.
 According to him, the United States would have no problem with adversaries being able to independently verify American intent because, as he sees it, American intent is benign.
@@
 
-% States that allow espionage are operating with faith that a freer flow of information is going to lead to a better outcome.
+\subsection{Hypothesis}
 
-% because it allows us to independently verify the intent and capabilities of other nations.
-Doing so actually reduces the likelihood that miscalculation will lead to open hostilities.
-As a result, both states allow it to happen, because they each benefit from a less-tense security environment.
-Stone aptly described this phenomenon as \enquote{reciprocally tolerated espionage.}
-
-\subsection{Introducing the empirical examples}
 Previous intelligence scholars have noted specific benefits of espionage that support the abstract theoretical justification I build here.
@@
 Limiting ballistic missile defense systems was a relatively minor concession for both sides, but in the highly risk-averse context of the Cold War, the signal was significant.\footcite[p.~66]{glaser_rational_2010} Facing the omnipresent threat of nuclear annihilation, \emph{d\'etente} provided a rational strategic choice for both states---made possible by their reciprocal tolerance of espionage.
+
+% States that allow espionage are operating with faith that a freer flow of information is going to lead to a better outcome.
+
+% because it allows us to independently verify the intent and capabilities of other nations.
+Doing so actually reduces the likelihood that miscalculation will lead to open hostilities.
+As a result, both states allow it to happen, because they each benefit from a less-tense security environment.
+Stone aptly described this phenomenon as \enquote{reciprocally tolerated espionage.}
+
+\section{Case Studies}
+\subsection{Research Design}
 
@@
 
-\newpage
-\section{Aerial espionage}
+\subsection{Aerial espionage}
 The line between a \enquote{cold} and a \enquote{hot} war is supposed to be obvious.
@@
 
-\subsection{Military reconnaissance flights}
+\subsubsection{Military reconnaissance flights}
 The tense battle between the US and USSR over surveillance of Soviet border installations was never formally acknowledged by either party---and neither were its combatants.
@@
 John Farquhar specializes in aerial reconnaissance and its diplomatic consequences.
-Many of the incidents analyzed in this section are taken from newspaper clippings that he compiled and analyzed.
+Many of the incidents analyzed in this subsection are taken from newspaper clippings that he compiled and analyzed.
 Second, former intelligence analyst and section chief Michael Peterson wrote an article for the CIA history magazine \emph{Cryptologic Quarterly} that analyzed the Soviet shootdowns of US reconnaissance aircraft in depth.
@@
 
-\subsection{Inventing the spy plane}
+\subsubsection{Inventing the spy plane}
 While Eisenhower was able to defuse some of the tension surrounding the military reconnaissance flights, even the most aggressive overflights could only hope to skirt the borders of Soviet territory; the United States couldn't just simply fly a B-52 over Moscow.
@@
 
-\subsection{Espionage norms in aerial reconnaissance}
+\subsubsection{Espionage norms in aerial reconnaissance}
 The U-2 Incident is often cited as a seminal failure of the Eisenhower administration.
@@
 
-\section{Spy satellites}
+\subsection{Spy satellites}
 Gary Powers had been in Soviet prison for less than three months when the United States launched an instrument of espionage far more advanced, and far more damaging, than the U-2 could ever be.
@@
 
-\subsection{Freedom of space}
+\subsubsection{Freedom of space}
 Spy satellites are not just one of the best edge cases for norm against shooting down a satellite, they are the reason it exists in the first place.
@@
 
-\subsection{Space is not a race}
+\subsubsection{Space is not a race}
 Everything about the administration's space policy was carefully crafted to set the stage for the eventual launch of a spy satellite.
@@
 
-\subsection{The Corona Project}
+\subsubsection{The Corona Project}
 The Soviet Union unintentionally set an early precedent for ``freedom of space,'' but that did not automatically justify using space for reconnaissance.
@@
 
-\subsection{Espionage norms in satellite reconnaissance}
+\subsubsection{Espionage norms in satellite reconnaissance}
 The legitimization of spy satellites dealt an indisputable blow to Soviet standing, so why did the Soviet Union let it happen?

### 2020-09-08 article.tex
Add hypothesis and new Baltic Incident analysis

@@
 If removing a quarter of the officers doesn't do lasting damage, then why not remove the full 200?
-The swift expulsion of all suspected intelligence officers from a country---and the inevitable removal of the country's own intelligence officers in retaliation---would deeply injure both their espionage capabilities indefinitely.
-This nuclear option is always available to both parties and yet, in modern history, it has never been exercised.
+The swift expulsion of all suspected intelligence officers from a country---and the inevitable removal of the country's own intelligence officers in retaliation---would deeply injure both their espionage capabilities indefinitely.\footnote{Nothing on this scale has ever been attempted, but Operation Foot, the largest diplomatic expulsion in the Cold War, \enquote{was a blow from which the Soviet intelligence effort in the UK never recovered.} For more, see: \cite{hughes_giving_2006-1}} This nuclear option is always available to both parties and yet, in modern history, it has never been exercised.
 We can reasonably infer that each nation's leadership is making an identical calculation: the benefits of continuing to practice espionage in my adversary's backyard outweigh the risks associated with allowing my adversary to do the same to me.
@@
   \multirow{3}*{State 1}  & Tolerate Espionage & \makecell{~\\Both sides gain intelligence \\~} & \makecell{~\\ Only State 2 continues spying \\ ~} \\\cline{3-4}
-  & Punish Espionage & \makecell{~\\ Only State 1 continues spying \\~} & \makecell{~\\ No intelligence either direction \\~} \\\cline{3-4}
+  & Punish Espionage & \makecell{~\\ Only State 1 continues spying \\~} & \makecell{~\\ Limited intelligence either direction \\~} \\\cline{3-4}
 \end{tabular}
@@
 To arrive at mutual cooperation, a security-seeking state needs to not only determine that its rival is a security seeker, it needs its rival to make that exact same judgment, because it is the adversary's response to securitization that ultimately makes the first state insecure.
-Defensive realism therefore introduces a counterintuitive conclusion: it is often logical for a security seeker to take actions that increase not only its own security, but its \emph{adversary's} security as well.\footcite[p.~7]{glaser_rational_2010} Such actions would make the adversary less likely to securitize, and also signal to the adversary that it is a security seeker hoping to reach a mutually agreeable outcome.
-If the adversary is indeed a security seeker, then, no longer concerned about its security, it will respond in kind and resolve the security dilemma for both of them.
+Defensive realism therefore introduces a counterintuitive conclusion: it is often logical for a security seeker to take actions that increase not only its own security, but its \emph{adversary's} security as well.\footcite[p.~7]{glaser_rational_2010} If the adversary is indeed a security seeker, then, no longer concerned about its security, it will respond by not building up arms, and resolve the security dilemma for both of them.
 The problem for a defensive realist is that unilaterally taking steps to increase the adversary's security is obviously quite risky.
@@
 Some defense capabilities would probably be fine to have an adversary independently verify, and some the US might want to keep secret because their secrecy confers a strategic advantage.
-And under almost no circumstances would the US want to have its strategies and warfighting plans revealed, so every measure should be taken to ensure their secrecy.\footcite{sullivan_personal_2019} States would prefer to reveal only the \enquote{good} information, but they can't---the credibility of the information obtained depends on the fact that the espionage subverted its target's access controls.
-If you knew your target had let you find something, you wouldn't trust it.
+And under almost no circumstances would the US want to have its strategies and warfighting plans revealed, so every measure should be taken to ensure their secrecy.\footcite{sullivan_personal_2019} States might prefer to reveal only the \enquote{good} information, but they can't.
+Espionage is by definition illicit---the fact the information was obtained without the target's permission lends it credibility.
 
@@
 \subsection{Hypothesis}
+In the following section, I will use empirical case studies to demonstrate how associating an operation with espionage alters its diplomatic response.
+In order to guide that research, it is important to establish a series of clear, testable hypotheses.
+I first present a null hypothesis:
 
-Previous intelligence scholars have noted specific benefits of espionage that support the abstract theoretical justification I build here.
+\begin{quote}
+$H_0 =$ Associating a state action with the traditional practice of espionage will not meaningfully alter the diplomatic response.
+\end{quote}
+
+The null hypothesis posits that there is insufficient evidence linking the perception that an act \enquote{is} espionage to the response that the affected state ultimately delivers.
+If this hypothesis holds, espionage may have some symbolic importance, but it isn't one that can be meaningfully tied to a diplomatic response.
+By contrast, here is the alternative hypothesis, the one that would prove true if the theory holds:
+
+\begin{quote}
+$H_1 =$ Associating a state action with espionage will result in a reduced diplomatic response when compared with equivalent actions that do not share that association.
+\end{quote}
+
+The argument presented by this article risks a tautology: perhaps we associate certain actions with espionage \emph{because} they received a reduced diplomatic response.
+For this reason, the relationship in $H_1$ is explicitly framed as causal.
+Unfortunately, defining the relationship that way makes it difficult to prove directly; no two state actions take place under equivalent circumstances, so no natural experiment could demonstrate that equivalent actions, perceived differently, received different responses.
+I therefore propose a corollary hypothesis to aid us in proving the alternative hypothesis:
+
+\begin{quote}
+$H_2 =$ States purposefully pursue measures to signal that their actions constitute espionage, because they apply reduced diplomatic responses to espionage and hope that their adversary will do the same.
+\end{quote}
+
+Proving $H_2$ is significantly easier than proving $H_1$.
+State actions---and the reasons policymakers provided for pursuing those actions---are often documented in the historical record.
+Careful research could demonstrate that \enquote{espionage} is an ontologically significant concept, not just for academics, but for the very people who engage in it.
+If $H_2$ holds, one would find evidence of policymakers being willing to accept the cost of enemy espionage in order to practice espionage themselves based on the value of mutual observation.
+Furthermore, when $H_2$ is proven to be true bidirectionally, it implies $H_1$.
+Two states that are both maintaining reduced responses to espionage (while pursuing espionage themselves) create a relationship in which $H_1$ holds: state actions successfully associated with espionage will receive a reduced response.
+
+% Previous intelligence scholars have noted specific benefits of espionage that support the abstract theoretical justification I build here.
 Christopher Baker explains how espionage makes functional cooperation possible through treaties.\footcite[p.~6]{baker_tolerance_2004} Charles Glaser uses the 1972 Anti-Ballistic missile treaty as an example of precisely the type of signal that could lead states to resolve the security dilemma Julius Stone, a professor of international law, foresaw this exact role for espionage in 1962: \enquote{If you do not have a system of international inspection and if you can't get one (and I think it is quite likely that we can't) then the function which international inspection is supposed to serve still needs fulfilling.
@@
 
-The best way to demonstrate that this theory has explanatory power for state decision-making is to find historical examples where states made decisions that showed they were willing to accept the cost of enemy espionage in order to practice espionage themselves.
-If the theory is correct, we would find evidence of policymakers justifying their decisions based on the value of mutual observation.
-Such decisions were made frequently during the Cold War, and with the explicit intent to increase the bilateral flow of information to create a more secure world environment.
+The following section will demonstrate that both the United States and the USSR were forced to contend with the benefits and risks of a mutually permissive espionage regime during the Cold War.
+When faced with opportunities to either shut down the espionage apparatus or expand it, they both chose to expand it.
+Policymakers on either side of the Iron Curtain repeatedly decided that preserving the bilateral flow of information ensured a more secure world environment.
 This article will present two such cases: aerial espionage and spy satellites.
 
-Both planes and satellites played an under-explored role in the formulation of some of the Cold War's landmark diplomatic achievements.
+% Both planes and satellites played an under-explored role in the formulation of some of the Cold War's landmark diplomatic achievements.
 SALT I (the ABM Treaty) is widely considered to be the landmark diplomatic achievement of Nixon's \emph{d\'etente} policy, which set the stage for a broader easing of relations between the US and the USSR, and spy satellites were crucial to reducing the cost that each side would pay to sign it.
@@
 % States that allow espionage are operating with faith that a freer flow of information is going to lead to a better outcome.
-
-% because it allows us to independently verify the intent and capabilities of other nations.
-Doing so actually reduces the likelihood that miscalculation will lead to open hostilities.
-As a result, both states allow it to happen, because they each benefit from a less-tense security environment.
-Stone aptly described this phenomenon as \enquote{reciprocally tolerated espionage.}
 
@@
 \subsection{Aerial espionage}
-The line between a \enquote{cold} and a \enquote{hot} war is supposed to be obvious.
-A superpower's own uniformed military personnel opening fire on the other superpower's military or territory would be \enquote{hot,} everything else would be \enquote{cold.} Broadly speaking, the Cold War is remembered as lacking direct military confrontation between the United States and Russia; though the two superpowers of the Cold War were open rivals, they supposedly competed for influence by funding, arming, or otherwise influencing proxies, rather than attacking each other directly.
-That impression is not exactly accurate.
-Not only did the US and USSR often come close to openly attacking each other, but a surprising number of times they actually did—--including one of the Cold War's most famous moments.
 
-% citation needed on the definition of cold
+Just a few weeks before the \enquote{Four Powers Summit} in Summer 1960, Soviet Premeir Nikita Khrushchev announced that the USSR had shot down an American plane.\footcite[p.~112]{powers_operation_2004} The CIA knew immediately that it had lost a U-2, an advanced spy plane that had bedeviled the Soviets for years, proving beyond the reach of their surface-to-air missiles.\footcite{orlov_u-2_2007} Though clearly the technological gap had now closed, the CIA was still confident that its cover story would hold, thanks to a morbid fact of the U-2's design: in the event of a shootdown, it should have been impossible for the pilot to survive.\footcite[p.~35]{lindgren_trust_2000} When the United States put out at a statement that the plane was part of a \enquote{weather reconnaissance} mission gone awry, Khrushchev sprung his trap and revealed that the U-2 pilot, Gary Powers, had been taken alive.
 
-Pilot Francis Gary Powers knew that U-2 overflights of the Soviet Union were incredibly risky.
-The Soviets were deeply unhappy that the US was violating their airspace in such a brazen fashion, but for years they had kept quiet about the U-2 missions, embarrassed that they were technologically incapable of intercepting the high-flying aircraft.\footcite[p.~59]{powers_operation_2004} When Soviet air defenses finally managed to shoot one down in 1960, Soviet leadership was equal parts triumphant and furious.
-At a planned summit in Paris only a few weeks later, Soviet Premier Khrushchev launched into a rant about the U-2 planes and stormed out on the first day, effectively ending the summit---and any hope that pressing issues like the fate of West Berlin would be resolved there.
+In a few days, the previously pleasant trend in US-Soviet relations completely reversed.
+President Dwight D.
+Eisenhower admitted that the Gary Powers had been flying a spy plane, declaring in a chagrined press conference that espionage was ``a distasteful but vital necessity'' to prevent another attack like Pearl Harbor.\footcite{eisenhower_news_1960} At the planned summit, Khrushchev launched into a tirade about U-2 planes and walked out on the first day, effectively ending the summit.
 He also canceled a planned visit by US President Dwight Eisenhower to the USSR that was scheduled to take place the next month.
@@
 
-The shootdown of Gary Powers, in what would later come to be known as \enquote{The U-2 Incident}, clearly violated the only \enquote{rule} of the Cold War.
+In the traditional telling, the U-2 Incident is often cited as a major failure of the Eisenhower administration, which significantly worsened US-Soviet relations at time when lasting peace appeared possible.
+Only in the under-studied context of Cold War reconnaissance flights does it become clear that the U-2 Incident actually shows both the US and USSR exhibiting restraint and sound judgment.
+Over a period of about 15 years, they each allowed provocative aerial intelligence operations to take place without military escalation.
+These operations were permitted because, consistent with $H_2$, the United States took great pains to signal that its various actions were within the bounds of espionage.
+The Soviet Union responded by honoring that signal and not escalating the situation beyond the immediately necessary measures to counteract it.
+The U-2 Incident should be understood as a uniquely embarrassing but otherwise typical failure of an intelligence operation---the maximal case for the minimal consequences typically applied to espionage.
+
+% The shootdown of Gary Powers clearly violated the only \enquote{rule} of the Cold War.
 Official Soviet air defenses shot down a US aircraft.
@@
 
-No single attribute defines a spy plane.
+% No single attribute defines a spy plane.
 A military aircraft can forgo weapons but, by virtue of its pilot, its mission, its timing, or its branding still conduct what is unquestionably a military operation.
-Inversely, intelligence missions can be flown by uniformed Air Force pilots in re-purposed B-52s and somehow still be understood as conducting espionage.
-A spy plane is an ontological designation, one that reveals itself to the adversary through a series of social and physical signifiers that are not replicable across time.
-Germany in 1937 might not have reacted to a U-2 overflight the same way that the Soviet Union did in 1960.
-But even if the way different states handled espionage was perfectly consistent, there still wouldn't be a checklist to guarantee that a mission would be treated as espionage; it is not enough make a military mission look like an intelligence operation.
-The reality is both simpler and more frustrating---it has \emph{to be one}.
-
-The U-2 incident was the most visible in a series of aerial intelligence operations that the United States undertook early in the Cold War.
-The US did not always employ civilian pilots nor it did always use obviously weaponless aircraft, yet it managed to run a shocking number of missions that violated Soviet airspace without significantly altering the relative level of tension between the two superpowers.
-On a surprisingly frequent basis, these missions would turn \enquote{hot} when Soviet air defenses shot them down.
-Not once, however, did these shootdowns lead to retaliation from the US, or significant pressure from the Soviets to stop the flights.
-Instead, each one was resolved as if the pilots who survived were simply spies, captured in the normal course of counterintelligence.
-This article seeks to answers how and why states ensure that certain operations reveal themselves as espionage; the decade of covert aerial espionage was one such successful operation.
+Inversely, intelligence missions can be flown by uniformed Air Force pilots in re-purposed B-52s and somehow still be understood as espionage.
+Espionage is an ontological designation, one that reveals itself to the adversary through a series of social and physical signifiers that are not replicable across time.
+Germany in 1937 would likely have reacted to a U-2 overflight differently than the Soviet Union did in 1960.
+But even if the way different states handled espionage was perfectly consistent, there still wouldn't be a checklist to guarantee that a mission would be treated as espionage.
+It is not enough make a military mission look like an intelligence operation; the reality is both simpler and more frustrating---it has \emph{to be one}.
 
 \subsubsection{Military reconnaissance flights}
-The tense battle between the US and USSR over surveillance of Soviet border installations was never formally acknowledged by either party---and neither were its combatants.
+Beginning in 1947, the US began a campaign of reconnaissance flights that routinely violated Soviet airspace.
+Converted Air Force bombers---called \enquote{ferrets}---were outfitted with advanced radio equipment and sent to intercept as much information about Soviet radio transmissions as possible.\footcite[p.~4]{peterson_maybe_1993} The pilots who flew these missions were conducting one of the earliest forms of Cold War peacetime espionage: gathering SIGINT (signals intelligence) that identified the location of critical radar installations along the Soviet border.
+Some of these flights only flew near the border, remaining over traditionally recognized international waters, while others were \enquote{overflights,} missions that deliberately violated Soviet airspace.
+
+The tense battle over Soviet border surveillance was never formally acknowledged by either state---and neither were its combatants.
 Over the course of the Cold War, more than 200 American pilots were shot down while spying on the Soviet Union.\footcite{glenshaw_secret_2017} These missions were so highly classified that in many cases the families of the pilots never learned how they died.
 126 of these American airmen remain unaccounted for today, casualties of missions whose existence both the US government and the Soviet Union collaborated for decades to deny.
+Not once did these shootdowns lead to retaliation from the US, or significant pressure from the Soviets to stop the flights.
+Instead, the cases resolved as if the pilots were spies, captured or killed in the course of counterintelligence, whereabouts undisclosed.
 
-Though they flew Air Force planes, these men were actually conducting one of the earliest forms of Cold War peacetime espionage.
-Their purpose was to gather SIGINT (signals intelligence) identifying the location of critical radar installations along the Soviet border.
-Converted bombers---called \enquote{ferrets} by the Air Force---were outfitted with advanced radio equipment and sent to intercept as much information about Soviet radio transmissions as possible.\footcite[p.~4]{peterson_maybe_1993} According to the CIA, these missions began as early as 1947.\footcite[p.~4]{peterson_maybe_1993} Some of these flights only flew near the border, remaining over traditionally recognized international waters, while others were \enquote{overflights,} missions that deliberately violated Soviet airspace.
-The earliest definitive postwar overflight of the Soviet Union took place on May 10, 1949, when two RF-80As surveilled the Kurile islands\footcite[p.~8]{peebles_shadow_2000}
+% The earliest definitive postwar overflight of the Soviet Union took place on May 10, 1949, when two RF-80As surveilled the Kurile islands\footcite[p.~8]{peebles_shadow_2000}
 
-Without access to classified archives it is impossible to state for certain exactly how many of these flights were conducted.
-Even if one did have unlimited access, some of the necessary information was so highly classified that it is likely lost forever.
-Nonetheless, the civilian researcher is aided by the declassified research of military archivists before them and the works of two scholars in particular are highlighted here.
-First, Air Force Academy historian Dr.
-John Farquhar specializes in aerial reconnaissance and its diplomatic consequences.
-Many of the incidents analyzed in this subsection are taken from newspaper clippings that he compiled and analyzed.
-Second, former intelligence analyst and section chief Michael Peterson wrote an article for the CIA history magazine \emph{Cryptologic Quarterly} that analyzed the Soviet shootdowns of US reconnaissance aircraft in depth.
-
-Peterson's article was written for a classified audience in 1993 and declassified in 2009, providing the public with a comprehensive (as far as we know) list of all American SIGINT flights shot down by the USSR (Table \ref{soviet-shootdowns}).
-Peterson uses two limiting criteria that perfectly suit the purposes of this article: the only flights he lists are those that were (a) shot down by Soviet attacks, not Chinese, North Korean, etc., and (b) explicitly US reconnaissance aircraft.\footcite[p.~4]{peterson_maybe_1993} Note that not all of these flights are necessarily overflights; many took place over international waters when they were shot down.
-In effect, the table catalogs the 13 times where the USSR shot down an American aircraft designed for collecting intelligence between 1950 and 1964---intercepted espionage attempts.\footcite[p.~5.
+Without access to classified archives it is impossible to state for certain exactly how many of these flights were conducted.\footnote{Some of the necessary information was so highly classified that it is likely lost forever.} Nonetheless, the civilian researcher is aided by the declassified research of military and intelligence archivists.
+Writing for the internal CIA publication \emph{Cryptologic Quarterly}, Michael Peterson compiled a now-declassified list of American SIGINT flights shot down by the USSR (Table \ref{soviet-shootdowns}).\footnote{Michael Peterson is a former intelligence analyst and section chief for the CIA.} Peterson uses two limiting criteria that perfectly suit the purposes of this article: the only flights he lists are those that were (a) shot down by Soviet attacks, not Chinese, North Korean, etc., and (b) involved US aircraft exclusively performing reconnaissance.\footcite[p.~4.
+Regardless of how thorough declassified CIA information appears, one cannot rule out the possibility that there were other, even more secret missions which, for whatever reason, the United States government has chosen to conceal to this day.
+Nonetheless, a few aspects of the \emph{Cryptologic Quarterly} article suggest that the information is relatively complete.
+First, the article was written for a classified audience, likely given that designation so that it could be comprehensive, and declassified after the fact.
+Additionally, when the article was declassified 16 years later, it appears to have been declassified without redactions or missing pages.
+Lastly, if there had been another mission so highly classified that it was omitted from this document, it is difficult to imagine how diplomatic consequences could have been imposed with a degree of secrecy that would prevent us from knowing about it today.]{peterson_maybe_1993} In effect, the table catalogs the 13 times where the USSR shot down an American aircraft designed for collecting intelligence between 1950 and 1964---intercepted espionage attempts.\footcite[p.~5.
 In the original document, this table lists the first incident as having taken place over the Barents Sea, not the Baltic Sea.
-Because the description of the mission---including a map of its route in the same document---takes place entirely over the Baltic sea, I have concluded that this has to be a typographical error, and corrected it here.]{peterson_maybe_1993}
+Because the description of the mission---including a map of its route in the same document---takes place entirely over the Baltic sea, I have concluded that this must be a typographical error, and corrected it here.]{peterson_maybe_1993}
 
@@
 4 September 1954  & USN P2V Neptune               & Sea of Japan                         \\
-7 November 1954   & USAD RB-29                    & East of Hokkaido.Kuril Is.           \\
+7 November 1954   & USAD RB-29                    & East of Hokkaido. Kuril Is.           \\
 18 April 1955     & USAF RB-47                    & Off Kamchatka Peninsula              \\
@@
 
-Regardless of how thorough declassified CIA information appears, one cannot rule out the possibility that there were other, even more secret missions which, for whatever reason, the United States government has chosen to conceal to this day.
-Nonetheless, a few aspects of the \emph{Cryptologic Quarterly} article are promising, and suggest that the information is relatively complete.
-First, the article was written for a classified audience as an overview of reconnaissance shootdowns during the 1950s, so it is likely intended to be comprehensive.
-Shootdowns are a sufficiently significant event such that it is unlikely the author would have just missed one.
-Additionally, when the article was declassified 16 years later, it appears to have been declassified without redactions or missing pages.
-Lastly, if there had been another mission so highly classified that it was omitted from this document, it is difficult to imagine how diplomatic consequences could have been imposed with a degree of secrecy that would prevent us from knowing about it today.
+Peterson's article represents the CIA's internal accounting for SIGNIT shootdowns, but it does not tell which missions were perceived as significant by the Soviet Union.
+For that, I cross-referenced the CIA's list with Air Force historian Dr.
+John Farquhar's analysis of Cold War shootdowns that had a diplomatic impact.\footcite[Dr.
+John T Farquhar is a retired Lieutenant Colonel in the United States Air Force, and currently an Associate Professor of Military \& Strategic Studies at the US Air Force Academy.]{farquhar_aerial_2015} The combination of these two lists is instructive---some incidents in the CIA list are missing from Farquhar's list because they did not have a diplomatic impact, and some incidents that Farquhar analyzes are not mentioned by Peterson because they were not SIGINT reconnaissance missions.
+Taken together, the full spectrum of aerial reconnaissance incidents demonstrates a clear pattern: shootdowns that were interpreted as espionage were quickly resolved by both parties, while the shootdown of a flight unrelated to espionage produced swift, significant backlash.
 
-To comprehensively analyze the diplomatic effects of aerial espionage, I cross-referenced the shootdowns that the CIA says took place with Farquhar's analysis of their media and diplomatic impact.
-A shootdown is either present in both lists, or just one of the two---some incidents in the CIA list are missing from Farquhar's list because they did not have a diplomatic impact, and some incidents that Farquhar analyzes are not mentioned by Peterson because they were not SIGINT reconnaissance missions and likely (though not definitively) served little espionage function.
-Together, the full spectrum of aerial reconnaissance incidents demonstrates a clear pattern in which shootdowns that were interpreted as espionage were quickly resolved by both parties, while the shootdown of a flight unrelated to espionage produced swift, significant backlash.
+% First, many of the incidents analyzed in this subsection are taken from clippings compiled by Air Force Academy historian Dr.
+John Farquhar, who specializes in aerial reconnaissance and its diplomatic consequences.
+Second,
 
-The first of these shootdowns, on April 1950, brought the previously-hidden aerial reconnaissance program to the attention of the American people.
-An unarmed Navy Patrol plane, a PB4Y2 Privateer, was intercepted and downed by Soviet fighters over the Baltic sea.
-The Soviet ambassador to the US lodged a formal note of protest, incorrectly alleging that the US aircraft which had been shot down was a B-29 bomber.\footcite{kirk_ambassador_1950} A few days later, the American ambassador issued a formal response, in which he correctly identified the model of plane and claimed that at no point did it cross into any territorial waters.
+The first of these shootdowns brought the aerial reconnaissance program to the attention of the American people on April 8, 1950.\footnote{Note that the flights analyzed here were not necessary overflights; many took place over international waters or dispute waters when they were shot down.} An unarmed Navy patrol plane was downed by Soviet fighters over the Baltic sea.
+The Soviet ambassador to the US lodged a formal note of protest.\footcite{kirk_ambassador_1950} A few days later, the American ambassador issued a formal response, claiming that at no point did the plane cross into any territorial waters.
 He demanded the Soviets conduct \enquote{a prompt and thorough investigation} and \enquote{see to it that those responsible for this action [were] promptly and severely punished.}\footcite{the_associated_press_text_1950} The Soviet Union recognized this for the toothless statement that it was, and rejected the demand in its entirety.\footcite{salisbury_kremlin_1950} No further action was taken.
 
-Both sides had reasons to make this a larger political issue.
-The Soviet Union had fired first on a US Navy aircraft---which could not have been the aggressor because its only armament was a .45-cal pistol carried by one of the crewmen.\footcite[p.~7]{peterson_maybe_1993} Given that the plane was never found, all available evidence suggests that the flight likely did not violate the traditional 12-mile airspace boundary.
-Based on the last received position report, today the CIA claims that the plane was flying 20-25 miles off the coast of Latvia at the time it was shot down.\footcite[p.~7]{peterson_maybe_1993} At the time, it was theorized that a theoretical breakdown in the plane's navigation system could have caused it to veer off-course into Latvia, but given its last known position, it would have taken \enquote{a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.}\footcite{the_new_york_times_soviet_1950} Meanwhile, the United States knew that the missing plane probably hadn't ended up in Soviet airspace, but the Privateer in question \emph{had} been performing a reconnaissance mission of questionable legality when it was shot down.
-No traditional boundaries were violated, but a repurposed Navy privateer loaded with electronic reconnaissance equipment flying close to Latvia looked, to use a less technical term, shady.
-And still, based on the initial exchange of diplomatic protests, neither side was inclined to magnify the issue if they could avoid it.
+Consistent with $H_2$, the United States took steps to code this flight as espionage, which was successful not just in the eyes of the Soviet Union, but to the American press as well.
+\emph{Washington Post} columnist Walter Lippmann wrote that \enquote{the known facts indicate that Soviet intelligence \textelp{} believed it was carrying important electronic equipment and that orders were given to the Soviet fighter command to intercept it.}\footcite{lippmann_baltic_1950} \enquote{Baltic Plane Mystery,} a \emph{Post} article written a few days later, was almost sympathetic to the Soviet commanders.
+\enquote{Electronics make the old delimitations for border coastal flights ridiculous.
+A plane flying on a course perfectly legal by standards accepted today might still be engaged in reconnaissance of the first importance that an unfriendly power would try to frustrate.}\footcite{childs_baltic_1950}
 
-% The American press had no such reservations.
-\emph{Washington Post} columnist Walter Lippmann wrote that it could not have been \enquote{a local incident brought on by a local commander but [rather] that it was an act of Soviet policy.
-The known facts indicate that Soviet intelligence \textelp{} believed it was carrying important electronic equipment and that orders were given to the Soviet fighter command to intercept it.}\footcite{lippmann_baltic_1950} He speculated that there was no way the plane could have violated Soviet airspace, because wreckage would have been recovered, the plane was too slow, and no commander would have sent it on such a dangerous mission intentionally.
-\enquote{If, when, and as the American command were reconnoitering the Soviet military establishments on the Baltic coast,} Lippmann wrote, \enquote{it would use a plane of a wholly different type.}\footcite{lippmann_baltic_1950}
-
-% A newspaper columnist is not a policymaker, but his argument was likely reflective of the political opinion that Americans held regarding espionage at the time: of course we do these things, we just do them \emph{better} than that---and if the US had been doing the thing that the Soviets claimed it was, then it would have been justified of them to open fire.
-\enquote{Baltic Plane Mystery,} a \emph{Post} article written a few days later, was almost sympathetic to the Soviet commanders.
-\enquote{Electronics make the old delimitations for border coastal flights ridiculous.
-A plane flying on a course perfectly legal by standards accepted today might still be engaged in reconnaissance of the first importance that an unfriendly power would try to frustrate.}\footcite{childs_baltic_1950} This is true, both in terms of the mission itself and the technological environment, but he went even further.
+% This is true, both in terms of the mission itself and the technological environment, but he went even further.
 \enquote{The only sound and safe assumption is that the Russians have a thorough and far-reaching espionage system.
@@
 
-The general attitudes displayed in response to the Baltic Incident became the normal course of response to Soviet flights.
-The shootdown of the Navy Privateer was not an isolated incident; from then until the U-2 Incident in 1960, the United States lost nine more aircraft to Soviet air defenses.
+The Baltic incident exemplifies the reduced consequences for espionage that $H_2$ predicts, in both directions.
+The Soviet Union opened fire on a US Navy aircraft, which based on all available evidence did not violate the traditional 12-mile airspace boundary.\footcite[p.~7.
+To be extra specific]{peterson_maybe_1993} Given its last known position, it would have taken ``a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.''\footcite{the_new_york_times_soviet_1950} Under other circumstances, this could have been a major incident, but the United States chose not to publicize it.
+The Soviet Union, for its part, chose not to escalate the issue either, either, despite the ample evidence that the US was engaging in military reconnaissance.
+
+The attitudes displayed in response to the Baltic Incident became the normal course of response to reconnaissance flights.
+From then until the U-2 Incident in 1960, the United States lost nine more aircraft to Soviet air defenses.
 Each time, the pattern of responses was the same---each side released dueling statements of protest, and the situation continued more or less as before.

### 2020-09-13 article.tex
Finish satellites introduction

@@
 
-% States that allow espionage are operating with faith that a freer flow of information is going to lead to a better outcome.
-
 \section{Case Studies}
@@
 
-Peterson's article represents the CIA's internal accounting for SIGNIT shootdowns, but it does not tell which missions were perceived as significant by the Soviet Union.
-For that, I cross-referenced the CIA's list with Air Force historian Dr.
+Peterson wrote the CIA's internal accounting for SIGNIT shootdowns, but he cannot tell us which missions were perceived as significant by the Soviet Union.
+For that, I cross-referenced the Peterson's list with historian Dr.
 John Farquhar's analysis of Cold War shootdowns that had a diplomatic impact.\footcite[Dr.
@@
 Consistent with $H_2$, the United States took steps to code this flight as espionage, which was successful not just in the eyes of the Soviet Union, but to the American press as well.
-\emph{Washington Post} columnist Walter Lippmann wrote that \enquote{the known facts indicate that Soviet intelligence \textelp{} believed it was carrying important electronic equipment and that orders were given to the Soviet fighter command to intercept it.}\footcite{lippmann_baltic_1950} \enquote{Baltic Plane Mystery,} a \emph{Post} article written a few days later, was almost sympathetic to the Soviet commanders.
+\emph{Washington Post} columnist Walter Lippmann wrote: \enquote{the known facts indicate that Soviet intelligence \textelp{} believed [the plane] was carrying important electronic equipment and that orders were given to the Soviet fighter command to intercept it.}\footcite{lippmann_baltic_1950} A \emph{Post} article written a few days later was almost sympathetic to the Soviet commanders.
 \enquote{Electronics make the old delimitations for border coastal flights ridiculous.
@@
 To be extra specific]{peterson_maybe_1993} Given its last known position, it would have taken ``a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.''\footcite{the_new_york_times_soviet_1950} Under other circumstances, this could have been a major incident, but the United States chose not to publicize it.
-The Soviet Union, for its part, chose not to escalate the issue either, either, despite the ample evidence that the US was engaging in military reconnaissance.
+The Soviet Union, for its part, chose not to escalate the issue either, despite the ample evidence that the US was engaging in military reconnaissance.
 
-The attitudes displayed in response to the Baltic Incident became the normal course of response to reconnaissance flights.
-From then until the U-2 Incident in 1960, the United States lost nine more aircraft to Soviet air defenses.
-Each time, the pattern of responses was the same---each side released dueling statements of protest, and the situation continued more or less as before.
+The resolution of the Baltic Incident set a precedent for reconnaissance flights.
+From then on until 1960, the United States lost nine more aircraft to Soviet air defenses.
+Each time, the pattern of responses was the same---each side released dueling statements of protest, then the situation continued more or less as before.
 A B-29 that went missing on October 7, 1952 hit the front page of the \emph{New York Times} two days later.
-This mission is on Peterson's list as a ferret flight.
-Farquhar notes that the incident happened during the US election season and that the media thought the attack was intended to lower American prestige, but he does not mention any diplomatic consequences.\footcite[p.~43-44]{farquhar_aerial_2015} The official US response, however, is preserved on the State Department website.
+This mission is listed by Peterson as a ferret flight, and Farquhar notes that the media thought the attack was intended to lower American prestige during election season, but does not mention any diplomatic consequences.\footcite[p.~43-44]{farquhar_aerial_2015} The official US response, however, is preserved on the State Department website.
 As with the Baltic Incident, the US simply demanded that the Soviets pay for the plane and return any survivors.\footcite{the_new_york_times_u.s._1952} Two more shootdowns in March 1953---one involving an American F–84 Thunderjet and the other an RAF Lincoln bomber---actually resulted in conciliatory responses from both sides, including a secret meeting to reduce aerial tensions \enquote{of which little is known.}\footcite[p.~45]{farquhar_aerial_2015}
@@
 The Soviet Union responded by protesting the alleged American shootdown of a Soviet passenger plane.
-Despite press outrage, no further action was taken.\footcite[p.~47]{farquhar_aerial_2015} A September 4, 1954 shootdown of a Navy P2V Neptune actually led to a call by one US Senator to break off diplomatic ties with the Soviet Union: \enquote{Just another note from our State Department to the Kremlin hierarchy will not impress these uncivilized rulers \textelp{} that this new attack upon an American plane confirms Communist arrogance and aggressiveness to a point where the breaking of diplomatic relations is justified.}\footcite{the_associated_press_ending_1954} Instead of a harsher punishment, President Eisenhower brought the issue to the UN with the full knowledge that an unfavorable judgment would be vetoed by the Soviets.
-The move demonstrated Eisenhower's ability and commitment to resisting the calls for harsher action while still placating the American public.\footcite[p.~47]{farquhar_aerial_2015} Besides, by late 1954, the \enquote{cycle of hostility} over aerial reconnaissance was already breaking down.\footcite[p.~49]{farquhar_aerial_2015} In terms of both media coverage and diplomatic significance, no shootdown that followed would require more significant consequences than the ones that had already happened.
+Despite press outrage, no further action was taken.\footcite[p.~47]{farquhar_aerial_2015} After the September 4, 1954 shootdown of a Navy P2V Neptune, a US Senator publicly denounced the continued anodyne response: \enquote{just another note from our State Department to the Kremlin hierarchy will not impress these uncivilized rulers.}\footcite{the_associated_press_ending_1954} Instead of a harsher punishment, President Eisenhower brought the issue to the UN with the full knowledge that an unfavorable judgment would be vetoed by the Soviets.
+A further signal of deescalation, the UN stunt demonstrated Eisenhower's ability to resist calls for harsher action while still placating the American public.\footcite[p.~47]{farquhar_aerial_2015} By late 1954, the \enquote{cycle of hostility} over aerial reconnaissance was already breaking down.\footcite[p.~49]{farquhar_aerial_2015} In terms of both media coverage and diplomatic significance, no shootdown that followed would require more significant consequences than the ones that had already happened.
 
-One incident in the period between 1950-1954 does not fit the broader pattern.
-On November 20, 1951, an American C-47 transport was forced down in Hungary.
-This incident is omitted from Peterson's list because, despite Soviet cries of \enquote{spies and saboteurs,} there is no evidence that it was a reconnaissance flight.\footnote{It also wasn't exactly shot down so much as forced to land.} Of all the cases that Farquhar examines, this is the only one that provoked a sustained diplomatic response from the United States---and it is the only case where the plane in question definitively carried no espionage equipment.
-While the ferret flights were perfectly calibrated to fit as much radio surveillance equipment as possible, the only evidence of espionage that the Soviets were able to produce from the C-47---a plane that they recovered intact---was a portable radio, two extra parachutes, and some packets of warm blankets.\footcite{the_united_press_soviet_1951}
-
-I contend that the lack of espionage equipment is precisely the reason the United States made a sustained public effort to recover the four airmen, which is something that almost never happened with ferret flights.\footnote{While some of the flights were shot down overseas, making recovery difficult, I am referring here to situations where the US was aware that the Soviets likely held their airmen in captivity and still made little effort to get them back.} Farquhar summarizes the American response to this incident as follows: \enquote{Responding to the press attention, the Truman Administration acted swiftly, attempting to gain the fliers release through diplomatic pressure.
+During this period of aerial reconnaissance, the United States also suggested that $H_1$ could be true by showcasing its inverse.\footnote{The inverse of $p \implies q$ is $\neg p \implies \neg q$ i.e.
+if an act is not espionage, then it will not receive the reduced consequences typically applied to espionage.} On November 20, 1951, an American C-47 transport was forced down in Hungary.
+Of all the cases that Farquhar examines, this is the only case that provoked a sustained diplomatic response from the United States---and the only case where the plane in question definitively carried no espionage equipment.\footcite[While the ferret flights were perfectly calibrated to fit as much radio surveillance equipment as possible, the only evidence of espionage that the Soviets were able to produce from the C-47---a plane that they recovered intact---was a portable radio, two extra parachutes, and some packets of warm blankets.
+Because it was not a ferret flight, the C-47 is omitted from the Peterson list as well.]{the_united_press_soviet_1951} He summarizes the American response to this incident as follows: \enquote{Responding to the press attention, the Truman Administration acted swiftly, attempting to gain the fliers release through diplomatic pressure.
 The President ordered the Hungarian consulates in New York and Cleveland closed and banned private travel to the country.
@@
 
-We also know that the United States conducted many more overflights than the failed ones that are listed here, using a variety of retrofitted planes.
-In 2001, the Air Force held a symposium honoring and recollecting the pilots who flew the overflights on Soviet territory in the early Cold War, an effort in which they were joined by the British in 1954.\footcite[p.~v]{hall_early_2003} Many of the presenters spoke about entire reconnaissance projects that apparently didn't even register diplomatically.
+The United States conducted many more overflights than the failed ones mentioned here, using a variety of retrofitted planes.
 Project Heart Throb, for example, outfitted RB-57s with surveillance equipment and flew between 15 and 19 missions over Eastern Europe from 1955 to 1956.\footcite[p.~194.
 Based on the recollection of Gerald E.
-Cooke, a Air Force pilot assigned to the project.]{hall_early_2003} Another project from around the same time, Slick Chick, set up F-100 jets with rapid-fire 20mm cameras.
-The pilot who presented this project at the symposium, Cecil Rigby, personally flew two such missions and estimated that there were six Slick Chick missions total.\footcite[p.~176]{hall_early_2003}
+Cooke, a Air Force pilot assigned to the project.]{hall_early_2003} Project Slick Chick set up F-100 jets with rapid-fire 20mm cameras.\footcite[p.~176]{hall_early_2003} Though we know a lot less about these missions than we do about the ferret flights, the basic contours are familiar: missions with no weaponry, operating at a high level of secrecy, subject to countermeasures (tracked by radar and often chased by MiG fighters) but no evidence of diplomatic pressure.
+In meetings between the State department and high-level Soviet diplomats, the subject of captured pilots \enquote{was broached only perfunctorily in relation to other things being discussed.}\footcite[p.~72]{brugioni_eyes_2010} Some family remembers of lost airmen received posthumous awards, but most simply received the pilot's personal effects and no explanation.
 
-The USSR knew about these missions, which were tracked by radar and often chased by MiG fighters.
-Unlike some of the flights that were shot down, these overflights were clear violation of their territorial sovereignty.
-We will likely never know exactly how many such missions were flown.
-The Truman and Eisenhower administrations were shockingly unsuccessful at recovering the pilots who were held as POWs by the Soviets.
-Dino Brugioni, a senior CIA officer involved in the aerial imagery program, paints a grim picture of the institutional accountability involved when the flights went bad.
-In meetings between the State department and high-level Soviet diplomats, the subject of captured pilots \enquote{was broached only perfunctorily in relation to other things being discussed.}\footcite[p.~72]{brugioni_eyes_2010} Some family remembers of lost airmen received posthumous awards, but most simply received the pilot's personal effects and no explanation.
-\enquote{And because these were secret missions conducted by a field command,} Brugioni writes, \enquote{a change of field commander meant that the fate of the men was soon forgotten.}\footcite[p.~72]{brugioni_eyes_2010} With missions so secret that we abandoned the men who flew them, we can have no expectation of finding a complete record today.
-
-Even without a complete list of flights, it is clear that the territorial boundaries of the USSR were routinely violated to generate critical intelligence for NATO powers in an early, relatively volatile stage of the Cold War.
-As time passed these flights appear to have become more accepted, not less.
-Diplomats of both nations were, as a matter of policy, able to paper over these issues in service of advancing other interests.
-When the shootdown caused unavoidable diplomatic consequences---such as when it was well-publicized---the countries exchanged the necessary diplomatic cables and engaged in the requisite saber rattling.
-These incidents undoubtedly altered the climate of Cold War diplomacy, but the fact remains that in 10 years of reconnaissance, the only action the Soviets took to keep the United States from conducting these incredibly provocative reconnaissance flights was to shoot them down.
+The quantity and consistency of these flights demonstrates that coding reconnaissance flights as espionage, prototyped by the Baltic Incident, had settled into a practiced routine.
+The Soviet Union understood that these planes only carried reconnaissance equipment, the United States understood that these planes might get shot down, and both sides understood that these clashes would not escalate into a larger diplomatic battle.
+As predicted by $H_2$, the US took steps to associate its activities with espionage, and it applied reduced consequences to the Soviet Union's counterespionage.
+And while we lack primary sources that betray the Soviet rationale, the circumstances suggest that, consistent with $H_1$, the Soviet Union applied reduced consequences to espionage over this period of time as well.
 
 \subsubsection{Inventing the spy plane}
-While Eisenhower was able to defuse some of the tension surrounding the military reconnaissance flights, even the most aggressive overflights could only hope to skirt the borders of Soviet territory; the United States couldn't just simply fly a B-52 over Moscow.
-What US intelligence-gathering capabilities needed was some way of photographing suspicious installations deep within Russia while still being able to claim the same benefit-of-the-doubt status that prevented even the most contentious overflights from escalating into a larger conflict.
+% Even the most aggressive overflights could only hope to skirt the borders of Soviet territory---the United States couldn't just simply fly a B-52 over Moscow.
+Despite the success of the ferret flights, the US intelligence community needed a way to photograph suspicious installations deep within Russia, one that would continue to benefit from espionage norms.
 What they needed was a spy plane.
 
-For the spy plane to serve its intended purpose as a minimally provocative agent of espionage two things must be true: its operation must be understood by adversaries to be non-military, and that distinction must meaningfully alter their associated response.
-To build such an aircraft, Eisenhower approved Project Aquatone---the codename for the U-2.
-From its conception, the U-2 was purpose-built to be read as a civilian aircraft.
-Eisenhower required that the pilots of U-2 aircraft be CIA employees, specifically forbidding uniformed Air Force officers.\footcite[p.~33, Though many of the pilots did have Air Force backgrounds.]{lindgren_trust_2000} The CIA also had operational control of the program instead of the Air Force.
+% For the spy plane to serve its intended purpose as a minimally provocative agent of espionage two things must be true: its operation must be understood by adversaries to be non-military, and that distinction must fulfill the promise of $H_1$ by meaningfully reducing the associated response.
+
+As its name suggests, a spy plane is the physical manifestation of a state's attempt to code aerial reconnaissance as espionage.
+The first recorded reconnaissance flight took place in 1911, but that was not the first spy plane.\footcite[Surveilling the Turkish infantry near Tripoli, \enquote{Lieut.
+Piazzi today, for the first time in the history of warfare, made from this place an aerial reconnaissance against a hostile power}]{special_cable_to_the_new_york_times_air_1911} Ontologically, the spy plane---distinct from a plane that is doing spying---was invented by the Eisenhower administration.
+For such a device, what it actually does is less important than what the adversary understands it to be.
+A spy plane must garner the protection of espionage norms while performing far more aggressive missions than a traditional reconnaissance flight.
+To create such an aircraft , President Eisenhower approved Project Aquatone: the U-2.
+
+$H_2$ posits that states will pursue measures to signal that their activities are espionage, and the US did everything possible to ensure that the U-2 would be read as a civilian aircraft.
+Eisenhower required that the pilots of U-2 aircraft be CIA officers, specifically forbidding uniformed Air Force personnel.\footcite[p.~33, Though many of the pilots did have Air Force backgrounds.]{lindgren_trust_2000} The CIA also had operational control of the program instead of the Air Force.
 \enquote{I want this whole thing to be a civilian operation,} Eisenhower wrote, when settling an operational dispute between the two departments.
@@
 
-The steps Eisenhower took to ensure the U-2 civilian's character went beyond mere deception.
+The steps that Eisenhower took to ensure the U-2 civilian's character went beyond mere deception.
 If his only goal had been to achieve a level of plausible deniability in the event of a shootdown, then simply not wearing an Air Force uniform during the mission likely would have been sufficient.
-The classic \enquote{it's a weather reconnaissance craft} excuse had been used before with the ferret flights.\footcite[p.~45]{farquhar_aerial_2015} At the highest levels of American government, however, policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation.\enquote{It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft} read a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} Both halves of that equation were equally important.
-The Russians must believe that these operations are peacetime civilian operations, and they must be correct about it.
+The classic \enquote{weather reconnaissance craft} cover had been used before with the ferret flights.\footcite[p.~45]{farquhar_aerial_2015} But policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation.
+\enquote{It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft} reads a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} No finite list of attributes could ensure that the Soviet Union wouldn't treat the U-2 plane---a massive leap in military surveillance technology---as a military development.
+But if the people in charge of Project Aquatone could convince \emph{themselves} that it was a civilian project, they stood a better chance of persuading the Soviet Union.
+Both halves of that equation were equally important: the Russians must believe that these operations are peacetime civilian operations, and they must be correct about it.
 
-At first, what made the U-2 flights \enquote{count} as espionage appears to be a deeply unsatisfying tautology---they counted as espionage because they were espionage!
-But Eisenhower's precautions were based on the United State's previous experience with reconnaissance flights.
-He had already been testing the limits of what the Soviet Union would be willing to accept; if he did everything possible to signal that this plane functioned like a spy, perhaps it would accept this new development as well.
-No finite list of attributes could ensure that the Soviet Union wouldn't treat the U-2 plane---a massive leap in military surveillance technology---as a military development.
-But if the people in charge of Project Aquatone could convince \emph{themselves} that it was a civilian project, they stood a better chance of persuading the Russians to interpret it the same way.
-The U-2 simply had to be a spy plane, in a manner both instantly recognizable and ontologically undefinable.
+These precautions had the intended effect, even though it initially seemed like the Soviets were attempting to extract significant concessions.
+A Soviet colonel in the USSR Air Defense later recalled that Khrushchev ``clearly viewed the violation of their nation's skies by a foreign reconnaissance aircraft \textelp{} as a political provocation,'' and yet few real consequences followed.\footcite{orlov_u-2_2007} No one disputes that the collapse of the Paris Summit was directly attributable to the Gary Powers shootdown but, in practice, summits are not where important issues of the day get resolved.\footnote{The main issue that did not get resolved in 1960 was the status of Berlin.
+It also did not get resolved at the next summit in 1961, a comparatively warmer affair between Khrushchev and Eisenhower's successor, John F.
+Kennedy.} From 1953 to the end of the Cold War in 1991, there were 23 total US-USSR summits.\footcite{fain_chronology_2011} With the surprise rollout of the captured pilot, the dramatic scene at the summit, and the cancellation of Eisenhower's visit, Khrushchev publicly embarrassed the United States without meaningfully altering Soviet policy towards it.
 
-These precautions had the intended effect, even though it initially seemed like the Soviets were going to extract significant concessions.
-A Soviet colonel in the USSR Air Defense later recalled that Khrushchev ``clearly viewed the violation of their nation's skies by a foreign reconnaissance aircraft \textelp{} as a political provocation.''\footcite{orlov_u-2_2007} Three days after NASA quietly reported that they had lost a U-2 type weather reconnaissance plane, Khrushchev announced that the USSR had shot down an American plane that had flown into its airspace in \enquote{an aggressive provocation.}\footcite[p.~112]{powers_operation_2004} When the Americans stuck to the \emph{weather craft} cover story, Khrushchev gleefully debunked it by revealing that Powers had been taken alive.
-In a press conference where Eisenhower memorably declared that espionage was a \enquote{a distasteful but vital necessity} to prevent another attack like Pearl Harbor, the President admitted that the Gary Powers had been flying a spy plane.\footcite{eisenhower_news_1960}
-
-And yet few real consequences followed.
-No one disputes that the collapse of the Paris Summit was directly attributable to the Gary Powers shootdown but, in practice, summits aren't all that important.
-From 1953 to the end of the Cold War in 1991, there were 23 total US-USSR summits.\footcite{fain_chronology_2011} The main issue that did not get resolved in 1960 was the status of Berlin.
-It also did not get resolved at the next summit in 1961, a comparatively warmer affair between Khrushchev and Eisenhower's successor, John F.
-Kennedy.
-If the worst that happened because of the U-2 incident was an unproductive summit meeting, then that is a relatively minor punishment.
-With the late rollout of the captured pilot, the dramatic scene at the summit, and the cancellation of Eisenhower's visit, Khrushchev publicly embarrassed the United States without meaningfully altering Soviet policy towards it.
-
-Nor did the U-2 Incident end all reconnaissance flights.
+Nor did the U-2 Incident mark the end of reconnaissance flights.
 As is characteristic of espionage, when the US really needed intelligence that only an overflight could provide, the diplomatic consequences that the USSR had imposed were not sufficient to discourage it.
@@
 While Powers was serving a 10-year sentence in Soviet prison, the two airmen were returned to the US in less than six months with only mild fanfare.
-Khrushchev wrote in his memoir that he wanted to continue \enquote{our general line of peaceful coexistence} and cited the resolution of this incident as an example of such, because the US was forced to make a formal request for the return of the airmen.\footcite[p.~256-257]{khrushchev_memoirs_2007} The only concessions the US made in return were to announce the discontinuation of its U-2 overflights (which Kennedy was already committed to) and to not make an issue of the illegal detention of the pilots.\footcite{time_cold_1961} For spy flights after Powers, the USSR simply took the necessary immediate countermeasures and shot down the planes that it could.
-Just as with traditional espionage, the countermeasures did not punish the attempt.
-
-\subsubsection{Espionage norms in aerial reconnaissance}
-The U-2 Incident is often cited as a seminal failure of the Eisenhower administration.
-In the traditional telling, it significantly worsened US-Soviet relations at time when it seemed like lasting peace might be possible.
-This interpretation is incomplete, if not inaccurate.
-Instead, the U-2 Incident should be understood as an embarrassing but otherwise typical failure of an intelligence operation, one consistent with the minimal consequences typically applied to espionage.\footnote{While I argue that the response to the U-2 flight was within the reasonable bounds for espionage, I acknowledge that it was still unprecedented in the context of reconnaissance flights.
+Khrushchev wrote in his memoir that he wanted to continue \enquote{our general line of peaceful coexistence} and cited the resolution of this incident as an example of such, because the US was forced to make a formal request for the return of the airmen.\footcite[p.~256-257]{khrushchev_memoirs_2007} The only concessions the US made in return were to announce the discontinuation of its U-2 overflights (which Kennedy was already committed to) and to not make an issue of the illegal detention of the pilots.\footcite{time_cold_1961} The precedents from the Baltic Incident still applied: the USSR simply took the necessary immediate countermeasures and shot down the planes that it could.\footnote{While I argue that the response to the U-2 flight was within the reasonable bounds for espionage, I acknowledge that it was still unprecedented in the context of reconnaissance flights.
 There are a number of reasonable explanations for why the U-2 shootdown escalated.
@@
 Eisenhower thought that Khrushchev might have been looking for an excuse to cancel his visit to Russia.
-See: \cite[p.~555]{eisenhower_waging_1965}.
-Others in his administration believed that the U-2 Incident was an excuse to justify an intensification of hostilities that had already been decided by hard-liners in the Kremlin.
-See: \cite[p.~328]{kistiakowsky_scientist_1976}.} What would a major punishment have looked like?
+See: \cite[p.~555]{eisenhower_waging_1965}}.
+
+With respect to the practice of aerial reconnaissance, the U-2 Incident changed very little.
+Khrushchev greatly exaggerated the event's importance until the day he died.
+He wrote in his memoir that \enquote{the commander of American forces in West Germany gave the order not to fly any closer than 50 kilometers from the border between East and West Germany.
+And no more incidents of that kind occurred.}\footcite[p.~256]{khrushchev_memoirs_2007} His son Sergei, who annotated the memoir, admitted that this statement was false: \enquote{In practice such incidents occurred again from 1961 onward, but instead of the U-2 the Americans now used various types of Phantom or SR-71.}\footcite[p.~258]{khrushchev_memoirs_2007} On at least two occasions after the U-2 Incident, an American overflight crossed over into Soviet territory and the Soviets responded with a diplomatic protest.
+The US quickly apologized, and that settled the matter.\footcite{orlov_u-2_2007} The United States continued to use the U-2 extensively in other regions as well.\footnote{A recently declassified CIA report details the many locations in which U-2 photography was employed to gather intelligence after 1960, including China, India, Indonesia, Thailand, Tibet, Laos, North Vietnam, and Venezuela.
+Lyndon Johnson even clarified that his predecessor's order to end U-2 flights over the Soviet Bloc was not indefinitely binding---it was valid only until countermanded.
+For more, see: \cite[p.~195]{pedlow_central_1992}}
+
+% What would a major punishment have looked like?
 One proportional response Khrushchev could have considered was to press harder for the remove of regional air bases.
@@
 The Soviets were incredibly preoccupied with removing the forward-deployed Jupiter missiles from Turkey.
-\footnote{Although the missiles were deployed to bases well within NATO territory, and arguably less threatening to Soviet security than American overflights of the Soviet bloc, the existence of these missile became a central bargaining chip in the Cuban Missile Crisis.} By contrast, the Soviets simply did not have the same level of interest in curtailing reconnaissance flights.
-That both sides of the Cold War had strong incentives to avoid military escalation is well-documented, but at least threatening military retaliation for launching provocative military operations would not seem to have been out of place.
+\footnote{Although the missiles were deployed to bases well within NATO territory, and arguably less threatening to Soviet security than American overflights of the Soviet bloc, the existence of these missile
 
-Most scholars who study this period argue that the cumulative tension created by these flights contributed to an overall chilly tone in US-Soviet relations.
-Farquhar, for instance, refers to this as the \enquote{cycle of hostility,} in which the series of aerial incidents increased suspicions of military buildup on both sides, leading to even more aerial incidents as both powers attempted to verify their suspicions.\footcite[p.~43]{farquhar_aerial_2015} While it's true that the there were many aerial incidents, it is not at all clear how this affected any other aspect of US-Soviet diplomacy.
-Consider the period of time over which these flights took place; if aerial reconnaissance had risen to anything more than a lingering irritant, then it should have to have been resolved in some way before 1960.
-The US-Soviet relationship saw huge shifts, including periods of optimism, over the 11 years preceding the U-2 incident.
-During this period, hundreds of flights took place in contentious territory---many of them purposefully violating Soviet boundaries, causing some of them to be shot down---and the most significant consequence was a failed summit meeting, after which the flights continued on almost as before.
-Both sides were so willing to contain the issue that, tragically, many of the airmen never saw their families again.
-
-Until the day he died, Khrushchev greatly exaggerated the lasting effect of the U-2 Incident.
-He wrote in his memoir that \enquote{the commander of American forces in West Germany gave the order not to fly any closer than 50 kilometers from the border between East and West Germany.
-And no more incidents of that kind occurred.}\footcite[p.~256]{khrushchev_memoirs_2007} His son Sergei, who annotated the memoir, admitted that this statement was false: \enquote{In practice such incidents occurred again from 1961 onward, but instead of the U-2 the Americans now used various types of Phantom or SR-71.}\footcite[p.~258]{khrushchev_memoirs_2007} The United States continued to use the U-2 extensively in other regions as well.\footnote{A recently declassified CIA report details the many locations in which U-2 photography was employed to gather intelligence after 1960, including China, India, Indonesia, Thailand, Tibet, Laos, North Vietnam, and Venezuela.
-Lyndon Johnson even clarified that his predecessor's order to end U-2 flights over the Soviet Bloc was not indefinitely binding---it was valid only until countermanded.
-For more, see: \cite[p.~195]{pedlow_central_1992}}.
-On at least two occasions after the U-2 Incident, an American overflight crossed over into Soviet territory and the Soviets responded with...
-a diplomatic protest.
-The US quickly apologized, and that settled the matter.\footcite{orlov_u-2_2007}
+% Here's where we get into why Khruschev kinda sorta demonstrates evidence of H2
 
@@
 
+\subsubsection{Espionage norms in aerial reconnaissance}
+
 % Add slight emphasis here that the U-2 was built as a spy plane because it would have a certain reaction, and it worked
 
+In this section, we see direct evidence of $H_2$ in the actions of the Unite States and somewhat less direct evidence from the Soviet Union.
+Up to President Eisenhower, the United States placed a clear value on coding their actions as espionage, even when undertaking military reconnaissance missions of unprecedented daring.
+At the same time, American diplomats were willing to resolve disputes that arose from these missions quietly---even when loss of life was involved---which demonstrates the reduced consequences associated with espionage missions.
+While evidence of intent is hard to come by across the Iron Curtain, Khrushchev appears to have downplayed the incidents as well.
+The delta between his notes and the actual events suggests he was aware that he would have been expected to respond to these incursions more aggressively.
+The espionage coding was significant to the Soviet apparatus, even if is unlikely to find evidence admitting such.
+
+That salutary neglect had the intended effect.
 Project Aquatone channeled the norms of espionage in the most crucial respect: its strategic value easily outweighed the consequences of being caught.
@@
 
+In the aggregate, the period from 1947-1962 provides evidence of $H_1$ as well.
+The few scholars who study this period argue that the cumulative tension created by these flights contributed to an overall chilly tone in US-Soviet relations.
+Farquhar, for instance, refers to this as the \enquote{cycle of hostility,} in which the series of aerial incidents increased suspicions of military buildup on both sides, leading to even more aerial incidents as both powers attempted to verify their suspicions.\footcite[p.~43]{farquhar_aerial_2015} While it is true that the there were many aerial incidents, it is not at all clear how this affected any other aspect of US-Soviet diplomacy.
+The US-Soviet relationship saw huge shifts, including periods of optimism, over the 11 years preceding the U-2 incident.
+During this period, hundreds of flights took place in contentious territory---many of them purposefully violating Soviet boundaries, causing some of them to be shot down---and the most significant consequence was a failed summit meeting, after which the flights continued on almost as before.
+What ended this era of aerial reconnaissance wasn't diplomatic pressure---it was an advancement in espionage so massive that it made reconnaissance flights seem quaint.
+
 \subsection{Spy satellites}
-Gary Powers had been in Soviet prison for less than three months when the United States launched an instrument of espionage far more advanced, and far more damaging, than the U-2 could ever be.
-The first generation of spy satellite had limitless airspace access, no countermeasures, and no technological antecedents.
-It was impossible to predict how the the international order would react; in 1955, when the US and the USSR both announced their plans to build an artificial satellite, Outer Space was a true blank slate for international norms.
+Gary Powers had been in Soviet prison for less than three months when the United States launched an instrument of espionage far more damaging than the U-2 could ever be.
+The first generation of spy satellite had no technological antecedents, no airspace restrictions, and no countermeasures.
+\emph{Sputnik} famously put the USSR at the forefront of the final frontier, but the satellite itself was totally innocuous; it collected data about the density of the atmosphere and propagation of radio waves.\footcite{nasa_sputnik_2019} Five years after \emph{Sputnik}, the United States launched a satellite that threatened to reveal the lie at the heart of the Soviet defense posture---and the Soviet Union that allowed it to happen, again and again.
+Even after developing a feasible anti-satellite (ASAT) weapon, the Soviet Union never made use of it.
+Instead, reconnaissance satellites were allowed to perform their missions unimpeded by every nation on Earth, and they continue to do so today.
 
-Only four years after the first U-2 overflight, a familiar dynamic could have played out when the first spy satellite launched in August 1960: the US develops a new photo-reconnaissance technology which its main adversary is unable to prevent; the USSR diplomatically objects to the technology's use; the USSR finally develops a weapon capable of counteracting the new technology; and the USSR employs it.
-Each one of these steps came to pass except the last---even after developing a feasible anti-satellite (ASAT) weapon, the Soviet Union never made use of it.
-And the onus was on the Soviets to push back against American efforts to normalize satellite reconnaissance.
-\emph{Sputnik} briefly put the USSR at the forefront of the final frontier, but the satellite itself was totally innocuous; it collected data about the density of the atmosphere and propagation of radio waves.\footcite{nasa_sputnik_2019} It was the United States that would first launch a satellite capable of photo-reconnaissance---and it was the Soviet Union that allowed it to happen.
+% Only four years after the first U-2 overflight, a familiar dynamic could have played out when the first spy satellite launched in August 1960: the US develops a new photo-reconnaissance technology which its main adversary is unable to prevent; the USSR diplomatically objects to the technology's use; the USSR finally develops a weapon capable of counteracting the new technology; and the USSR employs it
+
+Consistent with $H_2$, the spy satellite exists because the United States undertook measures to signal that its new technology should be understood within the context of espionage.
+The scale of this effort dwarfed the espionage-coding measures applied to Project Aquatone: the United States developed its entire space program around the lone goal of protecting its use of satellites for intelligence-gathering.
+As a direct result, the Soviet Union never shot down a spy satellite, and it only objected to their use until it was able to make its own.
+By attempting to \enquote{mark} these satellites as espionage to shield their use from severe diplomatic consequences, the spy satellite programs of both the US and USSR support $H_2$; because it was successful, we can infer that $H_1$ holds as well.
+
+% By protecting the peaceful uses of outer space, the United States finally gained unrestricted access to aerial intelligence.
+That all satellites, regardless of purpose, are immune to hostilities as a result is just a happy byproduct.
 
 \subsubsection{Freedom of space}
-Spy satellites are not just one of the best edge cases for norm against shooting down a satellite, they are the reason it exists in the first place.
-The reconnaissance satellite succeeded in its mission because the Eisenhower administration purposefully designed its space program to establish a norm of peaceful development in space---with the single goal of legitimizing satellites for intelligence-gathering.
-No military action was ever been taken against satellites, and the Soviet Union only objected to reconnaissance satellites until it was able to make its own.
-By protecting the peaceful uses of outer space, the United States finally gained unrestricted access to aerial intelligence.
-That all satellites, regardless of purpose, are immune to hostilities as a result is just a happy byproduct.
-
+% Spy satellites are not just one of the best edge cases for norm against shooting down a satellite, they are the reason it exists in the first place.
 In 1954, President Eisenhower established the Technological Capabilities Panel (TCP), a task force which would tackle the problem of America's vulnerability to a surprise attack.
@@
 
-The U-2 plane made the leap from concept to reality much faster than the artificial satellite, but even before the Soviet Union became capable of shooting it down, the intelligence that the U-2 provided had limitations.
-While the U-2 provided CIA imagery analysts with impossibly high-resolution photos of specific locations, it was less helpful in finding unknown military bases throughout the vast expanse of Soviet territory.
-The first generation of artificial satellites promised to change that.
-In the time it takes a satellite to complete a full orbit, the Earth has rotated slightly underneath it, altering the satellite's field of view---with each pass an artifical satellite launched into polar orbit will photograph an area that has been longitudinally shifted.\footnote{Think of a winding a piece of string around a ball with your right hand, while slowly rotating the ball with your left; after a series of passes, you've covered a broad slice of the ball with string.} With enough orbits, that satellite can photograph a wide slice of Earth.
-Together, planes and satellites could operate for complementary photo-reconnaissance purposes: a plane dispatched when high quality localized intelligence was required, and a satellite to form a comprehensive picture of the Soviet military posture.
-And of course, satellites could surveil dangerous territory without risking the life of a pilot.
-
-Preventing a surprise nuclear attack could not be done without better intelligence and better intelligence could not be achieved without comprehensive photo-reconnaissance---reconnaissance performed using a technique that the Soviet Union would accept.
-The solution was to spy from space.
+The U-2 plane made the leap from concept to reality much faster than the artificial satellite, but the intelligence that it provided had limitations.
+The U-2 can take high-resolution photographs of specific locations---as long as you already knew where to look.
+By contrast, artificial satellites promised to find unknown military bases throughout the vast expanse of Soviet territory; with enough orbits, a satellite can photograph a wide slice of Earth.\footnote{In the time it takes a satellite to complete a full orbit, the Earth has rotated slightly underneath it, altering the satellite's field of view.
+With each pass an artifical satellite launched into polar orbit will photograph an area that has been longitudinally shifted.
+Imagine winding a piece of string around a ball with your right hand, while slowly rotating the ball with your left; after winding it a number of times, you've covered a broad slice of the ball with string.} Together, planes and satellites operate for complementary photo-reconnaissance purposes: a plane dispatched when high quality localized intelligence was required, and a satellite to form a comprehensive picture of the Soviet military posture.
+And of course, satellites can surveil dangerous territory without risking the life of a pilot.
 \enquote{Camera-toting satellites,} historian Walter McDougall explains, \enquote{circling the earth south to north [sic] in a polar orbit, could view the entire surface of the earth as it rotated below, return to any location in a few days' time, hone in on suspicious areas, and do it all under the legal cover of freedom of space---if such legal cover could be established.}\footcite[p.~117]{mcdougall_heavens_1985}
 
-That legal cover could be established was not a given.
+Establishing such legal cover was not a given.
 The Soviets considered U-2 overflights to be a violation of their territorial sovereignty, and continued to register diplomatic protests about them; the overflight program was so risky that it was originally intended to last only a year or two.\footcite[p.~33]{lindgren_trust_2000} The undefined legal status of space presented an opportunity, but also a challenge.

### 2020-10-20 article.tex
Make last edits before conclusion

@@
 
-In 2015, federal authorities discovered that malicious actors had for months maintained access to databases within the United States Office of Personnel Management (OPM).
-The hackers downloaded reams of sensitive data, including a cache of applications for federal security clearance.
-These applications exposed not just the personal information of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses.
-The breach ultimately compromised the information of 22.1 million people, including personnel at the highest levels of government.\footcite{nakashima_hacks_2015} US officials feared that the stolen data could help foreign intelligence services unmask undercover CIA officers, or identify American officials who might be susceptible to blackmail, or worse, recruitment.
+In 2015, federal authorities discovered a massive data breach inside the United States Office of Personnel Management (OPM).
+Hackers had, over the course of months, downloaded reams of sensitive data, including a cache of applications for federal security clearance.
+These applications exposed not just the personal information of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses---compromising the information of 22.1 million people in total.\footcite{nakashima_hacks_2015} US officials feared that the stolen data could help foreign intelligence services unmask undercover CIA officers, and the information could identify American officials who might be susceptible to blackmail, or worse, recruitment.
 \enquote{[I]f you have my [application for security clearance],} FBI Director James Comey admitted, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.
-Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015}.
-In the case of the OPM hack, a vivid imagination wasn't necessary; the US had already established that the Chinese military was likely behind the sophisticated attack.\footcite{koerner_inside_2016}.
-But even though it possessed conclusive digital forensics, the Obama administration refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018.
+Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015} In the case of the OPM hack, a vivid imagination wasn't necessary; the US had already established that the Chinese military was likely behind the sophisticated attack.\footcite{koerner_inside_2016} But even though it possessed conclusive digital forensics, the Obama administration refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018.
 See \cite{sanger_trump_2018}}
 
-The US government's tepid response to the OPM hack illustrates a crucial fact about the relationship between cybersecurity and espionage: when a cyberattack looks like a traditional intelligence operation, little to no diplomatic consequences will follow.
-Within both political and academic circles, this phenomenon is often rationalized as a logical extension of existing policy; cyber-espionage is treated like \enquote{traditional} espionage because they are both forms of espionage.
-This article seeks to unravel that tautology by first by explaining how states determine what constitutes \enquote{traditional} espionage and then by analyzing what strategic purpose that distinction serves.
-The practice of espionage in the digital realm has made this task increasingly urgent, as the internet offers new vectors of attack that magnify the potential of espionage to do harm, and governments re-evaluate whether to employ decades-old norms.
-Nations like Brazil are openly pushing for espionage to be regulated through larger regimes of Internet governance.\footcite[p.~464]{abdenur_triggering_2015} ``This is one of those cases where you have to ask, `Does the size of the operation change the nature of it?'\thinspace'', said one anonymous senior intelligence official of the OPM hack.
-\enquote{Clearly, it does.}\footcite{sanger_u.s._2015} For now, however, even operations on the scale of China's OPM infiltration still do not merit significant retaliation.
+The US government's tepid response to the OPM hack illustrates a potentially surprising aspect of \nth{21} century diplomacy: when a cyberattack looks like a traditional intelligence operation, little to no diplomatic consequences will follow.
+Within political and academic circles, this phenomenon is often taken for granted; cyber-espionage is treated like \enquote{traditional} espionage because they are both forms of espionage.
+The scale of the OPM hack, however, stretches our understanding of the kind of damage that espionage can achieve.
+``This is one of those cases where you have to ask, `Does the size of the operation change the nature of it?'\thinspace'' said one anonymous senior intelligence official.
+\enquote{Clearly, it does.}\footcite{sanger_u.s._2015} The actions of the US government say otherwise: at least for now, even operations on the scale of China's OPM infiltration still do not merit significant retaliation.
+
+To evaluate whether espionage norms should apply in cyberspace, one first needs to understand espionage—--specifically, why espionage has traditionally been met with few diplomatic consequences.
+Though not codified in US policy, the light touch applied to espionage is an open secret.
+A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].}\footnote{In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets.
+See \cite{finklea_cyber_2015}} For activities that fall under the umbrella of intelligence-gathering, counterintelligence rather than prosecution is considered the appropriate response.
+CIA Director John Brennan once described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, was particularly frank: \enquote{You have to kind of salute the Chinese for what they did [to OPM].
+If we had the opportunity to do that, I don’t think we’d hesitate for a minute.}\footcite{pepitone_clapper_2015} This permissive attitude is norm that governs espionage diplomacy, and its strategic purpose is rarely questioned or justified.
+
+% This article seeks to unravel that tautology by first by explaining how states determine what constitutes \enquote{traditional} espionage and then by analyzing what strategic purpose that distinction serves.
 
@@
 
-To evaluate each state's incentives I first establish a theoretical framework for analyzing the role that espionage plays in international relations (IR) and the diplomatic norms that are associated with it.
+To evaluate each state's incentive, I first establish a theoretical framework for analyzing the role that espionage plays in international relations (IR) and the diplomatic norms that are associated with it.
 Through this framework, I define the bounds of traditional espionage and why rational states might calculate that it is in their best interests to continue tolerating it.
@@
 
-\subsection{Espionage norms}
+\subsection{Defining espionage norms}
 There are few extant definitions of peacetime espionage, and no satisfying ones.
-The problem that aspiring espionage scholars inevitably encounter is that \enquote{peacetime espionage} is universally understood as a term of art but almost never defined by legal codes or interpreted by case law.
+The problem that aspiring espionage scholars inevitably encounter is that \enquote{peacetime espionage} is universally understood as a term of art but almost never defined by legal codes, or interpreted by case law.
 Most scholars who tackle this question settle on a definition like the one proposed by I\~{n}aki Navarrete and Russell Buchan: \enquote{\enquote{Peacetime espionage} is a colloquial rather than a legal term [used] to describe different methods of collecting confidential information from closed as opposed to open sources \textelp{} generally conducted by States, against States.}\footcite[p.~901-902]{navarrete_out_2019} This definition purposefully excludes forms of extralegal government intervention that are often conflated with espionage, such as assassinations, election interference, and covert operations.\footnote{Often papers won't take a stand on defining espionage, but rather offer a working definition just to bound the scope of that particular paper.
-See \cite[p.~600]{radsan_unresolved_2007}, \cite[p.~2]{kapp_spying_2007})} Excluding these activities is important; because assassinations and similar covert operations are not considered espionage, they do not typically receive the protections of espionage norms.
+See \cite[p.~600]{radsan_unresolved_2007}, \cite[p.~2]{kapp_spying_2007})} Excluding these activities is important: assassinations and similar covert operations are not considered espionage, so they do not typically receive the protections of espionage norms.
 
-For the purposes of this paper, I define espionage \emph{norms} as the common understanding of how states are expected to respond to activities that remain within the bounds of espionage.\footnote{There are, of course, other \enquote{norms} that involve espionage.
+For the purposes of this paper, I define espionage norms as the common understanding of how states are expected to respond to activities that remain within the bounds of espionage.\footnote{There are, of course, other \enquote{norms} that involve espionage.
 The so-called \enquote{Moscow Rules} dictate an intelligence officer are supposed to conduct themselves in hostile territory.
@@
 
-Once again, we can use cyber-espionage as an analogue for traditional espionage.
-Though not codified in US policy, the informal regime of espionage norms is an open secret.
-A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].}\footnote{In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets.
-See \cite{finklea_cyber_2015}} For activities that fall under the umbrella of intelligence-gathering, counterintelligence rather than prosecution is considered the appropriate response.
-CIA Director John Brennan once described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, was particularly frank: \enquote{You have to kind of salute the Chinese for what they did [to OPM].
-If we had the opportunity to do that, I don’t think we’d hesitate for a minute.}\footcite{pepitone_clapper_2015}
-
-There is solid data to support the presumption that cyber-espionage is routinely treated as a lesser category of offense by all nations, not just the United States.
-Of the hundreds of state-sponsored cyberattacks known to the public, attacks classified as espionage are by far the most common.
-By percentages, however, cyber-espionage is by far the least likely form of cyberattack to garner an official response from the victim (Table \ref{response-pct}).\footnote{These numbers were compiled from the Council on Foreign Relations' Cyber Operations Tracker, which is a publicly-sourced record of state-sponsored cyberattacks, updated quarterly.
-Of the 323 listed incidents, 38 were excluded from this calculation because they lacked an identifiable Victim, Sponsor, or Type.
-For more, see \cite{council_on_foreign_relations_new_2019}}
+The limits of espionage norms are defined by what intelligence professionals call \enquote{the rules of the game.} When states do impose harsher penalties, they often justify it by saying that they are punishing an operation that did not abide by the rules.
+The poisoning of Sergei Skripal, allegedly by Russian operatives, was widely described a violation of the rules.\footcite{masters_has_2018} Assassinations are off-limits, especially against former spies who have been traded to safety, and the US and the EU have issued multiple rounds of sanctions in response.\footcite{reuters_e.u._2019} Assassinations, however, are not the typical intelligence operation, and when an operation is understood by both parties to have been routine espionage, the consequences are minimized accordingly.
 
@@
 
+While most of the information that we have on espionage norms is qualitative, there is some data to support the limited consequences of espionage norms as well.
+Once again we can use cyber-espionage as an analogue for traditional espionage.
+Of the hundreds of state-sponsored cyberattacks known to the public, attacks classified as espionage are by far the most common.
+Cyber-espionage is, however, easily the least likely form of cyberattack to garner an official response from the attacked state (Table \ref{response-pct}).\footnote{These numbers were compiled from the Council on Foreign Relations' Cyber Operations Tracker, which is a publicly-sourced record of state-sponsored cyberattacks, updated quarterly.
+Of the 323 listed incidents, 38 were excluded from this calculation because they lacked an identifiable Victim, Sponsor, or Type.
+For more, see \cite{council_on_foreign_relations_new_2019}} In the world of cyber-incidents, if it looks like espionage then it is espionage, and it will receive the customary diplomatic benefits associated with that characterization.\footnote{In programming, this is a principle called \enquote{Duck Typing.} If it looks like a duck, and it quacks like a duck, then it is a duck.}
+
 % The benefits of espionage have never been underrated---\enquote{it is only the enlightened ruler and the wise general,} reads Sun Tzu, \enquote{who will use the highest intelligence of the army for purposes of spying and thereby they achieve great results}---but the inevitability of espionage has long been taken for granted.
@@
 
-\subsection{Existing literature}
+\subsection{Extant explanations for espionage norms}
 Prior literature has attempted to explain espionage norms using international law, but in the absence of any significant treaties governing peacetime espionage, scholars are often left without definitive answers.\footnote{Because it is universally agreed that international law does not deal with peacetime espionage directly, many journal articles that deal with this question frame their work almost Socratically, devoting two sections to cases for and against the legality of espionage.
@@
 John Radsan employs a more philosophical approach.
-Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage generally reach the same conclusion: it would be almost impossible to argue that espionage is legal, but the international system happily leaves this tension in a state of salutary neglect.
+Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage generally reach the same conclusion: it would be almost impossible to argue that espionage is legal, but the international system leaves this tension in a state of salutary neglect.
 Taken to its logical endpoint, this branch of scholarship would simply argue that the espionage norms we observe today result from simple inertia.
@@
 His analysis is limited, however, by its focus on enabling specific policy outcomes, like treaties.
-It doesn't resolve the gap between the absolute illegality of espionage in domestic law and its widespread practice internationally.
+It doesn't resolve the gap between the absolute illegality of espionage in domestic law and its widespread practice internationally, nor does address why espionage would be tolerated in cases where no tangible outcome, like a treaty, is forthcoming.
 Responding to Baker, Radsan notes that one-off deals and treaties \enquote{are a very slow and indirect way to achieve international consensus on the legality of espionage.
@@
 Each time a new espionage technology is made available, policymakers are given a fresh opportunity to decide whether or not the norms of espionage ought to extend to its use.
-Critically, the research that follows will demonstrate that they were not obligated to do so by any pre-existing norms.
-When the US first introduced photoreconnaissance satellites, for example, it was not clear whether the Soviet Union would treat the satellites as spies or as weapons.
+The research that follows will demonstrate that they were not obligated to do so by any pre-existing norms.
+When the US first introduced photo-reconnaissance satellites, for example, it was not clear whether the Soviet Union would treat the satellites as spies or as weapons.
 Ultimately, the infrastructure of what we now call \enquote{spy satellites} was developed, and the norms of espionage prevailed to protect their use.
@@
 In recent years China has relentlessly targeted American military secrets.
-A United States Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} \enquote{Long-term, US future military advantage is being diminished by years of IP exfiltration from [US military agencies],} the report read, \enquote{all with little to no adverse consequences to the thieves.}\footcite[p.~6]{bayer_cybersecurity_2019} The forecasts were dire: \enquote{If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.}\footcite[p.~5]{bayer_cybersecurity_2019} The US Navy is not a disinterested party, of course, and we should be cautious about its conclusions; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see: \cite{gilli_why_2019}} Nonetheless, espionage presents alarming opportunities for rival states to gain a comparative advantage and the consequences for taking advantage of them remain minimal.
+A United States Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} \enquote{Long-term, US future military advantage is being diminished by years of IP exfiltration from [US military agencies], all with little to no adverse consequences to the thieves.}\footcite[p.~6]{bayer_cybersecurity_2019} The forecasts were dire: \enquote{If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.}\footcite[p.~5]{bayer_cybersecurity_2019} The US Navy is not a disinterested party, of course, and we should be cautious about its conclusions; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see: \cite{gilli_why_2019}} Nonetheless, espionage presents alarming opportunities for rival states to gain a comparative advantage and the consequences for taking advantage of them remain minimal.
 
@@
 Were it more common to respond to espionage by imposing financial sanctions or even inflicting a military cost, a state would have to consider the possibility that attempting espionage could adversely impact other aspects of its security, leaving it worse off than when it started.
-States often need to make these calculations in other high-stakes arenas: pursuing a nuclear weapons program, for instance, or pursuing protectionist trade policies, could lead to severe economic retaliation.
+States often need to make these calculations in other high-stakes arenas; pursuing a nuclear weapons program or protectionist trade policies could lead to severe economic retaliation.
 Instead, when an especially damaging espionage operation is uncovered, the traditional punishment is to declare certain embassy officials \emph{persona non grata}, ejecting them from the country.
@@
 Losing intelligence officers in an embassy temporarily makes it more difficult to conduct espionage, but it doesn't make a state less likely to try.
-According to John Sipher, former head of the CIA's Russia operations, Western expulsions have never crippled Russian intelligence collection.
+According to John Sipher, former head of the CIA's Russia operations, previous Western expulsions have never crippled Russian intelligence collection.
 \enquote{If you have 150-200 intelligence officers in-country, losing 50 is painful, but hardly debilitating.} \footcite{dettmer_united_2018}
@@
 If removing a quarter of the officers doesn't do lasting damage, then why not remove the full 200?
-The swift expulsion of all suspected intelligence officers from a country---and the inevitable removal of the country's own intelligence officers in retaliation---would deeply injure both their espionage capabilities indefinitely.\footnote{Nothing on this scale has ever been attempted, but Operation Foot, the largest diplomatic expulsion in the Cold War, \enquote{was a blow from which the Soviet intelligence effort in the UK never recovered.} For more, see: \cite{hughes_giving_2006-1}} This nuclear option is always available to both parties and yet, in modern history, it has never been exercised.
+The swift expulsion of all suspected intelligence officers from a country---and the inevitable removal of the country's own intelligence officers in retaliation---would deeply injure both their espionage capabilities indefinitely.\footnote{Nothing on this scale has ever been attempted, but Operation Foot, the largest diplomatic expulsion in the Cold War, expelled 105 Soviet diplomats from the UK.
+While still not the total shutdown proposed in this section, Operation Foot \enquote{was a blow from which the Soviet intelligence effort in the UK never recovered.} For more, see: \cite{hughes_giving_2006-1}} This nuclear option is always available to both parties and yet, in modern history, it has never been exercised.
 We can reasonably infer that each nation's leadership is making an identical calculation: the benefits of continuing to practice espionage in my adversary's backyard outweigh the risks associated with allowing my adversary to do the same to me.
@@
 I propose an iterated decision-making model to describe how states cooperate to preserve the liminal legal status of espionage, based on the same tit-for-tat strategy associated with diplomatic expulsions.
-In game theory, tit-for-tat is highly-successful strategy wherein the player simply mimics the choice that that their opponent made in the last round.
-By employing this strategy, the player never gets taken advantage of for more than one turn and can always return to cooperation if their opponent unilaterally signals a willingness to do the same.
+In game theory, tit-for-tat is highly-successful strategy wherein the player simply mimics the choice that that their opponent made in the last round.\footcite[p.~8]{axelrod_effective_1980} By employing this strategy, the player never gets taken advantage of for more than one turn and can always return to cooperation if their opponent unilaterally signals a willingness to do the same.
 Diplomatic expulsions play out in exactly the same fashion: when one country expels the diplomats of another, the other country will often respond in kind.
@@
 My model changes the parameters so that the decision to cooperate or defect is based not on the amount of punishment received, but rather the type.
-Each time a state is forced to respond to a successful intelligence operation by a rival---such as the reveal of a high-level spy or a major data breach---it can either limit its response to one which is traditionally associated with espionage, or it can impose unprecedented costs that might discourage such an operation in the future.
-For instance, it might choose to impose strict financial sanctions or expel every suspected intelligence officer from its embassy, both of which would be considered punishments well in excess of the norm.
-Table \ref{espionage-decision-matrix} describes a single instance of this choice.
+Each time a state is forced to respond to a successful intelligence operation by a rival---such as the reveal of a high-level spy or a major data breach---it can either limit its response to one which is traditionally associated with espionage, or it can impose unprecedented costs that might discourage such an operation in the future (Table \ref{espionage-decision-matrix}).
+If the state chose to impose strict financial sanctions or expel every suspected intelligence officer from its embassy, for instance, both of those would be considered punishments well in excess of the norm.
 In the status quo, State 1 signals its interest in cooperating by limiting its retaliation in response to espionage, no matter how damaging the operation is to their national security.
@@
 
-While there is unquestionably a cultural norm informing espionage norms, tradition cannot fully explain why every state remains in a state of mutual cooperation.
+While there is unquestionably a cultural aspect to espionage norms, tradition cannot fully explain why every state remains in a state of mutual cooperation.
 Espionage consequences are an iterated game that both parties play tit-for-tat, so, in the long run, the only two possible equilibria are mutual defection or mutual cooperation.
@@
 Consider a hypothetical security seeker interacting with its adversary, which it believes to be a security-seeking state as well.
-If the first state feels insecure, it will build up arms to increase its security.
+If the first state feels insecure, it will build up armaments to increase its security.
 The rival state, seeing that its adversary is building up arms, now feels less secure and builds up its own arsenal in response, a classic security dilemma.
@@
 Like espionage norms, competition between security-seeking states has two natural equilibria: mutual cooperation and mutual defection.
-The ideal outcome for both is when neither state builds up its armaments---with their reciprocal lack of ability to deal damage, both states are confident that no war will result.
+The ideal outcome for both is when neither state builds up arms---with their reciprocal lack of ability to deal damage, both states are confident that no war will result.
 Unilateral build-up leads to short-term relative gains in security, but in the next turn the other state will doubtless respond with a defection.
@@
 States take the risk that espionage will reveal more than just a state's (hopefully benign) intent---it might also reveal a state's capabilities and strategy.
-For example, I interviewed senior Obama Administration official Jake Sullivan about the benefits of allowing foreign states to spy on the US.
+I interviewed senior Obama Administration official Jake Sullivan about the benefits of allowing foreign states to spy on the US.
 According to him, the United States would have no problem with adversaries being able to independently verify American intent because, as he sees it, American intent is benign.
@@
 And under almost no circumstances would the US want to have its strategies and warfighting plans revealed, so every measure should be taken to ensure their secrecy.\footcite{sullivan_personal_2019} States might prefer to reveal only the \enquote{good} information, but they can't.
-Espionage is by definition illicit---the fact the information was obtained without the target's permission lends it credibility.
+Espionage is by definition illicit; the fact the information was obtained without the target's permission lends it credibility.
 
@@
 We can translate that conclusion to a payoff matrix (for two security-seeking states) that looks almost exactly like the  security dilemma Stag Hunt: the greatest rewards for both are when they cooperate.
-Non-reciprocal espionage would generate significant relative advantages, but wouldn't confer the benefit of easing an adversary's fears.
-And the worst situation for both is when neither allows access to the other's information.
-Yes, that equilibrium minimizes the risk of losing of military secrets, but it significantly heightens the potential for conflict and miscalculation.
+Non-reciprocal espionage would generate significant relative advantages, but wouldn't confer the benefit of easing an adversary's fears, and neither side would allow it to last.
+The worst situation for both is when neither allows access to the other's information---that equilibrium minimizes the risk of losing of military secrets, but it significantly heightens the potential for conflict and miscalculation.
 
@@
 
-\subsection{What about greed?}
+\subsection{Assumptions and alternate explanations}
 The analysis in this section relies on one crucial assumption: that the states in question are security-seekers.
 Analyses of the risks of espionage must contend with possibility that one of the states in question is \enquote{fundamentally dissatisfied with the status quo, desiring additional territory even when it is not required for security}---a greedy state.\footcite[p.~4]{glaser_rational_2010} Such a state would obviously want to protect both its motives and its plans, and therefore derive no benefit from having espionage conducted against it.
-Nonetheless, we know that essentially every major nation-state abides by the accepted norms of espionage and I have already addressed the possibility that states---including greedy states---are bound to these norms for reasons that extend beyond rational calculus, and do not consider that explanation to be sufficient.
+Nonetheless, we know that essentially every major nation-state abides by the accepted norms of espionage.
+I have already addressed the possibility that states---including greedy states---are bound to these norms for reasons that extend beyond rational calculus, and do not consider that explanation to be sufficient.
 To fully evaluate the implications of greedy states on this theory of mutual cooperation is beyond the scope of this paper, but I will propose potential answers to the question of why greedy states would tolerate espionage as a starting point for future analysis.
@@
 
-The international community is likely to view a sudden withdrawal from international espionage norms as suspicious.
+The international community is likely to view a sudden withdrawal from universal espionage norms as suspicious.
 States that are farther along on the greed spectrum might decide that it is better to risk the chance that they'll be discovered as greedy states rather send a signal that will all but confirm it.
-And it cannot be discounted that sometimes states might not even be entirely confident of whether its \emph{own} intent is greedy.
+And it cannot be discounted that sometimes a state might not even be entirely confident of whether its \emph{own} intent is greedy.
 So far this paper has used the nation-state as its base level of analysis.
-As I present the empirical evidence for this theory however, it becomes necessary to analyze the motives of individual actors that comprise the state.
+As I present the empirical evidence for this theory however, it becomes necessary to analyze the motives of individual actors that make up the state.
 At this level, even officials within the same administration have conflicting ideas about how to use intelligence, and those collecting the intelligence can't be confident whose purposes they are serving.
 
-\subsection{Hypothesis}
+\subsection{Research design}
 In the following section, I will use empirical case studies to demonstrate how associating an operation with espionage alters its diplomatic response.
@@
 If this hypothesis holds, espionage may have some symbolic importance, but it isn't one that can be meaningfully tied to a diplomatic response.
-By contrast, here is the alternative hypothesis, the one that would prove true if the theory holds:
+By contrast, this is the alternative hypothesis, the one that would prove true if the theory holds:
 
@@
 Careful research could demonstrate that \enquote{espionage} is an ontologically significant concept, not just for academics, but for the very people who engage in it.
-If $H_2$ holds, one would find evidence of policymakers being willing to accept the cost of enemy espionage in order to practice espionage themselves based on the value of mutual observation.
-Furthermore, when $H_2$ is proven to be true bidirectionally, it implies $H_1$.
+If $H_2$ holds, one would find evidence of policymakers being willing to accept the cost of enemy espionage in order to practice espionage themselves, based on the value of mutual observation.
+Furthermore, when $H_2$ is proven bidirectionally, it implies $H_1$.
 Two states that are both maintaining reduced responses to espionage (while pursuing espionage themselves) create a relationship in which $H_1$ holds: state actions successfully associated with espionage will receive a reduced response.
-
-% Previous intelligence scholars have noted specific benefits of espionage that support the abstract theoretical justification I build here.
-Christopher Baker explains how espionage makes functional cooperation possible through treaties.\footcite[p.~6]{baker_tolerance_2004} Charles Glaser uses the 1972 Anti-Ballistic missile treaty as an example of precisely the type of signal that could lead states to resolve the security dilemma Julius Stone, a professor of international law, foresaw this exact role for espionage in 1962: \enquote{If you do not have a system of international inspection and if you can't get one (and I think it is quite likely that we can't) then the function which international inspection is supposed to serve still needs fulfilling.
-You may not be able to fulfill it to the optimum extent by reciprocally tolerated espionage, but you may be able to reduce the dangers.}\footcite[p.~41]{stone_legal_1962}
 
@@
 
-% Both planes and satellites played an under-explored role in the formulation of some of the Cold War's landmark diplomatic achievements.
-SALT I (the ABM Treaty) is widely considered to be the landmark diplomatic achievement of Nixon's \emph{d\'etente} policy, which set the stage for a broader easing of relations between the US and the USSR, and spy satellites were crucial to reducing the cost that each side would pay to sign it.
-Limiting ballistic missile defense systems was a relatively minor concession for both sides, but in the highly risk-averse context of the Cold War, the signal was significant.\footcite[p.~66]{glaser_rational_2010} Facing the omnipresent threat of nuclear annihilation, \emph{d\'etente} provided a rational strategic choice for both states---made possible by their reciprocal tolerance of espionage.
-
 \section{Case Studies}
-\subsection{Research Design}
-
-% I have yet to find a quote from Nikita Khrushchev in which he states that it was good for the Soviet Union to be have its state secrets compromised.
-I don't expect that I ever will.
-The realist approach to espionage is difficult to acknowledge politically; it cuts against national pride and power in order to reach a pragmatic, somewhat emotionally unsatisfying conclusion.
-I asked a senior national security policymaker, Jake Sullivan, whether he thought American espionage was a net positive for global security.
-``I wouldn't work for the American government if I did not believe that it was a net positive actor,'' he told me, \enquote{so I want the United States to be able to make the most informed decisions possible.} When I asked whether it promoted global stability for a foreign nation---Russia for instance---to spy on the United States, he qualified his answer.
-
-% He split the types of intelligence that foreign state might gather on the United States into three types: intent, capabilities, and strategy.
-\enquote{I have no problem with the Russians knowing that we do not have aggressive intent,} he told me.
-Capabilities, however, are more complicated.
-Some defense capabilities would probably be fine to have another country independently verify, and some we might want to keep secret, such as the atomic bomb before the end of the Second World War.
-And under almost no circumstances would a state want to have its strategies and warfighting plans revealed, so those should be kept as secret as possible.\footcite{sullivan_personal_2019} Without invoking the theory itself, Sullivan described exactly the type of intelligence that a defensive realist would need to solve the security dilemma.\footnote{At the point in my research when I conducted this interview, I had never even heard the term \enquote{defensive realism,} lest you think I led him to it.}
-
-% The problem is that a state will never be able to control what type of information is being leaked.
-Espionage is an all-or-nothing proposition; either a state is willing to risk compromising some government secrets---including the ones that it is always disadvantageous to leak---in order to preserve the right to attempt it itself, or the state defects and begins imposing severe diplomatic consequences for espionage, inevitably inviting the same consequences upon itself, at which point the information pipeline tightens drastically.
-Unfailingly, states will choose to continue cooperating, so they must all be making the same calculus that the information they lose is worth the security they gain from getting to practice espionage themselves.
-
-% If you believe that American intent is benign, then tacitly permitting espionage is one way out of the security dilemma.
-Sullivan and Glaser did not conclusively explain, however, why \emph{every} major power continues to cooperate.
-More likely than not, some states today are not pure security-seekers.
-Glaser calls these greedy states, and when a state's motives are mixed between greed and security, those get assigned the label of greedy as well.
-For a greedy state, having another state correctly evaluate its intent could prove disastrous.
-The state might be pretending to have a benign intent in order to gain an advantage and espionage might reveal the truth.
-
-% That is only partially true.
-I certainly believe that elements of espionage and its place in diplomacy are socially constructed.
-The practice of expelling diplomats, for instance, predates all the events discussed in this thesis, and almost certainly evolved from some sort of customary protocol, rather than a rational calculation about the most effective way of responding to espionage.
-
-% An institutional theory of politics cannot explain the endurance of espionage norms, because there are no international institutions dedicated to governing espionage.
-As the legal section notes, peacetime espionage in most traditional sense---activities associated with human spies---has virtually no legal recognition at all.
-The first institutional recognition of espionage for conflict-reduction purposes is the SALT I treaty in 1972, which protected \enquote{national means of verification,} but espionage-as-verification is limited to the specific treaties in which it is in included.
-Espionage is probably just too uncouth to formally acknowledge.
-
-% Likewise, the only legal regime that applied to a spy plane was the one that applied to planes in general, and that explicitly prohibited the U-2 overflights; the Treaty on Open Skies was signed in 1992, well after the events discussed here.
-The clear illegality of the U-2 overflights should have given the Soviet Union more leverage to gain concessions, not less.
-As for space, the Outer Space Treaty entered into effect in 1967, eight years after \emph{Discoverer 14}; the strange timing of that treaty, right as the USSR was developing ASAT capabilities, was noted in Chapter 4.
-
-% I posit that the desire from all states to continue the practice of espionage writ large, even as they seek to stop particular instances of it against themselves, has an important psychological component.
-Every policymaker would like to their decisions to be as well-informed as possible.
-Without the benefit of hindsight, it is very difficult objectively gauge how much \enquote{security} is lost from a successful intelligence operation.
-By contrast, it is very easy for someone in charge of making an important decision to notice how much better off they are for having had access to some illicitly-obtained information.
-I don't mean to suggest that these trade-offs are always impossible to evaluate.
-Khrushchev was keenly aware of how the satellite photography would affect his missile bluff; preserving the missile bluff was just not worth impacting Khrushchevs's own ability to make use of satellite photography.
-In situations where it is difficult to evaluate relative gains, there exists a strong psychological bias towards preserving ones own view of the situation, even if it means allowing the adversary to have its view as well.
-
-% If espionage often leads to mutually agreeable outcomes, then why is not just completely legal?
-Well, in some cases it is.
-A true Open Skies Treaty followed soon after the collapse of the Soviet Union.
-In just over a decade, spy satellites went from \enquote{espionage in space} to \enquote{national technical means of verification.} But even as espionage gained some legitimacy in statecraft, the vagueness of that wording, \enquote{means of verification,} is indicative of what keeps espionage from true public acknowledgment.
-The intelligence business is often looked at---not entirely unfairly---as being shady.
-Reframing espionage as treaty verification procedure can help mitigate that, but if formally acknowledging spy satellites feels uncouth, its hard to imagine that human spies will ever receive a legal regime of their own.
-
-% Regardless of theoretical explanation, the cooperative outcomes of the Cold War is remarkable.
-The USSR had reason to believe that it would have benefited from a mutual prohibition on the technology that the United States was using.
-The next two chapters will go in-depth on the process that led to this decision---from both sides---to make the case that espionage norms have a fundamental appeal that the United States was able to exploit for its own benefit, and induce the Soviet Union to cooperate.
-
-% I use defensive realism to formulate a theoretical foundation to explain the existence of espionage norms generally, and the Soviet Union's extension of espionage norms to spy satellites in particular.
-Between two security-seeking states, both states will be less secure in the event of an arms race, so preventing one is a mutually-agreeable outcome.
-Resolving the security dilemma requires that the states signal their benign intentions, which states are more likely to do if they have good intelligence---intelligence that informs them about their adversary's capabilities and intent.
-Therefore, cooperating to preserve the practice of espionage as a semi-legitimate state institution makes rational sense, both because of the benefits espionage offers the state doing the spying, and the benefits the same state derives \emph{from being spied upon}.
 
 \subsection{Aerial espionage}
-
-Just a few weeks before the \enquote{Four Powers Summit} in Summer 1960, Soviet Premeir Nikita Khrushchev announced that the USSR had shot down an American plane.\footcite[p.~112]{powers_operation_2004} The CIA knew immediately that it had lost a U-2, an advanced spy plane that had bedeviled the Soviets for years, proving beyond the reach of their surface-to-air missiles.\footcite{orlov_u-2_2007} Though clearly the technological gap had now closed, the CIA was still confident that its cover story would hold, thanks to a morbid fact of the U-2's design: in the event of a shootdown, it should have been impossible for the pilot to survive.\footcite[p.~35]{lindgren_trust_2000} When the United States put out at a statement that the plane was part of a \enquote{weather reconnaissance} mission gone awry, Khrushchev sprung his trap and revealed that the U-2 pilot, Gary Powers, had been taken alive.
+Just a few weeks before the \enquote{Four Powers Summit} in Summer 1960, Soviet Premeir Nikita Khrushchev announced that the USSR had shot down an American plane.\footcite[p.~112]{powers_operation_2004} The CIA knew tgat we was referring to the U-2, an advanced spy plane that had bedeviled the USSR for years, proving beyond the reach of Soviet surface-to-air missiles.\footcite{orlov_u-2_2007} Nevertheless, the CIA was still confident that its cover story would hold, thanks to a morbid fact of the U-2's design: in the event of a shootdown, it should have been impossible for the pilot to survive.\footcite[p.~35]{lindgren_trust_2000} When the United States put out at a statement that the plane was part of a \enquote{weather reconnaissance} mission gone awry, Khrushchev sprung his trap---he revealed that the U-2 pilot, Gary Powers, had been taken alive.
 
@@
 President Dwight D.
-Eisenhower admitted that the Gary Powers had been flying a spy plane, declaring in a chagrined press conference that espionage was ``a distasteful but vital necessity'' to prevent another attack like Pearl Harbor.\footcite{eisenhower_news_1960} At the planned summit, Khrushchev launched into a tirade about U-2 planes and walked out on the first day, effectively ending the summit.
-He also canceled a planned visit by US President Dwight Eisenhower to the USSR that was scheduled to take place the next month.
-A contemporary newsreel about the conference said that \enquote{in the course of two hours, Khrushchev brought US-Soviet relations to their lowest point since the end of World War II.}\footcite{universal_studios_summit_1960}
+Eisenhower admitted that the Gary Powers had been flying a spy plane, declaring in a chagrined press conference that espionage was ``a distasteful but vital necessity'' to prevent another attack like Pearl Harbor.\footcite{eisenhower_news_1960} At the summit, Khrushchev launched into a tirade about U-2 planes and walked out on the first day, effectively ending the summit.
+He also canceled a trip to the USSR that President Eisenhower was scheduled to take the next month.
+A contemporary newsreel about the conference announced that \enquote{in the course of two hours, Khrushchev brought US-Soviet relations to their lowest point since the end of World War II.}\footcite{universal_studios_summit_1960}
 
-In the traditional telling, the U-2 Incident is often cited as a major failure of the Eisenhower administration, which significantly worsened US-Soviet relations at time when lasting peace appeared possible.
+In the traditional telling, the U-2 Incident is cited as a major failure of the Eisenhower administration, one which significantly worsened US-Soviet relations at time when lasting peace appeared possible.
 Only in the under-studied context of Cold War reconnaissance flights does it become clear that the U-2 Incident actually shows both the US and USSR exhibiting restraint and sound judgment.
@@
 Nonetheless, a few aspects of the \emph{Cryptologic Quarterly} article suggest that the information is relatively complete.
-First, the article was written for a classified audience, likely given that designation so that it could be comprehensive, and declassified after the fact.
-Additionally, when the article was declassified 16 years later, it appears to have been declassified without redactions or missing pages.
+First, the article was written for a classified audience, likely given that designation so that it could be comprehensive, and declassified after 16 years.
+Additionally, it appears to have been declassified without redactions or missing pages.
 Lastly, if there had been another mission so highly classified that it was omitted from this document, it is difficult to imagine how diplomatic consequences could have been imposed with a degree of secrecy that would prevent us from knowing about it today.]{peterson_maybe_1993} In effect, the table catalogs the 13 times where the USSR shot down an American aircraft designed for collecting intelligence between 1950 and 1964---intercepted espionage attempts.\footcite[p.~5.
@@
 
-Peterson wrote the CIA's internal accounting for SIGNIT shootdowns, but he cannot tell us which missions were perceived as significant by the Soviet Union.
-For that, I cross-referenced the Peterson's list with historian Dr.
+Peterson's writings represent the CIA's internal accounting for SIGNIT shootdowns, but he does not comprehensively investigate how those various missions were perceived by the Soviet Union.
+For that, I cross-referenced the Peterson list with historian Dr.
 John Farquhar's analysis of Cold War shootdowns that had a diplomatic impact.\footcite[Dr.
@@
 The first of these shootdowns brought the aerial reconnaissance program to the attention of the American people on April 8, 1950.\footnote{Note that the flights analyzed here were not necessary overflights; many took place over international waters or dispute waters when they were shot down.} An unarmed Navy patrol plane was downed by Soviet fighters over the Baltic sea.
-The Soviet ambassador to the US lodged a formal note of protest.\footcite{kirk_ambassador_1950} A few days later, the American ambassador issued a formal response, claiming that at no point did the plane cross into any territorial waters.
-He demanded the Soviets conduct \enquote{a prompt and thorough investigation} and \enquote{see to it that those responsible for this action [were] promptly and severely punished.}\footcite{the_associated_press_text_1950} The Soviet Union recognized this for the toothless statement that it was, and rejected the demand in its entirety.\footcite{salisbury_kremlin_1950} No further action was taken.
+The Soviet ambassador to the US lodged a formal note of protest, and a few days later, the American ambassador issued a formal response, claiming that at no point did the plane cross into any territorial waters.\footcite{kirk_ambassador_1950} He demanded the Soviets conduct \enquote{a prompt and thorough investigation} and \enquote{see to it that those responsible for this action [were] promptly and severely punished.}\footcite{the_associated_press_text_1950} The Soviet Union recognized this for the toothless statement that it was, and rejected the demand in its entirety.\footcite{salisbury_kremlin_1950} No further action was taken.
 
-Consistent with $H_2$, the United States took steps to code this flight as espionage, which was successful not just in the eyes of the Soviet Union, but to the American press as well.
+Consistent with $H_2$, the United States took steps to code this flight as espionage, which was successful not just in the eyes of the Soviet Union, but in the American press as well.
 \emph{Washington Post} columnist Walter Lippmann wrote: \enquote{the known facts indicate that Soviet intelligence \textelp{} believed [the plane] was carrying important electronic equipment and that orders were given to the Soviet fighter command to intercept it.}\footcite{lippmann_baltic_1950} A \emph{Post} article written a few days later was almost sympathetic to the Soviet commanders.
@@
 The resolution of the Baltic Incident set a precedent for reconnaissance flights.
-From then on until 1960, the United States lost nine more aircraft to Soviet air defenses.
+From then on until 1960, the United States lost nine more aircraft to Soviet air defenses under similar circumstances.
 Each time, the pattern of responses was the same---each side released dueling statements of protest, then the situation continued more or less as before.
@@
 Legislatively, Truman asked Congress to pass a \$100 million Mutual Security Act to aid \enquote{selected persons residing in Soviet bloc states or refugees who wanted to form armed units} in opposition to Communism.}\footcite[p.~43]{farquhar_aerial_2015} Truman immediately took action against material Hungarian interests in the United States and threatened to empower \emph{armed dissidents} if the prisoners were not released.
-The contrast between this response and the one from just a year prior, when the US demanded \enquote{a prompt and thorough investigation} of the Baltic Incident, is stark.
+The US government's response to an aviation incident when no espionage was involved stands in sharp relief to the \enquote{prompt and thorough investigation} that it demanded after the Baltic Incident.
 
@@
 Cooke, a Air Force pilot assigned to the project.]{hall_early_2003} Project Slick Chick set up F-100 jets with rapid-fire 20mm cameras.\footcite[p.~176]{hall_early_2003} Though we know a lot less about these missions than we do about the ferret flights, the basic contours are familiar: missions with no weaponry, operating at a high level of secrecy, subject to countermeasures (tracked by radar and often chased by MiG fighters) but no evidence of diplomatic pressure.
-In meetings between the State department and high-level Soviet diplomats, the subject of captured pilots \enquote{was broached only perfunctorily in relation to other things being discussed.}\footcite[p.~72]{brugioni_eyes_2010} Some family remembers of lost airmen received posthumous awards, but most simply received the pilot's personal effects and no explanation.
+In meetings between the State department and high-level Soviet diplomats, the subject of captured pilots \enquote{was broached only perfunctorily in relation to other things being discussed.}\footcite[p.~72]{brugioni_eyes_2010} Some family members of lost airmen received posthumous awards, but most simply received the pilot's personal effects and no explanation.
 
@@
 \subsubsection{Inventing the spy plane}
-% Even the most aggressive overflights could only hope to skirt the borders of Soviet territory---the United States couldn't just simply fly a B-52 over Moscow.
-Despite the success of the ferret flights, the US intelligence community needed a way to photograph suspicious installations deep within Russia, one that would continue to benefit from espionage norms.
-What they needed was a spy plane.
-
 % For the spy plane to serve its intended purpose as a minimally provocative agent of espionage two things must be true: its operation must be understood by adversaries to be non-military, and that distinction must fulfill the promise of $H_1$ by meaningfully reducing the associated response.
 
-As its name suggests, a spy plane is the physical manifestation of a state's attempt to code aerial reconnaissance as espionage.
 The first recorded reconnaissance flight took place in 1911, but that was not the first spy plane.\footcite[Surveilling the Turkish infantry near Tripoli, \enquote{Lieut.
 Piazzi today, for the first time in the history of warfare, made from this place an aerial reconnaissance against a hostile power}]{special_cable_to_the_new_york_times_air_1911} Ontologically, the spy plane---distinct from a plane that is doing spying---was invented by the Eisenhower administration.
-For such a device, what it actually does is less important than what the adversary understands it to be.
-A spy plane must garner the protection of espionage norms while performing far more aggressive missions than a traditional reconnaissance flight.
-To create such an aircraft , President Eisenhower approved Project Aquatone: the U-2.
+As its name suggests, a spy plane is the physical manifestation of a state's attempt to code aerial reconnaissance as espionage.
+For such a device, what it actually does is less important than what the adversary understands it to be; a spy plane must garner the protection of espionage norms while performing far more aggressive missions than a traditional reconnaissance flight.
+To create this aircraft, President Eisenhower approved Project Aquatone: the U-2.
 
@@
 Eisenhower required that the pilots of U-2 aircraft be CIA officers, specifically forbidding uniformed Air Force personnel.\footcite[p.~33, Though many of the pilots did have Air Force backgrounds.]{lindgren_trust_2000} The CIA also had operational control of the program instead of the Air Force.
-\enquote{I want this whole thing to be a civilian operation,} Eisenhower wrote, when settling an operational dispute between the two departments.
+\enquote{I want this whole thing to be a civilian operation,} Eisenhower wrote, to settle an operational dispute between the two departments.
 \enquote{If uniformed personnel of the armed services of the United States fly over Russia, it is an act of war---legally---and I don't want any part of it.}\footcite[p.~60.
@@
 The classic \enquote{weather reconnaissance craft} cover had been used before with the ferret flights.\footcite[p.~45]{farquhar_aerial_2015} But policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation.
-\enquote{It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft} reads a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} No finite list of attributes could ensure that the Soviet Union wouldn't treat the U-2 plane---a massive leap in military surveillance technology---as a military development.
+\enquote{It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft} reads a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} No finite set of attributes could ensure that the Soviet Union wouldn't treat the U-2 plane---a massive leap in military surveillance technology---as a military development.
 But if the people in charge of Project Aquatone could convince \emph{themselves} that it was a civilian project, they stood a better chance of persuading the Soviet Union.
@@
 Eisenhower thought that Khrushchev might have been looking for an excuse to cancel his visit to Russia.
-See: \cite[p.~555]{eisenhower_waging_1965}}.
+See: \cite[p.~555]{eisenhower_waging_1965}}
 
@@
 
-% Add slight emphasis here that the U-2 was built as a spy plane because it would have a certain reaction, and it worked
-
 In this section, we see direct evidence of $H_2$ in the actions of the Unite States and somewhat less direct evidence from the Soviet Union.
-Up to President Eisenhower, the United States placed a clear value on coding their actions as espionage, even when undertaking military reconnaissance missions of unprecedented daring.
+All the way up to President Eisenhower, the United States placed a clear value on coding their actions as espionage, even when undertaking military reconnaissance missions of unprecedented daring.
 At the same time, American diplomats were willing to resolve disputes that arose from these missions quietly---even when loss of life was involved---which demonstrates the reduced consequences associated with espionage missions.
@@
 The delta between his notes and the actual events suggests he was aware that he would have been expected to respond to these incursions more aggressively.
-The espionage coding was significant to the Soviet apparatus, even if is unlikely to find evidence admitting such.
+and his son's notes confirm that.
+The espionage coding was significant to the Soviet apparatus, even if is unlikely to find primary sources that admit it.
 
@@
 \subsection{Spy satellites}
-Gary Powers had been in Soviet prison for less than three months when the United States launched an instrument of espionage far more damaging than the U-2 could ever be.
+Gary Powers had been in Soviet prison for less than three months when the United States and launched an instrument of espionage far more damaging than the U-2 could ever be.
 The first generation of spy satellite had no technological antecedents, no airspace restrictions, and no countermeasures.
-\emph{Sputnik} famously put the USSR at the forefront of the final frontier, but the satellite itself was totally innocuous; it collected data about the density of the atmosphere and propagation of radio waves.\footcite{nasa_sputnik_2019} Five years after \emph{Sputnik}, the United States launched a satellite that threatened to reveal the lie at the heart of the Soviet defense posture---and the Soviet Union that allowed it to happen, again and again.
+The Soviet defense posture in the early 1960s relied on the impression that the Soviet Union possessed more ballistic missile capabilities than it actually did; a spy satellite would quickly reveal that to be a lie.
+But when the first spy satellite launched in 1962, the diplomatic objections were minimal.
 Even after developing a feasible anti-satellite (ASAT) weapon, the Soviet Union never made use of it.
-Instead, reconnaissance satellites were allowed to perform their missions unimpeded by every nation on Earth, and they continue to do so today.
 
-% Only four years after the first U-2 overflight, a familiar dynamic could have played out when the first spy satellite launched in August 1960: the US develops a new photo-reconnaissance technology which its main adversary is unable to prevent; the USSR diplomatically objects to the technology's use; the USSR finally develops a weapon capable of counteracting the new technology; and the USSR employs it
-
-Consistent with $H_2$, the spy satellite exists because the United States undertook measures to signal that its new technology should be understood within the context of espionage.
-The scale of this effort dwarfed the espionage-coding measures applied to Project Aquatone: the United States developed its entire space program around the lone goal of protecting its use of satellites for intelligence-gathering.
-As a direct result, the Soviet Union never shot down a spy satellite, and it only objected to their use until it was able to make its own.
+The United States and the Soviet Union normalized the use of spy satellites in exactly the manner predicted by $H_2$.
+First, the United States took extraordinary measures to ensure that its new technology would understood within the context of espionage norms.
+These measures signaled that the US would be willing to tolerate a reciprocal form of espionage, as long as the Soviet Union was willing to do the same.
+And once the Soviet Union was able to make its own spy satellites, it ceased objecting to their use.
 By attempting to \enquote{mark} these satellites as espionage to shield their use from severe diplomatic consequences, the spy satellite programs of both the US and USSR support $H_2$; because it was successful, we can infer that $H_1$ holds as well.
 
-% By protecting the peaceful uses of outer space, the United States finally gained unrestricted access to aerial intelligence.
-That all satellites, regardless of purpose, are immune to hostilities as a result is just a happy byproduct.
+\subsubsection{Freedom of space}
 
-\subsubsection{Freedom of space}
-% Spy satellites are not just one of the best edge cases for norm against shooting down a satellite, they are the reason it exists in the first place.
-In 1954, President Eisenhower established the Technological Capabilities Panel (TCP), a task force which would tackle the problem of America's vulnerability to a surprise attack.
+In 1954, President Eisenhower established the Technological Capabilities Panel (TCP), a task force to tackle the problem of America's vulnerability to a surprise attack.
 When the panel returned with its results, its report concluded that any great power conflict would result in heavy American losses, so all necessary steps should be taken to avoid one.\footcite[p.~67.
-The TCP report is also variously referred to as the \enquote{Killian Report,} \enquote{Surprise Attack Study,} or its formal title, \enquote{Meeting the Threat of a Surprise Attack.}]{killian_sputnik_1977} Its recommendations obviously included furthering American missile capabilities---traditional deterrence through force---but intelligence-gathering capabilities took comparable priority.
-``We \emph{must} find ways to increase the number of hard facts upon which our intelligence estimates are based,'' the TCP Report states, ``to provide better strategic warning, to minimize surprise in the kind of attack, and to reduce the danger of gross overestimation or gross underestimation of the threat.''\footcite{technological_capabilities_panel_meeting_1955} The TCP recommendations---reviewed personally by President Eisenhower---were hugely consequential across US government policy, but two projects that came out of them are especially relevant here: the U-2 plane and the artificial spy satellite.
+The TCP report is also variously referred to as the \enquote{Killian Report,} \enquote{Surprise Attack Study,} or its formal title, \enquote{Meeting the Threat of a Surprise Attack.}]{killian_sputnik_1977} Its recommendations included traditional deterrence through force---furthering American missile capabilities---but intelligence-gathering capabilities were prioritized as well.
+``We \emph{must} find ways to increase the number of hard facts upon which our intelligence estimates are based,'' the TCP Report states, ``to provide better strategic warning, to minimize surprise in the kind of attack, and to reduce the danger of gross overestimation or gross underestimation of the threat.''\footcite{technological_capabilities_panel_meeting_1955} The TCP recommendations---personally reviewed by Eisenhower---included two projects foundational to the US reconnaissance regime: the U-2 plane and the artificial spy satellite.
 
-The U-2 plane made the leap from concept to reality much faster than the artificial satellite, but the intelligence that it provided had limitations.
-The U-2 can take high-resolution photographs of specific locations---as long as you already knew where to look.
-By contrast, artificial satellites promised to find unknown military bases throughout the vast expanse of Soviet territory; with enough orbits, a satellite can photograph a wide slice of Earth.\footnote{In the time it takes a satellite to complete a full orbit, the Earth has rotated slightly underneath it, altering the satellite's field of view.
-With each pass an artifical satellite launched into polar orbit will photograph an area that has been longitudinally shifted.
+The U-2 can take incredibly high-resolution photographs---as long as you already know where to look.
+A satellite can cover far more ground, photographing wide slices of the planet, uncovering new military installations or verifying that purported ones do not exist.\footnote{In the time it takes a satellite to complete a full orbit, the Earth has rotated slightly underneath it, altering the satellite's field of view.
+With each pass an artificial satellite launched into polar orbit will photograph an area that has been longitudinally shifted.
 Imagine winding a piece of string around a ball with your right hand, while slowly rotating the ball with your left; after winding it a number of times, you've covered a broad slice of the ball with string.} Together, planes and satellites operate for complementary photo-reconnaissance purposes: a plane dispatched when high quality localized intelligence was required, and a satellite to form a comprehensive picture of the Soviet military posture.
@@
 
-Establishing such legal cover was not a given.
-The Soviets considered U-2 overflights to be a violation of their territorial sovereignty, and continued to register diplomatic protests about them; the overflight program was so risky that it was originally intended to last only a year or two.\footcite[p.~33]{lindgren_trust_2000} The undefined legal status of space presented an opportunity, but also a challenge.
-The TCP understood that the Soviet Union would recognize the enormous advantage a photo-reconnaissance satellite conferred on the US, and object to its deployment using any international norms possible.
-It was therefore necessary to present reconnaissance satellites in a manner that would limit the grounds for objection.
-By the time that a reconnaissance satellite launched, it would have to have already been established that any nation had the unquestioned right to launch a satellite of any kind; this principle would come to be called \enquote{freedom of space.}
+Such legal cover was not guaranteed.
+The undefined legal status of space presented both an opportunity and a challenge.
+The Soviet Union would likely recognize the enormous strategic advantage of a photo-reconnaissance satellite and object to its deployment using any international norms possible; even the U-2 overflight program was considered so risky that it was originally intended to last only a year or two.\footcite[p.~33]{lindgren_trust_2000} It was therefore necessary to present reconnaissance satellites in a manner that would limit the grounds for objection.
+By the time that a reconnaissance satellite launched, it would have to have already been established that any nation had the unquestioned right to launch a satellite of any kind: a principle called \enquote{freedom of space.}
 
-\subsubsection{Space is not a race}
-Everything about the administration's space policy was carefully crafted to set the stage for the eventual launch of a spy satellite.
-From the time that artificial satellites first appeared viable until the day he left office, Eisenhower used the civilian space program to divert attention from the security-related space programs he valued.\footcite[p.~119]{day_eye_2015} Acting on the recommendations of the TCP, the National Security Council (NSC) proposed a scientific satellite program in May 1955, which Eisenhower accepted.
-In this memo, NSC-5520, the NSC Planning Board argued that, when framed properly, a scientific satellite could be used as diplomatic tool to promote a norm establishing ``freedom of space.'' The proposal tracked closely with a similar one from the 1950 RAND report, which suggested that the US first launch an innocuous ``experimental satellite'' that did not cross over Russian territory, and gauge the diplomatic reaction.\footcite[p.~21]{kecskemetic_satellite_1950} The political framing of the satellite is discussed at the top of NSC-5520, in a section entitled \emph{General Considerations} (emphasis mine):
+Eisenhower accepted the recommendations of NSC-5520, a memo arguing that a scientific satellite could be used as diplomatic tool to promote a norm establishing ``freedom of space.'' The proposal tracked closely with an earlier report suggesting that the US first launch an innocuous ``experimental satellite'' that did not cross over Russian territory, and gauge the diplomatic reaction.\footcite[p.~21]{kecskemetic_satellite_1950} The political framing of the satellite is discussed at the top of NSC-5520, in a section entitled \emph{General Considerations} (emphasis mine):
 \newline
@@
 6.
-\textelp{} \textbf{Furthermore, a small scientific satellite will provide a test of the principle of “Freedom of Space”.
-The implications of this principle are being studied within the Executive Branch.} However, preliminary studies indicate that there is no obstacle under international law to the launching of such a satellite.
+\textelp{} \textbf{Furthermore, a small scientific satellite will provide a test of the principle of “Freedom of Space.” The implications of this principle are being studied within the Executive Branch.} However, preliminary studies indicate that there is no obstacle under international law to the launching of such a satellite.
 
@@
 
-Even though there existed no international law prohibiting the use of satellites, a theoretical ``freedom of space'' principle depended on the first satellite not having military purposes.
-What NSC-5520 did not anticipate was that the US wouldn't have to establish that principle on their own.
-On October 4, 1957, the Soviet Union launched the world's first artificial satellite, a small metal ball named \emph{Sputnik}.
+The plan was straightforward: use the civilian space program to promote the freedom of space principle, which, once established, would protect the American right to quietly launch a spy satellite.\footcite[p.~119]{day_eye_2015} In fact, since the Soviet Union had already announced plans to launch a satellite, the existence of the Soviet project could be use to blunt the inevitable Soviet protests against American photo-reconnaissance satellites.\footcite[p.~120]{mcdougall_heavens_1985} When the Soviet Union launched the world's first artificial satellite on October 4, 1957, the plan succeeded.
+The \emph{Sputnik} launch put the United States in an unbeatable negotiating position.
+The Soviet Union could hardly denounce the American satellite program as suspicious when it had just launched the world's first scientific satellite---without asking permission to orbit over the United States.\footcite[p.~40]{peebles_corona_1997} By allowing \emph{Sputnik} to orbit over American territory without diplomatic objection, the US got to appear magnanimous while fulfilling one of its highest-priority strategic goals.
 
-Getting beaten to orbit by \emph{Sputnik} put the United States at an advantage.
-The Soviets could hardly denounce the American satellite program as suspicious when they very publicly made the first move---and without asking permission to orbit over the United States.\footcite[p.~40]{peebles_corona_1997} The US didn't \emph{try} to lose first leg of the space race exactly, but policymakers were aware well in advance that being the second nation to launch an artificial satellite would further their strategic goals.
-Reviewing NSC-5520, Nelson Rockefeller wrote that since the Soviets had already announced plans to launch a satellite, the existence of the Soviet project could be used to good effect against inevitable Soviet protests against American photo-reconnaissance satellites.\footcite[p.~120]{mcdougall_heavens_1985} Though ideally the US would set the precedent for ``freedom of space'' itself, senior administration officials understood that if the USSR launched first, then the American satellite program would be almost automatically justified.
+The available evidence of Eisenhower's internal deliberations overwhelmingly suggests that his administration never intended to be first in space.
+Launching the world's first artificial satellite was assigned the lowest possible priority because he determined that other goals---finishing the ICBM, monitoring Soviet R\&D, and framing the satellite program as a civilian scientific endeavor---were of greater strategic importance.\footcite[p.~123-124]{mcdougall_heavens_1985} Eisenhower himself repeatedly dismissed concerns about the prestige of being the first nation to orbit an object in space, openly contemptuous of the idea that space was any kind of race at all.\footcite[p.~100]{lindgren_trust_2000} His complete disdain for these ``stunts'' would prove disastrous politically, but his restraint was crucial to establishing a strategically beneficial precedent.\footcite[p.~134]{day_eye_2015}
 
-Not only did the Eisenhower administration recognize that being second to orbit came with a significant consolation prize, it was not particularly focused on winning the space race to begin with.
-Launching the world's first artificial satellite was assigned the lowest possible priority because the administration correctly determined that its other goals---finishing the ICBM, monitoring Soviet R\&D, and framing the satellite program as a civilian scientific endeavor---were of greater strategic importance.\footcite[p.~123-124]{mcdougall_heavens_1985} Eisenhower himself repeatedly dismissed concerns about the prestige of being the first nation to orbit an object in space, openly contemptuous of the idea that space was any kind of race at all.\footcite[p.~100]{lindgren_trust_2000} His complete disdain for what he called ``stunts'' would prove to be disastrous politically, but his restraint was crucial to establishing a strategically beneficial precedent.\footcite[p.~134]{day_eye_2015}
+Just as Eisenhower limited the involvement of the Air Force in Aquatone to paint the U-2 as a civilian project, his guidance led to technical decisions that, consistent with $H_2$, were intended to make the satellite program seem as non-military as possible.
+The group tasked with implementing NSC-5520 made the surprising choice to commission the unproven Navy Research Laboratory (NRL) \emph{Vanguard} rocket, instead of the Army Ballistic Missile Agency (ABMA) proposal to launch the satellite with better-tested missile technology.
+Launching the first artificial satellite with a ballistic missile threatened to compromise two national security priorities: it might delay the development of the ICBM, an existential threat to American military superiority, and it would jeopardize the civilian character of the satellite program, which the committee members knew was necessary to establish ``freedom of space.''\footcite[p.~122]{mcdougall_heavens_1985} Led by famous rocketeer (and ex-Nazi) Wernher von Braun, the ABMA engineers vigorously protested the decision, because they believed---correctly---that their proposal would succeed first.
+But being first in space was not the goal, establishing a legal precedent was.
+Launching a satellite with no scientific purpose other than demonstrating superiority would compromise the moral high ground which the US desperately needed to secure.\footcite[p.~129-131.
+Though the NRL is run by the Navy, it functioned like a civilian scientific operation.
+All parties involved believed that the NRL proposal would be seen as a civilian project---especially when compared with the Army agency building nuclear delivery systems.]{day_eye_2015}
 
-In fact, the overriding American priority to legitimize reconnaissance satellites is probably the \emph{reason} that the Soviets started the Space Age instead.
-Just as Eisenhower limited the involvement of the Air Force in Aquatone to paint the U-2 as a civilian project, he made technical decisions that were intended to make the scientific satellite seem as non-threatening as possible---decisions that directly impacted the project timeline.
-The group tasked with implementing NSC-5520 had to choose between competing proposals from the Army, Navy, and Air Force for which rocket would launch the satellite.
-It made the somewhat surprising choice to go with the unproven Navy Research Laboratory (NRL) \emph{Vanguard} rocket, even though the Army Ballistic Missile Agency (ABMA) engineers, headed by famous rocketeer (and ex-Nazi) Wernher von Braun, had been developing their proposal for a year.
-Among the reasons the committee gave for their decision, two stand out.
-First, the NRL impressed the committee with its plans for scientific components and a radio tracking system.\footcite[p.~122]{mcdougall_heavens_1985} Second, the panel specifically wanted to avoid using a ballistic missile to launch a scientific satellite.\footcite[p.~129.
-Though the NRL is run by the Navy, it functioned like a civilian scientific operation and all parties involved believed that the NRL proposal would be seen as a civilian project---especially in comparison with its competetor, the Army agency building nuclear delivery systems.]{day_eye_2015}
+Operationally, \emph{Vanguard} was a disaster.
+The rocket experienced an embarrassing setback on December 6, 1957, when it failed to launch a three-and-a-half pound satellite, just after the USSR had successfully launched \emph{Sputnik II}.\footcite[p.~119.
+This time, the satellite carried a dog named Laika.
+The Soviets put a dog in space before the Americans even managed to get a satellite up there.]{killian_sputnik_1977} When then United States eventually succeeded in launching a satellite, it was thanks to the ABMA team that had been shunned by the administration.
+Finally given the chance, Von Braun and his engineers quickly accomplished what they had always said they would: modify a Jupiter-C ballistic missile and send a satellite into orbit.
+The Army launched \emph{Explorer 1} on January 31, and it carried a pioneering array of miniaturized electronics, including two micrometeoroid detectors and a Geiger counter.\footcite[p.~168]{mcdougall_heavens_1985} \emph{Vanguard 1} finally launched soon after.
 
-Launching the first artificial satellite with a ballistic missile threatened to compromise both of the principal national security priorities.
-It might delay the development of the ICBM, an existential threat to American military superiority, and it would jeopardize the civilian character of the satellite program, which the committee members knew was necessary to establish ``freedom of space.''\footcite[p.~122]{mcdougall_heavens_1985} The Von Braun engineers vigorously protested the decision, because they believed---correctly---that their proposal would succeed first.
-But being first in space was not the goal, establishing a legal precedent was.
-Von Braun's appeals were continually rejected, and always for the same reason: launching a satellite with no scientific purpose other than demonstrating superiority would compromise the moral high ground which the US desperately needed to secure.\footcite[p.~131]{day_eye_2015}
-
-Publicly, \emph{Vanguard} was a disaster.
-The rocket experienced an embarrassing setback on December 6, 1957, when it failed to launch a three-and-a-half pound satellite, just after the USSR had successfully launched \emph{Sputnik II}.\footcite[p.~119]{killian_sputnik_1977} This time, the satellite carried a dog---the Soviets put a dog in space before the Americans even managed to get a satellite up there.\footcite[Her name was Laika.]{george_sad_2018} When then United States finally succeeded in launching at satellite, it was not thanks to the \emph{Vanguard}, but rather the ABMA team that had been originally shunned by the administration.
-Finally given the chance, Von Braun and his engineers quickly accomplished what they had always said they would: modify a Jupiter-C ballistic missile and send a satellite into orbit.
-The Army launched \emph{Explorer 1} on January 31, and it carried a pioneering array of miniaturized electronics, including two micrometeoroid detectors and a Geiger counter.\footcite[p.~168]{mcdougall_heavens_1985} \emph{Vanguard 1}---which finally launched about six weeks later---and \emph{Explorer 1} provided an impressive amount of scientific data in comparison to the Sputniks.\footcite[p.~168]{mcdougall_heavens_1985} More importantly, even though the US did end up launching its first satellite using a ballistic missile, its visible defeat in the space race and clear scientific aims had the intended effect of blunting Soviet efforts to characterize the American satellite program as a military development.
-It was time to launch a spy satellite.
+Choosing the NRL proposal ensured the public humiliation of the American space program---and the private victory of its espionage program.
+Applying espionage-coding measures pursuant to $H_2$ can be costly: had Eisenhower not enforced the overriding priority to minimize military involvement it is highly likely that the ABMA could have launched a satellite before \emph{Sputnik}.
+Even though the US did end up launching its first satellite using a ballistic missile, its visible defeat in the space race and clear scientific aims effectively thwarted Soviet efforts to characterize the American satellite program as a military development.
+\enquote{Freedom of space} had been established---it was time to launch a spy satellite.
 
 \subsubsection{The Corona Project}
-The Soviet Union unintentionally set an early precedent for ``freedom of space,'' but that did not automatically justify using space for reconnaissance.
-As far as the international community was concerned, it had only been established that overflights were permissible for scientific purposes; military reconnaissance was still an open question.\footcite[p.~47-48]{peebles_corona_1997} Now that the space age was underway, the next step was to normalize photographic satellite overflights.
+
+% Now that the space age was underway, the next step was to normalize photographic satellite overflights.
 Success was going to require tight discipline across all levels of government, but information about WS-117L, the Air Force reconnaissance satellite program, had already begun to leak.\footcite[p.~96]{lindgren_trust_2000}
 
-Like many of Eisenhower's space policy decisions, his solution for reconnaissance satellites was a brilliant long-term success---but a complete mystery to the public and utterly baffling to most of the military establishment.
-WS-117L was originally intended to be a family of reconnaissance satellites.
-Eisenhower simply selected the most immediately promising project---a single-use film-retrieval satellite, codenamed Corona---and canceled it.
-Air Force personnel were ``thunderstruck.''\footcite[p.~45]{peebles_corona_1997} With public bewilderment instilled and American reconnaissance satellites now seemingly a few years out, the administration began a multi-pronged approach to legitimizing spy satellites.
+The Soviet Union unintentionally set an early precedent for ``freedom of space,'' but that did not automatically justify the use of outer space for reconnaissance.
+As far as the international community was concerned, it had only been established that overflights were permissible for scientific purposes; military reconnaissance was still an open question.\footcite[p.~47-48]{peebles_corona_1997} Like many of Eisenhower's space policy decisions, his solution for reconnaissance satellites was a brilliant long-term success---but a complete mystery to the public and utterly baffling to most of the military establishment.
+Eisenhower simply selected the most immediately promising reconnaissance satellite---a single-use film-retrieval satellite, codenamed Corona---and canceled it.
+Air Force personnel were ``thunderstruck.''\footcite[p.~45]{peebles_corona_1997} Now that American reconnaissance satellites now seemingly a few years out, the administration began a multi-pronged approach to legitimizing spy satellites.
 Publicly, the US government advocated for a ``freedom of space'' policy that explicitly included reconnaissance satellites; secretly, it resurrected the Corona project under deep-cover CIA administration.
 
-The public half of Eisenhower's strategy required international approval for space-based overflights, so the US advocated for a United Nations Ad Hoc Committee on the Peaceful Uses of Outer Space (COPUOUS) to deal with this issue.
-The Soviets, worried about having the limitations of their ICBM program exposed, refused to participate for years.\footcite[p.~140]{day_eye_2015} Meanwhile, the actual spy satellite program progressed at an accelerated pace, and with unmatched secrecy.
+The public half of Eisenhower's strategy required international approval for space-based overflights, so the US advocated for a United Nations Ad Hoc Committee on the Peaceful Uses of Outer Space (COPUOS) to deal with this issue.
+The Soviets, worried about having the limitations of their ICBM program exposed, refused to participate for years.\footcite[p.~140]{day_eye_2015} Meanwhile, the actual spy satellite program progressed at an accelerated pace, and with unmatched secrecy.\footcite[p.~51.
 The team started with only 30 personnel and eventually swelled to 300, almost none of whom knew the entire scope of the project.
-Team members had to take different routes to work to avoid being followed and were never allowed to say the word ``Corona'' on the telephone, or even the abbreviation ``C.''\footcite[p.~51]{peebles_shadow_2000} As far as the government was concerned, Corona's cover was bulletproof.
-When the satellite began making test launches, the US heavily promoted its new \emph{Discoverer} program as a scientific achievement---confident that its status as a front for the Corona project could not be proven false the way the U-2's cover had been.
+Team members had to take different routes to work to avoid being followed and were never allowed to say the word ``Corona'' on the telephone, or even the abbreviation ``C.'']{peebles_shadow_2000} When the satellite began making test launches, the US heavily promoted its new \emph{Discoverer} program as a scientific achievement---confident that its status as a front for the Corona project could not be proven false the way the U-2's cover had been.
 When it finally succeeded in recovering a \emph{Discoverer} satellite from orbit, the US government released photographs of the first successful capsule triumphantly floating in the water, and an American flag retrieved from within was presented to the president.\footcite[p.~83]{peebles_shadow_2000}
@@
 And there were a great many launches.
-From February to August 1960, right in the middle of the U-2 incident, twelve test launches failed before the US achieved the first successful recovery of a Corona satellite---of any object from orbit in the human history, for that matter---when it fished \emph{Discoverer 13} out of the sea.
-The US repeated that feat with \emph{Discoverer 14} a week later, and this time it carried a camera, bringing invaluable intelligence back from space that dealt a mortal blow to the myth of the missile gap.\footcite[p.~101-102.
-The previous failures had been so demoralizing that \emph{Discoverer 13} only carried diagnostic equipment, which made its success all the more exciting.
-Ships on standby were able to retrieve the satellite even though it fell outside the recovery zone---\emph{Discoverer 13} many not have landed in the ballpark or the outfield, but in many respects, the mission was a home run.]{lindgren_trust_2000} The launches kept coming---the next one to return high quality photos was \emph{Discoverer 18} in December.
+From February to August 1960, right in the middle of the U-2 incident, twelve test launches failed before the US achieved the first successful recovery of a Corona satellite---of any object from orbit in the human history, for that matter---when it fished \emph{Discoverer 13} out of the sea.\footnote{Ships on standby were able to retrieve \emph{Discoverer 13} even though it fell outside the ballpark.
+Diplomatically and operationally, the mission was a home run.} The US repeated that feat with \emph{Discoverer 14} a week later, and this time it carried a camera, bringing invaluable intelligence back from space that dealt a mortal blow to the myth of the missile gap.\footcite[p.~101-102.
+The previous launch failures had been so demoralizing that \emph{Discoverer 13} only carried diagnostic equipment, no cameras.]{lindgren_trust_2000} The launches kept coming---the next one to return high quality photos was \emph{Discoverer 18} in December.
 Eisenhower left office soon after, but between \emph{Discoverer 18} and \emph{Discoverer 38}, twenty launches were performed, two of which carried actual radiometry equipment (in support of Corona's cover story), and seven produced film of high enough quality for the mission to be considered a complete success.\footcite[p.~103]{lindgren_trust_2000}
 
-Like the U-2 overflights, this intricate process was immediately identified by the Soviets for what it was: a military reconnaissance effort.
-The Soviet Union protested the very first \emph{Discoverer} launch, correctly alleging its military utility even though the satellite had failed in its purpose, and would fail eleven more times.\footcite[p.~140]{day_eye_2015} Until 1963, Soviet propaganda continually railed against the Americans, in domestic media and international forums, for attempting to militarize space.\footcite[p.~271]{mcdougall_heavens_1985} The USSR sponsored a UN resolution to prohibit espionage from space, and Khrushchev vowed that his military would shoot down spy satellites, just as they had the U-2.\footcite[p.~166]{day_eye_2015} In the first few years after the United States developed satellite reconnaissance capability, it appeared that the USSR was acting exactly as the 1950 RAND report had predicted, and using the legal means at their disposal to end the satellite overflights.
+The Soviet Union protested the very first \emph{Discoverer} launch, correctly alleging its military utility even though the satellite had failed in its purpose (and would fail eleven more times).\footcite[p.~140]{day_eye_2015} Until 1963, Soviet propaganda continually railed against the Americans, in domestic media and international forums, for attempting to militarize space.\footcite[p.~271]{mcdougall_heavens_1985} The USSR sponsored a UN resolution to prohibit espionage from space, and Khrushchev vowed that his military would shoot down spy satellites, just as they had the U-2.\footcite[p.~166]{day_eye_2015} In the first few years after the United States developed satellite reconnaissance capability, it appeared that the USSR was using all the means at its disposal to end the satellite overflights.
 
-Khrushchev had every incentive to discourage satellite reconnaissance, because the intelligence that the US gained through Corona was absolutely devastating to Soviet claims about their military prowess.
-The U-2 overflights had sufficiently debunked the ``bomber gap,'' and seemed to suggest that no ``missile gap'' existed either.\footcite{goodpaster_cold_2003} But as the CIA report on U-2 photography showed, being unable to find ICBM sites was not conclusive proof that they did not exist, and the vast majority of possible missile sites remained un-photographed.
-Confusion about Soviet long-range attack capabilities still existed at various levels of government, confusion which ultimately led to the doomed flight of Gary Powers.\footcite[p.~344]{brugioni_eyes_2010} It was the \emph{Discoverer 14} and \emph{Discoverer 18} missions that successfully photographed the sites that intelligence analysts considered most likely to host operational ICBMs, and found nothing.\footcite[p.~379]{brugioni_eyes_2010}
+Khrushchev had every incentive to discourage satellite reconnaissance, because the intelligence that the US gained through Corona was absolutely devastating to Soviet security.
+One of Khrushchev's top national security concerns at this time was concealing the Soviet Union's relative nuclear weakness.\footcite[p.~133]{brugioni_eyes_2010} The moment at which the US discovered that the USSR had far fewer missiles than it claimed was the exact moment at which that disparity was greatest.
+The missile gap actually went in the other direction---in 1962, the US had more than double the ICBMs, maintained a substantial lead in manned bombers, and had just begun to deploy nuclear submarines.\footcite[p.~251]{mcdougall_heavens_1985} If Khrushchev genuinely feared an American attack, absolutely nothing would have been a greater threat to Soviet security than the collapse of the USSR's most significant deterrent argument---a collapse brought about by American spy satellites.
+\emph{Discoverer 14} and \emph{Discoverer 18} successfully photographed the sites that intelligence analysts considered most likely to host operational ICBMs, and found nothing.\footcite[p.~379]{brugioni_eyes_2010}
 
-To make matters worse for the Soviets, the moment at which the US discovered that the USSR had far fewer missiles than they claimed was the exact moment at which that disparity was greatest.
-The missile gap actually went in the other direction---in 1962, the US had more than double the ICBMs, maintained a substantial lead in manned bombers, and had just begun to deploy nuclear submarines.\footcite[p.~251]{mcdougall_heavens_1985} By the time of the 1960 presidential election, both sides of the Cold War had a generally accurate read on each others' military capabilities, and they knew that their adversary was properly appraised of the strategic situation as well.
-If Khrushchev genuinely feared an American attack, absolutely nothing would have been a greater threat to Soviet security than the collapse of the USSR's most significant deterrent argument---a collapse brought about by American spy satellites.
+% The U-2 overflights had sufficiently debunked the \enquote{bomber gap,} and seemed to suggest that no \enquote{missile gap} existed either.\footcite{goodpaster_cold_2003} But as the CIA report on U-2 photography showed, being unable to find ICBM sites was not conclusive proof that they did not exist, and the vast majority of possible missile sites remained un-photographed.
+Confusion about Soviet long-range attack capabilities still existed at various levels of government, confusion which ultimately led to the doomed flight of Gary Powers.\footcite[p.~344]{brugioni_eyes_2010}
 
@@
 No subsequent missile buildup or bunker hardening or second-strike capability could undo the damage done to the Soviet military posture when \emph{Discoverer XIV} and its successors sailed past the Iron Curtain.
-Aerial reconnaissance neutralized one of Soviet Union's asymmetric structural advantages---the inherent secrecy of Soviet society---and put a dent in their claims to an arms advantage as well.
+Aerial reconnaissance neutralized one of Soviet Union's asymmetric structural advantages---the inherent secrecy of Soviet society---and put a dent in its claims to an arms advantage as well.
 And yet the only thing the Soviet Union did in response was publicly protest the spy satellites, and refuse to accept American proposals for a legal regime that included the rights of reconnaissance overflights.
-Those protests continued for about 3 years, until the Soviets developed their own working spy satellites, ``freedom of space'' became a mutually-agreeable reality.\footcite[p.~271-275]{mcdougall_heavens_1985}
+Those protests continued for about 3 years, until the Soviets developed their own working spy satellites, and ``freedom of space'' became a mutually-agreeable reality.\footcite[p.~271-275]{mcdougall_heavens_1985}
 
 \subsubsection{Espionage norms in satellite reconnaissance}
-The legitimization of spy satellites dealt an indisputable blow to Soviet standing, so why did the Soviet Union let it happen?
-The partial explanation is that the US played a good hand very well.
-If the US had not been engaging in multinational talks about \enquote{scientific} freedom of space while secretly launching spy satellites, then it is very likely that upon developing ASAT capabilities, the USSR would have simply started shooting those satellites down.
-That would have been an almost identical situation to the U-2 saga, and it was essential that when ASATs became available, the USSR not presume to have the unquestioned right to use them.
-It was likewise crucial that the US built a successful reconnaissance satellite in secret.
-The Soviet Union was unlikely to launch the first spy satellite and give the US a perfect excuse to neutralize the intelligence gap, so the US had to do it on their own.
-The situation that resulted was a perfect storm for the normalization of satellite reconnaissance: the Soviet leadership found that they had the ability to develop a reliable spy satellite much sooner than they could develop a reliable anti-satellite weapon, of which their right to use would still be uncertain.
-The path of least resistance, but greatest preservation of pride, was to simply start spying themselves.
+The American pursuit of spy satellites cleanly demonstrates the core premise of $H_2$.
+President Eisenhower afforded an attention to espionage uncharacteristic even for the leader of a Great Power, and his intent is reflected in the actions of the state to an unusual degree.
+The United States also incurred significant costs for diverting resources to signal that its satellite activities constituted espionage, highlighting the contrast with the more direct, militaristic approach.
+Though there is less direct evidence of its intent, the Soviet Union functionally applied the same playbook when it launched its own satellites a few years later---Khrushchev even claimed in 1960 that the Soviet Union had already put cameras in some of its satellites.\footcite[p.~353.
+As far as we know, these camera-equipped satellites, if they actually existed, were not successful enough to be considered spy satellites.]{brugioni_eyes_2010} The effective cessation of diplomatic complaints regarding spy satellites, from either side, is the expected outcome from two sides that are both acting to code their actions as espionage.
+The $H_1$ hypothesis is implied by exactly this situation, in which $H_2$ is reciprocal.
 
-% But if the US had not been simultaneously running a covert program before the norm was established, then how would the norm have ever been established at all?
-At what point would the United States have determined that it was officially legal to launch a spy satellite?
-Soviet intransigence on COPUOS demonstrated that they were unwilling to formally acknowledge the use of spy satellites, so no legal solution was forthcoming.
+Alternate explanations for why the Soviet Union allowed the legitimization of spy satellites rely exclusively on one's faith in diplomatic norms.
+Most boil down to gamesmanship: the United States effectively leveraged scientific diplomacy and covert operations; \emph{Sputnik} made it much harder for the USSR to argue that reconnaissance satellites were an imperialist invasion of territorial sovereignty; Nikita Khrushchev, in a heated moment at the 1960 summit, was said to have claimed that ``any nation in the world who wanted to photograph the Soviet areas by satellite was completely free to do so.''\footcite[p.~556]{eisenhower_waging_1965} None of these explanations, on their own, fully capture why the Soviet Union would assent to an espionage regime that asymmetrically neutralized one of its most significant security advantages.
 
-Nikita Khrushchev played an unwitting role in legitimizing satellite reconnaissance as well.
-At the infamous four-powers summit in 1960, after watching Khrushchev vent about the U-2 Incident, French president Charles de Gaulle calmly told him: ``You're making too big a fuss about the matter.
-There is probably a ton of Russian iron [referring to Russian satellites] coming through French space every day without my permissions.
-I have no idea what is inside those satellites and you have not told me.
-But I am not making a big fuss.'' Khrushchev admitted that the Soviets had put cameras in some satellites, but not necessarily any that flew over France.\footcite[p.~353]{brugioni_eyes_2010} He went on to say that he only considered airplanes to be out of bounds, not satellites.
-In Eisenhower's later paraphrasing, Khrushchev said that ``any nation in the world who wanted to photograph the Soviet areas by satellite was completely free to do so.''\footcite[p.~556]{eisenhower_waging_1965} Even being the first nation to space played a role.
-Of course, it cannot be said that launching the first satellite in human history was a \emph{mistake}---Eisenhower underestimated the symbolic importance of igniting the Space Age, to his significant political detriment---but it is indisputable that \emph{Sputnik} made it much harder for the USSR to argue that reconnaissance satellites were an imperialist invasion of territorial sovereignty.
-
-And yet, American strategy and Khrushchev's missteps still do not fully explain why Soviet Union continued to permit photo-reconnaissance after ASATs were developed.
-The weapons were feasible; the USSR started testing its ASAT capabilities in 1968.\footcite[p.~149]{karas_new_1983} Given that the Soviet Union was using spy satellites as well, shooting down an American one would have been hard to justify, but it really only risked having some Soviet satellites shot down in response.
-Instead, the USSR choose to help write and ratify the Outer Space Treaty (OST) in 1967, the first legal regime codifying the peaceful uses of space.
-While the OST did not ban ASATs, it nonetheless set a precedent for the peaceful uses of space---one that validated the US policy stance that the main military value of space was as a harbor for reconnaissance satellites.\footcite[p~16]{mowthorpe_militarization_2004}
-
-% Took out Welton Chang, ask if he should come back?
-
-Some might argue that the USSR simply miscalculated the relative gains that they received from a norm protecting all satellites.
-This argument is unpersuasive.
-Khrushchev made it clear in his memoirs that he rejected the Open Skies treaty because he knew it would reveal the weakness at the heart of the Soviet state---that its missile program was nowhere near as developed as Khrushchev claimed.
-The USSR would not reach missile parity with the United States until 1969, after both sides had tested ASAT weapons.\footnote{There is obviously a lot more to relative military strength than just adding up the number of missiles that each side has access to---deployment vehicle, missile quality, and susceptibility to a first-strike are crucial.
-But number of missiles is an easily-digestible metric that politicians pay attention to, and helps visualize just how far behind the Soviet Union was when their actual strength was revealed.} After that, the USSR might have had less to hide, but the fact remains that it still had the opportunity to ``turn off the lights'' by taking action against American spy satellites, something the United States, with its free press and comparatively open borders, could not do.
-Instead, Open Skies began the moment the USSR launched its first spy satellite, and it never ended.
+Each of the circumstances surrounding the normalization of spy satellites takes on an elevated significance if one allows for $H_1$ to be true---that states place importance on the \emph{appearance} of espionage and adjust their diplomatic consequences accordingly.
+Applying reduced consequences to espionage, while itself certainly a norm, has realist implications as well.
+By explicitly accepting the American justification for their reconnaissance satellites, at great cost to it security, the Soviet Union was signaling that the reciprocal tolerance of espionage was a pact that it is willing to preserve.
+When the United States allowed the Soviet Union to launch its own spy satellites a few years later, the mutual flow of intelligence was assured, along with the security benefits that both states derive from it.
 
@@
 
-Satellites are the purest demonstration of the appeal of espionage norms, because unlike with planes, there were no previously existing norms of territorial integrity.
-Therefore the only established norm at the dawn of the space age was one from the world of espionage---that intelligence-gathering entities are not inherently aggressive.
-Space became an environment where a strike against a satellite would constitute the first act of aggression, even if the satellite was operating for a military purpose, gathering information that could significantly compromise a state's security position.
+\section{Conclusion}
 
-There is a fundamental appeal to a world with tacit mutual reconnaissance, an appeal that is rooted in defensive realism and manifested in espionage norms.
+% And yet, American strategy and Khrushchev's missteps still do not fully explain why Soviet Union continued to permit photo-reconnaissance after ASATs were developed.
+The weapons were feasible; the USSR started testing its ASAT capabilities in 1968.\footcite[p.~149]{karas_new_1983} Given that the Soviet Union was using spy satellites as well, shooting down an American one would have been hard to justify, but it really only risked having some Soviet satellites shot down in response.
+Instead, the USSR choose to help write and ratify the Outer Space Treaty (OST) in 1967, the first legal regime codifying the peaceful uses of space.
+While the OST did not ban ASATs, it nonetheless set a precedent for the peaceful uses of space---one that validated the US policy stance that the main military value of space was as a harbor for reconnaissance satellites.\footcite[p~16]{mowthorpe_militarization_2004}
+
+% Though there is no chance that Khrushchev intended this, the intelligence that satellite photo-reconnaissance provided resulted in the United States building \emph{fewer} missiles and resisting the temptation to engage in an arms race.
+The success of the SALT treaties was a significant factor in the stabilization of relations between the two countries---it created an atmosphere of trust and cooperation that, while strained by Reagan the and the invasion of Afghanistan, held for the remainder of the Cold War.\footcite[p.~179]{lindgren_trust_2000} ``Freedom of space'' exists because both sides preferred a world where they could each verify each others' intent over one where they were shooting blind.
+
+% There is a fundamental appeal to a world with tacit mutual reconnaissance, an appeal that is rooted in defensive realism and manifested in espionage norms.
 Spy satellites are the single best proof that even when the relative gains are lopsided, states will choose to live in a world with more information, rather than less.
@@
 
-Though there is no chance that Khrushchev intended this, the intelligence that satellite photo-reconnaissance provided resulted in the United States building \emph{fewer} missiles and resisting the temptation to engage in an arms race.
-The success of the SALT treaties was a significant factor in the stabilization of relations between the two countries---it created an atmosphere of trust and cooperation that, while strained by Reagan the and the invasion of Afghanistan, held for the remainder of the Cold War.\footcite[p.~179]{lindgren_trust_2000} ``Freedom of space'' exists because both sides preferred a world where they could each verify each others' intent over one where they were shooting blind.
-
-\section{Conclusion}
 \end{document}

### 2020-10-26 article.tex
Add title and remove some commented text
Add conclusion

@@
 
-\title{Draft Article}
+\title{Reciprocal Tolerance: Why States Allow Their Adversaries to Practice Espionage with Minimal Consequences}
 \author{Alexander W. Petros}
-\date{April 2020}
+\date{October 2020}
 
@@
 
-% The United States fully believed it had more to gain from a ``freedom of space'' norm---and I agree---but is it possible that the Soviet Union had some reason to believe the same?
-Perhaps, but given the indisputable advantage that the US gained from reciprocal reconnaissance, it is hard to imagine what non-reconnaissance military function a satellite could perform that would advantage the Soviet Union to the same degree.
-Satellites were also used to improve the accuracy of missile guidance systems, but since both sides relied on nuclear missiles as their main deterrent, satellite-based missile targeting likely provide no relative gains.
-Or perhaps the Soviet Union thought being the first nation to shoot down a satellite would adversely impact a key propaganda point: that the United States was responsible for the militarization of space.
-Both the US and the USSR were preoccupied with demonstrating that their respective ideology---capitalism or communism---had an inherent moral superiority.
-They each believed that this propaganda helped project their power in the third world, but that also raises a question: why would the world react adversely to Soviet defense of its own borders, if the United States did in fact break the glass on space militarization?
+\section{Conclusion}
+The promise of an international community devoid of secrets is, at its core, a utopian promise.
+Complete, universal verification of capabilities and intents would make it impossible for an adversary to maintain hidden military capabilities or furtive plans for war; theoretically, if applied to our current system of nation-states, bad actors would be rendered obsolete.
+While diplomacy based on perfect information is impossible right now---and likely always will be---espionage norms that reduce the diplomatic impact of intelligence gathering are one means to approach a truly open society.
 
-\section{Conclusion}
+In order to best demonstrate the novel theory of this paper, I focused on a uniquely evidence-rich context: two nuclear superpowers who both maintained robust intelligence services.
+This is, of course, not the situation in which most espionage norms live---the stakes might be lower, the capabilities different, or the overall balance of power more lopsided.
+For example, the relationship between the espionage services of Israel and Iran could look very different from the US-USSR espionage relationship examined here, as could the relationship between the US and present-day Russia.
+This paper does not claim to consider all these variations, rather, it lays a foundation for future analysis of how other variables impact the espionage norms observed.
 
-% And yet, American strategy and Khrushchev's missteps still do not fully explain why Soviet Union continued to permit photo-reconnaissance after ASATs were developed.
-The weapons were feasible; the USSR started testing its ASAT capabilities in 1968.\footcite[p.~149]{karas_new_1983} Given that the Soviet Union was using spy satellites as well, shooting down an American one would have been hard to justify, but it really only risked having some Soviet satellites shot down in response.
-Instead, the USSR choose to help write and ratify the Outer Space Treaty (OST) in 1967, the first legal regime codifying the peaceful uses of space.
-While the OST did not ban ASATs, it nonetheless set a precedent for the peaceful uses of space---one that validated the US policy stance that the main military value of space was as a harbor for reconnaissance satellites.\footcite[p~16]{mowthorpe_militarization_2004}
+At scale, espionage removes risk from the international system, and the rare instances where espionage is codified suggests that states generally view espionage this way as well.
+The Strategic Arms Limitation Treaty of 1972 (otherwise known as The ABM Treaty, or SALT 1) enshrined in law that signatories are not permitted to interfere with each others' ``national technical means of verification,'' a tacit ban on ASAT weapons.\footcite[p.~431]{mcdougall_heavens_1985} Without having to say ``spy satellites,'' SALT 1 made them not just legal, but essential.
+The peaceful uses of aerial reconnaissance were formalized at the end of the Cold War, when members of NATO and the Warsaw Pact signed the \enquote{Treaty on Open Skies,} which established guidelines for unarmed aerial surveillance flights in the spirit of open information and de-escalation.\footcite{organization_for_security_and_co-operation_in_europe_treaty_1992} This treaty was almost identical to the one first suggested by Eisenhower in 1955.\footcite{center_for_arms_control_and_non-proliferation_fact_2017} With certain restrictions, the type of flight that altered the course of the Cold War is now officially routine and permissible.
 
-% Though there is no chance that Khrushchev intended this, the intelligence that satellite photo-reconnaissance provided resulted in the United States building \emph{fewer} missiles and resisting the temptation to engage in an arms race.
-The success of the SALT treaties was a significant factor in the stabilization of relations between the two countries---it created an atmosphere of trust and cooperation that, while strained by Reagan the and the invasion of Afghanistan, held for the remainder of the Cold War.\footcite[p.~179]{lindgren_trust_2000} ``Freedom of space'' exists because both sides preferred a world where they could each verify each others' intent over one where they were shooting blind.
+Even amongst the superpowers, these norms are not inviolate, and they are always subject to defection.
+In May 2020, President Trump announced his intent to withdraw the United States from the Open Skies Treaty.\footcite{sanger_trump_2020} If the withdrawal takes place, we can expect that Russia will increasingly deny access to its airspace as well.
+China destroyed one of its own satellites with an ASAT weapon in 2007, which a Harvard Astronomer described as ``the first real escalation in the weaponization of space that we’ve seen in 20 years.''\footcite{broad_china_2007} India successfully tested its own ASAT in March 2019.\footcite{gettleman_india_2019} Though the threat of weaponization continues to escalate, the US still refuses to pursue an official ban on weapons on space.\footcite{oconnor_u.s._2019}
+
+The value of an international community based on perfect information is still under contention.
+Despite recent worrying developments, at the time of writing no ASAT has even been used against an adversary's satellite, and the Open Skies treaty remains precariously in effect.
+Maintaining these norms requires formulating a positive case for their utility, and quickly.
+The need to present this case in a compelling manner will only become more urgent.
+
+\end{document}
 
@@
 
-\end{document}
-
-% The breadth of information that a satellite could provide might also dispel the Eisenhower administration's political albatross---his supposedly dovish defense spending.
-Eisnhower faced constant pressure to increase the defense budget and close the alleged ``missile gap'' between the US and the USSR.
-While U-2 photography (codenamed ``Talent'') had been helpful for pushing back against these demands, the diplomatic volatility of each flight limited the amount of ground that it could cover.
-In July 1960, the CIA's Photographic Interpretation Center conducted a study to determine how much of the Soviet territory that it deemed suitable for ICBM deployment had been observed by Talent photography since January 1959.
-As it turned out, very little.
-The ``intelligence bonanza'' from Aquatone had still only observed roughly 13.6\% of the viable Soviet terrain.
-Visual observations from other sources added an additional 1.5\%.\footcite[A significant portion of this document is redacted, including, for some reason, the page numbers.]{cia_future_1960} Satellites showed enormous potential for covering that remaining 85\%, and in doing so, shut down discussion of the ``missile gap'' once and for all.
-
-% The peaceful uses of aerial reconnaissance were formalized at the end of the Cold War, when members of NATO and the Warsaw Pact signed the \enquote{Treaty on Open Skies,} which established guidelines for unarmed aerial surveillance flights in the spirit of open information and de-escalation.\footcite{organization_for_security_and_co-operation_in_europe_treaty_1992} This treaty was almost identical to the one first suggested by Eisenhower in 1955.\footcite{center_for_arms_control_and_non-proliferation_fact_2017} With certain restrictions, the type of flight that altered the course of the Cold War is now officially routine and permissible.
-In February 1962, Francis Gary Powers walked across a bridge into West Berlin at the same time as Russian Colonel Rudolph Abel, an American prisoner.
+% In February 1962, Francis Gary Powers walked across a bridge into West Berlin at the same time as Russian Colonel Rudolph Abel, an American prisoner.
 Powers' father had suggested to the State Department that they might be traded, spy for a spy.\footcite[p.~239]{powers_operation_2004} \enquote{This,} Eisenhower wrote, ``was a tacit admission by Khrushchev that our `outrageous' U-2 pilots have their opposite numbers operating within the borders of this country.''\footcite[p.~558]{eisenhower_waging_1965}
-
-% While I make the claim that both the spy plane and the spy satellite were purposefully connected to an intelligence tradition, it cannot be denied that the first time each one was deployed resulted in very different diplomatic outcomes.
-The first series of U-2 flights led to diplomatic protests, then a shootdown, then a dramatic summit meeting; the first series of Corona launches led to diplomatic protests, then...
-nothing.
-In absolute terms, Corona was dramatically more successful.
-Eisenhower gained the right to orbit a satellite over the Soviet Union whenever he wanted, and after the Gary Powers incident the U-2 was only used sparingly.
-
-% U-2 overflights could have caused a much more damaging reaction, but their framing helped prevent it from becoming an even larger incident than it already way.
-The norms defining how states should respond to espionage are ambiguous---the principle that states should not fly planes deep into each others' territory is not.
-When the U-2 was first developed, it was not at all clear which norm would supersede the other.
-As it happened, neither did---the interaction of existing norms was unique to both the craft itself and the way that it was marketed by the administration that invented it.
-The U-2 is not 70\% spy and 30\% plane.
-Instead, it is a complicated intersection of the two.
-The Soviet Union---and frankly, any nation over which the US chooses to fly a plane without permission---had the undisputed right to shoot down the American spy plane, which they did, but they also chose not to treat as an act of war, despite Eisenhower's explicit concern that they would.

### 2020-11-08 abstract.tex
Conform to Security Studies Style Guidelines

\begin{abstract}
\begin{em}
Espionage has long been an accepted part of international relations, but the norms might be changing. Cyberattacks vastly increase the potential blast radius of an intelligence operation, and the old rules of the game are looking increasingly out of date. In this article, I defend the longstanding tolerance of espionage in the international community as a rational calculation of each state's incentives, which are biased towards cooperate. I employ defensive realism, in which states use intelligence as a means to solve the security dilemma, to explain how these norms can persist among states of varying power. Then, I test two past technological advancements that revolutionized intelligence---the spy plane and the spy satellite---to demonstrate how the United States and the Soviet Union purposefully employed espionage norms to diminish the consequences for their use.
\end{em}
\end{abstract}

### 2020-11-08 article.tex
Conform to Security Studies Style Guidelines

@@
-\documentclass[14pt]{extarticle}
+\documentclass[12pt]{extarticle}
 \usepackage{style}
@@
 \title{Reciprocal Tolerance: Why States Allow Their Adversaries to Practice Espionage with Minimal Consequences}
-\author{Alexander W. Petros}
+\author{Alexander W.
+Petros\thanks{This article would not be possible without the unwavering attention of Professor William Wohlforth, who continued to provide gentle encouragement and thoughtful criticism throughout its creation.
+I would also like to thank the Dartmouth Government department for its support in developing the thesis on which this article is based, and from which much of this research derives.
+Finally, my first and most valued editor is my father, William Petros, whose unsparing attention to detail is deeply reflected in this article, as in all things he does.}}
 \date{October 2020}
@@
 \begin{document}
+
 \maketitle
+\newpage
+\input{abstract.tex}
+\newpage
 
@@
 \section{Conclusion}
-The promise of an international community devoid of secrets is, at its core, a utopian promise.
+The promise of an international community devoid of secrets is, at its core, utopian.
 Complete, universal verification of capabilities and intents would make it impossible for an adversary to maintain hidden military capabilities or furtive plans for war; theoretically, if applied to our current system of nation-states, bad actors would be rendered obsolete.
@@
 In order to best demonstrate the novel theory of this paper, I focused on a uniquely evidence-rich context: two nuclear superpowers who both maintained robust intelligence services.
-This is, of course, not the situation in which most espionage norms live---the stakes might be lower, the capabilities different, or the overall balance of power more lopsided.
+This is, of course, not the situation in which most espionage norms exist---the stakes might be lower, the capabilities different, or the overall balance of power more lopsided.
 For example, the relationship between the espionage services of Israel and Iran could look very different from the US-USSR espionage relationship examined here, as could the relationship between the US and present-day Russia.
@@
 
-The value of an international community based on perfect information is still under contention.
-Despite recent worrying developments, at the time of writing no ASAT has even been used against an adversary's satellite, and the Open Skies treaty remains precariously in effect.
-Maintaining these norms requires formulating a positive case for their utility, and quickly.
-The need to present this case in a compelling manner will only become more urgent.
+The viability of international community based on perfect information is a question that remains to be settled.
+Despite recent worrying developments, there are still reasons to be optimistic: at the time of writing no ASAT has even been used against an adversary's satellite, and the Open Skies treaty remains precariously in effect.
+Maintaining these norms urgently requires formulating a compelling case for their utility.
 

### 2020-11-08 article.tex
Remove "p" from page numbers

@@
 \newpage
-
+\section{Introduction}
 In 2015, federal authorities discovered a massive data breach inside the United States Office of Personnel Management (OPM).
@@
 The problem that aspiring espionage scholars inevitably encounter is that \enquote{peacetime espionage} is universally understood as a term of art but almost never defined by legal codes, or interpreted by case law.
-Most scholars who tackle this question settle on a definition like the one proposed by I\~{n}aki Navarrete and Russell Buchan: \enquote{\enquote{Peacetime espionage} is a colloquial rather than a legal term [used] to describe different methods of collecting confidential information from closed as opposed to open sources \textelp{} generally conducted by States, against States.}\footcite[p.~901-902]{navarrete_out_2019} This definition purposefully excludes forms of extralegal government intervention that are often conflated with espionage, such as assassinations, election interference, and covert operations.\footnote{Often papers won't take a stand on defining espionage, but rather offer a working definition just to bound the scope of that particular paper.
-See \cite[p.~600]{radsan_unresolved_2007}, \cite[p.~2]{kapp_spying_2007})} Excluding these activities is important: assassinations and similar covert operations are not considered espionage, so they do not typically receive the protections of espionage norms.
+Most scholars who tackle this question settle on a definition like the one proposed by I\~{n}aki Navarrete and Russell Buchan: \enquote{\enquote{Peacetime espionage} is a colloquial rather than a legal term [used] to describe different methods of collecting confidential information from closed as opposed to open sources \textelp{} generally conducted by States, against States.}\footcite[901-902]{navarrete_out_2019} This definition purposefully excludes forms of extralegal government intervention that are often conflated with espionage, such as assassinations, election interference, and covert operations.\footnote{Often papers won't take a stand on defining espionage, but rather offer a working definition just to bound the scope of that particular paper.
+See \cite[600]{radsan_unresolved_2007}, \cite[2]{kapp_spying_2007})} Excluding these activities is important: assassinations and similar covert operations are not considered espionage, so they do not typically receive the protections of espionage norms.
 
@@
 Jared Beim concludes that while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so there ought to be stronger enforcement mechanisms in place.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe.
-For more, see \cite{beim_enforcing_2018}} Beim acknowledges, however, that at best \enquote{it is perhaps possible to stop the more egregious violations.}\footcite[p.~672]{beim_enforcing_2018} A.
+For more, see \cite{beim_enforcing_2018}} Beim acknowledges, however, that at best \enquote{it is perhaps possible to stop the more egregious violations.}\footcite[672]{beim_enforcing_2018} A.
 John Radsan employs a more philosophical approach.
@@
 Responding to Baker, Radsan notes that one-off deals and treaties \enquote{are a very slow and indirect way to achieve international consensus on the legality of espionage.
-The gap is still there.}\footcite[p.~607]{radsan_unresolved_2007}
+The gap is still there.}\footcite[607]{radsan_unresolved_2007}
 
@@
 In recent years China has relentlessly targeted American military secrets.
-A United States Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} \enquote{Long-term, US future military advantage is being diminished by years of IP exfiltration from [US military agencies], all with little to no adverse consequences to the thieves.}\footcite[p.~6]{bayer_cybersecurity_2019} The forecasts were dire: \enquote{If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.}\footcite[p.~5]{bayer_cybersecurity_2019} The US Navy is not a disinterested party, of course, and we should be cautious about its conclusions; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see: \cite{gilli_why_2019}} Nonetheless, espionage presents alarming opportunities for rival states to gain a comparative advantage and the consequences for taking advantage of them remain minimal.
+A United States Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} \enquote{Long-term, US future military advantage is being diminished by years of IP exfiltration from [US military agencies], all with little to no adverse consequences to the thieves.}\footcite[6]{bayer_cybersecurity_2019} The forecasts were dire: \enquote{If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.}\footcite[5]{bayer_cybersecurity_2019} The US Navy is not a disinterested party, of course, and we should be cautious about its conclusions; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see: \cite{gilli_why_2019}} Nonetheless, espionage presents alarming opportunities for rival states to gain a comparative advantage and the consequences for taking advantage of them remain minimal.
 
@@
 I propose an iterated decision-making model to describe how states cooperate to preserve the liminal legal status of espionage, based on the same tit-for-tat strategy associated with diplomatic expulsions.
-In game theory, tit-for-tat is highly-successful strategy wherein the player simply mimics the choice that that their opponent made in the last round.\footcite[p.~8]{axelrod_effective_1980} By employing this strategy, the player never gets taken advantage of for more than one turn and can always return to cooperation if their opponent unilaterally signals a willingness to do the same.
+In game theory, tit-for-tat is highly-successful strategy wherein the player simply mimics the choice that that their opponent made in the last round.\footcite[8]{axelrod_effective_1980} By employing this strategy, the player never gets taken advantage of for more than one turn and can always return to cooperation if their opponent unilaterally signals a willingness to do the same.
 Diplomatic expulsions play out in exactly the same fashion: when one country expels the diplomats of another, the other country will often respond in kind.
@@
 To arrive at mutual cooperation, a security-seeking state needs to not only determine that its rival is a security seeker, it needs its rival to make that exact same judgment, because it is the adversary's response to securitization that ultimately makes the first state insecure.
-Defensive realism therefore introduces a counterintuitive conclusion: it is often logical for a security seeker to take actions that increase not only its own security, but its \emph{adversary's} security as well.\footcite[p.~7]{glaser_rational_2010} If the adversary is indeed a security seeker, then, no longer concerned about its security, it will respond by not building up arms, and resolve the security dilemma for both of them.
+Defensive realism therefore introduces a counterintuitive conclusion: it is often logical for a security seeker to take actions that increase not only its own security, but its \emph{adversary's} security as well.\footcite[7]{glaser_rational_2010} If the adversary is indeed a security seeker, then, no longer concerned about its security, it will respond by not building up arms, and resolve the security dilemma for both of them.
 The problem for a defensive realist is that unilaterally taking steps to increase the adversary's security is obviously quite risky.
@@
 Because the state receiving the signal is also a security-seeker, this is a welcome development; it could be empowered to signal its benign intentions as well, based on its own intelligence.
-For both states, this is the optimal outcome, contributing to both information variables that Glaser identifies as key to the security dilemma: \enquote{the state’s estimate of the adversary’s motives, and the state’s estimate of the adversary’s estimate of the state’s motives.}\footcite[p.~34-35]{glaser_rational_2010} Both states therefore benefit not just from spying, but from being spied \emph{upon}; their adversary's intelligence made that adversary more likely to propose or accept a risky signal for deescalation.
+For both states, this is the optimal outcome, contributing to both information variables that Glaser identifies as key to the security dilemma: \enquote{the state’s estimate of the adversary’s motives, and the state’s estimate of the adversary’s estimate of the state’s motives.}\footcite[34-35]{glaser_rational_2010} Both states therefore benefit not just from spying, but from being spied \emph{upon}; their adversary's intelligence made that adversary more likely to propose or accept a risky signal for deescalation.
 
@@
 The analysis in this section relies on one crucial assumption: that the states in question are security-seekers.
-Analyses of the risks of espionage must contend with possibility that one of the states in question is \enquote{fundamentally dissatisfied with the status quo, desiring additional territory even when it is not required for security}---a greedy state.\footcite[p.~4]{glaser_rational_2010} Such a state would obviously want to protect both its motives and its plans, and therefore derive no benefit from having espionage conducted against it.
+Analyses of the risks of espionage must contend with possibility that one of the states in question is \enquote{fundamentally dissatisfied with the status quo, desiring additional territory even when it is not required for security}---a greedy state.\footcite[4]{glaser_rational_2010} Such a state would obviously want to protect both its motives and its plans, and therefore derive no benefit from having espionage conducted against it.
 Nonetheless, we know that essentially every major nation-state abides by the accepted norms of espionage.
@@
 Perhaps greedy states simply do not exist.
-This is a very provocative assertion---one that is, in its most stringent sense, almost certainly wrong---but consider the possibility that no states which abide by espionage norms are greedy \emph{to a destabilizing degree.} The concept of a greedy state is one that naturally operates on a spectrum.\footcite[p.~39.
+This is a very provocative assertion---one that is, in its most stringent sense, almost certainly wrong---but consider the possibility that no states which abide by espionage norms are greedy \emph{to a destabilizing degree.} The concept of a greedy state is one that naturally operates on a spectrum.\footcite[39.
 Throughout his book, Glaser simplifies and complicates his theory as is necessary for the particular analysis he is performing.
@@
 \subsection{Aerial espionage}
-Just a few weeks before the \enquote{Four Powers Summit} in Summer 1960, Soviet Premeir Nikita Khrushchev announced that the USSR had shot down an American plane.\footcite[p.~112]{powers_operation_2004} The CIA knew tgat we was referring to the U-2, an advanced spy plane that had bedeviled the USSR for years, proving beyond the reach of Soviet surface-to-air missiles.\footcite{orlov_u-2_2007} Nevertheless, the CIA was still confident that its cover story would hold, thanks to a morbid fact of the U-2's design: in the event of a shootdown, it should have been impossible for the pilot to survive.\footcite[p.~35]{lindgren_trust_2000} When the United States put out at a statement that the plane was part of a \enquote{weather reconnaissance} mission gone awry, Khrushchev sprung his trap---he revealed that the U-2 pilot, Gary Powers, had been taken alive.
+Just a few weeks before the \enquote{Four Powers Summit} in Summer 1960, Soviet Premeir Nikita Khrushchev announced that the USSR had shot down an American plane.\footcite[112]{powers_operation_2004} The CIA knew tgat we was referring to the U-2, an advanced spy plane that had bedeviled the USSR for years, proving beyond the reach of Soviet surface-to-air missiles.\footcite{orlov_u-2_2007} Nevertheless, the CIA was still confident that its cover story would hold, thanks to a morbid fact of the U-2's design: in the event of a shootdown, it should have been impossible for the pilot to survive.\footcite[35]{lindgren_trust_2000} When the United States put out at a statement that the plane was part of a \enquote{weather reconnaissance} mission gone awry, Khrushchev sprung his trap---he revealed that the U-2 pilot, Gary Powers, had been taken alive.
 
@@
 Beginning in 1947, the US began a campaign of reconnaissance flights that routinely violated Soviet airspace.
-Converted Air Force bombers---called \enquote{ferrets}---were outfitted with advanced radio equipment and sent to intercept as much information about Soviet radio transmissions as possible.\footcite[p.~4]{peterson_maybe_1993} The pilots who flew these missions were conducting one of the earliest forms of Cold War peacetime espionage: gathering SIGINT (signals intelligence) that identified the location of critical radar installations along the Soviet border.
+Converted Air Force bombers---called \enquote{ferrets}---were outfitted with advanced radio equipment and sent to intercept as much information about Soviet radio transmissions as possible.\footcite[4]{peterson_maybe_1993} The pilots who flew these missions were conducting one of the earliest forms of Cold War peacetime espionage: gathering SIGINT (signals intelligence) that identified the location of critical radar installations along the Soviet border.
 Some of these flights only flew near the border, remaining over traditionally recognized international waters, while others were \enquote{overflights,} missions that deliberately violated Soviet airspace.
@@
 
-% The earliest definitive postwar overflight of the Soviet Union took place on May 10, 1949, when two RF-80As surveilled the Kurile islands\footcite[p.~8]{peebles_shadow_2000}
+% The earliest definitive postwar overflight of the Soviet Union took place on May 10, 1949, when two RF-80As surveilled the Kurile islands\footcite[8]{peebles_shadow_2000}
 
 Without access to classified archives it is impossible to state for certain exactly how many of these flights were conducted.\footnote{Some of the necessary information was so highly classified that it is likely lost forever.} Nonetheless, the civilian researcher is aided by the declassified research of military and intelligence archivists.
-Writing for the internal CIA publication \emph{Cryptologic Quarterly}, Michael Peterson compiled a now-declassified list of American SIGINT flights shot down by the USSR (Table \ref{soviet-shootdowns}).\footnote{Michael Peterson is a former intelligence analyst and section chief for the CIA.} Peterson uses two limiting criteria that perfectly suit the purposes of this article: the only flights he lists are those that were (a) shot down by Soviet attacks, not Chinese, North Korean, etc., and (b) involved US aircraft exclusively performing reconnaissance.\footcite[p.~4.
+Writing for the internal CIA publication \emph{Cryptologic Quarterly}, Michael Peterson compiled a now-declassified list of American SIGINT flights shot down by the USSR (Table \ref{soviet-shootdowns}).\footnote{Michael Peterson is a former intelligence analyst and section chief for the CIA.} Peterson uses two limiting criteria that perfectly suit the purposes of this article: the only flights he lists are those that were (a) shot down by Soviet attacks, not Chinese, North Korean, etc., and (b) involved US aircraft exclusively performing reconnaissance.\footcite[4.
 Regardless of how thorough declassified CIA information appears, one cannot rule out the possibility that there were other, even more secret missions which, for whatever reason, the United States government has chosen to conceal to this day.
@@
 Additionally, it appears to have been declassified without redactions or missing pages.
-Lastly, if there had been another mission so highly classified that it was omitted from this document, it is difficult to imagine how diplomatic consequences could have been imposed with a degree of secrecy that would prevent us from knowing about it today.]{peterson_maybe_1993} In effect, the table catalogs the 13 times where the USSR shot down an American aircraft designed for collecting intelligence between 1950 and 1964---intercepted espionage attempts.\footcite[p.~5.
+Lastly, if there had been another mission so highly classified that it was omitted from this document, it is difficult to imagine how diplomatic consequences could have been imposed with a degree of secrecy that would prevent us from knowing about it today.]{peterson_maybe_1993} In effect, the table catalogs the 13 times where the USSR shot down an American aircraft designed for collecting intelligence between 1950 and 1964---intercepted espionage attempts.\footcite[5.
 In the original document, this table lists the first incident as having taken place over the Barents Sea, not the Baltic Sea.
@@
 The Baltic incident exemplifies the reduced consequences for espionage that $H_2$ predicts, in both directions.
-The Soviet Union opened fire on a US Navy aircraft, which based on all available evidence did not violate the traditional 12-mile airspace boundary.\footcite[p.~7.
+The Soviet Union opened fire on a US Navy aircraft, which based on all available evidence did not violate the traditional 12-mile airspace boundary.\footcite[7.
 To be extra specific]{peterson_maybe_1993} Given its last known position, it would have taken ``a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.''\footcite{the_new_york_times_soviet_1950} Under other circumstances, this could have been a major incident, but the United States chose not to publicize it.
@@
 A B-29 that went missing on October 7, 1952 hit the front page of the \emph{New York Times} two days later.
-This mission is listed by Peterson as a ferret flight, and Farquhar notes that the media thought the attack was intended to lower American prestige during election season, but does not mention any diplomatic consequences.\footcite[p.~43-44]{farquhar_aerial_2015} The official US response, however, is preserved on the State Department website.
-As with the Baltic Incident, the US simply demanded that the Soviets pay for the plane and return any survivors.\footcite{the_new_york_times_u.s._1952} Two more shootdowns in March 1953---one involving an American F–84 Thunderjet and the other an RAF Lincoln bomber---actually resulted in conciliatory responses from both sides, including a secret meeting to reduce aerial tensions \enquote{of which little is known.}\footcite[p.~45]{farquhar_aerial_2015}
+This mission is listed by Peterson as a ferret flight, and Farquhar notes that the media thought the attack was intended to lower American prestige during election season, but does not mention any diplomatic consequences.\footcite[43-44]{farquhar_aerial_2015} The official US response, however, is preserved on the State Department website.
+As with the Baltic Incident, the US simply demanded that the Soviets pay for the plane and return any survivors.\footcite{the_new_york_times_u.s._1952} Two more shootdowns in March 1953---one involving an American F–84 Thunderjet and the other an RAF Lincoln bomber---actually resulted in conciliatory responses from both sides, including a secret meeting to reduce aerial tensions \enquote{of which little is known.}\footcite[45]{farquhar_aerial_2015}
 
@@
 The Soviet Union responded by protesting the alleged American shootdown of a Soviet passenger plane.
-Despite press outrage, no further action was taken.\footcite[p.~47]{farquhar_aerial_2015} After the September 4, 1954 shootdown of a Navy P2V Neptune, a US Senator publicly denounced the continued anodyne response: \enquote{just another note from our State Department to the Kremlin hierarchy will not impress these uncivilized rulers.}\footcite{the_associated_press_ending_1954} Instead of a harsher punishment, President Eisenhower brought the issue to the UN with the full knowledge that an unfavorable judgment would be vetoed by the Soviets.
-A further signal of deescalation, the UN stunt demonstrated Eisenhower's ability to resist calls for harsher action while still placating the American public.\footcite[p.~47]{farquhar_aerial_2015} By late 1954, the \enquote{cycle of hostility} over aerial reconnaissance was already breaking down.\footcite[p.~49]{farquhar_aerial_2015} In terms of both media coverage and diplomatic significance, no shootdown that followed would require more significant consequences than the ones that had already happened.
+Despite press outrage, no further action was taken.\footcite[47]{farquhar_aerial_2015} After the September 4, 1954 shootdown of a Navy P2V Neptune, a US Senator publicly denounced the continued anodyne response: \enquote{just another note from our State Department to the Kremlin hierarchy will not impress these uncivilized rulers.}\footcite{the_associated_press_ending_1954} Instead of a harsher punishment, President Eisenhower brought the issue to the UN with the full knowledge that an unfavorable judgment would be vetoed by the Soviets.
+A further signal of deescalation, the UN stunt demonstrated Eisenhower's ability to resist calls for harsher action while still placating the American public.\footcite[47]{farquhar_aerial_2015} By late 1954, the \enquote{cycle of hostility} over aerial reconnaissance was already breaking down.\footcite[49]{farquhar_aerial_2015} In terms of both media coverage and diplomatic significance, no shootdown that followed would require more significant consequences than the ones that had already happened.
 
@@
 The President ordered the Hungarian consulates in New York and Cleveland closed and banned private travel to the country.
-Legislatively, Truman asked Congress to pass a \$100 million Mutual Security Act to aid \enquote{selected persons residing in Soviet bloc states or refugees who wanted to form armed units} in opposition to Communism.}\footcite[p.~43]{farquhar_aerial_2015} Truman immediately took action against material Hungarian interests in the United States and threatened to empower \emph{armed dissidents} if the prisoners were not released.
+Legislatively, Truman asked Congress to pass a \$100 million Mutual Security Act to aid \enquote{selected persons residing in Soviet bloc states or refugees who wanted to form armed units} in opposition to Communism.}\footcite[43]{farquhar_aerial_2015} Truman immediately took action against material Hungarian interests in the United States and threatened to empower \emph{armed dissidents} if the prisoners were not released.
 The US government's response to an aviation incident when no espionage was involved stands in sharp relief to the \enquote{prompt and thorough investigation} that it demanded after the Baltic Incident.
@@
 The United States conducted many more overflights than the failed ones mentioned here, using a variety of retrofitted planes.
-Project Heart Throb, for example, outfitted RB-57s with surveillance equipment and flew between 15 and 19 missions over Eastern Europe from 1955 to 1956.\footcite[p.~194.
+Project Heart Throb, for example, outfitted RB-57s with surveillance equipment and flew between 15 and 19 missions over Eastern Europe from 1955 to 1956.\footcite[194.
 Based on the recollection of Gerald E.
-Cooke, a Air Force pilot assigned to the project.]{hall_early_2003} Project Slick Chick set up F-100 jets with rapid-fire 20mm cameras.\footcite[p.~176]{hall_early_2003} Though we know a lot less about these missions than we do about the ferret flights, the basic contours are familiar: missions with no weaponry, operating at a high level of secrecy, subject to countermeasures (tracked by radar and often chased by MiG fighters) but no evidence of diplomatic pressure.
-In meetings between the State department and high-level Soviet diplomats, the subject of captured pilots \enquote{was broached only perfunctorily in relation to other things being discussed.}\footcite[p.~72]{brugioni_eyes_2010} Some family members of lost airmen received posthumous awards, but most simply received the pilot's personal effects and no explanation.
+Cooke, a Air Force pilot assigned to the project.]{hall_early_2003} Project Slick Chick set up F-100 jets with rapid-fire 20mm cameras.\footcite[176]{hall_early_2003} Though we know a lot less about these missions than we do about the ferret flights, the basic contours are familiar: missions with no weaponry, operating at a high level of secrecy, subject to countermeasures (tracked by radar and often chased by MiG fighters) but no evidence of diplomatic pressure.
+In meetings between the State department and high-level Soviet diplomats, the subject of captured pilots \enquote{was broached only perfunctorily in relation to other things being discussed.}\footcite[72]{brugioni_eyes_2010} Some family members of lost airmen received posthumous awards, but most simply received the pilot's personal effects and no explanation.
 
@@
 $H_2$ posits that states will pursue measures to signal that their activities are espionage, and the US did everything possible to ensure that the U-2 would be read as a civilian aircraft.
-Eisenhower required that the pilots of U-2 aircraft be CIA officers, specifically forbidding uniformed Air Force personnel.\footcite[p.~33, Though many of the pilots did have Air Force backgrounds.]{lindgren_trust_2000} The CIA also had operational control of the program instead of the Air Force.
+Eisenhower required that the pilots of U-2 aircraft be CIA officers, specifically forbidding uniformed Air Force personnel.\footcite[33, Though many of the pilots did have Air Force backgrounds.]{lindgren_trust_2000} The CIA also had operational control of the program instead of the Air Force.
 \enquote{I want this whole thing to be a civilian operation,} Eisenhower wrote, to settle an operational dispute between the two departments.
-\enquote{If uniformed personnel of the armed services of the United States fly over Russia, it is an act of war---legally---and I don't want any part of it.}\footcite[p.~60.
+\enquote{If uniformed personnel of the armed services of the United States fly over Russia, it is an act of war---legally---and I don't want any part of it.}\footcite[60.
 The original source for this quote is an \emph{OSA History} that requires codeword clearance.
-It is quoted here by the History Staff of the CIA.]{pedlow_central_1992} This allowed the United States to, in the phrasing of the CIA, \enquote{truthfully deny} that any US \enquote{military planes} had flown over the USSR---which they had to do when the inevitable Soviet protest notes were filed after U-2 overflights began in 1956.\footcite[p.~109]{pedlow_central_1992}
+It is quoted here by the History Staff of the CIA.]{pedlow_central_1992} This allowed the United States to, in the phrasing of the CIA, \enquote{truthfully deny} that any US \enquote{military planes} had flown over the USSR---which they had to do when the inevitable Soviet protest notes were filed after U-2 overflights began in 1956.\footcite[109]{pedlow_central_1992}
 
@@
 If his only goal had been to achieve a level of plausible deniability in the event of a shootdown, then simply not wearing an Air Force uniform during the mission likely would have been sufficient.
-The classic \enquote{weather reconnaissance craft} cover had been used before with the ferret flights.\footcite[p.~45]{farquhar_aerial_2015} But policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation.
-\enquote{It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft} reads a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} No finite set of attributes could ensure that the Soviet Union wouldn't treat the U-2 plane---a massive leap in military surveillance technology---as a military development.
+The classic \enquote{weather reconnaissance craft} cover had been used before with the ferret flights.\footcite[45]{farquhar_aerial_2015} But policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation.
+\enquote{It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft} reads a top-secret CIA memo from 1956.\footcite[1]{miller_suggestions_1956} No finite set of attributes could ensure that the Soviet Union wouldn't treat the U-2 plane---a massive leap in military surveillance technology---as a military development.
 But if the people in charge of Project Aquatone could convince \emph{themselves} that it was a civilian project, they stood a better chance of persuading the Soviet Union.
@@
 While Powers was serving a 10-year sentence in Soviet prison, the two airmen were returned to the US in less than six months with only mild fanfare.
-Khrushchev wrote in his memoir that he wanted to continue \enquote{our general line of peaceful coexistence} and cited the resolution of this incident as an example of such, because the US was forced to make a formal request for the return of the airmen.\footcite[p.~256-257]{khrushchev_memoirs_2007} The only concessions the US made in return were to announce the discontinuation of its U-2 overflights (which Kennedy was already committed to) and to not make an issue of the illegal detention of the pilots.\footcite{time_cold_1961} The precedents from the Baltic Incident still applied: the USSR simply took the necessary immediate countermeasures and shot down the planes that it could.\footnote{While I argue that the response to the U-2 flight was within the reasonable bounds for espionage, I acknowledge that it was still unprecedented in the context of reconnaissance flights.
+Khrushchev wrote in his memoir that he wanted to continue \enquote{our general line of peaceful coexistence} and cited the resolution of this incident as an example of such, because the US was forced to make a formal request for the return of the airmen.\footcite[256-257]{khrushchev_memoirs_2007} The only concessions the US made in return were to announce the discontinuation of its U-2 overflights (which Kennedy was already committed to) and to not make an issue of the illegal detention of the pilots.\footcite{time_cold_1961} The precedents from the Baltic Incident still applied: the USSR simply took the necessary immediate countermeasures and shot down the planes that it could.\footnote{While I argue that the response to the U-2 flight was within the reasonable bounds for espionage, I acknowledge that it was still unprecedented in the context of reconnaissance flights.
 There are a number of reasonable explanations for why the U-2 shootdown escalated.
@@
 Eisenhower thought that Khrushchev might have been looking for an excuse to cancel his visit to Russia.
-See: \cite[p.~555]{eisenhower_waging_1965}}
+See: \cite[555]{eisenhower_waging_1965}}
 
@@
 He wrote in his memoir that \enquote{the commander of American forces in West Germany gave the order not to fly any closer than 50 kilometers from the border between East and West Germany.
-And no more incidents of that kind occurred.}\footcite[p.~256]{khrushchev_memoirs_2007} His son Sergei, who annotated the memoir, admitted that this statement was false: \enquote{In practice such incidents occurred again from 1961 onward, but instead of the U-2 the Americans now used various types of Phantom or SR-71.}\footcite[p.~258]{khrushchev_memoirs_2007} On at least two occasions after the U-2 Incident, an American overflight crossed over into Soviet territory and the Soviets responded with a diplomatic protest.
+And no more incidents of that kind occurred.}\footcite[256]{khrushchev_memoirs_2007} His son Sergei, who annotated the memoir, admitted that this statement was false: \enquote{In practice such incidents occurred again from 1961 onward, but instead of the U-2 the Americans now used various types of Phantom or SR-71.}\footcite[258]{khrushchev_memoirs_2007} On at least two occasions after the U-2 Incident, an American overflight crossed over into Soviet territory and the Soviets responded with a diplomatic protest.
 The US quickly apologized, and that settled the matter.\footcite{orlov_u-2_2007} The United States continued to use the U-2 extensively in other regions as well.\footnote{A recently declassified CIA report details the many locations in which U-2 photography was employed to gather intelligence after 1960, including China, India, Indonesia, Thailand, Tibet, Laos, North Vietnam, and Venezuela.
 Lyndon Johnson even clarified that his predecessor's order to end U-2 flights over the Soviet Bloc was not indefinitely binding---it was valid only until countermanded.
-For more, see: \cite[p.~195]{pedlow_central_1992}}
+For more, see: \cite[195]{pedlow_central_1992}}
 
@@
 government, and they made no attempt to conceal it.
-They didn't think we had the capability of \textelp{} acquiring irrefutable proof that the United States was using methods that were impermissible in peacetime.''\footcite[p.~239]{khrushchev_memoirs_2007} Sergei's annotations once again contradicted his father's claim that these methods were impermissible; more than forty American reconnaissance aircraft were shot down by the USSR during the Cold War.
+They didn't think we had the capability of \textelp{} acquiring irrefutable proof that the United States was using methods that were impermissible in peacetime.''\footcite[239]{khrushchev_memoirs_2007} Sergei's annotations once again contradicted his father's claim that these methods were impermissible; more than forty American reconnaissance aircraft were shot down by the USSR during the Cold War.
 Eisenhower, an insightful thinker on issues of intelligence, understood that beneath the grandstanding the fundamentals of international espionage still applied.
-\enquote{[Khrushchev's] government,} he wrote, \enquote{had been so notoriously involved in spying, especially in the United States, as to dwarf our activities, but by separating this particular type of espionage from all others he hoped to make convincing his charge of \enquote{warmongering.} To claim that because the equipment employed was an airplane with a camera, and therefore provocative of war, was plain silly, and I felt it necessary that the matter be put in perspective.} \footcite[p.~551]{eisenhower_waging_1965}
+\enquote{[Khrushchev's] government,} he wrote, \enquote{had been so notoriously involved in spying, especially in the United States, as to dwarf our activities, but by separating this particular type of espionage from all others he hoped to make convincing his charge of \enquote{warmongering.} To claim that because the equipment employed was an airplane with a camera, and therefore provocative of war, was plain silly, and I felt it necessary that the matter be put in perspective.} \footcite[551]{eisenhower_waging_1965}
 
@@
 Project Aquatone channeled the norms of espionage in the most crucial respect: its strategic value easily outweighed the consequences of being caught.
-No one understood this better than President Eisenhower, who, when he was occasionally questioned about the wisdom of the U-2 flights, would always reply: \enquote{Would you be ready to give back all of the information we secured from our U-2 flights over Russia if there had been no disaster to one of our planes in Russia?} In his telling, he never received an affirmative response.\footcite[p.~559]{eisenhower_waging_1965}
+No one understood this better than President Eisenhower, who, when he was occasionally questioned about the wisdom of the U-2 flights, would always reply: \enquote{Would you be ready to give back all of the information we secured from our U-2 flights over Russia if there had been no disaster to one of our planes in Russia?} In his telling, he never received an affirmative response.\footcite[559]{eisenhower_waging_1965}
 
@@
 The few scholars who study this period argue that the cumulative tension created by these flights contributed to an overall chilly tone in US-Soviet relations.
-Farquhar, for instance, refers to this as the \enquote{cycle of hostility,} in which the series of aerial incidents increased suspicions of military buildup on both sides, leading to even more aerial incidents as both powers attempted to verify their suspicions.\footcite[p.~43]{farquhar_aerial_2015} While it is true that the there were many aerial incidents, it is not at all clear how this affected any other aspect of US-Soviet diplomacy.
+Farquhar, for instance, refers to this as the \enquote{cycle of hostility,} in which the series of aerial incidents increased suspicions of military buildup on both sides, leading to even more aerial incidents as both powers attempted to verify their suspicions.\footcite[43]{farquhar_aerial_2015} While it is true that the there were many aerial incidents, it is not at all clear how this affected any other aspect of US-Soviet diplomacy.
 The US-Soviet relationship saw huge shifts, including periods of optimism, over the 11 years preceding the U-2 incident.
@@
 In 1954, President Eisenhower established the Technological Capabilities Panel (TCP), a task force to tackle the problem of America's vulnerability to a surprise attack.
-When the panel returned with its results, its report concluded that any great power conflict would result in heavy American losses, so all necessary steps should be taken to avoid one.\footcite[p.~67.
+When the panel returned with its results, its report concluded that any great power conflict would result in heavy American losses, so all necessary steps should be taken to avoid one.\footcite[67.
 The TCP report is also variously referred to as the \enquote{Killian Report,} \enquote{Surprise Attack Study,} or its formal title, \enquote{Meeting the Threat of a Surprise Attack.}]{killian_sputnik_1977} Its recommendations included traditional deterrence through force---furthering American missile capabilities---but intelligence-gathering capabilities were prioritized as well.
@@
 And of course, satellites can surveil dangerous territory without risking the life of a pilot.
-\enquote{Camera-toting satellites,} historian Walter McDougall explains, \enquote{circling the earth south to north [sic] in a polar orbit, could view the entire surface of the earth as it rotated below, return to any location in a few days' time, hone in on suspicious areas, and do it all under the legal cover of freedom of space---if such legal cover could be established.}\footcite[p.~117]{mcdougall_heavens_1985}
+\enquote{Camera-toting satellites,} historian Walter McDougall explains, \enquote{circling the earth south to north [sic] in a polar orbit, could view the entire surface of the earth as it rotated below, return to any location in a few days' time, hone in on suspicious areas, and do it all under the legal cover of freedom of space---if such legal cover could be established.}\footcite[117]{mcdougall_heavens_1985}
 
@@
 The undefined legal status of space presented both an opportunity and a challenge.
-The Soviet Union would likely recognize the enormous strategic advantage of a photo-reconnaissance satellite and object to its deployment using any international norms possible; even the U-2 overflight program was considered so risky that it was originally intended to last only a year or two.\footcite[p.~33]{lindgren_trust_2000} It was therefore necessary to present reconnaissance satellites in a manner that would limit the grounds for objection.
+The Soviet Union would likely recognize the enormous strategic advantage of a photo-reconnaissance satellite and object to its deployment using any international norms possible; even the U-2 overflight program was considered so risky that it was originally intended to last only a year or two.\footcite[33]{lindgren_trust_2000} It was therefore necessary to present reconnaissance satellites in a manner that would limit the grounds for objection.
 By the time that a reconnaissance satellite launched, it would have to have already been established that any nation had the unquestioned right to launch a satellite of any kind: a principle called \enquote{freedom of space.}
 
-Eisenhower accepted the recommendations of NSC-5520, a memo arguing that a scientific satellite could be used as diplomatic tool to promote a norm establishing ``freedom of space.'' The proposal tracked closely with an earlier report suggesting that the US first launch an innocuous ``experimental satellite'' that did not cross over Russian territory, and gauge the diplomatic reaction.\footcite[p.~21]{kecskemetic_satellite_1950} The political framing of the satellite is discussed at the top of NSC-5520, in a section entitled \emph{General Considerations} (emphasis mine):
+Eisenhower accepted the recommendations of NSC-5520, a memo arguing that a scientific satellite could be used as diplomatic tool to promote a norm establishing ``freedom of space.'' The proposal tracked closely with an earlier report suggesting that the US first launch an innocuous ``experimental satellite'' that did not cross over Russian territory, and gauge the diplomatic reaction.\footcite[21]{kecskemetic_satellite_1950} The political framing of the satellite is discussed at the top of NSC-5520, in a section entitled \emph{General Considerations} (emphasis mine):
 \newline
@@
 
-The plan was straightforward: use the civilian space program to promote the freedom of space principle, which, once established, would protect the American right to quietly launch a spy satellite.\footcite[p.~119]{day_eye_2015} In fact, since the Soviet Union had already announced plans to launch a satellite, the existence of the Soviet project could be use to blunt the inevitable Soviet protests against American photo-reconnaissance satellites.\footcite[p.~120]{mcdougall_heavens_1985} When the Soviet Union launched the world's first artificial satellite on October 4, 1957, the plan succeeded.
+The plan was straightforward: use the civilian space program to promote the freedom of space principle, which, once established, would protect the American right to quietly launch a spy satellite.\footcite[119]{day_eye_2015} In fact, since the Soviet Union had already announced plans to launch a satellite, the existence of the Soviet project could be use to blunt the inevitable Soviet protests against American photo-reconnaissance satellites.\footcite[120]{mcdougall_heavens_1985} When the Soviet Union launched the world's first artificial satellite on October 4, 1957, the plan succeeded.
 The \emph{Sputnik} launch put the United States in an unbeatable negotiating position.
-The Soviet Union could hardly denounce the American satellite program as suspicious when it had just launched the world's first scientific satellite---without asking permission to orbit over the United States.\footcite[p.~40]{peebles_corona_1997} By allowing \emph{Sputnik} to orbit over American territory without diplomatic objection, the US got to appear magnanimous while fulfilling one of its highest-priority strategic goals.
+The Soviet Union could hardly denounce the American satellite program as suspicious when it had just launched the world's first scientific satellite---without asking permission to orbit over the United States.\footcite[40]{peebles_corona_1997} By allowing \emph{Sputnik} to orbit over American territory without diplomatic objection, the US got to appear magnanimous while fulfilling one of its highest-priority strategic goals.
 
 The available evidence of Eisenhower's internal deliberations overwhelmingly suggests that his administration never intended to be first in space.
-Launching the world's first artificial satellite was assigned the lowest possible priority because he determined that other goals---finishing the ICBM, monitoring Soviet R\&D, and framing the satellite program as a civilian scientific endeavor---were of greater strategic importance.\footcite[p.~123-124]{mcdougall_heavens_1985} Eisenhower himself repeatedly dismissed concerns about the prestige of being the first nation to orbit an object in space, openly contemptuous of the idea that space was any kind of race at all.\footcite[p.~100]{lindgren_trust_2000} His complete disdain for these ``stunts'' would prove disastrous politically, but his restraint was crucial to establishing a strategically beneficial precedent.\footcite[p.~134]{day_eye_2015}
+Launching the world's first artificial satellite was assigned the lowest possible priority because he determined that other goals---finishing the ICBM, monitoring Soviet R\&D, and framing the satellite program as a civilian scientific endeavor---were of greater strategic importance.\footcite[123-124]{mcdougall_heavens_1985} Eisenhower himself repeatedly dismissed concerns about the prestige of being the first nation to orbit an object in space, openly contemptuous of the idea that space was any kind of race at all.\footcite[100]{lindgren_trust_2000} His complete disdain for these ``stunts'' would prove disastrous politically, but his restraint was crucial to establishing a strategically beneficial precedent.\footcite[134]{day_eye_2015}
 
@@
 The group tasked with implementing NSC-5520 made the surprising choice to commission the unproven Navy Research Laboratory (NRL) \emph{Vanguard} rocket, instead of the Army Ballistic Missile Agency (ABMA) proposal to launch the satellite with better-tested missile technology.
-Launching the first artificial satellite with a ballistic missile threatened to compromise two national security priorities: it might delay the development of the ICBM, an existential threat to American military superiority, and it would jeopardize the civilian character of the satellite program, which the committee members knew was necessary to establish ``freedom of space.''\footcite[p.~122]{mcdougall_heavens_1985} Led by famous rocketeer (and ex-Nazi) Wernher von Braun, the ABMA engineers vigorously protested the decision, because they believed---correctly---that their proposal would succeed first.
+Launching the first artificial satellite with a ballistic missile threatened to compromise two national security priorities: it might delay the development of the ICBM, an existential threat to American military superiority, and it would jeopardize the civilian character of the satellite program, which the committee members knew was necessary to establish ``freedom of space.''\footcite[122]{mcdougall_heavens_1985} Led by famous rocketeer (and ex-Nazi) Wernher von Braun, the ABMA engineers vigorously protested the decision, because they believed---correctly---that their proposal would succeed first.
 But being first in space was not the goal, establishing a legal precedent was.
-Launching a satellite with no scientific purpose other than demonstrating superiority would compromise the moral high ground which the US desperately needed to secure.\footcite[p.~129-131.
+Launching a satellite with no scientific purpose other than demonstrating superiority would compromise the moral high ground which the US desperately needed to secure.\footcite[129-131.
 Though the NRL is run by the Navy, it functioned like a civilian scientific operation.
@@
 Operationally, \emph{Vanguard} was a disaster.
-The rocket experienced an embarrassing setback on December 6, 1957, when it failed to launch a three-and-a-half pound satellite, just after the USSR had successfully launched \emph{Sputnik II}.\footcite[p.~119.
+The rocket experienced an embarrassing setback on December 6, 1957, when it failed to launch a three-and-a-half pound satellite, just after the USSR had successfully launched \emph{Sputnik II}.\footcite[119.
 This time, the satellite carried a dog named Laika.
@@
 Finally given the chance, Von Braun and his engineers quickly accomplished what they had always said they would: modify a Jupiter-C ballistic missile and send a satellite into orbit.
-The Army launched \emph{Explorer 1} on January 31, and it carried a pioneering array of miniaturized electronics, including two micrometeoroid detectors and a Geiger counter.\footcite[p.~168]{mcdougall_heavens_1985} \emph{Vanguard 1} finally launched soon after.
+The Army launched \emph{Explorer 1} on January 31, and it carried a pioneering array of miniaturized electronics, including two micrometeoroid detectors and a Geiger counter.\footcite[168]{mcdougall_heavens_1985} \emph{Vanguard 1} finally launched soon after.
 
@@
 % Now that the space age was underway, the next step was to normalize photographic satellite overflights.
-Success was going to require tight discipline across all levels of government, but information about WS-117L, the Air Force reconnaissance satellite program, had already begun to leak.\footcite[p.~96]{lindgren_trust_2000}
+Success was going to require tight discipline across all levels of government, but information about WS-117L, the Air Force reconnaissance satellite program, had already begun to leak.\footcite[96]{lindgren_trust_2000}
 
 The Soviet Union unintentionally set an early precedent for ``freedom of space,'' but that did not automatically justify the use of outer space for reconnaissance.
-As far as the international community was concerned, it had only been established that overflights were permissible for scientific purposes; military reconnaissance was still an open question.\footcite[p.~47-48]{peebles_corona_1997} Like many of Eisenhower's space policy decisions, his solution for reconnaissance satellites was a brilliant long-term success---but a complete mystery to the public and utterly baffling to most of the military establishment.
+As far as the international community was concerned, it had only been established that overflights were permissible for scientific purposes; military reconnaissance was still an open question.\footcite[47-48]{peebles_corona_1997} Like many of Eisenhower's space policy decisions, his solution for reconnaissance satellites was a brilliant long-term success---but a complete mystery to the public and utterly baffling to most of the military establishment.
 Eisenhower simply selected the most immediately promising reconnaissance satellite---a single-use film-retrieval satellite, codenamed Corona---and canceled it.
-Air Force personnel were ``thunderstruck.''\footcite[p.~45]{peebles_corona_1997} Now that American reconnaissance satellites now seemingly a few years out, the administration began a multi-pronged approach to legitimizing spy satellites.
+Air Force personnel were ``thunderstruck.''\footcite[45]{peebles_corona_1997} Now that American reconnaissance satellites now seemingly a few years out, the administration began a multi-pronged approach to legitimizing spy satellites.
 Publicly, the US government advocated for a ``freedom of space'' policy that explicitly included reconnaissance satellites; secretly, it resurrected the Corona project under deep-cover CIA administration.
@@
 The public half of Eisenhower's strategy required international approval for space-based overflights, so the US advocated for a United Nations Ad Hoc Committee on the Peaceful Uses of Outer Space (COPUOS) to deal with this issue.
-The Soviets, worried about having the limitations of their ICBM program exposed, refused to participate for years.\footcite[p.~140]{day_eye_2015} Meanwhile, the actual spy satellite program progressed at an accelerated pace, and with unmatched secrecy.\footcite[p.~51.
+The Soviets, worried about having the limitations of their ICBM program exposed, refused to participate for years.\footcite[140]{day_eye_2015} Meanwhile, the actual spy satellite program progressed at an accelerated pace, and with unmatched secrecy.\footcite[51.
 The team started with only 30 personnel and eventually swelled to 300, almost none of whom knew the entire scope of the project.
 Team members had to take different routes to work to avoid being followed and were never allowed to say the word ``Corona'' on the telephone, or even the abbreviation ``C.'']{peebles_shadow_2000} When the satellite began making test launches, the US heavily promoted its new \emph{Discoverer} program as a scientific achievement---confident that its status as a front for the Corona project could not be proven false the way the U-2's cover had been.
-When it finally succeeded in recovering a \emph{Discoverer} satellite from orbit, the US government released photographs of the first successful capsule triumphantly floating in the water, and an American flag retrieved from within was presented to the president.\footcite[p.~83]{peebles_shadow_2000}
+When it finally succeeded in recovering a \emph{Discoverer} satellite from orbit, the US government released photographs of the first successful capsule triumphantly floating in the water, and an American flag retrieved from within was presented to the president.\footcite[83]{peebles_shadow_2000}
 
@@
 A rival project that developed the film midair and transmitted the data back down failed to provide intelligence of sufficient quality.
-(\cite[p.~203-204]{brugioni_eyes_2010}) The last film-based orbiting reconnaissance camera developed for the U.S.
+(\cite[203-204]{brugioni_eyes_2010}) The last film-based orbiting reconnaissance camera developed for the U.S.
 Government was the KH-9 HEXAGON, which operated between 1971 and 1986.
@@
 What remained of the capsule would hopefully descend within a 200-by-60 mile recovery zone (``the ballpark''), but at the very least within the boundaries of an additional 400-mile section (``the outfield'').
-Ideally, it would be caught in midair by a cargo plane.\footcite[p.~56]{peebles_corona_1997}
+Ideally, it would be caught in midair by a cargo plane.\footcite[56]{peebles_corona_1997}
 
@@
 From February to August 1960, right in the middle of the U-2 incident, twelve test launches failed before the US achieved the first successful recovery of a Corona satellite---of any object from orbit in the human history, for that matter---when it fished \emph{Discoverer 13} out of the sea.\footnote{Ships on standby were able to retrieve \emph{Discoverer 13} even though it fell outside the ballpark.
-Diplomatically and operationally, the mission was a home run.} The US repeated that feat with \emph{Discoverer 14} a week later, and this time it carried a camera, bringing invaluable intelligence back from space that dealt a mortal blow to the myth of the missile gap.\footcite[p.~101-102.
+Diplomatically and operationally, the mission was a home run.} The US repeated that feat with \emph{Discoverer 14} a week later, and this time it carried a camera, bringing invaluable intelligence back from space that dealt a mortal blow to the myth of the missile gap.\footcite[101-102.
 The previous launch failures had been so demoralizing that \emph{Discoverer 13} only carried diagnostic equipment, no cameras.]{lindgren_trust_2000} The launches kept coming---the next one to return high quality photos was \emph{Discoverer 18} in December.
-Eisenhower left office soon after, but between \emph{Discoverer 18} and \emph{Discoverer 38}, twenty launches were performed, two of which carried actual radiometry equipment (in support of Corona's cover story), and seven produced film of high enough quality for the mission to be considered a complete success.\footcite[p.~103]{lindgren_trust_2000}
+Eisenhower left office soon after, but between \emph{Discoverer 18} and \emph{Discoverer 38}, twenty launches were performed, two of which carried actual radiometry equipment (in support of Corona's cover story), and seven produced film of high enough quality for the mission to be considered a complete success.\footcite[103]{lindgren_trust_2000}
 
-The Soviet Union protested the very first \emph{Discoverer} launch, correctly alleging its military utility even though the satellite had failed in its purpose (and would fail eleven more times).\footcite[p.~140]{day_eye_2015} Until 1963, Soviet propaganda continually railed against the Americans, in domestic media and international forums, for attempting to militarize space.\footcite[p.~271]{mcdougall_heavens_1985} The USSR sponsored a UN resolution to prohibit espionage from space, and Khrushchev vowed that his military would shoot down spy satellites, just as they had the U-2.\footcite[p.~166]{day_eye_2015} In the first few years after the United States developed satellite reconnaissance capability, it appeared that the USSR was using all the means at its disposal to end the satellite overflights.
+The Soviet Union protested the very first \emph{Discoverer} launch, correctly alleging its military utility even though the satellite had failed in its purpose (and would fail eleven more times).\footcite[140]{day_eye_2015} Until 1963, Soviet propaganda continually railed against the Americans, in domestic media and international forums, for attempting to militarize space.\footcite[271]{mcdougall_heavens_1985} The USSR sponsored a UN resolution to prohibit espionage from space, and Khrushchev vowed that his military would shoot down spy satellites, just as they had the U-2.\footcite[166]{day_eye_2015} In the first few years after the United States developed satellite reconnaissance capability, it appeared that the USSR was using all the means at its disposal to end the satellite overflights.
 
 Khrushchev had every incentive to discourage satellite reconnaissance, because the intelligence that the US gained through Corona was absolutely devastating to Soviet security.
-One of Khrushchev's top national security concerns at this time was concealing the Soviet Union's relative nuclear weakness.\footcite[p.~133]{brugioni_eyes_2010} The moment at which the US discovered that the USSR had far fewer missiles than it claimed was the exact moment at which that disparity was greatest.
-The missile gap actually went in the other direction---in 1962, the US had more than double the ICBMs, maintained a substantial lead in manned bombers, and had just begun to deploy nuclear submarines.\footcite[p.~251]{mcdougall_heavens_1985} If Khrushchev genuinely feared an American attack, absolutely nothing would have been a greater threat to Soviet security than the collapse of the USSR's most significant deterrent argument---a collapse brought about by American spy satellites.
-\emph{Discoverer 14} and \emph{Discoverer 18} successfully photographed the sites that intelligence analysts considered most likely to host operational ICBMs, and found nothing.\footcite[p.~379]{brugioni_eyes_2010}
+One of Khrushchev's top national security concerns at this time was concealing the Soviet Union's relative nuclear weakness.\footcite[133]{brugioni_eyes_2010} The moment at which the US discovered that the USSR had far fewer missiles than it claimed was the exact moment at which that disparity was greatest.
+The missile gap actually went in the other direction---in 1962, the US had more than double the ICBMs, maintained a substantial lead in manned bombers, and had just begun to deploy nuclear submarines.\footcite[251]{mcdougall_heavens_1985} If Khrushchev genuinely feared an American attack, absolutely nothing would have been a greater threat to Soviet security than the collapse of the USSR's most significant deterrent argument---a collapse brought about by American spy satellites.
+\emph{Discoverer 14} and \emph{Discoverer 18} successfully photographed the sites that intelligence analysts considered most likely to host operational ICBMs, and found nothing.\footcite[379]{brugioni_eyes_2010}
 
 % The U-2 overflights had sufficiently debunked the \enquote{bomber gap,} and seemed to suggest that no \enquote{missile gap} existed either.\footcite{goodpaster_cold_2003} But as the CIA report on U-2 photography showed, being unable to find ICBM sites was not conclusive proof that they did not exist, and the vast majority of possible missile sites remained un-photographed.
-Confusion about Soviet long-range attack capabilities still existed at various levels of government, confusion which ultimately led to the doomed flight of Gary Powers.\footcite[p.~344]{brugioni_eyes_2010}
+Confusion about Soviet long-range attack capabilities still existed at various levels of government, confusion which ultimately led to the doomed flight of Gary Powers.\footcite[344]{brugioni_eyes_2010}
 
@@
 Urban targets had gone from the majority to next to last in number.
-The unilateral British plan also underwent a similar change.\footcite[p.~139]{peebles_corona_1997}
+The unilateral British plan also underwent a similar change.\footcite[139]{peebles_corona_1997}
 \end{quoteblock}
@@
 And yet the only thing the Soviet Union did in response was publicly protest the spy satellites, and refuse to accept American proposals for a legal regime that included the rights of reconnaissance overflights.
-Those protests continued for about 3 years, until the Soviets developed their own working spy satellites, and ``freedom of space'' became a mutually-agreeable reality.\footcite[p.~271-275]{mcdougall_heavens_1985}
+Those protests continued for about 3 years, until the Soviets developed their own working spy satellites, and ``freedom of space'' became a mutually-agreeable reality.\footcite[271-275]{mcdougall_heavens_1985}
 
@@
 The United States also incurred significant costs for diverting resources to signal that its satellite activities constituted espionage, highlighting the contrast with the more direct, militaristic approach.
-Though there is less direct evidence of its intent, the Soviet Union functionally applied the same playbook when it launched its own satellites a few years later---Khrushchev even claimed in 1960 that the Soviet Union had already put cameras in some of its satellites.\footcite[p.~353.
+Though there is less direct evidence of its intent, the Soviet Union functionally applied the same playbook when it launched its own satellites a few years later---Khrushchev even claimed in 1960 that the Soviet Union had already put cameras in some of its satellites.\footcite[353.
 As far as we know, these camera-equipped satellites, if they actually existed, were not successful enough to be considered spy satellites.]{brugioni_eyes_2010} The effective cessation of diplomatic complaints regarding spy satellites, from either side, is the expected outcome from two sides that are both acting to code their actions as espionage.
@@
 Alternate explanations for why the Soviet Union allowed the legitimization of spy satellites rely exclusively on one's faith in diplomatic norms.
-Most boil down to gamesmanship: the United States effectively leveraged scientific diplomacy and covert operations; \emph{Sputnik} made it much harder for the USSR to argue that reconnaissance satellites were an imperialist invasion of territorial sovereignty; Nikita Khrushchev, in a heated moment at the 1960 summit, was said to have claimed that ``any nation in the world who wanted to photograph the Soviet areas by satellite was completely free to do so.''\footcite[p.~556]{eisenhower_waging_1965} None of these explanations, on their own, fully capture why the Soviet Union would assent to an espionage regime that asymmetrically neutralized one of its most significant security advantages.
+Most boil down to gamesmanship: the United States effectively leveraged scientific diplomacy and covert operations; \emph{Sputnik} made it much harder for the USSR to argue that reconnaissance satellites were an imperialist invasion of territorial sovereignty; Nikita Khrushchev, in a heated moment at the 1960 summit, was said to have claimed that ``any nation in the world who wanted to photograph the Soviet areas by satellite was completely free to do so.''\footcite[556]{eisenhower_waging_1965} None of these explanations, on their own, fully capture why the Soviet Union would assent to an espionage regime that asymmetrically neutralized one of its most significant security advantages.
 
@@
 At scale, espionage removes risk from the international system, and the rare instances where espionage is codified suggests that states generally view espionage this way as well.
-The Strategic Arms Limitation Treaty of 1972 (otherwise known as The ABM Treaty, or SALT 1) enshrined in law that signatories are not permitted to interfere with each others' ``national technical means of verification,'' a tacit ban on ASAT weapons.\footcite[p.~431]{mcdougall_heavens_1985} Without having to say ``spy satellites,'' SALT 1 made them not just legal, but essential.
+The Strategic Arms Limitation Treaty of 1972 (otherwise known as The ABM Treaty, or SALT 1) enshrined in law that signatories are not permitted to interfere with each others' ``national technical means of verification,'' a tacit ban on ASAT weapons.\footcite[431]{mcdougall_heavens_1985} Without having to say ``spy satellites,'' SALT 1 made them not just legal, but essential.
 The peaceful uses of aerial reconnaissance were formalized at the end of the Cold War, when members of NATO and the Warsaw Pact signed the \enquote{Treaty on Open Skies,} which established guidelines for unarmed aerial surveillance flights in the spirit of open information and de-escalation.\footcite{organization_for_security_and_co-operation_in_europe_treaty_1992} This treaty was almost identical to the one first suggested by Eisenhower in 1955.\footcite{center_for_arms_control_and_non-proliferation_fact_2017} With certain restrictions, the type of flight that altered the course of the Cold War is now officially routine and permissible.
@@
 The advent of mutual reconnaissance solved what had, until then, been the single most intractable problem of the Cold War---treaty verification.
-Rae Huffstuttler, later a director of the CIA's National Photographic Interpretation Center (the same institution that conducted the survey of U-2 photography), said that ``imagery set the stage for the arms limitation talks \textelp{} With imagery, we could go to the numbers-based strategic arms limitations negotiation with a high degree of confidence.''\footcite[p.~403]{brugioni_eyes_2010} With satellite imagery in hand, the United States could enter the negotiations with appropriate estimates and leave with a means of verifying that the treaty had been carried out.
-Khrushchev claimed for years to be in favor of arms control, but talks always collapsed because of the US demand for on-site inspections, which the Soviet Union always refused.\footcite[p.~255]{mcdougall_heavens_1985} Satellites served the same purpose, and the Strategic Arms Limitation Treaty of 1972 (otherwise known as The ABM Treaty, or SALT 1) enshrined in law that signatories are not permitted to interfere with each others' ``national technical means of verification,'' a tacit ban on ASAT weapons.\footcite[p.~431]{mcdougall_heavens_1985} Without having to say ``spy satellites,'' SALT 1 made them not just legal, but essential.
+Rae Huffstuttler, later a director of the CIA's National Photographic Interpretation Center (the same institution that conducted the survey of U-2 photography), said that ``imagery set the stage for the arms limitation talks \textelp{} With imagery, we could go to the numbers-based strategic arms limitations negotiation with a high degree of confidence.''\footcite[403]{brugioni_eyes_2010} With satellite imagery in hand, the United States could enter the negotiations with appropriate estimates and leave with a means of verifying that the treaty had been carried out.
+Khrushchev claimed for years to be in favor of arms control, but talks always collapsed because of the US demand for on-site inspections, which the Soviet Union always refused.\footcite[255]{mcdougall_heavens_1985} Satellites served the same purpose, and the Strategic Arms Limitation Treaty of 1972 (otherwise known as The ABM Treaty, or SALT 1) enshrined in law that signatories are not permitted to interfere with each others' ``national technical means of verification,'' a tacit ban on ASAT weapons.\footcite[431]{mcdougall_heavens_1985} Without having to say ``spy satellites,'' SALT 1 made them not just legal, but essential.
 
 % In February 1962, Francis Gary Powers walked across a bridge into West Berlin at the same time as Russian Colonel Rudolph Abel, an American prisoner.
-Powers' father had suggested to the State Department that they might be traded, spy for a spy.\footcite[p.~239]{powers_operation_2004} \enquote{This,} Eisenhower wrote, ``was a tacit admission by Khrushchev that our `outrageous' U-2 pilots have their opposite numbers operating within the borders of this country.''\footcite[p.~558]{eisenhower_waging_1965}
+Powers' father had suggested to the State Department that they might be traded, spy for a spy.\footcite[239]{powers_operation_2004} \enquote{This,} Eisenhower wrote, ``was a tacit admission by Khrushchev that our `outrageous' U-2 pilots have their opposite numbers operating within the borders of this country.''\footcite[558]{eisenhower_waging_1965}

### 2021-04-17 article.tex
Rewrite introduction

@@
 These applications exposed not just the personal information of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses---compromising the information of 22.1 million people in total.\footcite{nakashima_hacks_2015} US officials feared that the stolen data could help foreign intelligence services unmask undercover CIA officers, and the information could identify American officials who might be susceptible to blackmail, or worse, recruitment.
-\enquote{[I]f you have my [application for security clearance],} FBI Director James Comey admitted, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.
-Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015} In the case of the OPM hack, a vivid imagination wasn't necessary; the US had already established that the Chinese military was likely behind the sophisticated attack.\footcite{koerner_inside_2016} But even though it possessed conclusive digital forensics, the Obama administration refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018.
+FBI Director James Comey admitted that, with the kind of information leaked, a foreign intelligence service could ``know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.''\footcite[Comey is referring specifically to his SF-86, the form that one files with OPM to apply for a security clearance.]{nakashima_hacks_2015} In the case of the OPM hack, it wasn't necessary to imagine who was in possession of that information; the US had already established that the Chinese military was likely behind the sophisticated attack.\footcite{koerner_inside_2016} But even though it possessed conclusive digital forensics, the Obama administration refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018.
 See \cite{sanger_trump_2018}}
 
-The US government's tepid response to the OPM hack illustrates a potentially surprising aspect of \nth{21} century diplomacy: when a cyberattack looks like a traditional intelligence operation, little to no diplomatic consequences will follow.
-Within political and academic circles, this phenomenon is often taken for granted; cyber-espionage is treated like \enquote{traditional} espionage because they are both forms of espionage.
-The scale of the OPM hack, however, stretches our understanding of the kind of damage that espionage can achieve.
-``This is one of those cases where you have to ask, `Does the size of the operation change the nature of it?'\thinspace'' said one anonymous senior intelligence official.
-\enquote{Clearly, it does.}\footcite{sanger_u.s._2015} The actions of the US government say otherwise: at least for now, even operations on the scale of China's OPM infiltration still do not merit significant retaliation.
+The OPM hack was not the first time the United States Federal Government suffered a data breach at the hands of a foreign adversary, nor would it be last.
+Cases like the OPM hack, where sensitive data is stolen but no physical damage is done, are typically classified as \enquote{espionage;} according to the Council on Foreign Relations, there have been 73 such state-sponsored cyberattacks against the United States from 2010 through 2020 (Table \ref{response-pct}).\footnote{These numbers were compiled from the Council on Foreign Relations' Cyber Operations Tracker, which is a publicly-sourced record of state-sponsored cyberattacks, updated quarterly.
+The data chosen for this analysis spans from the beginning of 2010 through the end of 2020.
+For more, see \cite{council_on_foreign_relations_new_2019}} Of those 73 instance of cyber-espionage, only six received any public response at all.
+By contrast, both of the state-sponsored cyberattacks classified as \enquote{sabotage} received official responses from the US government.
 
-To evaluate whether espionage norms should apply in cyberspace, one first needs to understand espionage—--specifically, why espionage has traditionally been met with few diplomatic consequences.
-Though not codified in US policy, the light touch applied to espionage is an open secret.
-A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].}\footnote{In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets.
-See \cite{finklea_cyber_2015}} For activities that fall under the umbrella of intelligence-gathering, counterintelligence rather than prosecution is considered the appropriate response.
-CIA Director John Brennan once described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, was particularly frank: \enquote{You have to kind of salute the Chinese for what they did [to OPM].
-If we had the opportunity to do that, I don’t think we’d hesitate for a minute.}\footcite{pepitone_clapper_2015} This permissive attitude is norm that governs espionage diplomacy, and its strategic purpose is rarely questioned or justified.
+Espionage is by far the most common form of cyberattack and governments do not always reveal when they have been compromised, so it is not terribly convincing to say that most cyber-espionage was ignored by the US Government while all 2 out 2 incidents of cyber-sabotage were not.
+But if we expand our scope globally for just a moment, it becomes apparent that the same pattern broadly holds.
+Stripped of all other context, espionage is the single least likely form of cyberattack to generate any type of public response.
+
+% latex table generated in R 3.4.3 by xtable 1.8-2 package
+% Sat Apr  3 15:30:42 2021
+\begin{table}[ht]
+\centering
+\setlength{\extrarowheight}{2pt}
+\small
+\begin{tabular}{lrrr}
+  \hline
+Type & Incidents & Public responses & Response \% \\
+  \hline
+Defacement &   5 &   3 & 60.00 \\
+  Doxing &   6 &   3 & 50.00 \\
+  Data destruction &  11 &   5 & 45.45 \\
+  Sabotage &  22 &   9 & 40.91 \\
+  Financial Theft &   8 &   2 & 25.00 \\
+  Denial of service &  13 &   3 & 23.08 \\
+  Espionage & 332 &  58 & 17.47 \\
+   \hline
+\end{tabular}
+\caption{Global cyber-incident response rates, 2010-2020}
+\label{response-pct}
+\end{table}
+
+Based on this data, one could hypothesize that espionage does not always merit the same level of retaliation that sabotage or financial theft do because espionage is the least damaging form of cyberattack.
+While this makes sense in theory, it is not obviously true of the cyber-sabotage we've seen in practice.
+Consider how the OPM hack compares to a contemporary incident of state-sponsored sabotage.
+In 2014, hackers affiliated with the North Korean government leaked damaging emails from Sony executives and wiped some of Sony's servers.\footcite[This hack was allegedly in retaliation for the upcoming James Franco/Seth Rogen comedy about assassinating North Korean leader Kim Jong-Un.]{barnes_sony_2014} White House Press Secretary Josh Earnest alleged that the Sony hack was \enquote{destructive activity with malicious intent that was initiated by a sophisticated actor,} which he said merits \enquote{a proportional response.}\footcite{nakashima_white_2014} The Sony hack would be the first time a sitting US president publicly accused a foreign nation of launching a cyberattack and promised retaliation.\footcite{sanger_u.s._2016} The US made good on its threat and imposed sanctions on North Korea the following year.\footcite{lederman_us_2015} So why did the Sony hack---a newsworthy but far less widespread data breach---receive a significantly more serious response than a direct attack on a federal office that stole the private information of 22 million Americans?
+
+The key distinction between the two events is the way that they are characterized by policymakers.
+Characterizations are not necessarily a first principle---one could choose which response is most politically expedient, and then characterize the incident accordingly---but in this case the two incidents are different in ways that would not work as an \emph{ex post facto} justification.
+President Obama described the Sony hack as \enquote{cyber vandalism} and Republican senators described it as \enquote{terrorism.}\footcite{fung_obama_2014} The OPM hack was instead described by the administration as a form of \enquote{intelligence-gathering activities that all countries do.}\footcite{finklea_cyber_2015} John Brennan, Director of the CIA, described foreign spying on domestic political institutions as ``fair game.''\footcite{sanger_u.s._2016} Displaying a candor atypical of the intelligence community,  Director of National Intelligence James Clapper ruefully admitted: ``[Y]ou have to kind of salute the Chinese for what they did.
+If we had the opportunity to do that, I don't think we'd hesitate for a minute.''\footcite{pepitone_clapper_2015}
+
+To the US government, stealing classified data from a federal office looks like something a spy might do, but wiping a movie studio's servers does not.
+While this explanation might ring true---cyber-espionage is espionage, sure---it is also essentially a tautology.
+It raises the question: why does it matter whether something qualifies as espionage, and is that a label that continues to apply even when the scale of what espionage can accomplish is magnified by new technology?
+Like a traditional intelligence operation, the OPM hack \enquote{merely} involved stealing sensitive information, but the quantity of compromised data was orders of magnitude beyond what could be accomplished by photographing the contents of a file cabinet and passing it along in a briefcase.
+\enquote{This is one of those cases where you have to ask, `Does the size of the operation change the nature of it?'\thinspace} said one anonymous senior intelligence official.
+\enquote{Clearly, it does.}\footcite{sanger_u.s._2015} The actions of the US government still say otherwise: at least for now, even operations on the scale of China's OPM infiltration do not merit significant retaliation.
+
+This article seeks to explain how an old set of espionage norms continues to weather disruptive technological change.
+To do so first requires one to explain how it is possible that the routine minimization of espionage response happens at all.
+Extant espionage literature tends to focus on prescriptive questions---should espionage exist?
+is espionage legal?---rather than descriptive ones.
+There are few academic scholars who examine the theoretical implications of espionage in practice: what does it mean for existing theories of IR that a category of inter-state action routinely receives retaliation far less severe than the action's national security implications would seem to merit?
+In the first section of this paper I review the existing scholarship on norms and note the areas in which espionage does not neatly fit into the existing research.
+
+In the latter section of this paper I demonstrate how espionage norms have been deployed during previous periods of tectonic technological change.
+The historical record demonstrates that the idea of espionage is deeply meaningful to the individual actors that decide state policy; if an action is considered to be espionage, it exists in a self-contained ontological space, with rules and consequences unaffected by rational cost analysis.
+The people who are in charge of these matters---what operations to undertake, how much risk to absorb, whether to respond to an operation conducted against them---intuit whether something \emph{counts as} espionage, and adjust their own actions to fit within that paradigm.
+The benefits for doing so are clear: if the aggrieved party considers an operation to be espionage, it is much less likely to retaliate, even when the consequences for that operation are deeply damaging.
+Cyber-espionage is a novel challenge for the intelligence community, but it is not the first paradigmatic shift in intelligence capabilities.
+Spy planes and spy satellites radically re-defined what espionage could accomplish and changed the course of the Cold War---the consequences for using them did not evolve accordingly.
+
+\newpage
 
@@
 
-This article posits that the universal tolerance of espionage in the international system can be explained through a rational calculation of each state's incentives.
+% This article posits that the universal tolerance of espionage in the international system can be explained through a rational calculation of each state's incentives.
 On balance, all states derive benefits from preserving a system in which the consequences for espionage are minimized, so they refrain from imposing particularly harsh punishments for its practice.
@@
 
-To evaluate each state's incentive, I first establish a theoretical framework for analyzing the role that espionage plays in international relations (IR) and the diplomatic norms that are associated with it.
+% To evaluate each state's incentive, I first establish a theoretical framework for analyzing the role that espionage plays in international relations (IR) and the diplomatic norms that are associated with it.
 Through this framework, I define the bounds of traditional espionage and why rational states might calculate that it is in their best interests to continue tolerating it.
@@
 
-% latex table generated in R 3.4.3 by xtable 1.8-2 package
-% Sat Aug 24 13:46:46 2019
-\begin{table}[ht]
-\centering
-\begin{tabular}{lrrr}
-  \hline
-Type & Incidents & Official responses & Response \% \\
-  \hline
-Defacement &   3 &   2 & 66.67 \\
-  Data destruction &   8 &   5 & 62.50 \\
-  Doxing &   6 &   3 & 50.00 \\
-  Sabotage &  14 &   5 & 35.71 \\
-  Denial of service &  17 &   6 & 35.29 \\
-  Espionage & 231 &  44 & 19.05 \\
-   \hline
-\end{tabular}
-\caption{Global cyber-incident response rates}
-\label{response-pct}
-\end{table}
-
 While most of the information that we have on espionage norms is qualitative, there is some data to support the limited consequences of espionage norms as well.
@@
 Of the hundreds of state-sponsored cyberattacks known to the public, attacks classified as espionage are by far the most common.
-Cyber-espionage is, however, easily the least likely form of cyberattack to garner an official response from the attacked state (Table \ref{response-pct}).\footnote{These numbers were compiled from the Council on Foreign Relations' Cyber Operations Tracker, which is a publicly-sourced record of state-sponsored cyberattacks, updated quarterly.
-Of the 323 listed incidents, 38 were excluded from this calculation because they lacked an identifiable Victim, Sponsor, or Type.
-For more, see \cite{council_on_foreign_relations_new_2019}} In the world of cyber-incidents, if it looks like espionage then it is espionage, and it will receive the customary diplomatic benefits associated with that characterization.\footnote{In programming, this is a principle called \enquote{Duck Typing.} If it looks like a duck, and it quacks like a duck, then it is a duck.}
+Cyber-espionage is, however, easily the least likely form of cyberattack to garner an official response from the attacked state
+
+In the world of cyber-incidents, if it looks like espionage then it is espionage, and it will receive the customary diplomatic benefits associated with that characterization.\footnote{In programming, this is a principle called \enquote{Duck Typing.} If it looks like a duck, and it quacks like a duck, then it is a duck.}
 

### 2021-05-26 article.tex
Write legal section

@@
 These applications exposed not just the personal information of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses---compromising the information of 22.1 million people in total.\footcite{nakashima_hacks_2015} US officials feared that the stolen data could help foreign intelligence services unmask undercover CIA officers, and the information could identify American officials who might be susceptible to blackmail, or worse, recruitment.
-FBI Director James Comey admitted that, with the kind of information leaked, a foreign intelligence service could ``know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.''\footcite[Comey is referring specifically to his SF-86, the form that one files with OPM to apply for a security clearance.]{nakashima_hacks_2015} In the case of the OPM hack, it wasn't necessary to imagine who was in possession of that information; the US had already established that the Chinese military was likely behind the sophisticated attack.\footcite{koerner_inside_2016} But even though it possessed conclusive digital forensics, the Obama administration refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018.
+FBI Director James Comey admitted that, with the kind of information leaked, a foreign intelligence service could ``know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.''\footnote{\cite{nakashima_hacks_2015}.
+Comey is referring specifically to his SF-86, the form that one files with OPM to apply for a security clearance.} In the case of the OPM hack, it wasn't necessary to imagine who was in possession of that information; the US had already established that the Chinese military was likely behind the sophisticated attack.\footcite{koerner_inside_2016} But even though it possessed conclusive digital forensics, the Obama administration refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018.
 See \cite{sanger_trump_2018}}
@@
 
-Espionage is by far the most common form of cyberattack and governments do not always reveal when they have been compromised, so it is not terribly convincing to say that most cyber-espionage was ignored by the US Government while all 2 out 2 incidents of cyber-sabotage were not.
+Espionage is by far the most common form of cyberattack and governments do not always reveal when they have been compromised, so it is not terribly convincing to say that most cyber-espionage was ignored by the US Government while both of the two known incidents of cyber-sabotage were not.
+After all, each cyber-incident exists in the context of the American relationship with the nation that perpetrated it; the decision to respond or retaliate is influenced by countless factors and there simply are not enough incidents of cyber-sabotage against the United States to establish a clear contrast in the degree of retaliation.
 But if we expand our scope globally for just a moment, it becomes apparent that the same pattern broadly holds.
@@
 Consider how the OPM hack compares to a contemporary incident of state-sponsored sabotage.
-In 2014, hackers affiliated with the North Korean government leaked damaging emails from Sony executives and wiped some of Sony's servers.\footcite[This hack was allegedly in retaliation for the upcoming James Franco/Seth Rogen comedy about assassinating North Korean leader Kim Jong-Un.]{barnes_sony_2014} White House Press Secretary Josh Earnest alleged that the Sony hack was \enquote{destructive activity with malicious intent that was initiated by a sophisticated actor,} which he said merits \enquote{a proportional response.}\footcite{nakashima_white_2014} The Sony hack would be the first time a sitting US president publicly accused a foreign nation of launching a cyberattack and promised retaliation.\footcite{sanger_u.s._2016} The US made good on its threat and imposed sanctions on North Korea the following year.\footcite{lederman_us_2015} So why did the Sony hack---a newsworthy but far less widespread data breach---receive a significantly more serious response than a direct attack on a federal office that stole the private information of 22 million Americans?
+In 2014, hackers affiliated with the North Korean government leaked damaging emails from Sony executives and wiped some of Sony's servers.\footnote{\cite{barnes_sony_2014}.
+The Sony hack was allegedly in retaliation for the upcoming James Franco/Seth Rogen comedy about assassinating North Korean leader Kim Jong-Un.} White House Press Secretary Josh Earnest alleged that the Sony hack was \enquote{destructive activity with malicious intent that was initiated by a sophisticated actor,} which he said merits \enquote{a proportional response.}\footcite{nakashima_white_2014} The Sony hack would be the first time a sitting US president publicly accused a foreign nation of launching a cyberattack and promised retaliation.\footcite{sanger_u.s._2016} The US made good on its threat and imposed sanctions on North Korea the following year.\footcite{lederman_us_2015} So why did the Sony hack---a newsworthy but far less widespread data breach---receive a significantly more serious response than a direct attack on a federal office that stole the private information of 22 million Americans?
 
@@
 
-To the US government, stealing classified data from a federal office looks like something a spy might do, but wiping a movie studio's servers does not.
+According to the US government, stealing classified data from a federal office looks like something a spy might do, but wiping a movie studio's servers does not.
 While this explanation might ring true---cyber-espionage is espionage, sure---it is also essentially a tautology.
@@
 Like a traditional intelligence operation, the OPM hack \enquote{merely} involved stealing sensitive information, but the quantity of compromised data was orders of magnitude beyond what could be accomplished by photographing the contents of a file cabinet and passing it along in a briefcase.
-\enquote{This is one of those cases where you have to ask, `Does the size of the operation change the nature of it?'\thinspace} said one anonymous senior intelligence official.
-\enquote{Clearly, it does.}\footcite{sanger_u.s._2015} The actions of the US government still say otherwise: at least for now, even operations on the scale of China's OPM infiltration do not merit significant retaliation.
+Said one anonymous senior intelligence official: \enquote{This is one of those cases where you have to ask, \enquote{Does the size of the operation change the nature of it?} Clearly, it does.}\footcite{sanger_u.s._2015} The actions of the US government would ultimately say otherwise; at least for now, even operations on the scale of China's OPM infiltration do not merit significant retaliation.
 
@@
 is espionage legal?---rather than descriptive ones.
-There are few academic scholars who examine the theoretical implications of espionage in practice: what does it mean for existing theories of IR that a category of inter-state action routinely receives retaliation far less severe than the action's national security implications would seem to merit?
+There are few academic scholars who examine the theoretical implications of espionage as it exists in practice: what does it mean for existing theories of IR that a category of inter-state action routinely receives retaliation far less severe than the action's national security implications would seem to merit?
 In the first section of this paper I review the existing scholarship on norms and note the areas in which espionage does not neatly fit into the existing research.
@@
 In the latter section of this paper I demonstrate how espionage norms have been deployed during previous periods of tectonic technological change.
-The historical record demonstrates that the idea of espionage is deeply meaningful to the individual actors that decide state policy; if an action is considered to be espionage, it exists in a self-contained ontological space, with rules and consequences unaffected by rational cost analysis.
+The historical record demonstrates that the idea of espionage is deeply meaningful to the individual actors that decide state policy; if an action is considered to be espionage, it exists in a self-contained ontological space, with rules and consequences that operate outside rational cost analysis.
 The people who are in charge of these matters---what operations to undertake, how much risk to absorb, whether to respond to an operation conducted against them---intuit whether something \emph{counts as} espionage, and adjust their own actions to fit within that paradigm.
@@
 
-\newpage
+\section{Possible Theories}
+Few nations formally acknowledge that peacetime espionage is a core function of the state, so you rarely see treaties or statutes properly define it.
+The Espionage Act of 1917 implicitly defines espionage as providing \enquote{information relating to the national defense which information the possessor has reason to believe could be used to the injury of the United States or to the advantage of any foreign nation [\textellipsis] to any person not entitled to receive it.}\footnote{\cite{noauthor_notitle_nodate}.
+I say \enquote{implicitly} because the Espionage Act of 1917 never actually defines espionage, it's just a piece of legislation called \enquote{The Espionage Act} that criminializes the quoted behavior.} Passed right after US entry into World War I, The Espionage Act was written to address wartime espionage (each clause contains an exhaustive list of military installations, from torpedo stations to canals, within which one might purloin damaging information) but remarkably remains the sole legal tool for prosecuting peacetime espionage as well; almost a century later, the United States charged Edward Snowden with two counts brought under the Espionage Act for providing information about domestic surveillance programs to the press.\footnote{\cite{finn_us_2013}.
+Even though the text of the law does not specify that charges can only be levied during wartime, the Espionage Act precedes the modern concept of peacetime espionage by decades.
+The CIA was founded in 1497, and its main predecessor, the military Office of Strategic Services, did not exist prior to the Second World War.} Procedures governing the appropriate practice of wartime espionage are outlined in both the Geneva Convention and its precursor, the Hague Conventions of 1907, but the legal status of peacetime espionage remains likewise unresolved in international law.\footcite[651]{beim_enforcing_2018}
 
-% This article seeks to unravel that tautology by first by explaining how states determine what constitutes \enquote{traditional} espionage and then by analyzing what strategic purpose that distinction serves.
+The definition of espionage that best captures the phenomenon discussed here is the one proposed by I\~{n}aki Navarrete and Russell Buchan: \enquote{\enquote{Peacetime espionage} is a colloquial rather than a legal term [used] to describe different methods of collecting confidential information from closed as opposed to open sources \textelp{} generally conducted by States, against States.}\footcite[901-902]{navarrete_out_2019} This definition purposefully excludes covert operations that are often conflated with espionage, such as assassinations, election interference, and other forms of extralegal government intervention.\footnote{Some scholars also choose not to export their own definition of espionage, but rather offer a working definition just to bound the scope of that particular paper.
+See \cite[600]{radsan_unresolved_2007}; \cite[2]{kapp_spying_2007}} Covert operations are often conducted by intelligence agencies and by people who would call themselves spies, but do not universally receive the \enquote{you gotta hand it to them} light touch that more strictly-defined espionage does.\footnote{This article excludes covert action from espionage because covert action results in diplomatic consequences, but the question of whether \enquote{intelligence} includes covert action is disputed.
+While I employ the Navarrete and Buchan definition, for a contrasting definition that includes covert action, see: \cite[518]{stout_intelligence_2018}.} When Russian operatives were alleged to have poisoned Sergei Skripal, this was widely to considered to have violated \enquote{the rules of the game,} so the US and the EU issued multiple rounds of sanctions in response.\footnote{\cite{masters_has_2018}.
+Not only are assassinations considered in violation of the rules, but Skripal was a former British spy who had been traded back to the UK in exchange for a Soviet spy, which is especially off-limits.} Espionage is far more common than assassinations however, and when an operation is understood by both parties to have been classic espionage, both parties routinely minimize the consequences associated with it.
+
+% Rarely does the response amount to more than a \emph{pro forma} denunciation of the activity and implementation of the necessary countermeasures to end it---the consequences for espionage are borne by the agents who commit it, not the state that sent them.
+
+To attribute state decisions of that magnitude to violations of \enquote{the rules of the game} is painfully colloquial, but as Navarrete and Buchan smartly identify, the very business of peacetime espionage is colloquial.
+The challenge an espionage scholar faces is to impose some orderly framework of analysis on a field that lacks the key structures---laws, treaties, promises, wars---traditionally used to build such frameworks.
+The remainder of this section will explore whether existing theories of international relations can explain the disparity between the harm that espionage can accomplish and the reluctance of states to impose the sorts of consequences (economic or military) that would discourage its use.\footnote{If at this point in time you are not sufficiently convinced that espionage \emph{is} something that states treat with a different set of consequences, please be aware that the evidentiary section of this article will examine cases in which this disparity is especially stark.}
 
@@
 
-% Add information about my conclusions here
-
-\subsection{Defining espionage norms}
-There are few extant definitions of peacetime espionage, and no satisfying ones.
-The problem that aspiring espionage scholars inevitably encounter is that \enquote{peacetime espionage} is universally understood as a term of art but almost never defined by legal codes, or interpreted by case law.
-Most scholars who tackle this question settle on a definition like the one proposed by I\~{n}aki Navarrete and Russell Buchan: \enquote{\enquote{Peacetime espionage} is a colloquial rather than a legal term [used] to describe different methods of collecting confidential information from closed as opposed to open sources \textelp{} generally conducted by States, against States.}\footcite[901-902]{navarrete_out_2019} This definition purposefully excludes forms of extralegal government intervention that are often conflated with espionage, such as assassinations, election interference, and covert operations.\footnote{Often papers won't take a stand on defining espionage, but rather offer a working definition just to bound the scope of that particular paper.
-See \cite[600]{radsan_unresolved_2007}, \cite[2]{kapp_spying_2007})} Excluding these activities is important: assassinations and similar covert operations are not considered espionage, so they do not typically receive the protections of espionage norms.
-
-For the purposes of this paper, I define espionage norms as the common understanding of how states are expected to respond to activities that remain within the bounds of espionage.\footnote{There are, of course, other \enquote{norms} that involve espionage.
-The so-called \enquote{Moscow Rules} dictate an intelligence officer are supposed to conduct themselves in hostile territory.
-In order to avoid constantly specifying, espionage norms are more strictly defined to include only the ones that affect IR.} Espionage norms prescribe that when one state uncovers an espionage operation conducted against it, the consequences imposed on the spying country remain minimal.
-Rarely does the response amount to more than a \emph{pro forma} denunciation of the activity and implementation of the necessary countermeasures to end it---the consequences for espionage are borne by the agents who commit it, not the state that sent them.
-In rare cases when additional costs \emph{are} imposed on a state for conducting a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.
-
-The limits of espionage norms are defined by what intelligence professionals call \enquote{the rules of the game.} When states do impose harsher penalties, they often justify it by saying that they are punishing an operation that did not abide by the rules.
-The poisoning of Sergei Skripal, allegedly by Russian operatives, was widely described a violation of the rules.\footcite{masters_has_2018} Assassinations are off-limits, especially against former spies who have been traded to safety, and the US and the EU have issued multiple rounds of sanctions in response.\footcite{reuters_e.u._2019} Assassinations, however, are not the typical intelligence operation, and when an operation is understood by both parties to have been routine espionage, the consequences are minimized accordingly.
-
-While most of the information that we have on espionage norms is qualitative, there is some data to support the limited consequences of espionage norms as well.
-Once again we can use cyber-espionage as an analogue for traditional espionage.
-Of the hundreds of state-sponsored cyberattacks known to the public, attacks classified as espionage are by far the most common.
-Cyber-espionage is, however, easily the least likely form of cyberattack to garner an official response from the attacked state
-
-In the world of cyber-incidents, if it looks like espionage then it is espionage, and it will receive the customary diplomatic benefits associated with that characterization.\footnote{In programming, this is a principle called \enquote{Duck Typing.} If it looks like a duck, and it quacks like a duck, then it is a duck.}
-
 % The benefits of espionage have never been underrated---\enquote{it is only the enlightened ruler and the wise general,} reads Sun Tzu, \enquote{who will use the highest intelligence of the army for purposes of spying and thereby they achieve great results}---but the inevitability of espionage has long been taken for granted.
@@
 
-\subsection{Extant explanations for espionage norms}
-Prior literature has attempted to explain espionage norms using international law, but in the absence of any significant treaties governing peacetime espionage, scholars are often left without definitive answers.\footnote{Because it is universally agreed that international law does not deal with peacetime espionage directly, many journal articles that deal with this question frame their work almost Socratically, devoting two sections to cases for and against the legality of espionage.
-In fact, the first major work on law and peacetime espionage is a series of commentaries collected in a single publication that respond to each other's arguments.
-See \cite{wright_essays_1962}} Espionage is universally illegal but every state acknowledges that allies and adversaries alike are guaranteed to engage in it.\footnote{Further complicating matters, espionage is the rare crime that might not be a crime depending on who the victim is.
+\subsection{The curious case of espionage norms}
+The informal yet widely consistent set of espionage consequences is consistent with the manner in which international relations (IR) norms operate.
+The canonical framework for understanding how IR norms are created and promoted is the one outlined by Martha Finnemore and Kathryn Sikkink, who define a norm \enquote{as a standard of appropriate behavior for actors with a given identity.}\footcite[891]{finnemore_international_1998} In this case, espionage norms govern the standard of appropriate behavior for actors who wish to be considered spies; if said actor's behavior conforms to those standards, then the consequences for their actions will be minimized accordingly.
+They also note that because norms embody a sense of \enquote{oughtness,} norms are employed when justification for some action is required, and therefore \enquote{leave an extensive trail of communication among actors that we can study.}\footcite[892]{finnemore_international_1998} The paper trail that the Eisenhower spy program left behind---and the justifications that administration made for using its various devices---is the evidentiary focus of this article.
+
+What sets espionage apart from the typical IR norm is that there is little evidence of anyone arguing that espionage \emph{should} work in any particular way, only that it \emph{does} and therefore it makes sense to operate within those confines.
+Finnemore and Sikkink illustrate their theory of norm entrepreneurship using the women's rights movement and the development of the laws of war, both of which compare rather oddly to the norms of espionage.\footcite[894]{finnemore_international_1998} One struggles to imagine pro-espionage demonstrations analogous to the suffragette marches, and as we have already seen, peacetime espionage is entirely absent from the Geneva Convention and its antecedents.
+If norms embody \enquote{oughtness,} who is arguing that espionage ought to exist?
+
+\subsection{Espionage in international law1}
+
+One place you might look for evidence of norm entrepreneurship is international law.
+Norms are sometimes understood as an alternative to legal regimes, but laws and norms frequently reinforce each other: laws can establish new norms, or they can derive their legitimacy from preexisting norms.\footcite[441]{finnemore_constructing_2016} As Faesen et al.
+note, \enquote{one common goal of norm entrepreneurs is to construct ties to law that strengthen and refine the behavioral expectations of a given norm}\footcite[25]{faesen_blurred_2020} Therefore, if espionage norms exist, one might expect to find evidence that states have attempted to enshrine this set of norms into law, or referenced an existing law to defend their decision to treat an particular incident pursuant to espionage norms.
+
+Instead, legal scholarship has been divided for 80 years on whether international law takes a position on peacetime espionage, and, if so, whether that position endorses or prohibits it.
+The inaugural text on the subject, \emph{Essays on Espionage and International Law,} presents the views of two opposing law professors, Julius Stone and Quincy Wright, both of whom are widely cited in the subsequent literature.\footnote{These two essays are primarily concerned with addressing the implications of the U-2 Incident, a new form of peacetime espionage.
+The facts of that incident will be discussed in-depth in the Case Studies section of this article.} Wright argues that \enquote{[a]ny act by an agent of one state committed in another state’s territory, contrary to the laws of the latter, constitutes intervention.}\footcite[13]{wright_espionage_1962} In his view, if espionage were not legally wrong, then governments would not disclaim their spies.\footcite[17]{wright_espionage_1962} Stone proclaims his \enquote{almost complete dissent} from Wright's view, noting that when spies are caught, the aggrieved state never attempts \enquote{to shift responsibility for the espionage to the home state.}\footcite[33]{stone_legal_1962} If peacetime espionage were a delinquency, Stone argues, why would states continue to allow the transparent excuses of their adversaries to stand?\footcite[39]{stone_legal_1962}
+
+Modern literature on espionage largely builds on one of these two conclusions---either state tolerance of espionage is evidence for its legality, or state disavowal is evidence of its illegality.
+Taking after Wright, Jared Beim argues that while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so there ought to be stronger enforcement mechanisms in place.\footnote{\cite[672]{beim_enforcing_2018}.
+Beim suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe.
+He acknowledges, however, that at best \enquote{it is perhaps possible to stop the more egregious violations.}} Jeffrey Smith argues the opposite, that because espionage is so widespread, \enquote{it is fair to say that the practice of states recognizes espionage as a legitimate function of the state, and therefore it is legal as a matter of customary international law.}\footcite[544]{smith_keynote_2007} He notes that the routine practice of declaring suspected intelligence officers \emph{persona non grata} legitimizes the practice of espionage as one that merely violates diplomatic protocol, not international law.\footcite[544]{smith_keynote_2007} A.
+John Radsan avoids taking a position in either direction, arguing that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law at all.
+He concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.}\footcite[623]{radsan_unresolved_2007} Radsan's non-position is, of course, a position of its own; citing Radsan, Craig Forcese observes that \enquote{the very fact that there are three camps with such diametric positions itself suggests that \textelp{} there is no clear answer on the international legality of extraterritorial espionage.}\footcite[204-205]{forcese_spies_2011}
+
+As the Forcese quote suggests, what unites the literature (and meta-literature) on peacetime espionage is agreement that, regardless of the author's opinion on whether this is desirable, nations have left espionage jurisprudence in a state of salutary neglect.
+Forcese writes that \enquote{the international community seems content with an artful ambiguity on the question.}\footcite[205]{forcese_spies_2011} Darien Pun arrives at the same conclusion, noting that despite the years of debate reviewed here, the status of peacetime espionage has remained unclear.\footcite[260]{pun_rethinking_2017} Though this article is not interested in the legality of espionage, it is extraordinary to note that legal scholars have been completely unable to determine it, because no state has even attempted to resolve this question.
+At least within the context of international law, there is no evidence of any norm entrepreneurship for peacetime espionage whatsoever.
+
+\subsection{Espionage a framing device}
+Not only is peacetime espionage a practically unquestioned normative regime, the proximity of an action to espionage is a first principle against which other norms are judged.
+Two key strategies that norm entrepreneurs employ are \emph{framing}, contextualizing the facts of a case in the light most favorable to promoting a desired norm, and \emph{linking}, connecting the desired norm to an existing, more powerful one.\footcite[451]{finnemore_constructing_2016} While the boundary between these two concepts is blurry, policymakers characterize espionage incidents in a manner much closer to the former.
+The OPM hack was framed as an \enquote{espionage issue,} the way someone might frame landmines as a \enquote{war crimes issue} or women's suffrage as a \enquote{human rights issue,} invoked as a fundamental principle whose general desirability is assumed.\footnote{Obviously, the notion that war crimes should be discouraged is itself a norm.
+The distinction that I believe the authors of this concept intended is that one frames a prospective norm in a manner where its desirability does not need to be explained, whereas one links to an existing norm to borrow that norm's explanation for its utility.}
+
+The inverse framing is true as well: a norm that something is \emph{not} permissible can be pushed with the justification that it is \emph{not} espionage.
+The United States has long acted as a norm entrepreneur against economic espionage, the theft of commercial trade secrets for a nation's economic advantage, differentiating this particular form of espionage from the implied acceptability of traditional espionage.\footcite[19]{nye_normative_2018} The norm against economic espionage, as Samuel Rascoff describes it, \enquote{curiously gerrymandered} to forbid this form of market-state collaboration, while allowing other, functionally identical, ones.\footcite[250]{rascoff_norm_2016}
+
+% what's less clear is what keeps that norm in place
+
+% law is one place where it might be acknowledged
+
+% talk about institutionalization, finnemore + sikkink 901
+
+% Further complicating matters, espionage is the rare crime that might not be a crime depending on who the victim is.
 If an American government employee walks out of The Pentagon and hands classified documents to the KGB, they are at risk prosecuted as a traitor; if an American case officer hands classified KGB documents over to the CIA, they will be lauded as a hero.
-This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.} Because of that tension, scholars have typically treated espionage as \emph{sui generis}---starting from the assumption that espionage norms exist in a class of their own, then attempting to fit a legal doctrine around that fact.
-Jared Beim concludes that while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so there ought to be stronger enforcement mechanisms in place.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe.
-For more, see \cite{beim_enforcing_2018}} Beim acknowledges, however, that at best \enquote{it is perhaps possible to stop the more egregious violations.}\footcite[672]{beim_enforcing_2018} A.
-John Radsan employs a more philosophical approach.
-Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage generally reach the same conclusion: it would be almost impossible to argue that espionage is legal, but the international system leaves this tension in a state of salutary neglect.
-Taken to its logical endpoint, this branch of scholarship would simply argue that the espionage norms we observe today result from simple inertia.
+This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.
 
@@
 
-\section{Theory}
-\subsection{The nuclear option}
-% Add a poll about American knowledge of the CIA?
+% now do cybersec norms
 
-Civilian study of espionage is sometimes treated with skepticism because espionage agencies, by definition, operate with maximum possible secrecy.
+% \subsection{The nuclear option}
+% % Add a poll about American knowledge of the CIA?
+
+% Civilian study of espionage is sometimes treated with skepticism because espionage agencies, by definition, operate with maximum possible secrecy.
 That secrecy does not inoculate these agencies from critical analysis.
@@
 
-Using traditional methods of analysis, espionage certainly seems like something that states should be doing more to discourage.
+% Using traditional methods of analysis, espionage certainly seems like something that states should be doing more to discourage.
 The results of intelligence operations can have devastating, even existential effects on national security.
@@
 
-% Since this generally doesn't happen between major world powers, one of two things must be true: either a state is miscalculating the relative gains it achieves through espionage, or the mutual tolerance of espionage has benefits that are sufficiently important to maintain even if one state gains less from it than its adversary.
+% % Since this generally doesn't happen between major world powers, one of two things must be true: either a state is miscalculating the relative gains it achieves through espionage, or the mutual tolerance of espionage has benefits that are sufficiently important to maintain even if one state gains less from it than its adversary.
 
-% Certainly there are factors that the go into diplomatic decisions that the general public will likely never know.
+% % Certainly there are factors that the go into diplomatic decisions that the general public will likely never know.
 There are few reasons to believe, however, that this is more true for intelligence-related diplomatic matters than for generic ones.
@@
 
-Not only are the risks associated with practicing espionage minimal, espionage norms limit the acceptable repercussions to ones that cannot possibly threaten the structure of civilian intelligence services.
+% Not only are the risks associated with practicing espionage minimal, espionage norms limit the acceptable repercussions to ones that cannot possibly threaten the structure of civilian intelligence services.
 Were it more common to respond to espionage by imposing financial sanctions or even inflicting a military cost, a state would have to consider the possibility that attempting espionage could adversely impact other aspects of its security, leaving it worse off than when it started.
@@
 
-If removing a quarter of the officers doesn't do lasting damage, then why not remove the full 200?
+% If removing a quarter of the officers doesn't do lasting damage, then why not remove the full 200?
 The swift expulsion of all suspected intelligence officers from a country---and the inevitable removal of the country's own intelligence officers in retaliation---would deeply injure both their espionage capabilities indefinitely.\footnote{Nothing on this scale has ever been attempted, but Operation Foot, the largest diplomatic expulsion in the Cold War, expelled 105 Soviet diplomats from the UK.
@@
 
-\subsection{A model for maintaining espionage norms}
-I propose an iterated decision-making model to describe how states cooperate to preserve the liminal legal status of espionage, based on the same tit-for-tat strategy associated with diplomatic expulsions.
+% \subsection{A model for maintaining espionage norms}
+% I propose an iterated decision-making model to describe how states cooperate to preserve the liminal legal status of espionage, based on the same tit-for-tat strategy associated with diplomatic expulsions.
 In game theory, tit-for-tat is highly-successful strategy wherein the player simply mimics the choice that that their opponent made in the last round.\footcite[8]{axelrod_effective_1980} By employing this strategy, the player never gets taken advantage of for more than one turn and can always return to cooperation if their opponent unilaterally signals a willingness to do the same.
@@
 
-% Inevitably, both sides return to a cooperative state while they each still have \emph{some} diplomats left.
+% % Inevitably, both sides return to a cooperative state while they each still have \emph{some} diplomats left.
 Because neither side can gain an a decisive advantage, the back-and-forth punishment never ends in a mutual defection.
 
-My model changes the parameters so that the decision to cooperate or defect is based not on the amount of punishment received, but rather the type.
+% My model changes the parameters so that the decision to cooperate or defect is based not on the amount of punishment received, but rather the type.
 Each time a state is forced to respond to a successful intelligence operation by a rival---such as the reveal of a high-level spy or a major data breach---it can either limit its response to one which is traditionally associated with espionage, or it can impose unprecedented costs that might discourage such an operation in the future (Table \ref{espionage-decision-matrix}).
@@
 
-% https://tex.stackexchange.com/questions/249480/creating-a-payoff-matrix-using-latex-tabular-environment
-\begin{table}[ht]
-\centering
-\setlength{\extrarowheight}{2pt}
-\small
-\begin{tabular}{cc|c|c|}
-  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{State 2}\\
-  & \multicolumn{1}{c}{} & \multicolumn{1}{c}{Tolerate Espionage}  & \multicolumn{1}{c}{Punish Espionage} \\\cline{3-4}
-  \multirow{3}*{State 1}  & Tolerate Espionage & \makecell{~\\Both sides gain intelligence \\~} & \makecell{~\\ Only State 2 continues spying \\ ~} \\\cline{3-4}
-  & Punish Espionage & \makecell{~\\ Only State 1 continues spying \\~} & \makecell{~\\ Limited intelligence either direction \\~} \\\cline{3-4}
-\end{tabular}
-\caption{Espionage consequences decision matrix}
-\label{espionage-decision-matrix}
-\end{table}
+% % https://tex.stackexchange.com/questions/249480/creating-a-payoff-matrix-using-latex-tabular-environment
+% \begin{table}[ht]
+% \centering
+% \setlength{\extrarowheight}{2pt}
+% \small
+% \begin{tabular}{cc|c|c|}
+%   & \multicolumn{1}{c}{} & \multicolumn{2}{c}{State 2}\\
+%   & \multicolumn{1}{c}{} & \multicolumn{1}{c}{Tolerate Espionage}  & \multicolumn{1}{c}{Punish Espionage} \\\cline{3-4}
+%   \multirow{3}*{State 1}  & Tolerate Espionage & \makecell{~\\Both sides gain intelligence \\~} & \makecell{~\\ Only State 2 continues spying \\ ~} \\\cline{3-4}
+%   & Punish Espionage & \makecell{~\\ Only State 1 continues spying \\~} & \makecell{~\\ Limited intelligence either direction \\~} \\\cline{3-4}
+% \end{tabular}
+% \caption{Espionage consequences decision matrix}
+% \label{espionage-decision-matrix}
+% \end{table}
 
-While there is unquestionably a cultural aspect to espionage norms, tradition cannot fully explain why every state remains in a state of mutual cooperation.
+% While there is unquestionably a cultural aspect to espionage norms, tradition cannot fully explain why every state remains in a state of mutual cooperation.
 Espionage consequences are an iterated game that both parties play tit-for-tat, so, in the long run, the only two possible equilibria are mutual defection or mutual cooperation.
@@
 
-% First, even though either state has the opportunity to defect by imposing harsher punishments in response to espionage, its gains would be short-lived because the other state is guaranteed to immediately do the same---tit-for-tat.
+% % First, even though either state has the opportunity to defect by imposing harsher punishments in response to espionage, its gains would be short-lived because the other state is guaranteed to immediately do the same---tit-for-tat.
 
-One possible explanation is that some states are simply miscalculating.
+% One possible explanation is that some states are simply miscalculating.
 Theoretically, the basic premise is persuasive.
@@
 
-The surprisingly straightforward conclusion is that if only two worlds are possible---one with widespread espionage and one without---then all the relevant nations must be concluding that the world without espionage is completely unacceptable.
+% The surprisingly straightforward conclusion is that if only two worlds are possible---one with widespread espionage and one without---then all the relevant nations must be concluding that the world without espionage is completely unacceptable.
 Even if one nation correctly concludes that its rival gains more from a world with a free flow of information than it does, the alternative---a world with significantly harsher punishments for espionage and a staunched flow of intelligence---still leaves the former worse off.
 
-\subsection{Defensive realism}
-To explain why states might be willing to accept relative losses in security, I employ a theory called defensive realism, particularly the version formulated in Charles Glaser's \emph{Rational Theory of International Politics}.
+% \subsection{Defensive realism}
+% To explain why states might be willing to accept relative losses in security, I employ a theory called defensive realism, particularly the version formulated in Charles Glaser's \emph{Rational Theory of International Politics}.
 Glaser describes his theory as integrating defensive realism with neoclassical realism.\footnote{For the sake of simplicity, I am going to refer to it as defensive realism, because they both put the security dilemma at the center of their explanation for international competition and cooperation, which I believe to be the key observation.
@@
 
-Like espionage norms, competition between security-seeking states has two natural equilibria: mutual cooperation and mutual defection.
+% Like espionage norms, competition between security-seeking states has two natural equilibria: mutual cooperation and mutual defection.
 The ideal outcome for both is when neither state builds up arms---with their reciprocal lack of ability to deal damage, both states are confident that no war will result.
@@
 
-\begin{table}[ht]
-\centering
-\setlength{\extrarowheight}{2pt}
-\small
-\begin{tabular}{cc|c|c|}
-  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{State 2}\\
-  & \multicolumn{1}{c}{} & \multicolumn{1}{c}{Retain current arms}  & \multicolumn{1}{c}{Build up arms} \\\cline{3-4}
-  \multirow{3}*{State 1}  & Retain current arms & \makecell{~\\(5,5)\\~} & \makecell{~\\ (1,3) \\ ~} \\\cline{3-4}
-  & Build up arms & \makecell{~\\ (3,1) \\~} & \makecell{~\\ (2,2) \\~} \\\cline{3-4}
-\end{tabular}
-\caption{Defensive Realism payoff matrix}
-\label{defensive-realism-payoff-matrix}
-\end{table}
+% \begin{table}[ht]
+% \centering
+% \setlength{\extrarowheight}{2pt}
+% \small
+% \begin{tabular}{cc|c|c|}
+%   & \multicolumn{1}{c}{} & \multicolumn{2}{c}{State 2}\\
+%   & \multicolumn{1}{c}{} & \multicolumn{1}{c}{Retain current arms}  & \multicolumn{1}{c}{Build up arms} \\\cline{3-4}
+%   \multirow{3}*{State 1}  & Retain current arms & \makecell{~\\(5,5)\\~} & \makecell{~\\ (1,3) \\ ~} \\\cline{3-4}
+%   & Build up arms & \makecell{~\\ (3,1) \\~} & \makecell{~\\ (2,2) \\~} \\\cline{3-4}
+% \end{tabular}
+% \caption{Defensive Realism payoff matrix}
+% \label{defensive-realism-payoff-matrix}
+% \end{table}
 
-To arrive at mutual cooperation, a security-seeking state needs to not only determine that its rival is a security seeker, it needs its rival to make that exact same judgment, because it is the adversary's response to securitization that ultimately makes the first state insecure.
+% To arrive at mutual cooperation, a security-seeking state needs to not only determine that its rival is a security seeker, it needs its rival to make that exact same judgment, because it is the adversary's response to securitization that ultimately makes the first state insecure.
 Defensive realism therefore introduces a counterintuitive conclusion: it is often logical for a security seeker to take actions that increase not only its own security, but its \emph{adversary's} security as well.\footcite[7]{glaser_rational_2010} If the adversary is indeed a security seeker, then, no longer concerned about its security, it will respond by not building up arms, and resolve the security dilemma for both of them.
@@
 
-Espionage creates the conditions that make solving the security dilemma possible.
+% Espionage creates the conditions that make solving the security dilemma possible.
 Consider a situation in which one state is a genuine security-seeker and so is its adversary.
@@
 
-States take the risk that espionage will reveal more than just a state's (hopefully benign) intent---it might also reveal a state's capabilities and strategy.
+% States take the risk that espionage will reveal more than just a state's (hopefully benign) intent---it might also reveal a state's capabilities and strategy.
 I interviewed senior Obama Administration official Jake Sullivan about the benefits of allowing foreign states to spy on the US.
@@
 
-\begin{table}[ht]
-\centering
-\setlength{\extrarowheight}{2pt}
-\small
-\begin{tabular}{cc|c|c|}
-  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{State 2}\\
-  & \multicolumn{1}{c}{} & \multicolumn{1}{c}{Tolerate Espionage}  & \multicolumn{1}{c}{Punish Espionage} \\\cline{3-4}
-  \multirow{3}*{State 1}  & Tolerate Espionage & \makecell{~\\(5,5) \\~} & \makecell{~\\ (0,4) \\ ~} \\\cline{3-4}
-  & Punish Espionage & \makecell{~\\ (4,0) \\~} & \makecell{~\\ (2,2) \\~} \\\cline{3-4}
-\end{tabular}
-\caption{Espionage payoff matrix for two security-seeking states}
-\label{espionage-payoff-matrix}
-\end{table}
+% \begin{table}[ht]
+% \centering
+% \setlength{\extrarowheight}{2pt}
+% \small
+% \begin{tabular}{cc|c|c|}
+%   & \multicolumn{1}{c}{} & \multicolumn{2}{c}{State 2}\\
+%   & \multicolumn{1}{c}{} & \multicolumn{1}{c}{Tolerate Espionage}  & \multicolumn{1}{c}{Punish Espionage} \\\cline{3-4}
+%   \multirow{3}*{State 1}  & Tolerate Espionage & \makecell{~\\(5,5) \\~} & \makecell{~\\ (0,4) \\ ~} \\\cline{3-4}
+%   & Punish Espionage & \makecell{~\\ (4,0) \\~} & \makecell{~\\ (2,2) \\~} \\\cline{3-4}
+% \end{tabular}
+% \caption{Espionage payoff matrix for two security-seeking states}
+% \label{espionage-payoff-matrix}
+% \end{table}
 
-Therefore, states face a binary choice whether to tolerate espionage or punish it.
+% Therefore, states face a binary choice whether to tolerate espionage or punish it.
 We can translate that conclusion to a payoff matrix (for two security-seeking states) that looks almost exactly like the  security dilemma Stag Hunt: the greatest rewards for both are when they cooperate.
@@
 
-%%% %%%
-% A second thing that is a little bit confusing to me about the theory is whether the theory is intelligence revealing information about intentions, or capabilities, or about the other actors "type."  Most of the examples you have in the empirical section seem to be really about gathering intelligence on capabilities rather than intentions.
+% %%% %%%
+% % A second thing that is a little bit confusing to me about the theory is whether the theory is intelligence revealing information about intentions, or capabilities, or about the other actors "type."  Most of the examples you have in the empirical section seem to be really about gathering intelligence on capabilities rather than intentions.
 When you get to the cases it's never clear whether the US is discovering that the Soviet Union is truly a security seeker and hence promises escaping from the security dilemma, or whether the United States is actually gaining intelligence on Soviet capabilities that might, at the extreme, be used aggressively to “win” in a war with the Soviet Union.
 It is unclear from your theory why actors would tolerate this espionage if it is unclear whether it is in fact making both secure or actually is making one side better able to plan an offensive operation against the other.
-%%% %%%
+% %%% %%%
 
-\subsection{Assumptions and alternate explanations}
-The analysis in this section relies on one crucial assumption: that the states in question are security-seekers.
+% \subsection{Assumptions and alternate explanations}
+% The analysis in this section relies on one crucial assumption: that the states in question are security-seekers.
 Analyses of the risks of espionage must contend with possibility that one of the states in question is \enquote{fundamentally dissatisfied with the status quo, desiring additional territory even when it is not required for security}---a greedy state.\footcite[4]{glaser_rational_2010} Such a state would obviously want to protect both its motives and its plans, and therefore derive no benefit from having espionage conducted against it.
@@
 
-Perhaps greedy states simply do not exist.
+% Perhaps greedy states simply do not exist.
 This is a very provocative assertion---one that is, in its most stringent sense, almost certainly wrong---but consider the possibility that no states which abide by espionage norms are greedy \emph{to a destabilizing degree.} The concept of a greedy state is one that naturally operates on a spectrum.\footcite[39.
@@
 
-The international community is likely to view a sudden withdrawal from universal espionage norms as suspicious.
+% The international community is likely to view a sudden withdrawal from universal espionage norms as suspicious.
 States that are farther along on the greed spectrum might decide that it is better to risk the chance that they'll be discovered as greedy states rather send a signal that will all but confirm it.
@@
 
-\subsection{Research design}
-In the following section, I will use empirical case studies to demonstrate how associating an operation with espionage alters its diplomatic response.
+% \subsection{Research design}
+% In the following section, I will use empirical case studies to demonstrate how associating an operation with espionage alters its diplomatic response.
 In order to guide that research, it is important to establish a series of clear, testable hypotheses.
@@
 
-\begin{quote}
-$H_0 =$ Associating a state action with the traditional practice of espionage will not meaningfully alter the diplomatic response.
-\end{quote}
+% \begin{quote}
+% $H_0 =$ Associating a state action with the traditional practice of espionage will not meaningfully alter the diplomatic response.
+% \end{quote}
 
-The null hypothesis posits that there is insufficient evidence linking the perception that an act \enquote{is} espionage to the response that the affected state ultimately delivers.
+% The null hypothesis posits that there is insufficient evidence linking the perception that an act \enquote{is} espionage to the response that the affected state ultimately delivers.
 If this hypothesis holds, espionage may have some symbolic importance, but it isn't one that can be meaningfully tied to a diplomatic response.
@@
 
-\begin{quote}
-$H_1 =$ Associating a state action with espionage will result in a reduced diplomatic response when compared with equivalent actions that do not share that association.
-\end{quote}
+% \begin{quote}
+% $H_1 =$ Associating a state action with espionage will result in a reduced diplomatic response when compared with equivalent actions that do not share that association.
+% \end{quote}
 
-The argument presented by this article risks a tautology: perhaps we associate certain actions with espionage \emph{because} they received a reduced diplomatic response.
+% The argument presented by this article risks a tautology: perhaps we associate certain actions with espionage \emph{because} they received a reduced diplomatic response.
 For this reason, the relationship in $H_1$ is explicitly framed as causal.
@@
 
-\begin{quote}
-$H_2 =$ States purposefully pursue measures to signal that their actions constitute espionage, because they apply reduced diplomatic responses to espionage and hope that their adversary will do the same.
-\end{quote}
+% \begin{quote}
+% $H_2 =$ States purposefully pursue measures to signal that their actions constitute espionage, because they apply reduced diplomatic responses to espionage and hope that their adversary will do the same.
+% \end{quote}
 
-Proving $H_2$ is significantly easier than proving $H_1$.
+% Proving $H_2$ is significantly easier than proving $H_1$.
 State actions---and the reasons policymakers provided for pursuing those actions---are often documented in the historical record.
@@
 
-The following section will demonstrate that both the United States and the USSR were forced to contend with the benefits and risks of a mutually permissive espionage regime during the Cold War.
+% The following section will demonstrate that both the United States and the USSR were forced to contend with the benefits and risks of a mutually permissive espionage regime during the Cold War.
 When faced with opportunities to either shut down the espionage apparatus or expand it, they both chose to expand it.
